### 2025-03-28 _drafts/EDA.qmd
Initial commit :rocket:

---
title: "Procrastination & Dementia"
subtitle: "Markov Model Proposal"
author: 
  - name: "Cormac Monaghan"
    url: "https://c-monaghan.github.io/"
    orcid: "https://orcid.org/0000-0001-9012-3060"
    affiliation:
      - "Hamilton Institute, Maynooth University, Ireland"
      - "Department of Psychology, Maynooth University, Ireland"
  - name: "Rafael de Andrade Moral"
    url: "https://rafamoral.github.io/"
    orcid: "https://orcid.org/0000-0002-0875-3563"
    affiliation:
      - "Department of Mathematics and Statistics, Maynooth University, Ireland"
  - name: "Joanna McHugh Power"
    orcid: "https://orcid.org/0000-0002-7387-3107"
    affiliation: 
      - "Department of Psychology, Maynooth University, Ireland"
toc: true
toc-depth: 4
toc-title: On this page
bibliography: references.bib
csl: apa.csl
format: 
  html:
    theme: 
      light: united
    page-layout: full
    html-math-method: katex
    code-tools: true
    self-contained: true
    code-fold: true
    code-summary: "Show the code"
    include-in-header: 
      - file: ./html/github-corner.html
include-after-body: ./html/footer.html
execute: 
  warning: false
editor: source
---

```{r setting-up}
rm(list = ls())

# Packages
suppressMessages(library(dplyr))
library(ggplot2)
library(ggalluvial)

source(here::here("./02__Models/00__Markov_Functions.R"))

# Paths -----------------------------------------------------------------------
path_data <- "./01__Data/02__Processed/"
```

# Overview

The Health and Retirement Study (HRS) collects an extensive array of data related to cognitive health, physical well-being, economic status, and psychosocial factors. Among the information gathered, specific variables are used to assess and classify individuals into distinct categories of cognitive functioning. These classifications help to better understand the progression of cognitive decline, enabling researchers to track trends and identify factors that may influence cognitive health over time.

## Cognitive Function Categories

- **Normal Cognition:** Individuals in this category show no significant signs of cognitive impairment and are able to function independently in daily activities.
- **Mild Cognitive Impairment (MCI):** This stage reflects a slight but noticeable decline in cognitive abilities, such as memory or thinking skills, that goes beyond what would be expected with normal aging but does not yet interfere significantly with daily life.
- **Dementia:** This classification is marked by more severe cognitive deficits, impacting memory, reasoning, and the ability to perform everyday tasks. Dementia encompasses various neurodegenerative conditions, with Alzheimer's disease being the most common.

By tracking transitions between these categories, researchers can gain insights into the factors that contribute to cognitive decline and identify potential interventions to promote healthy aging.

# Langa-Weir Classifications

For previous waves of HRS data (1995 - 2020) there is a researcher contributed dataset of dementia classifications [@langa2023]. Researchers have used this dataset to study the trajectories of cognitive aging, dementia risk, and related health outcomes in older adults. However, with the recent release of the 2022 HRS data, **these classifications have yet to be updated**

## LWC2022 Package

In order to address this we developed the `LWC2022` package [@Monaghan2024] to replicate the methodology of @langa2023. This package automates the classification process, ensuring consistency with previous waves and providing researchers with an up-to-date resource for studying cognitive decline and dementia across all available HRS waves. @fig-lwc illustrates the LWC workflow.

![LWC Workflow](./images/LWC_Workflow.png){#fig-lwc}

# Dementia Dataset

Our updated dataset contains 13 columns of cognitive classifications spanning from 1996 to 2022. These columns reflect dementia status across each wave, using the same criteria and structure as the original Langa-Weir dataset, now extended to include the most recent 2022 wave.

```{r data}
data <- readxl::read_xlsx(here::here(path_data, "data.xlsx"))

dementia_data <- data |>
  count_transitions(years = seq(1996, 2022, by = 2))

glimpse(dementia_data)
```

## Missing data

Since the participant pool is drawn from the 2020 wave of the HRS data, any analysis that looks back at previous waves will inevitably encounter missing data. Not all participants were included in earlier waves, leading to gaps in the data. @fig-missing-data displays the frequency of missing cognitive test data across each HRS wave.

```{r}
#| fig-width: 8
#| label: fig-missing-data
#| fig-cap: Occurance of missing data across HRS waves (participants were gathered from the 2020 HRS wave [in red])

# Dynamic colours
colours <- c(rep("#2e2e2e", times = 12), "red", "#2e2e2e")

data |>
  select(any_of(paste0("cogtot27_imp", 1996:2022))) |>
  rename_with( ~ stringr::str_replace(., "cogtot27_imp", ""), starts_with("cogtot27")) |>
  visdat::vis_dat(palette = "cb_safe") +
  labs(title = "Missing data across HRS Waves (1996 - 2022)",
       subtitle = "Total Cognitive Scores") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold", colour = "#2e2e2e"),
    plot.subtitle = element_text(hjust = 0.5, size = 10, colour = "#2e2e2e"),
    axis.text.x = element_text(colour = colours)) +
  ggeasy::easy_remove_legend() +
  ggeasy::easy_x_axis_labels_size(size = 9)

```

The HRS had its most recent participant intake in 2016, which explains the notable decline in missing data occurrences from that point onward. As new participants were not added after 2016, the dataset becomes more complete in subsequent waves, with fewer missing values.

Given this shift, we will conduct our analysis on both:

- A full dataset (1996 - 2022) 
- A reduced dataset (2016 - 2022).

## Exploratory Data Analysis

::: {.panel-tabset}

# 1996 - 2022

## Cognitive Test Scores

Initially, we will plot a distribution of the cognitive test scores (ranging from 0 - 27) across time for all HRS participants.

```{r}
#| fig-width: 12
#| label: fig-cognitive-test-scores_full
#| fig-cap: Cognitive test scores (1996 - 2022)

plot_cognitive_scores(data = data, year = 1996)
```

## Classification proportion per year

@fig-dementia-proportions-full shows the proportion of dementia classifications per HRS wave. The variable **Missing** represents the procrastination HRS respondents who were not yet interviewed by the HRS. Since the analysis focuses on participants included in the 2020 HRS wave, any retroactive analysis of prior waves may result in missing data for certain individuals

```{r}
#| fig-width: 12
#| label: fig-dementia-proportions-full
#| fig-cap: Proportion of dementia classifications (1996 - 2022)

# Plotting proportions --------------------------------------------------------
dementia_data %>%
  ggplot(aes(x = Wave, fill = Status)) +
  geom_bar(position = "fill", 
           alpha = 0.5,
           width = 0.5,
           colour = "black") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Proportion of Cognitive Status Classifications (1996 - 2022)", 
       x = "", y = "Percentage") +
  scale_fill_viridis_d(direction = -1) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.title = element_text(size = 14, hjust = 0.5, face = "bold"),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 10),
        text = element_text(size = 10)) +
  ggeasy::easy_move_legend("bottom") +
  ggeasy::easy_remove_legend_title()
```

## Dementia transitions per year

@fig-dementia-transitions-full is an alluvial graph illustrating the transitions in cognitive classifications from one HRS wave to the next.

```{r}
#| fig-width: 12
#| label: fig-dementia-transitions-full
#| fig-cap: Dementia transitions (1996 - 2022)

dementia_data |>
  plot_transitions(size = 2.5)
```

# 2016 - 2022

## Cognitive Test Scores

Initially, we will plot a distribution of the cognitive test scores (ranging from 0 - 27) across time for all HRS participants.

```{r}
#| fig-width: 12
#| label: fig-cognitive-test-scores_reduced
#| fig-cap: Cognitive test scores (2016 - 2022)

plot_cognitive_scores(data = data, year = 2016)
```

## Classification proportion per year

@fig-dementia-proportions-reduced shows the proportion of dementia classifications from HRS 2016 - 2022. The variable **Missing** represents the procrastination HRS respondents who were not yet interviewed by the HRS. Since the analysis focuses on participants included in the 2020 HRS wave, any retroactive analysis of prior waves may result in missing data for certain individuals

```{r}
#| fig-width: 12
#| label: fig-dementia-proportions-reduced
#| fig-cap: Proportion of dementia classifications per year

# Plotting proportions --------------------------------------------------------
data |>
  count_transitions(years = seq(2016, 2022, by = 2)) |>
  ggplot(aes(x = Wave, fill = Status)) +
  geom_bar(position = "fill", alpha = 0.5, colour = "black") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Proportion of Cognitive Status Classifications (2016 - 2022)", 
       x = "", y = "Percentage") +
  scale_fill_viridis_d(direction = -1) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.title = element_text(size = 14, hjust = 0.5, face = "bold"),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 10),
        text = element_text(size = 10)) +
  ggeasy::easy_move_legend("bottom") +
  ggeasy::easy_remove_legend_title()
```

## Dementia transitions per year

@fig-dementia-transitions-reduced is an alluvial graph illustrating the transitions in cognitive classifications from one HRS wave to the next.

```{r}
#| fig-width: 12
#| label: fig-dementia-transitions-reduced
#| fig-cap: Dementia transitions (2016 - 2022)

data |>
  count_transitions(years = seq(2016, 2022, by = 2)) |>
  plot_transitions(size = 2.5)
```

:::

# Markov Modelling

In this section, we apply Markov modelling to analyze the transitions between cognitive states (Normal Cognition, Mild Cognitive Impairment [MCI], and Dementia) using longitudinal data from the Health and Retirement Study (HRS) from 1996 to 2022. By modeling these transitions, we aim to understand the probabilities of moving between different cognitive states over time.

::: {.panel-tabset}

# 1996 - 2022

## Data preparation

The first step is to prepare the transition dataset. We start by transforming the data to create a record of transitions from one cognitive state to another between consecutive waves (1996 - 2022).

Once the transitions are identified, we calculate the probabilities of each state-to-state transition. These probabilities are computed by counting the occurrences of each transition and dividing by the total number of transitions.

```{r}
tran_prob_full <- data |>
  extract_years(years = seq(1996, 2022, by = 2)) |>
  create_transitions(absorbing = TRUE) |>
  calculate_probabilties()

tran_prob_full |>
  mutate(prop = round(prop, digits = 3)) |>
  head(n = 9) |>
  gt::gt() |>
  gtExtras::gt_theme_538() |>
  gt::cols_align(align = "center") |>
  gt::tab_header(title = "Overall Transition Probabilties (1996 - 2022)") |>
  gt::tab_options(table.width = gt::pct(100)) |>
  gt::opt_align_table_header(align = "center")
```

## Creating transition matrix

Next, we organize the transition probabilities into a matrix format, where rows represent the "from" states and columns represent the "to" states. The matrix entries contain the transition probabilities for each state-to-state pair

```{r}
tran_matrix_full <- tran_prob_full |>
  transition_matrix() |>
  normalise()

tran_matrix_full
```

This code creates a $3 \times 3$ matrix (for the three states: Normal Cognition, MCI, and Dementia) and populates it with the transition probabilities.

## Visualising the matrix

To make the transition matrix more interpretable, we visualize the probabilities using a heat map (@fig-transition-matrix-full)

```{r}
#| fig-width: 8
#| label: fig-transition-matrix-full
#| fig-cap: Transition Matrix

plot_matrix(data = t(tran_matrix_full), year = 1996)

```

# 2016 - 2022

## Data preparation

The first step is to prepare the transition dataset. We start by transforming the data to create a record of transitions from one cognitive state to another between consecutive waves (2016 - 2022).

Once the transitions are identified, we calculate the probabilities of each state-to-state transition. These probabilities are computed by counting the occurrences of each transition and dividing by the total number of transitions.

```{r}
tran_prob_reduced <- data |>
  extract_years(years = seq(2016, 2022, by = 2)) |>
  create_transitions(absorbing = TRUE) |>
  calculate_probabilties()

tran_prob_reduced |>
  mutate(prop = round(prop, digits = 3)) |>
  head(n = 9) |>
  gt::gt() |>
  gtExtras::gt_theme_538() |>
  gt::cols_align(align = "center") |>
  gt::tab_header(title = "Overall Transition Probabilties (2016 - 2022)") |>
  gt::tab_options(table.width = gt::pct(100)) |>
  gt::opt_align_table_header(align = "center")

```

## Creating transition matrix

Next, we organize the transition probabilities into a matrix format, where rows represent the "from" states and columns represent the "to" states. The matrix entries contain the transition probabilities for each state-to-state pair. Additionally, we normalise the matrix so

```{r}
tran_matrix_reduced <- tran_prob_reduced |>
  transition_matrix() |>
  normalise()

tran_matrix_reduced
```

This code creates a $3 \times 3$ matrix (for the three states: Normal Cognition, MCI, and Dementia) and populates it with the transition probabilities.

## Visualising the matrix

To make the transition matrix more interpretable, we visualize the probabilities using a heat map (@fig-transition-matrix-reduced)

```{r}
#| fig-width: 8
#| label: fig-transition-matrix-reduced
#| fig-cap: Transition Matrix

plot_matrix(data = t(tran_matrix_reduced), year = 2016)

```

:::

# Per Wave Probabilities

Instead of looking at the overall probabilities of transitioning between cognitive states, we now examine how these transition probabilities change across multiple waves of data collection. This allows us to understand how cognitive states (Normal Cognition, Mild Cognitive Impairment (MCI), and Dementia) evolve over time within the study population.

## Creating per wave transition matrices

We begin we create transition matrices for each wave of data. Each matrix will represent the probabilities of transitioning from one cognitive state to another within a given wave.

```{r}
create_transition_matrix <- function(wave_data) {
  # Aggregate by 'from' and 'to' states to sum transition probabilities
  wave_data_agg <- wave_data |>
    dplyr::group_by(from, to) |>
    dplyr::summarise(transition_prob = sum(prop), .groups = "drop")

  wave_matrix <- wave_data_agg |>
    tidyr::pivot_wider(
      names_from = to,
      values_from = transition_prob,
      values_fill = 0
    ) |>
    tibble::column_to_rownames("from") |>
    as.matrix()

  return(wave_matrix)
}

# Calculating transition matrix
tran_matrices_reduced <- data |>
  extract_years(years = seq(1998, 2022, by = 2)) |>
  create_transitions(absorbing = TRUE) |>
  group_by(Wave) |>
  calculate_probabilties() |>
  group_map( ~ create_transition_matrix(.x), .keep = TRUE)

tran_matrices_reduced[[13]] <- NULL

head(tran_matrices_reduced, n = 3)
```

## Reshaping matrix

Once we have the transition matrices, we need to reshape them into a long format for easier plotting and further analysis. In the long format, each row represents a single transition, along with the corresponding transition probability and the wave of data collection.

```{r}
# Function to reshape each matrix and add the Wave information
reshape_matrix <- function(matrix, wave) {
  matrix |>
    as.data.frame() |>
    tibble::rownames_to_column(var = "from") |>
    tidyr::pivot_longer(cols = -from,
                        names_to = "to",
                        values_to = "transition_prob") |>
    mutate(Wave = wave)
}

# I want my matrix in a certain order
tran_matrices_reduced <- lapply(tran_matrices_reduced, function(mat) {
  mat[c("Normal Cognition", "MCI", "Dementia"), c("Normal Cognition", "MCI", "Dementia")] |>
    normalise()
})

# Creating a singular dataframe of transition probabilities --------------------
wave_names <- c(seq(1998, 2020, by = 2))

# Apply the reshaping function to each matrix and combine the results
transition_probabilties_long <- purrr::map2_dfr(tran_matrices_reduced, wave_names, reshape_matrix) |>
  mutate(
    transition_prob = round(transition_prob, digits = 3),
    from = factor(from, levels = c("Normal Cognition", "MCI", "Dementia"))
  )

# Outputting table
DT::datatable(
  transition_probabilties_long,
  filter = "top",
  colnames = c("Prob" = "transition_prob"),
  options = list(pageLength = 6, autoWidth = TRUE)
)
```

## Plotting transition probabilities

Finally, we plot the transition probabilities as heatmaps to visualize how they evolve across different waves. We will create an animated GIF that cycles through the waves, showing the changes in transition probabilities over time.

```{r fig.width=8, fig.show="animate", ffmpeg.format="gif", dev="png"}

plot_transition_wave <- function(data, wave) {
  # Main transitions per wave
  filter_data <- data.frame(
    from = c(
      "Normal Cognition",
      "Normal Cognition",
      "Normal Cognition",
      "MCI",
      "MCI",
      "MCI",
      "Dementia",
      "Dementia",
      "Dementia"
    ),
    to = c(
      "Normal Cognition",
      "MCI",
      "Dementia",
      "Normal Cognition",
      "MCI",
      "Dementia",
      "Normal Cognition",
      "MCI",
      "Dementia"
    )
  )

  data |>
    semi_join(filter_data, by = c("from", "to")) |>
    filter(Wave == wave) |>
    ggplot(aes(x = to, y = from, fill = transition_prob)) +
    geom_tile() +
    geom_text(aes(label = transition_prob), size = 5) +
    scale_fill_viridis_c(alpha = 0.8) +
    labs(
      title = paste0("Transition Probabilities Across Cognitive States (", wave, ")"),
      x = "To",
      y = "From",
      fill = "Prob"
    ) +
    theme_light() +
    theme(
      plot.title = element_text(
        hjust = 0.5,
        size = 12,
        face = "bold"
      ),
      axis.title = element_text(size = 10),
      axis.text = element_text(size = 8),
      panel.grid = element_blank(),
      panel.border = element_blank()
    )
}

for (wave in wave_names) {
  print(plot_transition_wave(data = transition_probabilties_long, wave = wave))
}

```

# Cognitive Scores

In this section, we incorporate participants' cognitive scores (from 2016 - 2022) into our analysis. The cognitive scores are visualized in @fig-cognitive-test-scores_reduced, which provides a summary of the data.

```{r}
# Data processing --------------------------------------------------------------
ed_level <- c(
  "No_degree", "GED", "High_school", "College_2_years", 
  "College_4_years", "Masters", "PhD", "Other")

ed_label <- c(
  "No degree", "GED", "High school", "College (2 years)", 
  "College (4 years)", "Masters", "PhD", "Other")

# Variables to add to datafile
adding <- data |>
  dplyr::select(ID, Age_2022, Education, Total_p) |>
  dplyr::mutate(
    Education = dplyr::case_when(
      Education == 0 ~ "No_degree",
      Education == 1 ~ "GED",
      Education == 2 ~ "High_school",
      Education == 3 ~ "College_2_years",
      Education == 4 ~ "College_4_years",
      Education == 5 ~ "Masters",
      Education == 6 ~ "PhD",
      Education == 9 ~ "Other"
    ),
    Education = factor(Education, levels = ed_level, labels = ed_label)
  )

cog_data <- data |>
  extract_years(years = seq(2016, 2022, by = 2), scores = TRUE) |>
  process_data(absorbing = TRUE) |>
  dplyr::inner_join(adding, by = "ID") |>
  backtrack_age() |>
  dplyr::group_by(ID) |>
  dplyr::filter(all(!is.na(Score))) |>
  dplyr::ungroup() |>
  dplyr::filter(!is.na(Total_p)) |>
  dplyr::mutate(Total_p = ifelse(Year == 2022, Total_p, NA))

# Outputting table
cog_data |>
  dplyr::relocate(c(Age, Education), .after = Year) |>
  dplyr::rename(Procrastination = Total_p) |>
  head(n = 12) |>
  gt::gt() |>
  gtExtras::gt_theme_538() |>
  gt::cols_align(align = "center") |>
  gt::tab_header(title = "Cognition Scores (2016 - 2022)") |>
  gt::tab_options(table.width = gt::pct(100)) |>
  gt::opt_align_table_header(align = "center")
```

## Visualisations

To explore the relationships between cognitive scores and other variables, we create a series of visualizations. These plots help us identify trends, correlations, and potential areas for further investigation.

::: {.panel-tabset}

# Age

The first visualization (@fig-scatter-age) examines the relationship between cognitive scores and age. A scatter plot is used to display the data, with points jittered to reduce overlap.

```{r}
#| fig-width: 12
#| label: fig-scatter-age
#| fig-cap: Scatter plot of age vs cognitive scores

cog_data |>
  ggplot(aes(x = Age, y = Score, colour = Class)) +
  geom_jitter(size = 2, width = 0.2, height = 0.2) +
  scale_x_continuous(breaks = seq(0, 100, by = 10)) +
  scale_y_continuous(breaks = seq(0, 27, by = 5)) +
  facet_wrap(~ Year) +
  theme_bw() +
  ggeasy::easy_remove_legend_title() +
  ggeasy::easy_move_legend(to = "bottom")
```

# Education

The second visualization (@fig-education) explores the relationship between cognitive scores and education levels. A bar plot is used to display the mean cognitive score for each education level, faceted by year.

```{r}
#| fig-width: 12
#| label: fig-education
#| fig-cap: Mean cognitive score by education level

cog_data |>
  group_by(Education, Year) |>
  summarise(Score = mean(Score), .groups = "keep") |>
  ggplot(aes(x = Education, y = Score, fill = forcats::fct_rev(Education))) +
  geom_bar(stat = "identity", width = 0.75, alpha = 0.6) +
  scale_y_continuous(breaks = seq(0, 27, by = 3)) +
  labs(title = "Mean Cognitive Score by Education Level",
       subtitle = "Across Time", x = "", y = "Total Cognitive Score") +
  facet_wrap( ~ Year) +
  theme_minimal() +
  coord_flip() +
  ggeasy::easy_remove_legend() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 15, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 15, face = "bold"),
    axis.text.y = element_text(size = 12),
    strip.text = element_text(size = 10, face = "bold"))

```

# Procrastination

The final visualization (@fig-scatter-score-procrastination) investigates the relationship between cognitive scores and procrastination tendencies. A scatter plot is used to display the data for the year 2022.

```{r}
#| fig-width: 12
#| label: fig-scatter-score-procrastination
#| fig-cap: Scatter plot of procrastination vs cognitive scores

cog_data |>
  filter(Year == 2022) |>
  ggplot(aes(x = Total_p, y = Score)) +
  geom_jitter(width = 0.25, size = 1.5, alpha = 0.5) +
  labs(
    title = "Procrastination vs. Cognitive Classification Score",
    x = "Total Procrastination", 
    y = "Cognitive Classification Score") +
  scale_x_continuous(breaks = seq(0, 60, by = 10)) +
  scale_y_continuous(breaks = seq(0, 27, by = 5)) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10))

```

:::

### 2025-03-28 _drafts/Markov.qmd
Initial commit :rocket:

---
title: "Procrastination & Dementia: A Markov Model"
author: 
  - name: "Cormac Monaghan"
    url: "https://c-monaghan.github.io/"
    orcid: "https://orcid.org/0000-0001-9012-3060"
    affiliation:
      - "Hamilton Institute, Maynooth University, Ireland"
      - "Department of Psychology, Maynooth University, Ireland"
  - name: "Rafael de Andrade Moral"
    url: "https://rafamoral.github.io/"
    orcid: "https://orcid.org/0000-0002-0875-3563"
    affiliation:
      - "Department of Mathematics and Statistics, Maynooth University, Ireland"
  - name: "Joanna McHugh Power"
    orcid: "https://orcid.org/0000-0002-7387-3107"
    affiliation: 
      - "Department of Psychology, Maynooth University, Ireland"
toc: true
toc-depth: 4
toc-title: On this page
bibliography: references.bib
csl: apa.csl
format: 
  html:
    theme: 
      light: united
    page-layout: full
    html-math-method: katex
    code-tools: true
    self-contained: true
    code-fold: true
    code-summary: "Show the code"
    include-in-header: 
      - file: ./html/github-corner.html
include-after-body: ./html/footer.html
execute: 
  warning: false
editor: source
---

```{r setting-up}
rm(list = ls())

set.seed(43561)

# Packages
suppressMessages(library(dplyr))
library(tidyr)
library(ggplot2)
library(ggalluvial)

source(here::here("./02__Models/00__Functions.R"))

# Paths -----------------------------------------------------------------------
path_data <- "./01__Data/02__Processed/"
```

# Overview

The Health and Retirement Study (HRS) collects an extensive array of data related to cognitive health, physical well-being, economic status, and psychosocial factors. Among the information gathered, specific variables are used to assess and classify individuals into distinct categories of cognitive functioning. These classifications help to better understand the progression of cognitive decline, enabling researchers to track trends and identify factors that may influence cognitive health over time.

## Cognitive Function Categories

- **Normal Cognition:** Individuals in this category show no significant signs of cognitive impairment and are able to function independently in daily activities.
- **Mild Cognitive Impairment (MCI):** This stage reflects a slight but noticeable decline in cognitive abilities, such as memory or thinking skills, that goes beyond what would be expected with normal aging but does not yet interfere significantly with daily life.
- **Dementia:** This classification is marked by more severe cognitive deficits, impacting memory, reasoning, and the ability to perform everyday tasks. Dementia encompasses various neurodegenerative conditions, with Alzheimer's disease being the most common.

By tracking transitions between these categories, researchers can gain insights into the factors that contribute to cognitive decline and identify potential interventions to promote healthy aging.

# Langa-Weir Classifications

For previous waves of HRS data (1995 - 2020) there is a researcher contributed dataset of dementia classifications [@langa2023]. Researchers have used this dataset to study the trajectories of cognitive aging, dementia risk, and related health outcomes in older adults. However, with the recent release of the 2022 HRS data, **these classifications have yet to be updated**

## LWC2022 Package

In order to address this we developed the `LWC2022` package [@Monaghan2024] to replicate the methodology of @langa2023. This package automates the classification process, ensuring consistency with previous waves and providing researchers with an up-to-date resource for studying cognitive decline and dementia across all available HRS waves. @fig-lwc illustrates the LWC workflow.

![LWC Workflow](./images/LWC_Workflow.png){#fig-lwc}

# Dementia Dataset

Our updated dataset contains 13 columns of cognitive classifications spanning from 1996 to 2022. These columns reflect dementia status across each wave, using the same criteria and structure as the original Langa-Weir dataset, now extended to include the most recent 2022 wave.

```{r data-import}
data <- readxl::read_xlsx(here::here(path_data, "data.xlsx"))

dementia_data <- data |>
  count_transitions(years = seq(1996, 2022, by = 2))

glimpse(dementia_data)
```

## Missing data

Since the participant pool is drawn from the 2020 wave of the HRS data, any analysis that looks back at previous waves will inevitably encounter missing data. Not all participants were included in earlier waves, leading to gaps in the data. @fig-missing-data displays the frequency of missing cognitive test data across each HRS wave.

```{r missing-data}
#| fig-width: 8
#| label: fig-missing-data
#| fig-cap: Occurance of missing data across HRS waves (participants were gathered from the 2020 HRS wave [in red])

# Dynamic colours
colours <- c(rep("#2e2e2e", times = 12), "red", "#2e2e2e")

data |>
  select(any_of(paste0("cogtot27_imp", 1996:2022))) |>
  rename_with( ~ stringr::str_replace(., "cogtot27_imp", ""), starts_with("cogtot27")) |>
  visdat::vis_dat(palette = "cb_safe") +
  labs(title = "Missing data across HRS Waves (1996 - 2022)",
       subtitle = "Total Cognitive Scores") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold", colour = "#2e2e2e"),
    plot.subtitle = element_text(hjust = 0.5, size = 10, colour = "#2e2e2e"),
    axis.text.x = element_text(colour = colours)) +
  ggeasy::easy_remove_legend() +
  ggeasy::easy_x_axis_labels_size(size = 9)

```

The HRS had its most recent participant intake in 2016, which explains the notable decline in missing data occurrences from that point onward. As new participants were not added after 2016, the dataset becomes more complete in subsequent waves, with fewer missing values.

Given this shift, we will conduct our analysis **the reduced dataset (2016 - 2022)** (@fig-complete-data) along with removing the missing values from 2016 and 2018.

```{r complete-data}
#| fig-width: 8
#| label: fig-complete-data
#| fig-cap: HRS dataset after processing

data |>
  select(any_of(paste0("cogtot27_imp", 2016:2022))) |>
  rename_with( ~ stringr::str_replace(., "cogtot27_imp", ""), starts_with("cogtot27")) |>
  na.omit() |>
  visdat::vis_dat(palette = "cb_safe") +
  labs(title = "Complete dataset across HRS Waves (2016 - 2022)",
       subtitle = "Total Cognitive Scores") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 12, face = "bold", colour = "#2e2e2e"),
    plot.subtitle = element_text(hjust = 0.5, size = 10, colour = "#2e2e2e")) +
  ggeasy::easy_remove_legend() +
  ggeasy::easy_x_axis_labels_size(size = 9)
```

## Exploratory Data Analysis

### Cognitive Test Scores

Initially, we will plot a distribution of the cognitive test scores (ranging from 0 - 27) across time for all HRS participants.

```{r cog-scores}
#| fig-width: 12
#| label: fig-cognitive-test-scores
#| fig-cap: Cognitive test scores (2016 - 2022)

plot_cognitive_scores(data = data, year = 2016)
```

### Classification proportion per year

@fig-dementia-proportions shows the proportion of dementia classifications from HRS 2016 - 2022. The variable **Missing** represents the procrastination HRS respondents who were not yet interviewed by the HRS. Since the analysis focuses on participants included in the 2020 HRS wave, any retroactive analysis of prior waves may result in missing data for certain individuals

```{r cog-class}
#| fig-width: 12
#| label: fig-dementia-proportions
#| fig-cap: Proportion of dementia classifications per year

# Plotting proportions --------------------------------------------------------
data |>
  count_transitions(years = seq(2016, 2022, by = 2), absorbing = TRUE) |>
  ggplot(aes(x = Wave, fill = Classification)) +
  geom_bar(position = "fill", alpha = 0.5, colour = "black") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Proportion of Cognitive Status Classifications (2016 - 2022)", 
       x = "", y = "Percentage") +
  scale_fill_viridis_d(direction = -1) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.title = element_text(size = 14, hjust = 0.5, face = "bold"),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 10),
        text = element_text(size = 10)) +
  ggeasy::easy_move_legend("bottom") +
  ggeasy::easy_remove_legend_title()
```

### Dementia transitions per year

@fig-dementia-transitions is an alluvial graph illustrating the transitions in cognitive classifications from one HRS wave to the next.

```{r cog-transition}
#| fig-width: 12
#| label: fig-dementia-transitions
#| fig-cap: Dementia transitions (2016 - 2022)

data |>
  count_transitions(years = seq(2016, 2022, by = 2)) |>
  group_by(ID, Wave) |>
  reframe(plyr::count(Classification)) |>
  rename(Classification = x, n = freq) |>
  plot_transitions(size = 2.5)
```

From this figure we can see that some individuals classified with dementia in one wave transition out of dementia in a subsequent wave. Let's fix this by adding an additional parameter to our `count_transitions()` function to turn dementia into an absorbing state (@fig-dementia-transitions-absorbing). We can do this by using the `cumany()` function from the `dplyr` [@Wickham2023] package.

```{r cog-transition-absorbing}
#| fig-width: 12
#| label: fig-dementia-transitions-absorbing
#| fig-cap: Dementia transitions with dementia as an absorbing state (2016 - 2022)

data |>
  count_transitions(years = seq(2016, 2022, by = 2), absorbing = TRUE) |>
  group_by(ID, Wave) |>
  reframe(plyr::count(Classification)) |>
  rename(Classification = x, n = freq) |>
  plot_transitions(size = 2.5)
```

# Markov Modelling

In this section, we apply Markov modelling to analyze the transitions between cognitive states (Normal Cognition, Mild Cognitive Impairment [MCI], and Dementia) using longitudinal data from the Health and Retirement Study (HRS) from 2016 to 2022. By modeling these transitions, we aim to understand the probabilities of moving between different cognitive states over time.

## Data preparation

The first step is to prepare the transition dataset. We start by transforming the data to create a record of transitions from one cognitive state to another between consecutive waves.

Once the transitions are identified, we calculate the probabilities of each state-to-state transition. These probabilities are computed by counting the occurrences of each transition and dividing by the total number of transitions.

```{r transition-probability}
tran_prob <- data |>
  extract_years(years = seq(2016, 2022, by = 2)) |>
  create_transitions(absorbing = TRUE) |>
  calculate_probabilties()

tran_prob |>
  mutate(prop = round(prop, digits = 3)) |>
  head(n = 9) |>
  gt::gt() |>
  gtExtras::gt_theme_538() |>
  gt::cols_align(align = "center") |>
  gt::tab_header(title = "Overall Transition Probabilties (2016 - 2022)") |>
  gt::tab_options(table.width = gt::pct(100)) |>
  gt::opt_align_table_header(align = "center")
```

## Creating transition matrix

The probability distribution of transitions from one state to another can be represented into a transition matrix $P = (p_{ij})_{i,j}$ where each element of position $(i, j)$ represents the transition probability $p_{ij}$.

$$
P = \begin{bmatrix} 
p_{11} & p_{12} & p_{13} \\
p_{21} & p_{22} & p_{23} \\
p_{31} & p_{32} & p_{33} \\
\end{bmatrix}
$$

```{r transition-matrix}
tran_matrix <- tran_prob |>
  transition_matrix() |>
  normalise() |>
  round(digits = 3)

tran_matrix
```

This code creates a $3 \times 3$ matrix (for the three states: Normal Cognition, MCI, and Dementia) and populates it with the transition probabilities.

## Visualising the matrix

To make the transition matrix more interpretable we visualize the probabilities using a heat map (@fig-transition-matrix)

```{r transition-matrix-plot}
#| fig-width: 8
#| label: fig-transition-matrix
#| fig-cap: Heat map of transition probabilities (2016 - 2022)

plot_matrix(data = t(tran_matrix), year = 2016)
```

We can also create and plot transition matrices for each wave separately allowing us to see how these transition probabilities change across multiple waves of data collection

```{r transition-matrix-longitudinal}
tran_prob_long <- data |>
  extract_years(years = seq(2016, 2022, by = 2)) |>
  create_transitions(absorbing = TRUE) |>
  group_by(Wave) |>
  calculate_probabilties() |>
  group_map(~ transition_matrix(.x, longitudinal = TRUE), .keep = TRUE) |>
  lapply(function(mat) {
    mat[c("Normal Cognition", "MCI", "Dementia"), c("Normal Cognition", "MCI", "Dementia")] |>
        normalise()
    }) |>
  purrr::map2_df(.y = c(seq(2016, 2020, by = 2)), .f = reshape_matrix) |>
  mutate(
    transition_prob = round(transition_prob, digits = 3),
    from = factor(from, levels = c("Normal Cognition", "MCI", "Dementia"))
  )

DT::datatable(
  tran_prob_long,
  filter = "top",
  colnames = c("Probability" = "transition_prob"),
  options = list(pageLength = 6, autoWidth = TRUE)
)
```

```{r transition-matrix-longitudinal-plot, fig.width=8, fig.show="animate", ffmpeg.format="gif", dev="png"}
for(wave in c(seq(2016, 2020, by = 2))){
  print(plot_transition_wave(data = tran_prob_long, wave = wave))
}
```

# Multi-State Markov Model

Now that we have an idea of the disease progression across time we are interested in the effect of different covariates on these transitions. For this, we can fit a multi-state Markov model. Multi-state models (MSMs) provide a comprehensive view of the disease process and make risk-factors analysis and long-term predictions more easily. The conditional distribution of the status of an individual participant at an arbitrary examination given their status at previous examinations was assumed to have the **Markov property**. That is the distribution of the forthcoming state $X_{n+1}$ depends only on the current state $X_n$ and doesn't depend on the previous ones $X_{n-1}, X_{n-2}, \dots, X_1$.

$$Pr(X_{n+1} = x_{n+1} \; \vert \; X_1 = x_1, X_2 = x_2, \dots, X_n = x_n) = Pr(X_{n+1} = x_{n+1} \; \vert \; X_n = x_n)$$

## Data preperation

For our model we are going to look at several different covariates:

- **Gender**: $0 = \text{Male;} \; 1 = \text{Female}$ 
- **Age**: $\text{Range} \; = 50 - 91$ 
- **Education**: $0 = \text{No Degree;} \; 1 = \text{High School Degree;} \; 2 = \text{Further Education}$ 
- **Number of cardiovascular risk factors**: $\text{Range} \; = 0 - 4$
- **Procrastination**: $\text{Range} \; = 4 - 60$

```{r markov-prep}
cols <- c(
  "ID", "Gender", "Age", "Education_tri", 
  paste0("Cardio_risk_", seq(16, 22, by = 2)), 
  paste0("Total_dep_20", seq(16, 22, by = 2)), "Total_p")

markov_data <- data |>
  extract_years(seq(2016, 2022, by = 2)) |>
  na.omit() |>
  inner_join(data[, cols], by = "ID") |>
  pivot_data() |>
  backtrack_age(base_year = 2016, current_year = 2022) |>
  group_by(ID) |>
  mutate(
    Age = min(Age), # Baseline age
    Cardio_risk = Cardio_risk[Wave == 1],
    Depression = Depression[Wave == 1]
    ) |>
  ungroup() |>
  filter(Age >= 50) # Only looking at older adults

DT::datatable(
  markov_data,
  filter = "top",
  options = list(pageLength = 6, autoWidth = TRUE)
)
```

Let's look at a state table to visualize the amount of transitions over time

```{r state-table}
markov_data |>
  select(ID, Wave, Status) |>
  mutate(Status = case_when(
    Status == 1 ~ "Normal Cognition", 
    Status == 2 ~ "MCI", 
    Status == 3 ~ "Dementia")) |>
  msm::statetable.msm(state = Status, subject = ID)
```

## Modelling

### Defining a q-matrix

The Q matrix (also known as the transition intensity matrix or generator matrix) defines the rates at which transitions occur between states in the model.

It is a square matrix where:

- Each row corresponds to a **from** state in the model.
- Each column corresponds to a **to** state in the model.
- The off-diagonal elements $q_{ij}$ represent the transition rate from state $i$ to state $j$.

For a model with $n$ states, the Q matrix has the following structure

$$
Q = \begin{bmatrix} 
q_{11} & q_{12} & \cdots & q_{1n} \\
q_{21} & q_{22} & \cdots & q_{2n} \\
\vdots & \vdots & \ddots &\vdots  \\
q_{n1} & q_{n2} & \cdots & q_{nn}
\end{bmatrix}
$$

```{r q-matrix}
# Creating transition matrix
# Define possible transitions
states <- c("Normal Cognition", "MCI", "Dementia")

# Theoretical transition probabilities
q_matrix <- rbind(
  c(0.75, 0.20, 0.05),  # From Normal, can transition to MCI or Dementia
  c(0.25, 0.50, 0.25),  # From MCI, can transition to Normal or Dementia
  c(0.00, 0.00, 1.00)         # From Dementia, no transitions (absorbing state)
)

rownames(q_matrix) <- states
colnames(q_matrix) <- states

q_matrix
```

## Fitting

We will fit three different multi-state models

1. A model with no covariates
2. A multivariate model with covariates acting on all transitions
3. A multivariate model with covariates (and interactions) acting on all transitions
4. A multivariate model with covariates acting on only NC <-> MCI & MCI - Dementia
5. A multivariate model with covariates (and interactions) acting on only NC <-> MCI & MCI - Dementia

```{r model-fitting}
# Fitting model with no covariates
model_1 <- msm::msm(
  formula = Status ~ Wave,
  subject = ID,
  data = markov_data,
  qmatrix = q_matrix,
  obstype = 1,
  method = "BFGS", 
  gen.inits = TRUE,
  control = list(
    fnscale = 5000, 
    maxit = 2000 # To reach convergence
  ))

# Multivariate model: covariates acting on all transitions
model_2 <- msm::msm(
  formula = Status ~ Wave,
  subject = ID,
  data = markov_data,
  qmatrix = q_matrix,
  covariates = ~ Gender + Age + Education_tri + Cardio_risk + Depression + Total_p,
  obstype = 1,
  gen.inits = TRUE,
  method = "BFGS",
  control = list(
    fnscale = 5000, 
    maxit = 2000 # To reach convergence
  ))

# Multivariate model: covariates acting on all transitions (with interaction)
model_3 <- msm::msm(
  formula = Status ~ Wave,
  subject = ID,
  data = markov_data,
  qmatrix = q_matrix,
  covariates = ~ Gender + Age + Education_tri + Cardio_risk + (Depression * Total_p),
  obstype = 1,
  gen.inits = TRUE,
  method = "BFGS",
  control = list(
    fnscale = 5000, 
    maxit = 2000 # To reach convergence
  ))

# Multivariate model: covariates acting on NC <-> MCI and MCI -> D
model_4 <- msm::msm(
  formula = Status ~ Wave,
  subject = ID,
  data = markov_data,
  qmatrix = q_matrix,
  covariates = list(
    "1-2" = ~ Gender + Age + Education_tri + Cardio_risk + Depression + Total_p,
    "2-1" = ~ Gender + Age + Education_tri + Cardio_risk + Depression + Total_p,
    "2-3" = ~ Gender + Age + Education_tri + Cardio_risk + Depression + Total_p
  ),
  obstype = 1,
  gen.inits = TRUE,
  method = "BFGS",
  control = list(
    fnscale = 5000, 
    maxit = 2000 # To reach convergence
  ))

# Multivariate model: covariates acting on NC <-> MCI and MCI -> D (with interaction)
model_5 <- msm::msm(
  formula = Status ~ Wave,
  subject = ID,
  data = markov_data,
  qmatrix = q_matrix,
  covariates = list(
    "1-2" = ~ Gender + Age + Education_tri + Cardio_risk + (Depression * Total_p),
    "2-1" = ~ Gender + Age + Education_tri + Cardio_risk + (Depression * Total_p),
    "2-3" = ~ Gender + Age + Education_tri + Cardio_risk + (Depression * Total_p)
  ),
  obstype = 1,
  gen.inits = TRUE,
  method = "BFGS",
  control = list(
    fnscale = 5000, 
    maxit = 20002 # To reach convergence
  ))

```

### Model Checking

After fitting each of our models we can extract some goodness of fit criteria

```{r model-goodness}
log_l <- c(logLik(model_1), 
           logLik(model_2), 
           logLik(model_3), 
           logLik(model_4), 
           logLik(model_5))

aic <- AIC(model_1, model_2, model_3, model_4, model_5)

# Very small difference between model 2 and 3 (makes sense)
cbind(log_l, aic) |>
  tibble::rownames_to_column("Model") |>
  mutate(Model = stringr::str_remove(Model, "model_"))

# Likelihood ratio test
lmtest::lrtest(model_1, model_2, model_3, model_4, model_5)
```

We can also look at a measure of how well the model fits the data using a test proposed by [@gentleman1994]. The function plots the observed numbers of individuals occupying a state at a series of times against forecasts from the fitted model, for each state (@fig-gentleman).

```{r gentleman_test}
#| fig-width: 12
#| fig-height: 12
#| label: fig-gentleman
#| fig-cap: Fitted values versus observed values

# Plotting fitted vs. observed values
g1 <- gentleman_test(data = msm::prevalence.msm(model_1), 
                     model = "Model 1")
g2 <- gentleman_test(data = msm::prevalence.msm(model_2), 
                     model = "Model 2")
g3 <- gentleman_test(data = msm::prevalence.msm(model_3), 
                     model = "Model 3")
g4 <- gentleman_test(data = msm::prevalence.msm(model_4), 
                     model = "Model 4")
g5 <- gentleman_test(data = msm::prevalence.msm(model_5), 
                     model = "Model 5")

ggpubr::ggarrange(g1, g2, g3, g4, g5, 
                  common.legend = TRUE, 
                  nrow = 5,
                  heights = c(2, 2, 2, 2, 2), 
                  legend = "bottom")
```

### Hazard ratios

We now present hazard ratios for each of our fitted models. These ratios represent how the instantaneous risk of making a particular transition is modified by the covariate.

::: {.panel-tabset}

## Model 2

@fig-hazard-ratio-1 presents hazard ratios from model 2 (our multivariate model with covariates acting on all transitions) for each covariate.

```{r}
#| fig-width: 12
#| fig-height: 12
#| label: fig-hazard-ratio-1
#| fig-cap: Ouputted hazard ratios from Markov model. Green bars represent a decreased risk of transition while red bars indicate an increased risk of transition.

# Plotting hazard ratios
model_2 |>
  msm::hazard.msm() |>
  process_hr() |>
  calculate_p_value() |>
  plot_hr()
```

## Model 3

@fig-hazard-ratio-2 presents hazard ratios from model 3 (our multivariate model with covariates **(and interactions)** acting on all transitions) for each covariate.

```{r}
#| fig-width: 12
#| fig-height: 12
#| label: fig-hazard-ratio-2
#| fig-cap: Ouputted hazard ratios from Markov model. Green bars represent a decreased risk of transition while red bars indicate an increased risk of transition.

# Plotting hazard ratios
model_3 |>
  msm::hazard.msm() |>
  process_hr() |>
  calculate_p_value() |>
  plot_hr()
```

## Model 4

@fig-hazard-ratio-3 presents hazard ratios for model 4 (our multivariate model with covariates acting on only NC <-> MCI & MCI -> Dementia) for each covariate

```{r}
#| fig-width: 12
#| fig-height: 12
#| label: fig-hazard-ratio-3
#| fig-cap: Ouputted hazard ratios from Markov model. Green bars represent a decreased risk of transition while red bars indicate an increased risk of transition.

# Plotting hazard ratios
model_4 |>
  msm::hazard.msm() |>
  process_hr() |>
  calculate_p_value() |>
  plot_hr()
```

## Model 5

@fig-hazard-ratio-4 presents hazard ratios for model 5 (our multivariate model with covariates **(and interactions)** acting on only NC <-> MCI & MCI -> Dementia) for each covariate

```{r}
#| fig-width: 12
#| fig-height: 12
#| label: fig-hazard-ratio-4
#| fig-cap: Ouputted hazard ratios from Markov model. Green bars represent a decreased risk of transition while red bars indicate an increased risk of transition.

# Plotting hazard ratios
model_5 |>
  msm::hazard.msm() |>
  process_hr() |>
  calculate_p_value() |>
  plot_hr()
```

:::

### 2025-03-28 reports/discrete_model.qmd
Initial commit :rocket:

---
title: "Discrete Time Markov Model Analysis of Cognitive Decline"
subtitle: "Examining the Association Between Procrastination and Cognitive Transitions"
author: 
  - name: "Cormac Monaghan"
    url: "https://c-monaghan.github.io/"
    orcid: "https://orcid.org/0000-0001-9012-3060"
    affiliation:
      - "Hamilton Institute, Maynooth University, Ireland"
      - "Department of Psychology, Maynooth University, Ireland"
abstract: |
  This analysis examines transitions between cognitive states (Normal Cognition, MCI, Dementia) using discrete-time Markov models, with particular focus on procrastination as a risk factor.
toc: true
toc-depth: 3
toc-title: On this page
number-sections: true
format: 
  html:
    theme: 
      light: cosmo
    page-layout: full
    shift-heading-level-by: 1
    html-math-method: katex
    code-tools: true
    code-copy: true
    code-block-bg: "#333A3F"
    highlight-style: monokai
    fig-align: center
    self-contained: true
---

# Setting up {.unnumbered}

## Loading packages {.unnumbered}

```{r packages}
#| code-fold: true
#| code-summary: "Check out my packages"
#| warning: false

# Packages ---------------------------------------------------------------------
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
```

## Loading functions {.unnumbered}

```{r functions}
#| code-fold: true
#| code-summary: "Check out my functions"
#| warning: false

# Functions --------------------------------------------------------------------
extract_years <- function(data, years, impute = TRUE, cog_total = FALSE, absorbing = TRUE) {
  # Extracts cognitive function data for specified years from a dataset.
  # Converts numeric cognitive status codes (1, 2, 3) into descriptive labels
  # ("Normal Cognition", "MCI", "Dementia") for easier interpretation.
  # Arguments:
  #   - data: The input dataset containing cognitive function data.
  #   - years: A vector of years for which data should be extracted.
  # Returns:
  #   - A dataset with ID and cognitive status columns for the specified years.

  # Create dynamic column names based on the years provided
  cogfunction_cols <- paste0("cogfunction", years)
  cogtotal_cols    <- paste0("cogtot27_imp", years)

  if(cog_total == FALSE) {
    data <- data |>
      # Select only the ID column and cognitive function columns for the specified years
      select(ID, any_of(cogfunction_cols)) |>
      mutate(across(!c(ID), ~ case_when(
        .x == 1 ~ "Normal Cognition",
        .x == 2 ~ "MCI",
        .x == 3 ~ "Dementia",
        TRUE ~ NA_character_  # To handle missing/other cases
      )))
  } else {
    data <- data |>
      # Select only the ID column and cognitive function columns for the specified years
      select(ID, any_of(cogtotal_cols)) |>
      rename_with( .cols = !ID, .fn = ~ stringr::str_replace(
        string = .x,
        pattern = "cogtot27_imp", 
        replacement = "cog_score_"))
  }

  if(impute == TRUE){
    data <- data |>
      tidyr::pivot_longer(
        cols = !ID,
        names_to = "Wave",
        values_to = "Status") |>
      group_by(ID) |>
      tidyr::fill(Status, .direction = "down") |>
      ungroup() |>
      tidyr::pivot_wider(names_from = "Wave", values_from = "Status")
  }

  if(absorbing == TRUE) {
    data <- data |>
      tidyr::pivot_longer(
        cols = !ID,
        names_to = "Wave",
        values_to = "Status") |>
      group_by(ID) |>
      mutate(Status = ifelse(cumany(Status == "Dementia"), "Dementia", Status)) |>
      ungroup() |>
      tidyr::pivot_wider(names_from = "Wave", values_from = "Status")
  }

  return(data)
}

pivot_and_factorise <- function(data) {
  # Converts cognitive function data from wide to long format and factorizes key variables.
  # Pivots multiple cogfunction columns into wave/status pairs and converts categorical
  # variables (Gender, Education, status) to factors with meaningful labels.
  # Arguments:
  #   - data: Dataset containing cognitive function variables in wide format
  # Returns:
  #   - Long-format dataset with factorized variables, ordered by ID and wave
  # 
  data |>
    tidyr::pivot_longer(
      cols = starts_with("cogfunction"), names_to = "wave",
      names_prefix = "cogfunction", values_to = "status") |>
    mutate(
      Gender = factor(Gender, levels = c(0, 1)),
      Education_tri = factor(Education_tri, levels = c(0, 1, 2)),
      wave = factor(wave),
      status = factor(status, 
                      levels = c("Normal Cognition", "MCI", "Dementia"),
                      labels = c(1, 2, 3))) |>
    relocate(wave, .after = ID) |>
    relocate(status, .after = wave)
}

create_transition_table <- function(year_to, year_from) {
  # Creates a transition frequency table between two specified years of cognitive status data.
  # Calculates row sums and total observations for transition analysis.
  # Arguments:
  #   - year_to: Target year for transitions (character or numeric)
  #   - year_from: Origin year for transitions (character or numeric)
  # Returns:
  #   - List containing: 
  #     - transition frequency table
  #     - row sums
  #     - total observations
  # 
  tbl <- table(
    table_data[[paste0("HRS_", year_to)]],
    table_data[[paste0("HRS_", year_from)]],
    dnn = c(year_to, year_from)
  )

  # Calculate row sums and total
  row_sums <- rowSums(tbl)
  total <- sum(row_sums)

  # Return as a list with both the table and summary stats
  list(
    table = tbl,
    row_sums = row_sums,
    total = total
  )
}

create_transition_dataset <- function(data, transition_results) {
  # Combines multiple transition tables into a single analysis-ready dataset.
  # Formats period labels, ensures consistent factor levels, and structures data for visualization.
  # Arguments:
  #   - data: List of year pairs to process (e.g., list(c(2016,2018)))
  #   - transition_results: List containing transition tables from create_transition_table()
  # Returns:
  #   - Tidy dataset with transition frequencies between all specified periods
  # 
  data |>
    purrr::map_dfr(~ {
      period_name <- paste(.x[2], .x[1], sep = " - ")
      tbl <- transition_results[[paste(.x[2], .x[1], sep = "-")]]$table

      as.data.frame(tbl) |>
        rename(t_minus_1 = 1, t = 2) %>%  # Positional renaming
        mutate(Period = period_name, .before = t_minus_1)
    }) |>
    mutate(
      Period = stringr::str_replace(Period, "(\\d+) - (\\d+)", "\\2 - \\1"),
      Period = factor(Period, levels = c("2016 - 2018", "2018 - 2020", "2020 - 2022")),
      t_minus_1 = factor(t_minus_1, levels = c("Normal Cognition", "MCI", "Dementia")),
      t = factor(t, levels = c("Normal Cognition", "MCI", "Dementia"))
    )
}

create_transitions <- function(data, absorbing = FALSE){
  # Reshapes data from wide to long format to track cognitive status transitions over time.
  # Calculates the next wave's cognitive status for each individual and creates a transition column.
  # Optionally treats "Dementia" as an absorbing state, meaning once an individual is classified
  # with dementia, their status cannot change in subsequent waves.
  # Arguments:
  #   - data: The dataset containing cognitive status data.
  #   - absorbing: A logical flag indicating whether "Dementia" should be treated 
  #   as an absorbing state.
  # Returns:
  #   - A dataset with transition information, including current and next wave statuses.

  # Reshape the data from wide to long format to track cognitive status over waves
  data <- data |>
    select(ID, starts_with("cogfunction")) |>
    tidyr::pivot_longer(cols = !ID,
                        names_to = "Wave",
                        values_to = "Status") |>
    mutate(Wave = as.factor(stringr::str_replace(Wave, "cogfunction", ""))) |>
    # Arrange by ID and Wave to prepare for transition calculation
    arrange(ID, Wave) |>
    group_by(ID) |>
    # Get the next wave's cognitive status for each person
    mutate(next_wave_status = lead(Status)) |>
    ungroup()

  # We can optionally specify dementia as an absorbing state
  # Once an individual is classified with dementia they cannot be classified
  # with anything else 
  if(absorbing == TRUE) {
    data <- data |>
      group_by(ID) |>
      mutate(
        Status = ifelse(cumany(Status == "Dementia"), "Dementia", Status),
        next_wave_status = ifelse(cumany(Status == "Dementia"), "Dementia", next_wave_status),
        transition = paste(Status, next_wave_status, sep = " to ")
      ) |>
      filter(Wave %in% c(2016, 2018, 2020))
  }

  # Filter out rows where either the current or next status is missing
  data <- data |>
    group_by(ID) |>
    filter(!is.na(Status), !is.na(next_wave_status)) |>
    ungroup() |>
    # Create a new column representing the transition from one status to the next
    mutate(
      transition = paste(Status, next_wave_status, sep = " to "),
      transition = factor(
        transition, 
        levels = c("Normal Cognition to Normal Cognition", "Normal Cognition to MCI",
                   "Normal Cognition to Dementia", "MCI to Normal Cognition", "MCI to MCI",
                   "MCI to Dementia", "Dementia to Dementia") 
      ))

  return(data)
}

observed_transition_matrix <- function(data) {
  # Converts transition data into a properly formatted probability transition matrix.
  # Ensures consistent state ordering and converts proportions to matrix format suitable
  # for multi-state modeling and visualization.
  # Arguments:
  #   - data: Transition dataset from create_transitions()
  # Returns:
  #   - Square transition probability matrix with states as row/column names
  # 
  # The order I want my matrix in
  state_order <- c("Normal Cognition", "MCI", "Dementia")

  data |>
    group_by(Status, next_wave_status) |>
    summarise(freq = n(),  .groups = "drop") |>
    group_by(Status) |>
    mutate(freq = round(proportions(freq), 3)) |>
    ungroup() |>
    complete(Status, next_wave_status, fill = list(freq = 0)) |>
    tidyr::pivot_wider(names_from = next_wave_status, values_from = freq, names_sort = TRUE) |>
    tibble::column_to_rownames("Status") |>
    select(all_of(state_order)) |>
    _[state_order, ] |>
    as.matrix()
}

normalise <- function(x) {
  # Normalizes a matrix so that each row sums to 1.
  # Arguments:
  #   - x: The input matrix to normalize.
  # Returns:
  #   - A normalized matrix where each row sums to 1.
  x / rowSums(x)
}

tran_stack_graph <- function(data) {
  # Creates stacked bar charts visualizing cognitive state transitions across time periods.
  # Shows composition of current states by previous state, faceted by observation period.
  # Arguments:
  #   - data: Transition dataset from create_transition_dataset()
  # Returns:
  #   - ggplot object showing stacked transition proportions
  # 
  data |>
    ggplot(aes(x = t_minus_1, y = Freq, fill = t)) +
    geom_col(position = "stack", colour = "black") +
    facet_wrap(~ Period, ncol = 3) +
    labs(
      x = "Previous State (t-1)", 
      y = "Count",
      title = "Outcomes by Prior Cognitive State",
      fill = "Current State (t)"
    ) +
    ggokabeito::scale_fill_okabe_ito() +
    ggeasy::easy_move_legend("bottom")
}

tran_heat_map <- function(data) {
  # Generates heatmap visualization of transition frequencies between cognitive states.
  # Uses color intensity and labeled values to show transition patterns across time periods.
  # Arguments:
  #   - data: Transition dataset from create_transition_dataset()
  # Returns:
  #   - ggplot heatmap with state transitions as cells
  # 
  data |>
    ggplot(aes(x = t_minus_1, y = t, fill = Freq)) +
    geom_tile(color = "white") +
    geom_text(aes(label = Freq), color = "black", size = 3.5) +  # Add counts
    scale_fill_gradient(low = "white", high = "steelblue") +
    facet_wrap(~ Period, ncol = 3) +  # Split by time period
    labs(
      x = "Previous State (t-1)", 
      y = "Current State (t)",
      title = "Cognitive State Transitions Between Time Periods",
      fill = "Frequency"
    ) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    ggeasy::easy_move_legend("bottom")
}

tidy_predictions <- function(predictions) {
  # Restructures model prediction matrices into tidy format for visualization.
  # Converts numeric codes to factor labels and reshapes multiple prediction columns
  # into key-value pairs suitable for ggplot.
  # Arguments:
  #   - predictions: Raw prediction matrix from model output
  # Returns:
  #   - Long-format dataset with probabilities for each cognitive state
  #   
  predictions |>
    as.matrix() |>
    as_tibble() |>
    mutate(
      Gender = factor(ifelse(Gender == 0, "Male", "Female"), levels = c("Male", "Female")),
      status_prev = case_when(
        status_prev == 1 ~ "Normal Cognition",
        status_prev == 2 ~ "MCI",
        status_prev == 3 ~ "Dementia",
      ),
      status_prev = factor(status_prev, levels = c("Normal Cognition", "MCI", "Dementia")),
      Total_p = as.numeric(Total_p),
      across(c(pred.1:pred.3), as.numeric)) |>
    tidyr::pivot_longer(cols = c(pred.1:pred.3), names_to = "status", values_to = "prob") |>
    mutate(status = case_when(
      status == "pred.1" ~ "Normal Cognition",
      status == "pred.2" ~ "MCI",
      status == "pred.3" ~ "Dementia"
    ),
    status = factor(status, levels = c("Normal Cognition", "MCI", "Dementia")),
    status_prev = factor(status_prev, levels = c("Normal Cognition", "MCI", "Dementia")))
}

plot_predictions_stationary <- function(predictions) {
  # Visualizes predicted transition probabilities from stationary multi-state models.
  # Shows probability curves by procrastination level, stratified by baseline state.
  # Arguments:
  #   - predictions: Tidy predictions from tidy_predictions()
  # Returns:
  #   - Faceted ggplot showing predicted probability curves
  # 
  predictions |>
    ggplot(aes(x = Total_p, y = prob, colour = status_prev)) +
    geom_line(linewidth = 1) +
    ggokabeito::scale_colour_okabe_ito() +
    labs(
      title = "Predicted transition probabilities (stationary model)",
      # subtitle = "According to previous cognitive state",
      x = "Total Procrastination", y = "Probability", colour = "Previous State") +
    facet_wrap(~ status, labeller = labeller(
      status = function(x){paste0("Transition to: ", x)}
    ))
}

get_time_varying_matrix <- function(model, time_point) {
  expand.grid(
    Gender = factor(0),
    Total_p = mean(data_stack$Total_p),
    status_prev = factor(1:3),
    wave = time_point) |> 
    modelr::add_predictions(model, var = "prob", type = "probs") |> 
    select(status_prev, starts_with("prob")) |> 
    mutate(
      status_prev = factor(status_prev, labels = state_names),
      across(starts_with("prob"), ~round(., 3))) |>
    tibble::column_to_rownames("status_prev") |>
    as.matrix() |> `colnames<-`(state_names)
}
```

## Setting theme {.unnumbered}

```{r set-theme}
#| code-fold: true
#| code-summary: "Check out my theme"

colour <- "#212427"

theme_set(
  theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14, colour = colour, face = "bold"),
      plot.subtitle = element_text(hjust = 0.5, size = 12, colour = colour),
      axis.title = element_text(size = 10, colour = colour, face = "bold"),
      strip.text = element_text(size = 10, colour = colour, face = "bold"),
      legend.title = element_text(hjust = 0.5, colour = colour, face = "bold"),
      ))
```

## Reading in data {.unnumbered}

```{r data}
data <- readxl::read_xlsx(here::here("data/data.xlsx"))
```

# Research Questions and Hypotheses

## Primary Hypothesis

We hypothesize that:

- **H1:** Higher levels of procrastination are associated with:
  - Increased risk of transitioning from Normal Cognition to MCI $(\beta = x, p < .05)$
  - Increased risk of transitioning from MCI to Dementia $(\beta = y, p < .05)$

- **H0:** There is no association between procrastination and cognitive decline transitions

In order to test this hypothesis we will be using a **discrete time Markov Model**

# Data Preparation

Before preparing the data, we want to create a vector of covariates that we are interested in using in our analysis

```{r covariates}
cols <- c("ID", "Gender", "Age", "Education_tri", "Cardio_risk_16", "Total_dep_2016", "Total_p")
```

Following this, we can now use the `extract_years()` function to extract the cognitive function data for the years 2016, 2018, and 2020.

- We will also impute some missing values `impute = TRUE` using logical reasoning (if a respondent has an NA value in 2018, but has a classification of "normal cognition" in 2020, then the missing 2018 value becomes "normal cognition").
- We will also treat dementia as an absorbing state `absorbing = TRUE`.

```{r extract-years}
data_stack <- data |> extract_years(years = seq(2016, 2022, by = 2), impute = TRUE, absorbing = TRUE) |> na.omit()

head(data_stack)
```

## Creating a stacked dataset

For each respondent we will now add in their relevant covariate data. Following this, we transform the data to long format and convert categorical variables to factors using the `pivot_and_factorise()` function. Additionally, we will fix the `Age` column to properly represent the age of the respondent at each time point

```{r long_data}
data_stack <- data_stack |>
  inner_join(data[, cols], by = "ID") |>
  pivot_and_factorise() |>
  group_by(ID) |>
  mutate(Age = Age - (2022 - as.numeric(as.character(wave))))

head(data_stack)
```

Finally, we will create a new variable `status_prev` that notes the cognitive status of the respondent in the previous wave $(t - 1)$. This will be done by using the `lag` function from the `dplyr` package.

```{r stack-data}
data_stack <- data_stack |>
  mutate(status_prev = lag(status), .after = status) |>
  ungroup() |> 
  filter(wave != 2016, !is.na(Total_p))

head(data_stack)
```

# Transition frequencies

We will now calculate the transition frequencies between cognitive states for each time period. We will use the `create_transition_table()` function to create a transition table for each time period. This will be done with the help of the `map()` function from the `purrr` package. Finally, to combine all the transition tables into one dataset, we will use the `create_transition_dataset()` function.

```{r transition-tables}
# Creating a table dataset
table_data <- data |>
  extract_years(seq(2016, 2022, by = 2)) |>
  rename_with(~ gsub("cogfunction", "HRS_", .)) |>
  mutate(
    across(c(HRS_2016:HRS_2022), ~ factor(.x, levels = c("Normal Cognition", "MCI", "Dementia")))
  )

# Creating transition frequencies ---------------------------------------------
## These are the time periods we are interested in
time_periods <- list(
  c("2016", "2018"),
  c("2018", "2020"), 
  c("2020", "2022")
)

# Applying function
transition_results <- purrr::map(time_periods, ~ create_transition_table(.x[1], .x[2]))
names(transition_results) <- purrr::map_chr(time_periods, ~ paste(.x[2], .x[1], sep = "-"))

### Creating one dataset
transition_frequencies <- time_periods |>
  create_transition_dataset(transition_results = transition_results)

head(transition_frequencies)
```

## Observed transition matrix

The probability distribution of transitions from one state to another can be represented into a transition matrix $P = (p_{ij})_{i,j}$ where each element of position $(i, j)$ represents the transition probability $p_{ij}$.

In order to create this matrix we will use both the `create_transitions()` and `observed_transition_matrix()` functions.

```{r observed-matrix}
## Creating observed transition matrix -----------------------------------------
transition_matrix_observed <- data |>
  extract_years(seq(2016, 2022, by = 2)) |>
  create_transitions() |>
  observed_transition_matrix()

transition_matrix_observed
```

## Visualisation

Let's visualize both the transition frequencies (@fig-frequencies) and matrix (@fig-matrix).

```{r visualise-transition}
#| code-fold: true
#| code-summary: "Check out my code"
#| fig-width: 12
#| fig-height: 10
#| label: fig-frequencies
#| fig-cap: "Transition frequencies between cognitive states for each time period."

fig_1 <- transition_frequencies |> tran_stack_graph()
fig_2 <- transition_frequencies |> tran_heat_map()

fig_1 / fig_2
```

```{r visualise-matrix}
#| code-fold: true
#| code-summary: "Check out my code"
#| fig-width: 12
#| label: fig-matrix
#| fig-cap: "Observed transition matrix between cognitive states for the years 2016 - 2022."

tran_df <- transition_matrix_observed |>
  as.data.frame() |>
  mutate(
    from_state = factor(c("Normal cognition", "MCI", "Dementia"),
                        levels = c("Normal cognition", "MCI", "Dementia"))
    ) |>
  reshape2::melt(
    id.vars = "from_state", variable.name = "to_state", value.name = "probability") |>
  mutate(
    to_state = factor(to_state, levels = rev(levels(to_state)))
    )

fig_2 <- tran_df |>
  ggplot(aes(x = from_state, y = to_state, fill = probability)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(
    aes(label = format(round(probability, 3), nsmall = 3)),
    size = 4.5, 
    color = "#212427",
    fontface = "bold") +
  colorspace::scale_fill_continuous_diverging(
    palette = "Blue-Red 3", mid = 0.50, alpha = 0.5, 
    limits = c(0, 1), name = "Transition \nProbability") +
  labs(title = "Transition Probabilities Across Cognitive States",
       subtitle = "Observed state transitions between assessment waves",
       x = "Previous State (t - 1)", 
       y = "Current State (t)") +
  theme(
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    legend.position = "right",
    legend.text = element_text(size = 9),
    panel.grid = element_blank()
  )

fig_2
```

# Modelling

## Markov Process Fundamentals

Discrete-time Markov models belong to a class of stochastic processes that satisfy the **Markov property**, which can be formally expressed as:

$$
P(X_{t+1} = j \vert X_t = i, X_{t-1} = i_{t-1}, \dots X_0 = i_0) = P(X_{t+1} = j \vert X_t = i)
$$ {#eq-markov}

This property establishes that the future state $X_{t+1}$ depends only on the current state $X_t$, not on the entire history of states.

### Transition Probability Matrix

For our three-state system (Normal Cognition, Mild Cognitive Impairment [MCI], Dementia), the transition matrix $P$ captures all possible transition probabilities:

$$
P = \begin{bmatrix} 
p_{11} & p_{12} & p_{13} \\
p_{21} & p_{22} & p_{23} \\
p_{31} & p_{32} & p_{33} \\
\end{bmatrix}
$$ {#eq-t-matrix}

where:

- $p_{ij} = P(X_{t+1} = j \vert X_t = i)$ represents the probability of transitioning from state $i$ to state $j$.
- Each row sums to 1, $\sum^3_{j=1} p_{ij} = 1 \quad  \forall_i \in \{1, 2, 3\}$

### Multinomial Logistic Regression Formulation

We model the transition probabilities using **multinomial logistic regression**, where the log-odds of each transition relative to a reference state are linear functions of covariates.

For a system with $K$ states (using state $K$ as reference), we have:

$$
log \left( \frac{P(Y = j \vert x)}{P(Y = k \vert x)} \right) = \beta_{j0} + \beta_j^Tx \qquad \text{for } j = 1, \dots K-1
$$ {#eq-multinomial}

For non-reference states $j = 1, \dots, K - 1$

$$
P(Y = j \vert x) = \frac{e^{\beta_{0j} + \beta_j^Tx}}{1 + \sum^{k - 1}_{k = 1} e^{\beta_{0k} + \beta_k^Tx}}
$$ {#eq-non-reference}

For the reference state $K$:

$$
P(Y = k \vert x) = \frac{1}{1 + \sum^{k - 1}_{k = 1} e^{\beta_{0k} + \beta_j^Tx }}
$$ {#eq-reference}

### Time-Homogeneous Approach

Our implementation assumes time-homogeneous transitions, but the framework can be extended to time-varying probabilities:

$$
P^{(t)} = \begin{bmatrix} 
p_{11}^{(t)} & p_{12}^{(t)} & \cdots & p_{1n}^{(t)} \\
p_{21}^{(t)} & p_{22}^{(t)} & \cdots & p_{2n}^{(t)} \\
\vdots & \vdots & \ddots & \vdots \\
p_{n1}^{(t)} & \cdots & \cdots & p_{nn}^{(t)} \\
\end{bmatrix}
$$ {#eq-non-stationary}

Our primary analysis assumes the transition matrix remains fixed:

$$
P^{(t)} \equiv P \qquad \forall \; t
$$ {#eq-assumption}

### Absorbing State Specification

We model Dementia as an absorbing state:

$$
p_{3j} = \begin{cases} 
1 \qquad \text{if } j = 3 \\
0 \qquad \text{otherwise} \end{cases}
$$ {#eq-dementia}

yielding the constrained transition matrix:

$$
P = \begin{bmatrix} 
p_{11} & p_{12} & p_{13} \\
p_{21} & p_{22} & p_{23} \\
0      & 0      & 1      \\
\end{bmatrix}
$$ {#eq-t-matrix-constrained}

## Stationary model

We estimate three progressively complex stationary models using `nnet::multinom()`

```{r model-fitting}
#| code-fold: true
#| code-summary: "Check out my models"
#| 
# Baseline model (only gender)
fit_1 <- nnet::multinom(
  status ~ Gender, family = multinomial, 
  data = data_stack, trace = FALSE)

# With procrastination
fit_2 <- nnet::multinom(
  status ~ Gender + Total_p, family = multinomial, 
  data = data_stack, trace = FALSE)

# Full model with previous state
fit_3 <- nnet::multinom(
  status ~ Gender + Total_p + status_prev, family = multinomial, 
  data = data_stack, trace = FALSE)
```

### Model comparison

We evaluate model improvement using likelihood ratio tests:

$$
D = -2 \times \ell_{\text{reduced}} - \ell_{\text{full}}
$$ {#eq-liklihood}

```{r model-comparison}
anova(fit_1, fit_2, fit_3) |>
  knitr::kable(caption = "Nested Model Comparison") |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE)
```

### Estimated parameters

```{r model-output}
#| code-fold: true
#| code-summary: "Check out my code"
#| 
broom::tidy(fit_3) |>
  filter(term != "(Intercept)") |>
  mutate(
    y.level = case_when(
      y.level == "2" ~ "NC - MCI",
      y.level == "3" ~ "NC - Dementia"),
    term = case_when(
      term == "Gender1" ~ "Being female",
      term == "Total_p" ~ "Procrastination",
      term == "status_prev2" ~ "Previous state: MCI",
      term == "status_prev3" ~ "Previous state: Dementia",
      TRUE ~ term)
    ) |>
  mutate(across(c(estimate:p.value), ~ round(x = ., digits = 3)))
```

### Model visualisation

Let's visualize the predicted transition probabilities from the stationary models. We will show probability curves by procrastination level

```{r}
#| code-fold: true
#| code-summary: "Check out my code"
#| fig-width: 12
#| label: fig-model-predictions
#| fig-cap: "Predicted transition probabilities from the stationary model."
#| 
## Making a prediction dataset -------------------------------------------------
pred_data <- expand.grid(
  Gender = factor(0),
  status_prev = levels(data_stack$status_prev),
  Total_p = seq(0, 60, length = 200))

## Plotting predictions
fig_3 <- pred_data |>
  modelr::add_predictions(model = fit_3, var = "pred", type = "probs") |>
  tidy_predictions() |>
  plot_predictions_stationary()

fig_3
```

### Estimated transition matrix

Finally, we can compare the observed and predicted transition matrices. We will use the estimated probabilities from the model to fill in the transition matrix.

```{r estimated-matrix}
# Get all unique states
names <- c("Normal Cognition", "MCI", "Dementia")
states <- sort(unique(data_stack$status))
n_states <- length(states)

# Create empty transition matrix
transition_matrix_estimated <- matrix(
  0, nrow = n_states, ncol = n_states,
  dimnames = list(paste("From", names), paste("To", names))
  )

# Getting estimated probabilities
estimated_probs <- expand.grid(
  Gender = factor(0),
  Total_p = mean(data_stack$Total_p),
  status_prev = states) |>
  modelr::add_predictions(model = fit_3, var = "pred", type = "probs")

# Filling in matrix
for(i in 1:n_states) {
  transition_matrix_estimated[i, ] <- estimated_probs$pred[i, ]
}

transition_matrix_estimated |> round(digits = 3)
```

Let's plot this and then compare with our observed matrix

```{r compare-matrix}
#| code-fold: true
#| code-summary: "Check out my code"
#| fig-width: 12
#| label: fig-matrix-comparison
#| fig-cap: "Comparison of observed and estimated transition matrices."

tran_df_estimate <- transition_matrix_estimated |>
  as.data.frame() |>
  mutate(from_state = factor(
    c("Normal cognition", "MCI", "Dementia"),
    levels = c("Normal cognition", "MCI", "Dementia")
  )) |>
  reshape2::melt(id.vars = "from_state",
                 variable.name = "to_state",
                 value.name = "probability") |>
  mutate(to_state = factor(to_state, levels = rev(levels(to_state))))

fig_4 <- tran_df_estimate |>
  ggplot(aes(x = from_state, y = to_state, fill = probability)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(
    aes(label = format(round(probability, 3), nsmall = 3)),
    size = 4.5, 
    color = "#212427",
    fontface = "bold") +
  colorspace::scale_fill_continuous_diverging(
    palette = "Blue-Red 3", mid = 0.50, alpha = 0.5, 
    limits = c(0, 1), name = "Transition \nProbability") +
  labs(title = "Transition Probabilities Across Cognitive States",
       subtitle = "Estimated state transitions between assessment waves",
       x = "Previous State (t - 1)", 
       y = "Current State (t)") +
  theme(
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    legend.position = "right",
    legend.text = element_text(size = 9),
    panel.grid = element_blank()
  )

fig_2 + fig_4 + plot_layout(axis_titles = "collect", guides = "collect")
```

## Non-stationary model

We can also estimate a non-stationary model (@eq-non-stationary) by incorporating time `(wave)` into our model. This model allows for time-varying transition probabilities $p^{(t)}_{ij}$.

We estimate three progressively complex non-stationary models:

- Additive time effects `(fit_4)`: Baseline covariates + wave
- State-specific time effects `(fit_5)`: Interaction between previous state and wave
- Full time interactions `(fit_6)`: All covariates interacting with wave

```{r non-stationary}
#| code-fold: true
#| code-summary: "Check out my models"

## Small processing to fix wave column
data_stack <- data_stack |>
  mutate(wave = case_when(
    wave == "2016" ~ 1,
    wave == "2018" ~ 2,
    wave == "2020" ~ 3,
    wave == "2022" ~ 4
  ))

# Model 4: Additive time effects
fit_4 <- nnet::multinom(
  status ~ Gender + Total_p + status_prev + wave, family = multinomial, 
  data = data_stack, trace = FALSE)

# Model 5: State-specific time effects
fit_5 <- nnet::multinom(
  status ~ Gender + Total_p + (status_prev * wave), family = multinomial, 
  data = data_stack, trace = FALSE)

# Model 6: Full time interactions
fit_6 <- nnet::multinom(
  status ~ (Gender + Total_p + status_prev) * wave, family = multinomial, 
  data = data_stack, trace = FALSE)
```

### Model comparison

Again, we will evaluate model improvement using likelihood ratio tests (@eq-liklihood). However, this time, we will also compare the best fitted non-stationary models `fit_3`.

```{r}
#| code-fold: true
#| code-summary: "Check out my code"
#| layout-nrow: 2

# Fit statistics
models <- list(
  "Stationary" = fit_3,
  "Additive Time" = fit_4,
  "State-Time Interaction" = fit_5,
  "Full Interactions" = fit_6
)

# Create comparison table
comparison_table <- purrr::map_dfr(models, broom::glance, .id = "Model") |>
  select(Model:AIC) |>
  mutate(across(where(is.numeric), \(x) round(x, digits = 1)))

# Likelihood ratio tests
lr_tests <- anova(fit_3, fit_4, fit_5, fit_6)

# Outputting tables
knitr::kable(lr_tests, caption = "Likelihood Ratio Tests") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))

knitr::kable(comparison_table, caption = "Model Fit Statistics") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
```

**Looks like adding in time as an additive term improved things**

### Estimated parameters

Key coefficients from the selected non-stationary model `fit_4`

```{r non-stationary-output}
#| code-fold: true
#| code-summary: "Check out my code"

broom::tidy(fit_4) |>
  filter(term != "(Intercept)") |>
  mutate(
    y.level = case_when(
      y.level == "2" ~ "NC - MCI",
      y.level == "3" ~ "NC - Dementia"),
    term = case_when(
      term == "Gender1" ~ "Being female",
      term == "Total_p" ~ "Procrastination",
      term == "status_prev2" ~ "Previous state: MCI",
      term == "status_prev3" ~ "Previous state: Dementia",
      term == "wave" ~ "Time",
      TRUE ~ term)
    ) |>
  mutate(across(c(estimate:p.value), ~ round(x = ., digits = 3)))
```

### Model visualisation

Let's visualize the predicted transition probabilities from the non-stationary model @fig-non-stationary. We will show probability curves by procrastination level

```{r non-stationary-visualise}
#| code-fold: true
#| code-summary: "Check our my code"
#| fig-width: 12
#| fig-height: 10
#| label: fig-non-stationary
#| fig-cap: "Predicted transition probabilities from the non-stationary model."
#| 
expand.grid(
  Gender = factor(0),
  status_prev = levels(data_stack$status_prev),
  Total_p = seq(0, 60, length = 200),
  wave = unique(data_stack$wave)) |>
  modelr::add_predictions(model = fit_4, var = "pred", type = "probs") |>
  tidy_predictions() |>
  mutate(wave = factor(wave)) |>
  ggplot(aes(x = Total_p, y = prob, color = status_prev)) +
  geom_line(linewidth = 1) +
  ggokabeito::scale_color_okabe_ito() +
  labs(
    title = "Time-Varying Transition Probabilities",
    x = "Total Procrastination", 
    y = "Predicted Probability", color = "Previous State") +
  facet_grid(wave ~ status, labeller = labeller(
    wave = function(x) paste("Years since baseline:", x),
    status = function(x) paste("Transition to:", x)
  )) +
  theme(panel.spacing = unit(1, "lines"))

```

### Estimated transition matrices

Let's visualize the estimated time-varying transition matrices @fig-non-stationary-matrix-plot

```{r non-stationary-matrix-visualise}
#| code-fold: true
#| code-summary: "Check out my code"
#| fig-width: 12
#| label: fig-non-stationary-matrix-plot
#| fig-cap: "Estimated time-varying transition matrices."

# Create prediction grid for each time point
time_points <- unique(data_stack$wave)
state_names <- c("Normal Cognition", "MCI", "Dementia")

# Get matrices for all time points
time_varying_matrices <- purrr::map(
  setNames(time_points, paste(time_points, "Years")),
  ~ get_time_varying_matrix(fit_4, .x)
)

# Converting into a tidy data frame
non_stationary_matrices <- purrr::imap_dfr(
  time_varying_matrices,
  ~ as.data.frame(.x) |> 
    tibble::rownames_to_column("From") |> 
    pivot_longer(-From, names_to = "To", values_to = "probability") |> 
    mutate(Time = .y),
  .id = "Time_point"
) |> 
  mutate(
    From = factor(From, levels = state_names),
    To = factor(To, levels = rev(state_names)),
    Time_point = forcats::fct_inorder(Time_point)
  )

# Plotting 
non_stationary_matrices |>
ggplot(aes(x = From, y = To, fill = probability)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(
    aes(label = format(round(probability, 3), nsmall = 3)),
    size = 4.5, 
    color = "#212427",
    fontface = "bold") +
    colorspace::scale_fill_continuous_diverging(
    palette = "Blue-Red 3", mid = 0.50, alpha = 0.5, 
    limits = c(0, 1), name = "Transition \nProbability") +
  labs(
    title = "Estimated Time-Varying Transition Matrices",
    subtitle = "Showing changes in transition probabilities over time",
    x = "Previous State (t-1)",
    y = "Next State (t)",
    fill = "Probability"
  ) +
  facet_wrap(~ Time_point, ncol = 3, labeller = labeller(
    Time_point = function(x){
      case_when(x == "2 Years" ~ "2018",
                x == "3 Years" ~ "2020",
                x == "4 Years" ~ "2022")
    })) +
  theme(
    axis.text = element_text(size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    legend.position = "right",
    legend.text = element_text(size = 9),
    panel.grid = element_blank()
  )
```

# Matrix distance metrics

Given two matrices $P = (p_{ij})$ and $\hat{P} = (\hat{p}_{ij})$ of size $m \times n$, we define the following distance measures

**Frobenius Norm (Matrix Euclidean Distance)**

$$D_{\text{Frobenius}}(P, \hat{P}) = \|P - \hat{P}\|_F = \sqrt{\sum_{i=1}^{m}\sum_{j=1}^{n} |p_{ij} - \hat{p}_{ij}|^2}$$

- Measures the Euclidean distance in matrix space
- Sensitive to large differences due to squaring
- Equivalent to the vector 2-norm of the flattened matrix

**Manhattan Distance (L1 Norm)**

$$D_{\text{Manhattan}}(P, \hat{P}) = \sum_{i=1}^{m}\sum_{j=1}^{n} |p_{ij} - \hat{p}_{ij}|$$

- Sum of absolute differences between corresponding elements
- Less sensitive to outliers than Frobenius norm
- Useful when sparse differences are important

**Maximum Difference (Infinity Norm)**

$$D_{\text{Max}}(A, B) = \max_{\substack{1 \leq i \leq m \\ 1 \leq j \leq n}} |a_{ij} - b_{ij}|$$

- Captures the single largest difference between elements
- Useful for worst-case analysis
- Ignores the distribution of other differences

**Mean Absolute Difference**

$$D_{\text{MeanAbs}}(P, \hat{P}) = \frac{1}{mn}\sum_{i=1}^{m}\sum_{j=1}^{n} |p_{ij} - \hat{p}_{ij}|$$

- Average absolute difference across all elements
- Scales with matrix size (unlike Manhattan distance)
- Easy to interpret as average error per element

**Root Mean Square Error (RMSE)**

$$D_{\text{RMSE}}(P, \hat{P}) = \sqrt{\frac{1}{mn}\sum_{i=1}^{m}\sum_{j=1}^{n} (p_{ij} - \hat{p}_{ij})^2}$$

- Similar to Frobenius but normalized by matrix size
-  Sensitive to large errors due to squaring
- Common in statistical and machine learning applications

**Correlation-Based Distance**

$$D_{\text{Corr}}(P, \hat{P}) = 1 - \frac{\sum_{i=1}^{m}\sum_{j=1}^{n} (p_{ij} - \bar{P})(\hat{p}_{ij} - \bar{\hat{P}})}{\sqrt{\sum_{i=1}^{m}\sum_{j=1}^{n} (p_{ij} - \bar{P})^2 \sum_{i=1}^{m}\sum_{j=1}^{n} (\hat{p}_{ij} - \bar{\hat{p}})^2}}$$

- Measures linear relationship between matrix elements
- $\bar{P} = \frac{1}{mn}\sum_{i,j}p_{ij}$ (mean of all elements in $P$)
- $\bar{\hat{P}} = \frac{1}{mn}\sum_{i,j}\hat{p}_{ij}$ (mean of all elements in $\hat{P}$)
- Range: [0,2] where 0 = perfect positive correlation

**Kullback-Leibler Divergence**

For matrices where each row sums to 1 ($\sum_j p_{ij} = \sum_j \hat{P}_{ij} = 1$) and $p_{ij}, \hat{p}_{ij} > 0$:

$$
D_{\text{KL}}(P \parallel \hat{P}) = \sum_{i=1}^{m}\sum_{j=1}^{n} p_{ij} \log\left(\frac{p_{ij}}{\hat{p}_{ij}}\right)
$$

- Measures information loss when $\hat{P}$ approximates $P$
- Asymmetric: $D_{\text{KL}}(P \parallel \hat{P}) \neq D_{\text{KL}}(\hat{P} \parallel P)$
- $0\log0 = 0$ by convention
- In practice, add $\epsilon > 0$ (e.g., $10^{-10}$) to avoid zeros

## Implementation

Let's calculate the distance metrics between the observed $P$ and estimated transition matrices $\hat{P}$.

```{r matrix-distance}
p       <- transition_matrix_observed
p_hat   <- transition_matrix_estimated
epsilon <- 1e-10

# Creating tibble of distance metrics
distances <- tibble(
  Metric = c("Frobenius", "Manhattan", "Max", "MeanAbs", "RMSE", "Correlation", "KL"),
  Value = c(
    norm(p - p_hat, type = "F"),
    sum(abs(p - p_hat)),
    max(abs(p - p_hat)),
    mean(abs(p - p_hat)),
    sqrt(mean((p - p_hat)^2)),
    1 - cor(c(p), c(p_hat)),
    sum((p + epsilon) * log((p + epsilon) / (p_hat + epsilon)))
  )) |>
  mutate(Value = round(Value, 4))

distances |>
  mutate(Metric = case_when(
    Metric == "Frobenius" ~ "Frobenius Distance",
    Metric == "Manhattan" ~ "Manhattan Distance",
    Metric == "Max" ~ "Max Difference",
    Metric == "MeanAbs" ~ "Mean Absolute Difference",
    Metric == "RMSE" ~ "Root Mean Square Error",
    Metric == "Correlation" ~ "Correlation Distance",
    Metric == "KL" ~ "Kullback-Leibler Divergence"
  )) |>
  knitr::kable(caption = "Distance based metrics", align = "c") |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE)

```

## Visualisation

Let's visualize the distance metrics between the observed and estimated transition matrices (@fig-metrics).

```{r matrix-distance-visualise}
#| code-fold: true
#| code-summary: "Check out my code"
#| fig-width: 12
#| label: fig-metrics
#| fig-cap: "Distance metrics between observed and estimated transition matrices."

caption = stringr::str_glue(
  "**RMSE:** Root Mean Squared Error\n
   **KL:** Kullback-Leibler Divergence\n
   **Correlation:** 1 - Pearson Correlation Coefficient")

distances |>
  ggplot(aes(x = reorder(Metric, -Value), y = Value, fill = Metric)) +
  geom_col(colour = "black") +
  geom_text(aes(label = Value), vjust = -0.5) +
  ggokabeito::scale_fill_okabe_ito() +
  scale_y_continuous(expand = expansion(mult = c(0.075, 0.075))) +
  labs(
    title = "Distance Between Transition Matrices",
    x = "Distance Metric",
    y = "Value",
    caption = caption) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(face = "bold", size = 12),
    plot.caption = ggtext::element_markdown(size = 10),
    legend.position = "none")
```

### 2025-03-28 R/00__Functions.R

@@ -1,0 +1,810 @@
+# A list of functions used throughout analysis
+# Used in 02__Aggregation.R ----------------------------------------------------
+
+process_health_data <- function(health_data, prefix) {
+  health_data |>
+    select(HHID, PN, 
+           Weight = paste0(prefix, "139"),
+           H_feet = paste0(prefix, "141"),
+           H_inches = paste0(prefix, "142"),
+           Hypertension = paste0(prefix, "005"),
+           Stroke = paste0(prefix, "053"),
+           H_problem = paste0(prefix, "036")) |>
+    mutate(across(Hypertension:H_problem, 
+                  ~ ifelse(. %in% c(4, 5, 6), 0, 1))) 
+}
+
+height_weight <- function(data) {
+  data |>
+    mutate(H_inches = round(H_inches, digits = 0)) |>
+    transform(Height = interaction(H_feet, H_inches)) |>
+    mutate(
+      Height = as.numeric(as.character(Height)),
+      Height = round((Height * 0.3048), digits = 2),
+      Weight = round((Weight * 0.453592), digits = 2),
+    ) |>
+    select(!c(H_feet, H_inches)) |>
+    relocate(Height, .before = Weight)
+}
+
+impute_height <- function(data) {
+  data |>
+    select(HHID, PN, starts_with("Height")) |>
+    mutate(ID = row_number()) |>
+    pivot_longer(cols = starts_with("Height"), names_to = "Year", values_to = "Height") |>
+    group_by(ID) |>
+    fill(Height, .direction = "downup") |>
+    ungroup() |>
+    pivot_wider(names_from = "Year", values_from = "Height") |>
+    select(-ID)
+}
+
+process_cognition_data <- function(cog_data, prefix, focus_total = FALSE){
+  cog_data <- cog_data |>
+    select(HHID, PN, any_of(paste0(prefix, 110:117))) |>
+    mutate(
+      across(starts_with(prefix), ~ case_when(
+        .x == 1 ~ 1,
+        .x == 5 ~ 0,
+        TRUE ~ NA_real_
+      )),
+      
+      # Invert values for specific columns
+      across(all_of(paste0(prefix, c(113, 115))), ~ ifelse(.x == 1, 0, 1)),
+      
+      # Calculate the total
+      Total_dep = rowSums(across(starts_with(prefix)), na.rm = TRUE)
+    )
+  
+  if(focus_total == TRUE) {
[…]

### 2025-03-28 R/01__LWC.R

@@ -1,0 +1,43 @@
+# Creating the LW classifications for the 2022 HRS dataset
+
+rm(list = ls())
+
+# Custom made package
+# devtools::install_github("C-Monaghan/lwc2022")
+
+path_data <- "data-raw"
+
+# Reading in data --------------------------------------------------------------
+data <- haven::read_sav(here::here(path_data, "2022/Cognition_2022.sav"))
+proc <- haven::read_sav(here::here(path_data, "2020/Procrastination.sav"))
+
+# Filter only to procrastination participants 
+proc <- proc |>
+  dplyr::filter(!is.na(RV155))
+
+lw_2022_data <- data |>
+  # Getting procrastination participants
+  dplyr::semi_join(proc, by = c("HHID", "PN")) |>
+  # Removing (potential) web respondents
+  dplyr::filter(!is.na(SD182M1), is.na(SD182WM1)) |>
+  lwc2022::extract() |>
+  lwc2022::score() |>
+  lwc2022::classify()
+
+# Joining classifications to original LW datafile -------------------------------
+lw_data <-  haven::read_dta(here::here(path_data, "03__LW_Dementia.dta"))
+
+# Joining 2022 classifications
+lw_data_full <- lw_data |>
+  dplyr::rename(HHID = hhid, PN = pn) |>
+  dplyr::semi_join(lw_2022_data, by = c("HHID", "PN")) |>
+  cbind(lw_2022_data) |>
+  dplyr::distinct() |> # HHID & PN duplicate so this removes the duplication
+  dplyr::select(!c(HHID.1, PN.1))
+
+# Exporting --------------------------------------------------------------------
+export_path <- "data"
+
+writexl::write_xlsx(
+  path = here::here(export_path, "lw_data_full.xlsx"),
+  x = lw_data_full)

### 2025-03-28 R/02__Aggregation.R

@@ -1,0 +1,178 @@
+# Aggregating procrastination data with LWC data
+
+rm(list = ls())
+
+library(dplyr)
+library(tidyr)
+
+source(here::here("R/00__Functions.R"))
+
+path_data <- "data-raw"
+
+# Reading in data --------------------------------------------------------------
+# General data
+tracker    <- haven::read_sav(here::here(path_data, "02__Tracker.sav"))
+lw_data    <- readxl::read_xlsx(here::here(path_data, "../02__Processed/lw_data_full.xlsx"))
+
+# 2016 data
+health_16  <- haven::read_sav(here::here(path_data, "2016/Health_2016.sav"))
+cog_16     <- haven::read_sav(here::here(path_data, "2016/Cognition_2016.sav"))
+
+# 2018 data
+health_18  <- haven::read_sav(here::here(path_data, "2018/Health_2018.sav"))
+cog_18     <- haven::read_sav(here::here(path_data, "2018/Cognition_2018.sav"))
+
+
+# 2020 data
+health_20  <- haven::read_sav(here::here(path_data, "2020/Health_2020.sav"))
+proc       <- haven::read_sav(here::here(path_data, "2020/Procrastination.sav"))
+cog_20     <- haven::read_sav(here::here(path_data, "2020/Cognition_2020.sav"))
+
+# 2022 data
+health_22  <- haven::read_sav(here::here(path_data, "2022/Health_2022.sav"))
+cog_22     <- haven::read_sav(here::here(path_data, "2022/Cognition_2022.sav"))
+
+
+# Filter to procrastination participants ---------------------------------------
+proc_red <- proc |>
+  filter(!is.na(RV155)) |>
+  select(HHID, PN, RV156:RV167) |>
+  rename(
+    P_1 = "RV156",
+    P_2 = "RV157",
+    P_3 = "RV158",
+    P_4 = "RV159",
+    P_5 = "RV160",
+    P_6 = "RV161",
+    P_7 = "RV162",
+    P_8 = "RV163",
+    P_9 = "RV164",
+    P_10 = "RV165",
+    P_11 = "RV166",
+    P_12 = "RV167"
+  ) |>
+  # Recoding missing data values
+  mutate(
+    across(starts_with("P_"), ~ ifelse(. %in% c(-8, 8, 9), NA, .))
+    ) |>
+  # Removing participants not in the LW data file
[…]

### 2025-03-28 R/03__EDA.R

@@ -1,0 +1,146 @@
+# Data exploration -------------------------------------------------------------
+
+rm(list = ls())
+
+library(dplyr)
+library(ggplot2)
+library(ggalluvial)
+
+# source(here::here("02__Models/00__Functions.R"))
+
+count_transitions <- function(data, years) {
+  
+  # Create dynamic column names based on the years provided
+  cogfunction_cols <- paste0("cogfunction", years)
+  
+  data |>
+    select(ID, Total_p, all_of(cogfunction_cols)) |>
+    mutate(across(all_of(cogfunction_cols), ~ case_when(
+      .x == 1 ~ "Normal Cognition",
+      .x == 2 ~ "MCI",
+      .x == 3 ~ "Dementia",
+      is.na(.x) ~ "Missing",  # Handle missing values
+      TRUE ~ "Other"          # Handle other unexpected cases
+    ))) |>
+    count(across(all_of(cogfunction_cols))) |>
+    mutate(ID = row_number()) |>
+    tidyr::pivot_longer(cols = all_of(cogfunction_cols),
+                        names_to = "Wave",
+                        values_to = "Status") |>
+    mutate(Status = factor(Status,
+                           levels = c("Normal Cognition", "MCI", "Dementia", "Missing"))) |>
+    mutate(Wave = stringr::str_replace(Wave, "cogfunction", "HRS "))
+}
+
+# Plotting dementia transitions ------------------------------------------------
+plot_transitions <- function(data, size) {
+  
+  data |>
+    ggplot(aes(x = Wave, y = n, stratum = Status, fill = Status, alluvium = ID)) +
+    geom_stratum(alpha = 0.5, width = 0.5) +
+    geom_flow(width = 0.5) +
+    geom_text(stat = "stratum", aes(label = stringr::str_wrap(Status, width = 10)), 
+              hjust = 0.5, vjust = 0.5, size = size) +
+    labs(title = "Dementia transitions across time", x = "", y = "Frequency") +
+    scale_fill_viridis_d(direction = -1) +
+    theme_minimal() +
+    theme(panel.grid = element_blank(),
+          plot.title = element_text(size = 14, hjust = 0.5, face = "bold"),
+          axis.text = element_text(size = 10),
+          axis.title = element_text(size = 10),
+          text = element_text(size = 10)) +
+    ggeasy::easy_remove_legend()
+}
+
+path_data <- "data/"
+
[…]

### 2025-03-28 _drafts/04a__Markov_Transitions.R

@@ -1,0 +1,113 @@
+rm(list = ls())
+  
+set.seed(34521)
+  
+library(dplyr)
+library(ggplot2)
+library(patchwork)
+
+source(here::here("./02__Models/00__Markov_Functions.R"))
+
+# Reading in data --------------------------------------------------------------
+path_data <- "./01__Data/02__Processed/"
+  
+data <- readxl::read_xlsx(here::here(path_data, "data.xlsx"))
+
+# Overall transition matrix ----------------------------------------------------
+# For full dataset (1996 - 2022)
+tran_matrix_full <- data |>
+  extract_years(years = seq(1996, 2022, by = 2)) |>
+  # na.omit() |>
+  create_transitions(absorbing = TRUE) |>
+  calculate_probabilties() |>
+  transition_matrix(longitudinal = FALSE) |>
+  normalise() |>
+  round(digits = 2)
+
+# For a reduced dataset (2016 - 2022)
+tran_matrix_reduced <- data |>
+  extract_years(years = seq(2016, 2022, by = 2)) |>
+  na.omit() |> # Removing some people who hadn't been recruited in 2016 / 2018
+  create_transitions(absorbing = TRUE) |>
+  calculate_probabilties() |>
+  transition_matrix(longitudinal = FALSE) |>
+  normalise() |>
+  round(digits = 2)
+
+# Visualizing ------------------------------------------------------------------
+plot_matrix(data = t(tran_matrix_full), year = 1996, ts = FALSE)
+
+plot_matrix(data = t(tran_matrix_reduced), year = 2016, ts = FALSE)
+
+# Creating transition matrices for each year ----------------------------------- 
+tran_matrices_reduced <- data |>
+  extract_years(years = seq(1998, 2022, by = 2)) |>
+  create_transitions() |>
+  group_by(Wave) |>
+  calculate_probabilties() |>
+  group_map(~ transition_matrix(.x, longitudinal = TRUE), .keep = TRUE)
+
+# I want my matrix in a certain order
+tran_matrices_reduced <- lapply(tran_matrices_reduced, function(mat) {
+  mat[c("Normal Cognition", "MCI", "Dementia"), c("Normal Cognition", "MCI", "Dementia")] |>
+    normalise()
+})
+
+# Creating a singular dataframe of transition probabilities --------------------
+wave_names <- c(seq(1998, 2020, by = 2))
+
+# Apply the reshaping function to each matrix and combine the results
[…]

### 2025-03-28 _drafts/04b__Markov_Model.R

@@ -1,0 +1,227 @@
+rm(list = ls())
+
+set.seed(34521)
+
+library(dplyr)
+library(ggplot2)
+library(patchwork)
+library(msm)
+
+source(here::here("./02__Models/00__Functions.R"))
+
+# Reading in data --------------------------------------------------------------
+path_data <- "./01__Data/02__Processed/"
+
+data <- readxl::read_xlsx(here::here(path_data, "data.xlsx"))
+
+# Applying Markov Model (MSM Package) ------------------------------------------
+# Creating suitable data set
+cols <- c(
+  "ID", "Gender", "Age", "Education_tri", 
+  paste0("Cardio_risk_", seq(16, 22, by = 2)), 
+  paste0("Total_dep_20", seq(16, 22, by = 2)),
+  "Total_p")
+
+markov_data <- data |>
+    extract_years(seq(2016, 2022, by = 2), impute = TRUE) |>
+    na.omit() |>
+    inner_join(data[, cols], by = "ID") |>
+    pivot_data() |>
+  backtrack_age(base_year = 2016, current_year = 2022) |>
+  group_by(ID) |>
+  mutate(
+    # Baseline covariates (measured in 2016)
+    Age = min(Age),
+    Cardio_risk = Cardio_risk[Wave == 1],
+    Depression = Depression[Wave == 1]
+    ) |>
+  ungroup() |>
+  # Only focusing on older adults
+  filter(Age > 50)
+
+# Let's look at a state table for the number of different transitions
+markov_data |>
+  select(ID, Wave, Status) |>
+  mutate(Status = case_when(
+    Status == 1 ~ "Normal Cognition", 
+    Status == 2 ~ "MCI", 
+    Status == 3 ~ "Dementia")) |>
+  msm::statetable.msm(state = Status, subject = ID)
+
+# Creating q matrix to specify allowed transitions
+# Define possible transitions
+states <- c("Normal Cognition", "MCI", "Dementia")
+
+# Theoretical transition probabilities
+q_matrix <- rbind(
+  c(0.00, 0.10, 0.05),  # From Normal, can transition to MCI or Dementia
+  c(0.15, 0.00, 0.15),  # From MCI, can transition to Normal or Dementia
+  c(0.00, 0.00, 0.00)         # From Dementia, no transitions (absorbing state)
+)
+
+rownames(q_matrix) <- states
+colnames(q_matrix) <- states
+
+q_matrix
+
[…]

### 2025-03-28 _drafts/05__Testing.R

@@ -1,0 +1,230 @@
+rm(list = ls())
+
+library(dplyr)
+library(easystats)
+library(ggplot2)
+
+extract_years <- function(data, years) {
+  cogtotal_cols    <- paste0("cogtot27_imp", years)
+  cogfunction_cols <- paste0("cogfunction", years)
+  
+  data |>
+    select(ID, any_of(cogtotal_cols), any_of(cogfunction_cols))
+}
+
+process_data <- function(data, absorbing = FALSE){
+  data <- data |>
+    tidyr::pivot_longer(
+      cols = matches("cogtot27_imp|cogfunction"),
+      names_to = c("Measure", "Year"),
+      names_pattern = "(cogtot27_imp|cogfunction)(\\d+)",
+      values_to = "Value"
+    ) |>
+    mutate(
+      Measure = ifelse(Measure == "cogtot27_imp", "Score", "Class"),
+      Year = factor(Year, levels = c("2016", "2018", "2020", "2022"))
+    ) |>
+    tidyr::pivot_wider(names_from = "Measure", values_from = "Value") |>
+    mutate(
+      Class = case_when(
+        Class == 1 ~ "Normal Cognition",
+        Class == 2 ~ "MCI",
+        Class == 3 ~ "Dementia",
+        TRUE ~ NA_character_
+      )
+    )
+  
+  # We can optionally specify dementia as an absorbing state
+  # Once an individual is classified with dementia they cannot be classified
+  # with anything else 
+  # Additionally, we change their score to be 6 to reflect the highest possible
+  # value you can get while still being classified with dementia
+  if(absorbing == TRUE){
+    data <- data |>
+      group_by(ID) |>
+      mutate(
+        Class = ifelse(cumany(Class == "Dementia"), "Dementia", Class),
+        Score = ifelse(Class == "Dementia" & Score > 6, 6, Score)) |>
+      ungroup()
+  }
+  
+  data <- data |>
+    mutate(Class = factor(Class, levels = c("Normal Cognition", "MCI", "Dementia")))
+  
+  return(data)
+}
+
+backtrack_age <- function(data) {
+  data |>
+    rename(Age = Age_2022) |>
+    mutate(
+      Age = Age - case_when(
+        Year == "2016" ~ 6,
+        Year == "2018" ~ 4,
+        Year == "2020" ~ 2,
+        TRUE ~ 0
+      )
+    )
+}
+
[…]

### 2025-03-28 _drafts/06__Cog_Scores.R

@@ -1,0 +1,30 @@
+rm(list = ls())
+
+set.seed(34521)
+
+library(dplyr)
+library(ggplot2)
+library(patchwork)
+library(msm)
+
+source(here::here("./02__Models/00__Functions.R"))
+
+# Reading in data --------------------------------------------------------------
+path_data <- "./01__Data/02__Processed/"
+
+data <- readxl::read_xlsx(here::here(path_data, "data.xlsx"))
+
+# Applying Markov Model (MSM Package) ------------------------------------------
+# Creating suitable data set
+cols <- c(
+  "ID", "Gender", "Age", "Education_tri", 
+  paste0("Cardio_risk_", seq(16, 22, by = 2)), 
+  paste0("Total_dep_20", seq(16, 22, by = 2)), "Total_p")
+
+data |>
+  extract_years(seq(2016, 2022, by = 2), cog_total = TRUE) |>
+  inner_join(data[, cols], by = "ID") |>
+  pivot_data(cog_total = TRUE) |>
+  backtrack_age(2016, 2022) |>
+  mutate(Total_p = ifelse(Wave == 4, Total_p, NA))
+

### 2025-03-28 _drafts/06__Discrete_Model.R

@@ -1,0 +1,357 @@
+rm(list = ls())
+
+set.seed(34521)
+
+library(dplyr)
+library(ggplot2)
+library(patchwork)
+
+source(here::here("./02__Models/00__Functions.R"))
+
+# For creating transition tables
+create_transition_table <- function(year_to, year_from) {
+  tbl <- table(
+    table_data[[paste0("HRS_", year_to)]],
+    table_data[[paste0("HRS_", year_from)]],
+    dnn = c(year_to, year_from)
+  )
+  
+  # Calculate row sums and total
+  row_sums <- rowSums(tbl)
+  total <- sum(row_sums)
+  
+  # Return as a list with both the table and summary stats
+  list(
+    table = tbl,
+    row_sums = row_sums,
+    total = total
+  )
+}
+
+# Reading in data --------------------------------------------------------------
+path_data <- "./01__Data/02__Processed/"
+
+data <- readxl::read_xlsx(here::here(path_data, "data.xlsx"))
+
+cols <- c(
+  "ID", "Gender", "Age", "Education_tri", 
+  "Cardio_risk_16", "Total_dep_2016", "Total_p")
+
+# Creating a stacked dataset (making a previous response category)
+data_stack <- data |>
+  extract_years(seq(2016, 2022, by = 2)) |>
+  na.omit() |> 
+  inner_join(data[, cols], by = "ID") |>
+  pivot_and_factorise() |>
+  group_by(ID) |>
+  mutate(Age = Age - (2022 - as.numeric(as.character(wave)))) |>
+  # Creating previous cognition column
+  mutate(status_prev = lag(status)) |>
+  ungroup() |>
+  # Removing wave 1
+  filter(wave != 2016, !is.na(Total_p))
+
+# Creating transition tables ---------------------------------------------------
+table_data <- data |>
+  extract_years(seq(2016, 2022, by = 2)) |>
+  rename_with(~ gsub("cogfunction", "HRS_", .)) |>
+  mutate(
+    across(c(HRS_2016:HRS_2022), ~ factor(.x, levels = c("Normal Cognition", "MCI", "Dementia")))
+    )
+
+# Define the time periods to analyze
+time_periods <- list(
+  c("2016", "2018"),
+  c("2018", "2020"), 
+  c("2020", "2022")
+)
+
+# Apply the function to all time periods
[…]

### 2025-03-28 _drafts/07__Discrete.R

@@ -1,0 +1,611 @@
+rm(list = ls())
+
+# Packages ---------------------------------------------------------------------
+library(dplyr)
+library(tidyr)
+library(ggplot2)
+library(patchwork)
+
+# Functions --------------------------------------------------------------------
+extract_years <- function(data, years, impute = TRUE, cog_total = FALSE, absorbing = TRUE) {
+  # Extracts cognitive function data for specified years from a dataset.
+  # Converts numeric cognitive status codes (1, 2, 3) into descriptive labels
+  # ("Normal Cognition", "MCI", "Dementia") for easier interpretation.
+  # Arguments:
+  #   - data: The input dataset containing cognitive function data.
+  #   - years: A vector of years for which data should be extracted.
+  # Returns:
+  #   - A dataset with ID and cognitive status columns for the specified years.
+  
+  # Create dynamic column names based on the years provided
+  cogfunction_cols <- paste0("cogfunction", years)
+  cogtotal_cols    <- paste0("cogtot27_imp", years)
+  
+  if(cog_total == FALSE) {
+    data <- data |>
+      # Select only the ID column and cognitive function columns for the specified years
+      select(ID, any_of(cogfunction_cols)) |>
+      mutate(across(!c(ID), ~ case_when(
+        .x == 1 ~ "Normal Cognition",
+        .x == 2 ~ "MCI",
+        .x == 3 ~ "Dementia",
+        TRUE ~ NA_character_  # To handle missing/other cases
+      )))
+  } else {
+    data <- data |>
+      # Select only the ID column and cognitive function columns for the specified years
+      select(ID, any_of(cogtotal_cols)) |>
+      rename_with( .cols = !ID, .fn = ~ stringr::str_replace(
+        string = .x,
+        pattern = "cogtot27_imp", 
+        replacement = "cog_score_"))
+  }
+  
+  if(impute == TRUE){
+    data <- data |>
+      tidyr::pivot_longer(
+        cols = !ID,
+        names_to = "Wave",
+        values_to = "Status") |>
+      group_by(ID) |>
+      tidyr::fill(Status, .direction = "down") |>
+      ungroup() |>
+      tidyr::pivot_wider(names_from = "Wave", values_from = "Status")
+  }
+  
+  if(absorbing == TRUE) {
+    data <- data |>
+      tidyr::pivot_longer(
+        cols = !ID,
+        names_to = "Wave",
+        values_to = "Status") |>
+      group_by(ID) |>
[…]

### 2025-03-28 _drafts/04a__Markov_Transitions.R
Update

@@ -1,113 +1,0 @@
-rm(list = ls())
-  
-set.seed(34521)
-  
-library(dplyr)
-library(ggplot2)
-library(patchwork)
-
-source(here::here("./02__Models/00__Markov_Functions.R"))
-
-# Reading in data --------------------------------------------------------------
-path_data <- "./01__Data/02__Processed/"
-  
-data <- readxl::read_xlsx(here::here(path_data, "data.xlsx"))
-
-# Overall transition matrix ----------------------------------------------------
-# For full dataset (1996 - 2022)
-tran_matrix_full <- data |>
-  extract_years(years = seq(1996, 2022, by = 2)) |>
-  # na.omit() |>
-  create_transitions(absorbing = TRUE) |>
-  calculate_probabilties() |>
-  transition_matrix(longitudinal = FALSE) |>
-  normalise() |>
-  round(digits = 2)
-
-# For a reduced dataset (2016 - 2022)
-tran_matrix_reduced <- data |>
-  extract_years(years = seq(2016, 2022, by = 2)) |>
-  na.omit() |> # Removing some people who hadn't been recruited in 2016 / 2018
-  create_transitions(absorbing = TRUE) |>
-  calculate_probabilties() |>
-  transition_matrix(longitudinal = FALSE) |>
-  normalise() |>
-  round(digits = 2)
-
-# Visualizing ------------------------------------------------------------------
-plot_matrix(data = t(tran_matrix_full), year = 1996, ts = FALSE)
-
-plot_matrix(data = t(tran_matrix_reduced), year = 2016, ts = FALSE)
-
-# Creating transition matrices for each year ----------------------------------- 
-tran_matrices_reduced <- data |>
-  extract_years(years = seq(1998, 2022, by = 2)) |>
-  create_transitions() |>
-  group_by(Wave) |>
-  calculate_probabilties() |>
-  group_map(~ transition_matrix(.x, longitudinal = TRUE), .keep = TRUE)
-
-# I want my matrix in a certain order
-tran_matrices_reduced <- lapply(tran_matrices_reduced, function(mat) {
-  mat[c("Normal Cognition", "MCI", "Dementia"), c("Normal Cognition", "MCI", "Dementia")] |>
-    normalise()
-})
-
-# Creating a singular dataframe of transition probabilities --------------------
-wave_names <- c(seq(1998, 2020, by = 2))
-
-# Apply the reshaping function to each matrix and combine the results
[…]

### 2025-03-28 _drafts/04b__Markov_Model.R

@@ -1,227 +1,0 @@
-rm(list = ls())
-
-set.seed(34521)
-
-library(dplyr)
-library(ggplot2)
-library(patchwork)
-library(msm)
-
-source(here::here("./02__Models/00__Functions.R"))
-
-# Reading in data --------------------------------------------------------------
-path_data <- "./01__Data/02__Processed/"
-
-data <- readxl::read_xlsx(here::here(path_data, "data.xlsx"))
-
-# Applying Markov Model (MSM Package) ------------------------------------------
-# Creating suitable data set
-cols <- c(
-  "ID", "Gender", "Age", "Education_tri", 
-  paste0("Cardio_risk_", seq(16, 22, by = 2)), 
-  paste0("Total_dep_20", seq(16, 22, by = 2)),
-  "Total_p")
-
-markov_data <- data |>
-    extract_years(seq(2016, 2022, by = 2), impute = TRUE) |>
-    na.omit() |>
-    inner_join(data[, cols], by = "ID") |>
-    pivot_data() |>
-  backtrack_age(base_year = 2016, current_year = 2022) |>
-  group_by(ID) |>
-  mutate(
-    # Baseline covariates (measured in 2016)
-    Age = min(Age),
-    Cardio_risk = Cardio_risk[Wave == 1],
-    Depression = Depression[Wave == 1]
-    ) |>
-  ungroup() |>
-  # Only focusing on older adults
-  filter(Age > 50)
-
-# Let's look at a state table for the number of different transitions
-markov_data |>
-  select(ID, Wave, Status) |>
-  mutate(Status = case_when(
-    Status == 1 ~ "Normal Cognition", 
-    Status == 2 ~ "MCI", 
-    Status == 3 ~ "Dementia")) |>
-  msm::statetable.msm(state = Status, subject = ID)
-
-# Creating q matrix to specify allowed transitions
-# Define possible transitions
-states <- c("Normal Cognition", "MCI", "Dementia")
-
-# Theoretical transition probabilities
-q_matrix <- rbind(
-  c(0.00, 0.10, 0.05),  # From Normal, can transition to MCI or Dementia
-  c(0.15, 0.00, 0.15),  # From MCI, can transition to Normal or Dementia
-  c(0.00, 0.00, 0.00)         # From Dementia, no transitions (absorbing state)
-)
-
-rownames(q_matrix) <- states
-colnames(q_matrix) <- states
-
-q_matrix
-
[…]

### 2025-03-28 _drafts/05__Testing.R

@@ -1,230 +1,0 @@
-rm(list = ls())
-
-library(dplyr)
-library(easystats)
-library(ggplot2)
-
-extract_years <- function(data, years) {
-  cogtotal_cols    <- paste0("cogtot27_imp", years)
-  cogfunction_cols <- paste0("cogfunction", years)
-  
-  data |>
-    select(ID, any_of(cogtotal_cols), any_of(cogfunction_cols))
-}
-
-process_data <- function(data, absorbing = FALSE){
-  data <- data |>
-    tidyr::pivot_longer(
-      cols = matches("cogtot27_imp|cogfunction"),
-      names_to = c("Measure", "Year"),
-      names_pattern = "(cogtot27_imp|cogfunction)(\\d+)",
-      values_to = "Value"
-    ) |>
-    mutate(
-      Measure = ifelse(Measure == "cogtot27_imp", "Score", "Class"),
-      Year = factor(Year, levels = c("2016", "2018", "2020", "2022"))
-    ) |>
-    tidyr::pivot_wider(names_from = "Measure", values_from = "Value") |>
-    mutate(
-      Class = case_when(
-        Class == 1 ~ "Normal Cognition",
-        Class == 2 ~ "MCI",
-        Class == 3 ~ "Dementia",
-        TRUE ~ NA_character_
-      )
-    )
-  
-  # We can optionally specify dementia as an absorbing state
-  # Once an individual is classified with dementia they cannot be classified
-  # with anything else 
-  # Additionally, we change their score to be 6 to reflect the highest possible
-  # value you can get while still being classified with dementia
-  if(absorbing == TRUE){
-    data <- data |>
-      group_by(ID) |>
-      mutate(
-        Class = ifelse(cumany(Class == "Dementia"), "Dementia", Class),
-        Score = ifelse(Class == "Dementia" & Score > 6, 6, Score)) |>
-      ungroup()
-  }
-  
-  data <- data |>
-    mutate(Class = factor(Class, levels = c("Normal Cognition", "MCI", "Dementia")))
-  
-  return(data)
-}
-
-backtrack_age <- function(data) {
-  data |>
-    rename(Age = Age_2022) |>
-    mutate(
-      Age = Age - case_when(
-        Year == "2016" ~ 6,
-        Year == "2018" ~ 4,
-        Year == "2020" ~ 2,
-        TRUE ~ 0
-      )
-    )
-}
-
[…]

### 2025-03-28 _drafts/06__Cog_Scores.R

@@ -1,30 +1,0 @@
-rm(list = ls())
-
-set.seed(34521)
-
-library(dplyr)
-library(ggplot2)
-library(patchwork)
-library(msm)
-
-source(here::here("./02__Models/00__Functions.R"))
-
-# Reading in data --------------------------------------------------------------
-path_data <- "./01__Data/02__Processed/"
-
-data <- readxl::read_xlsx(here::here(path_data, "data.xlsx"))
-
-# Applying Markov Model (MSM Package) ------------------------------------------
-# Creating suitable data set
-cols <- c(
-  "ID", "Gender", "Age", "Education_tri", 
-  paste0("Cardio_risk_", seq(16, 22, by = 2)), 
-  paste0("Total_dep_20", seq(16, 22, by = 2)), "Total_p")
-
-data |>
-  extract_years(seq(2016, 2022, by = 2), cog_total = TRUE) |>
-  inner_join(data[, cols], by = "ID") |>
-  pivot_data(cog_total = TRUE) |>
-  backtrack_age(2016, 2022) |>
-  mutate(Total_p = ifelse(Wave == 4, Total_p, NA))
-

### 2025-03-28 _drafts/06__Discrete_Model.R

@@ -1,357 +1,0 @@
-rm(list = ls())
-
-set.seed(34521)
-
-library(dplyr)
-library(ggplot2)
-library(patchwork)
-
-source(here::here("./02__Models/00__Functions.R"))
-
-# For creating transition tables
-create_transition_table <- function(year_to, year_from) {
-  tbl <- table(
-    table_data[[paste0("HRS_", year_to)]],
-    table_data[[paste0("HRS_", year_from)]],
-    dnn = c(year_to, year_from)
-  )
-  
-  # Calculate row sums and total
-  row_sums <- rowSums(tbl)
-  total <- sum(row_sums)
-  
-  # Return as a list with both the table and summary stats
-  list(
-    table = tbl,
-    row_sums = row_sums,
-    total = total
-  )
-}
-
-# Reading in data --------------------------------------------------------------
-path_data <- "./01__Data/02__Processed/"
-
-data <- readxl::read_xlsx(here::here(path_data, "data.xlsx"))
-
-cols <- c(
-  "ID", "Gender", "Age", "Education_tri", 
-  "Cardio_risk_16", "Total_dep_2016", "Total_p")
-
-# Creating a stacked dataset (making a previous response category)
-data_stack <- data |>
-  extract_years(seq(2016, 2022, by = 2)) |>
-  na.omit() |> 
-  inner_join(data[, cols], by = "ID") |>
-  pivot_and_factorise() |>
-  group_by(ID) |>
-  mutate(Age = Age - (2022 - as.numeric(as.character(wave)))) |>
-  # Creating previous cognition column
-  mutate(status_prev = lag(status)) |>
-  ungroup() |>
-  # Removing wave 1
-  filter(wave != 2016, !is.na(Total_p))
-
-# Creating transition tables ---------------------------------------------------
-table_data <- data |>
-  extract_years(seq(2016, 2022, by = 2)) |>
-  rename_with(~ gsub("cogfunction", "HRS_", .)) |>
-  mutate(
-    across(c(HRS_2016:HRS_2022), ~ factor(.x, levels = c("Normal Cognition", "MCI", "Dementia")))
-    )
-
-# Define the time periods to analyze
-time_periods <- list(
-  c("2016", "2018"),
-  c("2018", "2020"), 
-  c("2020", "2022")
-)
-
-# Apply the function to all time periods
[…]

### 2025-03-28 _drafts/07__Discrete.R

@@ -1,611 +1,0 @@
-rm(list = ls())
-
-# Packages ---------------------------------------------------------------------
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-library(patchwork)
-
-# Functions --------------------------------------------------------------------
-extract_years <- function(data, years, impute = TRUE, cog_total = FALSE, absorbing = TRUE) {
-  # Extracts cognitive function data for specified years from a dataset.
-  # Converts numeric cognitive status codes (1, 2, 3) into descriptive labels
-  # ("Normal Cognition", "MCI", "Dementia") for easier interpretation.
-  # Arguments:
-  #   - data: The input dataset containing cognitive function data.
-  #   - years: A vector of years for which data should be extracted.
-  # Returns:
-  #   - A dataset with ID and cognitive status columns for the specified years.
-  
-  # Create dynamic column names based on the years provided
-  cogfunction_cols <- paste0("cogfunction", years)
-  cogtotal_cols    <- paste0("cogtot27_imp", years)
-  
-  if(cog_total == FALSE) {
-    data <- data |>
-      # Select only the ID column and cognitive function columns for the specified years
-      select(ID, any_of(cogfunction_cols)) |>
-      mutate(across(!c(ID), ~ case_when(
-        .x == 1 ~ "Normal Cognition",
-        .x == 2 ~ "MCI",
-        .x == 3 ~ "Dementia",
-        TRUE ~ NA_character_  # To handle missing/other cases
-      )))
-  } else {
-    data <- data |>
-      # Select only the ID column and cognitive function columns for the specified years
-      select(ID, any_of(cogtotal_cols)) |>
-      rename_with( .cols = !ID, .fn = ~ stringr::str_replace(
-        string = .x,
-        pattern = "cogtot27_imp", 
-        replacement = "cog_score_"))
-  }
-  
-  if(impute == TRUE){
-    data <- data |>
-      tidyr::pivot_longer(
-        cols = !ID,
-        names_to = "Wave",
-        values_to = "Status") |>
-      group_by(ID) |>
-      tidyr::fill(Status, .direction = "down") |>
-      ungroup() |>
-      tidyr::pivot_wider(names_from = "Wave", values_from = "Status")
-  }
-  
-  if(absorbing == TRUE) {
-    data <- data |>
-      tidyr::pivot_longer(
-        cols = !ID,
-        names_to = "Wave",
-        values_to = "Status") |>
-      group_by(ID) |>
[…]

### 2025-03-28 analysis/data-final.qmd
First site push

---
title: "Download final data"
format:
  html:
    code-tools: false
    toc: false
---

The cleaned data is available in three formats:

- CSV file for any program
- `.rds` file for R <span class="small">(load with `df <- readRDS("weekly_panel.rds")`)</span>
- `.dta` file for Stata <span class="small">(load with `use "weekly_panel.dta"`)</span>

## Data

<a class="btn btn-primary btn-sm" target="_blank" href="data/data.csv">{{< fa table >}} &ensp;<span class="font-monospace">data.csv</span></a> &emsp; <a class="btn btn-success btn-sm" target="_blank" href="data/data.rds">{{< fa brands r-project >}} &ensp;<span class="font-monospace">data.rds</span></a> &emsp; <a class="btn btn-warning btn-sm" target="_blank" href="data/data.dta">{{< fa table-cells >}} &ensp;<span class="font-monospace">data.dta</span></a>

### 2025-03-28 analysis/discrete-markov.qmd
First site push

--- 2025-03-28 reports/discrete_model.qmd
@@
 ```{r data}
-data <- readxl::read_xlsx(here::here("data/data.xlsx"))
-```
-
-# Research Questions and Hypotheses
-
-## Primary Hypothesis
-
-We hypothesize that:
-
-- **H1:** Higher levels of procrastination are associated with:
-  - Increased risk of transitioning from Normal Cognition to MCI $(\beta = x, p < .05)$
-  - Increased risk of transitioning from MCI to Dementia $(\beta = y, p < .05)$
-
-- **H0:** There is no association between procrastination and cognitive decline transitions
-
-In order to test this hypothesis we will be using a **discrete time Markov Model**
+data <- read.csv(here::here("analysis/data/data.csv"))
+```
 

### 2025-03-28 analysis/hypothesis.qmd
First site push

--- 2025-03-28 reports/discrete_model.qmd
@@
 ---
-title: "Discrete Time Markov Model Analysis of Cognitive Decline"
-subtitle: "Examining the Association Between Procrastination and Cognitive Transitions"
-author: 
-  - name: "Cormac Monaghan"
-    url: "https://c-monaghan.github.io/"
-    orcid: "https://orcid.org/0000-0001-9012-3060"
-    affiliation:
-      - "Hamilton Institute, Maynooth University, Ireland"
-      - "Department of Psychology, Maynooth University, Ireland"
-abstract: |
-  This analysis examines transitions between cognitive states (Normal Cognition, MCI, Dementia) using discrete-time Markov models, with particular focus on procrastination as a risk factor.
-toc: true
-toc-depth: 3
-toc-title: On this page
-number-sections: true
+title: "Hypotheses"
 format: 
   html:
-    theme: 
-      light: cosmo
-    page-layout: full
-    shift-heading-level-by: 1
-    html-math-method: katex
-    code-tools: true
-    code-copy: true
-    code-block-bg: "#333A3F"
-    highlight-style: monokai
-    fig-align: center
-    self-contained: true
+    code-tools: false
 ---
-
-# Setting up {.unnumbered}
-
-## Loading packages {.unnumbered}
-
-```{r packages}
-#| code-fold: true
-#| code-summary: "Check out my packages"
-#| warning: false
-
-# Packages ---------------------------------------------------------------------
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-library(patchwork)
-```
-
-## Loading functions {.unnumbered}
-
-```{r functions}
-#| code-fold: true
-#| code-summary: "Check out my functions"
-#| warning: false
-
-# Functions --------------------------------------------------------------------
-extract_years <- function(data, years, impute = TRUE, cog_total = FALSE, absorbing = TRUE) {
-  # Extracts cognitive function data for specified years from a dataset.
-  # Converts numeric cognitive status codes (1, 2, 3) into descriptive labels
-  # ("Normal Cognition", "MCI", "Dementia") for easier interpretation.
-  # Arguments:
-  #   - data: The input dataset containing cognitive function data.
-  #   - years: A vector of years for which data should be extracted.
-  # Returns:
-  #   - A dataset with ID and cognitive status columns for the specified years.
-
-  # Create dynamic column names based on the years provided
-  cogfunction_cols <- paste0("cogfunction", years)
-  cogtotal_cols    <- paste0("cogtot27_imp", years)
-
-  if(cog_total == FALSE) {
-    data <- data |>
-      # Select only the ID column and cognitive function columns for the specified years
-      select(ID, any_of(cogfunction_cols)) |>
-      mutate(across(!c(ID), ~ case_when(
-        .x == 1 ~ "Normal Cognition",
-        .x == 2 ~ "MCI",
-        .x == 3 ~ "Dementia",
-        TRUE ~ NA_character_  # To handle missing/other cases
-      )))
-  } else {
-    data <- data |>
-      # Select only the ID column and cognitive function columns for the specified years
-      select(ID, any_of(cogtotal_cols)) |>
-      rename_with( .cols = !ID, .fn = ~ stringr::str_replace(
-        string = .x,
-        pattern = "cogtot27_imp", 
-        replacement = "cog_score_"))
-  }
-
-  if(impute == TRUE){
-    data <- data |>
-      tidyr::pivot_longer(
-        cols = !ID,
-        names_to = "Wave",
-        values_to = "Status") |>
-      group_by(ID) |>
-      tidyr::fill(Status, .direction = "down") |>
-      ungroup() |>
-      tidyr::pivot_wider(names_from = "Wave", values_from = "Status")
-  }
-
-  if(absorbing == TRUE) {
-    data <- data |>
-      tidyr::pivot_longer(
-        cols = !ID,
-        names_to = "Wave",
-        values_to = "Status") |>
-      group_by(ID) |>
-      mutate(Status = ifelse(cumany(Status == "Dementia"), "Dementia", Status)) |>
-      ungroup() |>
-      tidyr::pivot_wider(names_from = "Wave", values_from = "Status")
-  }
-
-  return(data)
-}
-
-pivot_and_factorise <- function(data) {
-  # Converts cognitive function data from wide to long format and factorizes key variables.
-  # Pivots multiple cogfunction columns into wave/status pairs and converts categorical
-  # variables (Gender, Education, status) to factors with meaningful labels.
-  # Arguments:
-  #   - data: Dataset containing cognitive function variables in wide format
-  # Returns:
-  #   - Long-format dataset with factorized variables, ordered by ID and wave
-  # 
-  data |>
-    tidyr::pivot_longer(
-      cols = starts_with("cogfunction"), names_to = "wave",
-      names_prefix = "cogfunction", values_to = "status") |>
-    mutate(
-      Gender = factor(Gender, levels = c(0, 1)),
-      Education_tri = factor(Education_tri, levels = c(0, 1, 2)),
-      wave = factor(wave),
-      status = factor(status, 
-                      levels = c("Normal Cognition", "MCI", "Dementia"),
-                      labels = c(1, 2, 3))) |>
-    relocate(wave, .after = ID) |>
-    relocate(status, .after = wave)
-}
-
-create_transition_table <- function(year_to, year_from) {
-  # Creates a transition frequency table between two specified years of cognitive status data.
-  # Calculates row sums and total observations for transition analysis.
-  # Arguments:
-  #   - year_to: Target year for transitions (character or numeric)
-  #   - year_from: Origin year for transitions (character or numeric)
-  # Returns:
-  #   - List containing: 
-  #     - transition frequency table
-  #     - row sums
-  #     - total observations
-  # 
-  tbl <- table(
-    table_data[[paste0("HRS_", year_to)]],
-    table_data[[paste0("HRS_", year_from)]],
-    dnn = c(year_to, year_from)
-  )
-
-  # Calculate row sums and total
-  row_sums <- rowSums(tbl)
-  total <- sum(row_sums)
-
-  # Return as a list with both the table and summary stats
-  list(
-    table = tbl,
-    row_sums = row_sums,
-    total = total
-  )
-}
-
-create_transition_dataset <- function(data, transition_results) {
-  # Combines multiple transition tables into a single analysis-ready dataset.
-  # Formats period labels, ensures consistent factor levels, and structures data for visualization.
-  # Arguments:
-  #   - data: List of year pairs to process (e.g., list(c(2016,2018)))
-  #   - transition_results: List containing transition tables from create_transition_table()
-  # Returns:
-  #   - Tidy dataset with transition frequencies between all specified periods
-  # 
-  data |>
-    purrr::map_dfr(~ {
-      period_name <- paste(.x[2], .x[1], sep = " - ")
-      tbl <- transition_results[[paste(.x[2], .x[1], sep = "-")]]$table
-
-      as.data.frame(tbl) |>
-        rename(t_minus_1 = 1, t = 2) %>%  # Positional renaming
-        mutate(Period = period_name, .before = t_minus_1)
-    }) |>
-    mutate(
-      Period = stringr::str_replace(Period, "(\\d+) - (\\d+)", "\\2 - \\1"),
-      Period = factor(Period, levels = c("2016 - 2018", "2018 - 2020", "2020 - 2022")),
-      t_minus_1 = factor(t_minus_1, levels = c("Normal Cognition", "MCI", "Dementia")),
-      t = factor(t, levels = c("Normal Cognition", "MCI", "Dementia"))
-    )
-}
-
-create_transitions <- function(data, absorbing = FALSE){
-  # Reshapes data from wide to long format to track cognitive status transitions over time.
-  # Calculates the next wave's cognitive status for each individual and creates a transition column.
-  # Optionally treats "Dementia" as an absorbing state, meaning once an individual is classified
-  # with dementia, their status cannot change in subsequent waves.
-  # Arguments:
-  #   - data: The dataset containing cognitive status data.
-  #   - absorbing: A logical flag indicating whether "Dementia" should be treated 
-  #   as an absorbing state.
-  # Returns:
-  #   - A dataset with transition information, including current and next wave statuses.
-
-  # Reshape the data from wide to long format to track cognitive status over waves
-  data <- data |>
-    select(ID, starts_with("cogfunction")) |>
-    tidyr::pivot_longer(cols = !ID,
-                        names_to = "Wave",
-                        values_to = "Status") |>
-    mutate(Wave = as.factor(stringr::str_replace(Wave, "cogfunction", ""))) |>
-    # Arrange by ID and Wave to prepare for transition calculation
-    arrange(ID, Wave) |>
-    group_by(ID) |>
-    # Get the next wave's cognitive status for each person
-    mutate(next_wave_status = lead(Status)) |>
-    ungroup()
-
-  # We can optionally specify dementia as an absorbing state
-  # Once an individual is classified with dementia they cannot be classified
-  # with anything else 
-  if(absorbing == TRUE) {
-    data <- data |>
-      group_by(ID) |>
-      mutate(
-        Status = ifelse(cumany(Status == "Dementia"), "Dementia", Status),
-        next_wave_status = ifelse(cumany(Status == "Dementia"), "Dementia", next_wave_status),
-        transition = paste(Status, next_wave_status, sep = " to ")
-      ) |>
-      filter(Wave %in% c(2016, 2018, 2020))
-  }
-
-  # Filter out rows where either the current or next status is missing
-  data <- data |>
-    group_by(ID) |>
-    filter(!is.na(Status), !is.na(next_wave_status)) |>
-    ungroup() |>
-    # Create a new column representing the transition from one status to the next
-    mutate(
-      transition = paste(Status, next_wave_status, sep = " to "),
-      transition = factor(
-        transition, 
-        levels = c("Normal Cognition to Normal Cognition", "Normal Cognition to MCI",
-                   "Normal Cognition to Dementia", "MCI to Normal Cognition", "MCI to MCI",
-                   "MCI to Dementia", "Dementia to Dementia") 
-      ))
-
-  return(data)
-}
-
-observed_transition_matrix <- function(data) {
-  # Converts transition data into a properly formatted probability transition matrix.
-  # Ensures consistent state ordering and converts proportions to matrix format suitable
-  # for multi-state modeling and visualization.
-  # Arguments:
-  #   - data: Transition dataset from create_transitions()
-  # Returns:
-  #   - Square transition probability matrix with states as row/column names
-  # 
-  # The order I want my matrix in
-  state_order <- c("Normal Cognition", "MCI", "Dementia")
-
-  data |>
-    group_by(Status, next_wave_status) |>
-    summarise(freq = n(),  .groups = "drop") |>
-    group_by(Status) |>
-    mutate(freq = round(proportions(freq), 3)) |>
-    ungroup() |>
-    complete(Status, next_wave_status, fill = list(freq = 0)) |>
-    tidyr::pivot_wider(names_from = next_wave_status, values_from = freq, names_sort = TRUE) |>
-    tibble::column_to_rownames("Status") |>
-    select(all_of(state_order)) |>
-    _[state_order, ] |>
-    as.matrix()
-}
-
-normalise <- function(x) {
-  # Normalizes a matrix so that each row sums to 1.
-  # Arguments:
-  #   - x: The input matrix to normalize.
-  # Returns:
-  #   - A normalized matrix where each row sums to 1.
-  x / rowSums(x)
-}
-
-tran_stack_graph <- function(data) {
-  # Creates stacked bar charts visualizing cognitive state transitions across time periods.
-  # Shows composition of current states by previous state, faceted by observation period.
-  # Arguments:
-  #   - data: Transition dataset from create_transition_dataset()
-  # Returns:
-  #   - ggplot object showing stacked transition proportions
-  # 
-  data |>
-    ggplot(aes(x = t_minus_1, y = Freq, fill = t)) +
-    geom_col(position = "stack", colour = "black") +
-    facet_wrap(~ Period, ncol = 3) +
-    labs(
-      x = "Previous State (t-1)", 
-      y = "Count",
-      title = "Outcomes by Prior Cognitive State",
-      fill = "Current State (t)"
-    ) +
-    ggokabeito::scale_fill_okabe_ito() +
-    ggeasy::easy_move_legend("bottom")
-}
-
-tran_heat_map <- function(data) {
-  # Generates heatmap visualization of transition frequencies between cognitive states.
-  # Uses color intensity and labeled values to show transition patterns across time periods.
-  # Arguments:
-  #   - data: Transition dataset from create_transition_dataset()
-  # Returns:
-  #   - ggplot heatmap with state transitions as cells
-  # 
-  data |>
-    ggplot(aes(x = t_minus_1, y = t, fill = Freq)) +
-    geom_tile(color = "white") +
-    geom_text(aes(label = Freq), color = "black", size = 3.5) +  # Add counts
-    scale_fill_gradient(low = "white", high = "steelblue") +
-    facet_wrap(~ Period, ncol = 3) +  # Split by time period
-    labs(
-      x = "Previous State (t-1)", 
-      y = "Current State (t)",
-      title = "Cognitive State Transitions Between Time Periods",
-      fill = "Frequency"
-    ) +
-    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-    ggeasy::easy_move_legend("bottom")
-}
-
-tidy_predictions <- function(predictions) {
-  # Restructures model prediction matrices into tidy format for visualization.
-  # Converts numeric codes to factor labels and reshapes multiple prediction columns
-  # into key-value pairs suitable for ggplot.
-  # Arguments:
-  #   - predictions: Raw prediction matrix from model output
-  # Returns:
-  #   - Long-format dataset with probabilities for each cognitive state
-  #   
-  predictions |>
-    as.matrix() |>
-    as_tibble() |>
-    mutate(
-      Gender = factor(ifelse(Gender == 0, "Male", "Female"), levels = c("Male", "Female")),
-      status_prev = case_when(
-        status_prev == 1 ~ "Normal Cognition",
-        status_prev == 2 ~ "MCI",
-        status_prev == 3 ~ "Dementia",
-      ),
-      status_prev = factor(status_prev, levels = c("Normal Cognition", "MCI", "Dementia")),
-      Total_p = as.numeric(Total_p),
-      across(c(pred.1:pred.3), as.numeric)) |>
-    tidyr::pivot_longer(cols = c(pred.1:pred.3), names_to = "status", values_to = "prob") |>
-    mutate(status = case_when(
-      status == "pred.1" ~ "Normal Cognition",
-      status == "pred.2" ~ "MCI",
-      status == "pred.3" ~ "Dementia"
-    ),
-    status = factor(status, levels = c("Normal Cognition", "MCI", "Dementia")),
-    status_prev = factor(status_prev, levels = c("Normal Cognition", "MCI", "Dementia")))
-}
-
-plot_predictions_stationary <- function(predictions) {
-  # Visualizes predicted transition probabilities from stationary multi-state models.
-  # Shows probability curves by procrastination level, stratified by baseline state.
-  # Arguments:
-  #   - predictions: Tidy predictions from tidy_predictions()
-  # Returns:
-  #   - Faceted ggplot showing predicted probability curves
-  # 
-  predictions |>
-    ggplot(aes(x = Total_p, y = prob, colour = status_prev)) +
-    geom_line(linewidth = 1) +
-    ggokabeito::scale_colour_okabe_ito() +
-    labs(
-      title = "Predicted transition probabilities (stationary model)",
-      # subtitle = "According to previous cognitive state",
-      x = "Total Procrastination", y = "Probability", colour = "Previous State") +
-    facet_wrap(~ status, labeller = labeller(
-      status = function(x){paste0("Transition to: ", x)}
-    ))
-}
-
-get_time_varying_matrix <- function(model, time_point) {
-  expand.grid(
-    Gender = factor(0),
-    Total_p = mean(data_stack$Total_p),
-    status_prev = factor(1:3),
-    wave = time_point) |> 
-    modelr::add_predictions(model, var = "prob", type = "probs") |> 
-    select(status_prev, starts_with("prob")) |> 
-    mutate(
-      status_prev = factor(status_prev, labels = state_names),
-      across(starts_with("prob"), ~round(., 3))) |>
-    tibble::column_to_rownames("status_prev") |>
-    as.matrix() |> `colnames<-`(state_names)
-}
-```
-
-## Setting theme {.unnumbered}
-
-```{r set-theme}
-#| code-fold: true
-#| code-summary: "Check out my theme"
-
-colour <- "#212427"
-
-theme_set(
-  theme_minimal() +
-    theme(
-      plot.title = element_text(hjust = 0.5, size = 14, colour = colour, face = "bold"),
-      plot.subtitle = element_text(hjust = 0.5, size = 12, colour = colour),
-      axis.title = element_text(size = 10, colour = colour, face = "bold"),
-      strip.text = element_text(size = 10, colour = colour, face = "bold"),
-      legend.title = element_text(hjust = 0.5, colour = colour, face = "bold"),
-      ))
-```
-
-## Reading in data {.unnumbered}
-
-```{r data}
-data <- readxl::read_xlsx(here::here("data/data.xlsx"))
-```
-
-# Research Questions and Hypotheses
 
@@
 In order to test this hypothesis we will be using a **discrete time Markov Model**
-
-# Data Preparation
-
-Before preparing the data, we want to create a vector of covariates that we are interested in using in our analysis
-
-```{r covariates}
-cols <- c("ID", "Gender", "Age", "Education_tri", "Cardio_risk_16", "Total_dep_2016", "Total_p")
-```
-
-Following this, we can now use the `extract_years()` function to extract the cognitive function data for the years 2016, 2018, and 2020.
-
-- We will also impute some missing values `impute = TRUE` using logical reasoning (if a respondent has an NA value in 2018, but has a classification of "normal cognition" in 2020, then the missing 2018 value becomes "normal cognition").
-- We will also treat dementia as an absorbing state `absorbing = TRUE`.
-
-```{r extract-years}
-data_stack <- data |> extract_years(years = seq(2016, 2022, by = 2), impute = TRUE, absorbing = TRUE) |> na.omit()
-
-head(data_stack)
-```
-
-## Creating a stacked dataset
-
-For each respondent we will now add in their relevant covariate data.
-Following this, we transform the data to long format and convert categorical variables to factors using the `pivot_and_factorise()` function.
-Additionally, we will fix the `Age` column to properly represent the age of the respondent at each time point
-
-```{r long_data}
-data_stack <- data_stack |>
-  inner_join(data[, cols], by = "ID") |>
-  pivot_and_factorise() |>
-  group_by(ID) |>
-  mutate(Age = Age - (2022 - as.numeric(as.character(wave))))
-
-head(data_stack)
-```
-
-Finally, we will create a new variable `status_prev` that notes the cognitive status of the respondent in the previous wave $(t - 1)$.
-This will be done by using the `lag` function from the `dplyr` package.
-
-```{r stack-data}
-data_stack <- data_stack |>
-  mutate(status_prev = lag(status), .after = status) |>
-  ungroup() |> 
-  filter(wave != 2016, !is.na(Total_p))
-
-head(data_stack)
-```
-
-# Transition frequencies
-
-We will now calculate the transition frequencies between cognitive states for each time period.
-We will use the `create_transition_table()` function to create a transition table for each time period.
-This will be done with the help of the `map()` function from the `purrr` package.
-Finally, to combine all the transition tables into one dataset, we will use the `create_transition_dataset()` function.
-
-```{r transition-tables}
-# Creating a table dataset
-table_data <- data |>
-  extract_years(seq(2016, 2022, by = 2)) |>
-  rename_with(~ gsub("cogfunction", "HRS_", .)) |>
-  mutate(
-    across(c(HRS_2016:HRS_2022), ~ factor(.x, levels = c("Normal Cognition", "MCI", "Dementia")))
-  )
-
-# Creating transition frequencies ---------------------------------------------
-## These are the time periods we are interested in
-time_periods <- list(
-  c("2016", "2018"),
-  c("2018", "2020"), 
-  c("2020", "2022")
-)
-
-# Applying function
-transition_results <- purrr::map(time_periods, ~ create_transition_table(.x[1], .x[2]))
-names(transition_results) <- purrr::map_chr(time_periods, ~ paste(.x[2], .x[1], sep = "-"))
-
-### Creating one dataset
-transition_frequencies <- time_periods |>
-  create_transition_dataset(transition_results = transition_results)
-
-head(transition_frequencies)
-```
-
-## Observed transition matrix
-
-The probability distribution of transitions from one state to another can be represented into a transition matrix $P = (p_{ij})_{i,j}$ where each element of position $(i, j)$ represents the transition probability $p_{ij}$.
-
-In order to create this matrix we will use both the `create_transitions()` and `observed_transition_matrix()` functions.
-
-```{r observed-matrix}
-## Creating observed transition matrix -----------------------------------------
-transition_matrix_observed <- data |>
-  extract_years(seq(2016, 2022, by = 2)) |>
-  create_transitions() |>
-  observed_transition_matrix()
-
-transition_matrix_observed
-```
-
-## Visualisation
-
-Let's visualize both the transition frequencies (@fig-frequencies) and matrix (@fig-matrix).
-
-```{r visualise-transition}
-#| code-fold: true
-#| code-summary: "Check out my code"
-#| fig-width: 12
-#| fig-height: 10
-#| label: fig-frequencies
-#| fig-cap: "Transition frequencies between cognitive states for each time period."
-
-fig_1 <- transition_frequencies |> tran_stack_graph()
-fig_2 <- transition_frequencies |> tran_heat_map()
-
-fig_1 / fig_2
-```
-
-```{r visualise-matrix}
-#| code-fold: true
-#| code-summary: "Check out my code"
-#| fig-width: 12
-#| label: fig-matrix
-#| fig-cap: "Observed transition matrix between cognitive states for the years 2016 - 2022."
-
-tran_df <- transition_matrix_observed |>
-  as.data.frame() |>
-  mutate(
-    from_state = factor(c("Normal cognition", "MCI", "Dementia"),
-                        levels = c("Normal cognition", "MCI", "Dementia"))
-    ) |>
-  reshape2::melt(
-    id.vars = "from_state", variable.name = "to_state", value.name = "probability") |>
-  mutate(
-    to_state = factor(to_state, levels = rev(levels(to_state)))
-    )
-
-fig_2 <- tran_df |>
-  ggplot(aes(x = from_state, y = to_state, fill = probability)) +
-  geom_tile(color = "white", linewidth = 0.5) +
-  geom_text(
-    aes(label = format(round(probability, 3), nsmall = 3)),
-    size = 4.5, 
-    color = "#212427",
-    fontface = "bold") +
-  colorspace::scale_fill_continuous_diverging(
-    palette = "Blue-Red 3", mid = 0.50, alpha = 0.5, 
-    limits = c(0, 1), name = "Transition \nProbability") +
-  labs(title = "Transition Probabilities Across Cognitive States",
-       subtitle = "Observed state transitions between assessment waves",
-       x = "Previous State (t - 1)", 
-       y = "Current State (t)") +
-  theme(
-    axis.text = element_text(size = 10),
-    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
-    legend.position = "right",
-    legend.text = element_text(size = 9),
-    panel.grid = element_blank()
-  )
-
-fig_2
-```
-
-# Modelling
-
-## Markov Process Fundamentals
-
-Discrete-time Markov models belong to a class of stochastic processes that satisfy the **Markov property**, which can be formally expressed as:
-
-$$
-P(X_{t+1} = j \vert X_t = i, X_{t-1} = i_{t-1}, \dots X_0 = i_0) = P(X_{t+1} = j \vert X_t = i)
-$$ {#eq-markov}
-
-This property establishes that the future state $X_{t+1}$ depends only on the current state $X_t$, not on the entire history of states.
-
-### Transition Probability Matrix
-
-For our three-state system (Normal Cognition, Mild Cognitive Impairment [MCI], Dementia), the transition matrix $P$ captures all possible transition probabilities:
-
-$$
-P = \begin{bmatrix} 
-p_{11} & p_{12} & p_{13} \\
-p_{21} & p_{22} & p_{23} \\
-p_{31} & p_{32} & p_{33} \\
-\end{bmatrix}
-$$ {#eq-t-matrix}
-
-where:
-
-- $p_{ij} = P(X_{t+1} = j \vert X_t = i)$ represents the probability of transitioning from state $i$ to state $j$.
-- Each row sums to 1, $\sum^3_{j=1} p_{ij} = 1 \quad  \forall_i \in \{1, 2, 3\}$
-
-### Multinomial Logistic Regression Formulation
-
-We model the transition probabilities using **multinomial logistic regression**, where the log-odds of each transition relative to a reference state are linear functions of covariates.
-
-For a system with $K$ states (using state $K$ as reference), we have:
-
-$$
-log \left( \frac{P(Y = j \vert x)}{P(Y = k \vert x)} \right) = \beta_{j0} + \beta_j^Tx \qquad \text{for } j = 1, \dots K-1
-$$ {#eq-multinomial}
-
-For non-reference states $j = 1, \dots, K - 1$
-
-$$
-P(Y = j \vert x) = \frac{e^{\beta_{0j} + \beta_j^Tx}}{1 + \sum^{k - 1}_{k = 1} e^{\beta_{0k} + \beta_k^Tx}}
-$$ {#eq-non-reference}
-
-For the reference state $K$:
-
-$$
-P(Y = k \vert x) = \frac{1}{1 + \sum^{k - 1}_{k = 1} e^{\beta_{0k} + \beta_j^Tx }}
-$$ {#eq-reference}
-
-### Time-Homogeneous Approach
-
-Our implementation assumes time-homogeneous transitions, but the framework can be extended to time-varying probabilities:
-
-$$
-P^{(t)} = \begin{bmatrix} 
-p_{11}^{(t)} & p_{12}^{(t)} & \cdots & p_{1n}^{(t)} \\
-p_{21}^{(t)} & p_{22}^{(t)} & \cdots & p_{2n}^{(t)} \\
-\vdots & \vdots & \ddots & \vdots \\
-p_{n1}^{(t)} & \cdots & \cdots & p_{nn}^{(t)} \\
-\end{bmatrix}
-$$ {#eq-non-stationary}
-
-Our primary analysis assumes the transition matrix remains fixed:
-
-$$
-P^{(t)} \equiv P \qquad \forall \; t
-$$ {#eq-assumption}
-
-### Absorbing State Specification
-
-We model Dementia as an absorbing state:
-
-$$
-p_{3j} = \begin{cases} 
-1 \qquad \text{if } j = 3 \\
-0 \qquad \text{otherwise} \end{cases}
-$$ {#eq-dementia}
-
-yielding the constrained transition matrix:
-
-$$
-P = \begin{bmatrix} 
-p_{11} & p_{12} & p_{13} \\
-p_{21} & p_{22} & p_{23} \\
-0      & 0      & 1      \\
-\end{bmatrix}
-$$ {#eq-t-matrix-constrained}
-
-## Stationary model
-
-We estimate three progressively complex stationary models using `nnet::multinom()`
-
-```{r model-fitting}
-#| code-fold: true
-#| code-summary: "Check out my models"
-#| 
-# Baseline model (only gender)
-fit_1 <- nnet::multinom(
-  status ~ Gender, family = multinomial, 
-  data = data_stack, trace = FALSE)
-
-# With procrastination
-fit_2 <- nnet::multinom(
-  status ~ Gender + Total_p, family = multinomial, 
-  data = data_stack, trace = FALSE)
-
-# Full model with previous state
-fit_3 <- nnet::multinom(
-  status ~ Gender + Total_p + status_prev, family = multinomial, 
-  data = data_stack, trace = FALSE)
-```
-
-### Model comparison
-
-We evaluate model improvement using likelihood ratio tests:
-
-$$
-D = -2 \times \ell_{\text{reduced}} - \ell_{\text{full}}
-$$ {#eq-liklihood}
-
-```{r model-comparison}
-anova(fit_1, fit_2, fit_3) |>
-  knitr::kable(caption = "Nested Model Comparison") |>
-  kableExtra::kable_styling(
-    bootstrap_options = c("striped", "hover"),
-    full_width = FALSE)
-```
-
-### Estimated parameters
-
-```{r model-output}
-#| code-fold: true
-#| code-summary: "Check out my code"
-#| 
-broom::tidy(fit_3) |>
-  filter(term != "(Intercept)") |>
-  mutate(
-    y.level = case_when(
-      y.level == "2" ~ "NC - MCI",
-      y.level == "3" ~ "NC - Dementia"),
-    term = case_when(
-      term == "Gender1" ~ "Being female",
-      term == "Total_p" ~ "Procrastination",
-      term == "status_prev2" ~ "Previous state: MCI",
-      term == "status_prev3" ~ "Previous state: Dementia",
-      TRUE ~ term)
-    ) |>
-  mutate(across(c(estimate:p.value), ~ round(x = ., digits = 3)))
-```
-
-### Model visualisation
-
-Let's visualize the predicted transition probabilities from the stationary models. We will show probability curves by procrastination level
-
-```{r}
-#| code-fold: true
-#| code-summary: "Check out my code"
-#| fig-width: 12
-#| label: fig-model-predictions
-#| fig-cap: "Predicted transition probabilities from the stationary model."
-#| 
-## Making a prediction dataset -------------------------------------------------
-pred_data <- expand.grid(
-  Gender = factor(0),
-  status_prev = levels(data_stack$status_prev),
-  Total_p = seq(0, 60, length = 200))
-
-## Plotting predictions
-fig_3 <- pred_data |>
-  modelr::add_predictions(model = fit_3, var = "pred", type = "probs") |>
-  tidy_predictions() |>
-  plot_predictions_stationary()
-
-fig_3
-```
-
-### Estimated transition matrix
-
-Finally, we can compare the observed and predicted transition matrices. We will use the estimated probabilities from the model to fill in the transition matrix.
-
-```{r estimated-matrix}
-# Get all unique states
-names <- c("Normal Cognition", "MCI", "Dementia")
-states <- sort(unique(data_stack$status))
-n_states <- length(states)
-
-# Create empty transition matrix
-transition_matrix_estimated <- matrix(
-  0, nrow = n_states, ncol = n_states,
-  dimnames = list(paste("From", names), paste("To", names))
-  )
-
-# Getting estimated probabilities
-estimated_probs <- expand.grid(
-  Gender = factor(0),
-  Total_p = mean(data_stack$Total_p),
-  status_prev = states) |>
-  modelr::add_predictions(model = fit_3, var = "pred", type = "probs")
-
-# Filling in matrix
-for(i in 1:n_states) {
-  transition_matrix_estimated[i, ] <- estimated_probs$pred[i, ]
-}
-
-transition_matrix_estimated |> round(digits = 3)
-```
-
-Let's plot this and then compare with our observed matrix
-
-```{r compare-matrix}
-#| code-fold: true
-#| code-summary: "Check out my code"
-#| fig-width: 12
-#| label: fig-matrix-comparison
-#| fig-cap: "Comparison of observed and estimated transition matrices."
-
-tran_df_estimate <- transition_matrix_estimated |>
-  as.data.frame() |>
-  mutate(from_state = factor(
-    c("Normal cognition", "MCI", "Dementia"),
-    levels = c("Normal cognition", "MCI", "Dementia")
-  )) |>
-  reshape2::melt(id.vars = "from_state",
-                 variable.name = "to_state",
-                 value.name = "probability") |>
-  mutate(to_state = factor(to_state, levels = rev(levels(to_state))))
-
-fig_4 <- tran_df_estimate |>
-  ggplot(aes(x = from_state, y = to_state, fill = probability)) +
-  geom_tile(color = "white", linewidth = 0.5) +
-  geom_text(
-    aes(label = format(round(probability, 3), nsmall = 3)),
-    size = 4.5, 
-    color = "#212427",
-    fontface = "bold") +
-  colorspace::scale_fill_continuous_diverging(
-    palette = "Blue-Red 3", mid = 0.50, alpha = 0.5, 
-    limits = c(0, 1), name = "Transition \nProbability") +
-  labs(title = "Transition Probabilities Across Cognitive States",
-       subtitle = "Estimated state transitions between assessment waves",
-       x = "Previous State (t - 1)", 
-       y = "Current State (t)") +
-  theme(
-    axis.text = element_text(size = 10),
-    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
-    legend.position = "right",
-    legend.text = element_text(size = 9),
-    panel.grid = element_blank()
-  )
-
-fig_2 + fig_4 + plot_layout(axis_titles = "collect", guides = "collect")
-```
-
-## Non-stationary model
-
-We can also estimate a non-stationary model (@eq-non-stationary) by incorporating time `(wave)` into our model. This model allows for time-varying transition probabilities $p^{(t)}_{ij}$.
-
-We estimate three progressively complex non-stationary models:
-
-- Additive time effects `(fit_4)`: Baseline covariates + wave
-- State-specific time effects `(fit_5)`: Interaction between previous state and wave
-- Full time interactions `(fit_6)`: All covariates interacting with wave
-
-```{r non-stationary}
-#| code-fold: true
-#| code-summary: "Check out my models"
-
-## Small processing to fix wave column
-data_stack <- data_stack |>
-  mutate(wave = case_when(
-    wave == "2016" ~ 1,
-    wave == "2018" ~ 2,
-    wave == "2020" ~ 3,
-    wave == "2022" ~ 4
-  ))
-
-# Model 4: Additive time effects
-fit_4 <- nnet::multinom(
-  status ~ Gender + Total_p + status_prev + wave, family = multinomial, 
-  data = data_stack, trace = FALSE)
-
-# Model 5: State-specific time effects
-fit_5 <- nnet::multinom(
-  status ~ Gender + Total_p + (status_prev * wave), family = multinomial, 
-  data = data_stack, trace = FALSE)
-
-# Model 6: Full time interactions
-fit_6 <- nnet::multinom(
-  status ~ (Gender + Total_p + status_prev) * wave, family = multinomial, 
-  data = data_stack, trace = FALSE)
-```
-
-### Model comparison
-
-Again, we will evaluate model improvement using likelihood ratio tests (@eq-liklihood). However, this time, we will also compare the best fitted non-stationary models `fit_3`.
-
-```{r}
-#| code-fold: true
-#| code-summary: "Check out my code"
-#| layout-nrow: 2
-
-# Fit statistics
-models <- list(
-  "Stationary" = fit_3,
-  "Additive Time" = fit_4,
-  "State-Time Interaction" = fit_5,
-  "Full Interactions" = fit_6
-)
-
-# Create comparison table
-comparison_table <- purrr::map_dfr(models, broom::glance, .id = "Model") |>
-  select(Model:AIC) |>
-  mutate(across(where(is.numeric), \(x) round(x, digits = 1)))
-
-# Likelihood ratio tests
-lr_tests <- anova(fit_3, fit_4, fit_5, fit_6)
-
-# Outputting tables
-knitr::kable(lr_tests, caption = "Likelihood Ratio Tests") |>
-  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
-
-knitr::kable(comparison_table, caption = "Model Fit Statistics") |>
-  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
-```
-
-**Looks like adding in time as an additive term improved things**
-
-### Estimated parameters
-
-Key coefficients from the selected non-stationary model `fit_4`
-
-```{r non-stationary-output}
-#| code-fold: true
-#| code-summary: "Check out my code"
-
-broom::tidy(fit_4) |>
-  filter(term != "(Intercept)") |>
-  mutate(
-    y.level = case_when(
-      y.level == "2" ~ "NC - MCI",
-      y.level == "3" ~ "NC - Dementia"),
-    term = case_when(
-      term == "Gender1" ~ "Being female",
-      term == "Total_p" ~ "Procrastination",
-      term == "status_prev2" ~ "Previous state: MCI",
-      term == "status_prev3" ~ "Previous state: Dementia",
-      term == "wave" ~ "Time",
-      TRUE ~ term)
-    ) |>
-  mutate(across(c(estimate:p.value), ~ round(x = ., digits = 3)))
-```
-
-### Model visualisation
-
-Let's visualize the predicted transition probabilities from the non-stationary model @fig-non-stationary. We will show probability curves by procrastination level
-
-```{r non-stationary-visualise}
-#| code-fold: true
-#| code-summary: "Check our my code"
-#| fig-width: 12
-#| fig-height: 10
-#| label: fig-non-stationary
-#| fig-cap: "Predicted transition probabilities from the non-stationary model."
-#| 
-expand.grid(
-  Gender = factor(0),
-  status_prev = levels(data_stack$status_prev),
-  Total_p = seq(0, 60, length = 200),
-  wave = unique(data_stack$wave)) |>
-  modelr::add_predictions(model = fit_4, var = "pred", type = "probs") |>
-  tidy_predictions() |>
-  mutate(wave = factor(wave)) |>
-  ggplot(aes(x = Total_p, y = prob, color = status_prev)) +
-  geom_line(linewidth = 1) +
-  ggokabeito::scale_color_okabe_ito() +
-  labs(
-    title = "Time-Varying Transition Probabilities",
-    x = "Total Procrastination", 
-    y = "Predicted Probability", color = "Previous State") +
-  facet_grid(wave ~ status, labeller = labeller(
-    wave = function(x) paste("Years since baseline:", x),
-    status = function(x) paste("Transition to:", x)
-  )) +
-  theme(panel.spacing = unit(1, "lines"))
-
-```
-
-### Estimated transition matrices
-
-Let's visualize the estimated time-varying transition matrices @fig-non-stationary-matrix-plot
-
-```{r non-stationary-matrix-visualise}
-#| code-fold: true
-#| code-summary: "Check out my code"
-#| fig-width: 12
-#| label: fig-non-stationary-matrix-plot
-#| fig-cap: "Estimated time-varying transition matrices."
-
-# Create prediction grid for each time point
-time_points <- unique(data_stack$wave)
-state_names <- c("Normal Cognition", "MCI", "Dementia")
-
-# Get matrices for all time points
-time_varying_matrices <- purrr::map(
-  setNames(time_points, paste(time_points, "Years")),
-  ~ get_time_varying_matrix(fit_4, .x)
-)
-
-# Converting into a tidy data frame
-non_stationary_matrices <- purrr::imap_dfr(
-  time_varying_matrices,
-  ~ as.data.frame(.x) |> 
-    tibble::rownames_to_column("From") |> 
-    pivot_longer(-From, names_to = "To", values_to = "probability") |> 
-    mutate(Time = .y),
-  .id = "Time_point"
-) |> 
-  mutate(
-    From = factor(From, levels = state_names),
-    To = factor(To, levels = rev(state_names)),
-    Time_point = forcats::fct_inorder(Time_point)
-  )
-
-# Plotting 
-non_stationary_matrices |>
-ggplot(aes(x = From, y = To, fill = probability)) +
-  geom_tile(color = "white", linewidth = 0.5) +
-  geom_text(
-    aes(label = format(round(probability, 3), nsmall = 3)),
-    size = 4.5, 
-    color = "#212427",
-    fontface = "bold") +
-    colorspace::scale_fill_continuous_diverging(
-    palette = "Blue-Red 3", mid = 0.50, alpha = 0.5, 
-    limits = c(0, 1), name = "Transition \nProbability") +
-  labs(
-    title = "Estimated Time-Varying Transition Matrices",
-    subtitle = "Showing changes in transition probabilities over time",
-    x = "Previous State (t-1)",
-    y = "Next State (t)",
-    fill = "Probability"
-  ) +
-  facet_wrap(~ Time_point, ncol = 3, labeller = labeller(
-    Time_point = function(x){
-      case_when(x == "2 Years" ~ "2018",
-                x == "3 Years" ~ "2020",
-                x == "4 Years" ~ "2022")
-    })) +
-  theme(
-    axis.text = element_text(size = 10),
-    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
-    legend.position = "right",
-    legend.text = element_text(size = 9),
-    panel.grid = element_blank()
-  )
-```
-
-# Matrix distance metrics
-
-Given two matrices $P = (p_{ij})$ and $\hat{P} = (\hat{p}_{ij})$ of size $m \times n$, we define the following distance measures
-
-**Frobenius Norm (Matrix Euclidean Distance)**
-
-$$D_{\text{Frobenius}}(P, \hat{P}) = \|P - \hat{P}\|_F = \sqrt{\sum_{i=1}^{m}\sum_{j=1}^{n} |p_{ij} - \hat{p}_{ij}|^2}$$
-
-- Measures the Euclidean distance in matrix space
-- Sensitive to large differences due to squaring
-- Equivalent to the vector 2-norm of the flattened matrix
-
-**Manhattan Distance (L1 Norm)**
-
-$$D_{\text{Manhattan}}(P, \hat{P}) = \sum_{i=1}^{m}\sum_{j=1}^{n} |p_{ij} - \hat{p}_{ij}|$$
-
-- Sum of absolute differences between corresponding elements
-- Less sensitive to outliers than Frobenius norm
-- Useful when sparse differences are important
-
-**Maximum Difference (Infinity Norm)**
-
-$$D_{\text{Max}}(A, B) = \max_{\substack{1 \leq i \leq m \\ 1 \leq j \leq n}} |a_{ij} - b_{ij}|$$
-
-- Captures the single largest difference between elements
-- Useful for worst-case analysis
-- Ignores the distribution of other differences
-
-**Mean Absolute Difference**
-
-$$D_{\text{MeanAbs}}(P, \hat{P}) = \frac{1}{mn}\sum_{i=1}^{m}\sum_{j=1}^{n} |p_{ij} - \hat{p}_{ij}|$$
-
-- Average absolute difference across all elements
-- Scales with matrix size (unlike Manhattan distance)
-- Easy to interpret as average error per element
-
-**Root Mean Square Error (RMSE)**
-
-$$D_{\text{RMSE}}(P, \hat{P}) = \sqrt{\frac{1}{mn}\sum_{i=1}^{m}\sum_{j=1}^{n} (p_{ij} - \hat{p}_{ij})^2}$$
-
-- Similar to Frobenius but normalized by matrix size
--  Sensitive to large errors due to squaring
-- Common in statistical and machine learning applications
-
-**Correlation-Based Distance**
-
-$$D_{\text{Corr}}(P, \hat{P}) = 1 - \frac{\sum_{i=1}^{m}\sum_{j=1}^{n} (p_{ij} - \bar{P})(\hat{p}_{ij} - \bar{\hat{P}})}{\sqrt{\sum_{i=1}^{m}\sum_{j=1}^{n} (p_{ij} - \bar{P})^2 \sum_{i=1}^{m}\sum_{j=1}^{n} (\hat{p}_{ij} - \bar{\hat{p}})^2}}$$
-
-- Measures linear relationship between matrix elements
-- $\bar{P} = \frac{1}{mn}\sum_{i,j}p_{ij}$ (mean of all elements in $P$)
-- $\bar{\hat{P}} = \frac{1}{mn}\sum_{i,j}\hat{p}_{ij}$ (mean of all elements in $\hat{P}$)
-- Range: [0,2] where 0 = perfect positive correlation
-
-**Kullback-Leibler Divergence**
-
-For matrices where each row sums to 1 ($\sum_j p_{ij} = \sum_j \hat{P}_{ij} = 1$) and $p_{ij}, \hat{p}_{ij} > 0$:
-
-$$
-D_{\text{KL}}(P \parallel \hat{P}) = \sum_{i=1}^{m}\sum_{j=1}^{n} p_{ij} \log\left(\frac{p_{ij}}{\hat{p}_{ij}}\right)
-$$
-
-- Measures information loss when $\hat{P}$ approximates $P$
-- Asymmetric: $D_{\text{KL}}(P \parallel \hat{P}) \neq D_{\text{KL}}(\hat{P} \parallel P)$
-- $0\log0 = 0$ by convention
-- In practice, add $\epsilon > 0$ (e.g., $10^{-10}$) to avoid zeros
-
-## Implementation
-
-Let's calculate the distance metrics between the observed $P$ and estimated transition matrices $\hat{P}$.
-
-```{r matrix-distance}
-p       <- transition_matrix_observed
-p_hat   <- transition_matrix_estimated
-epsilon <- 1e-10
-
-# Creating tibble of distance metrics
-distances <- tibble(
-  Metric = c("Frobenius", "Manhattan", "Max", "MeanAbs", "RMSE", "Correlation", "KL"),
-  Value = c(
-    norm(p - p_hat, type = "F"),
-    sum(abs(p - p_hat)),
-    max(abs(p - p_hat)),
-    mean(abs(p - p_hat)),
-    sqrt(mean((p - p_hat)^2)),
-    1 - cor(c(p), c(p_hat)),
-    sum((p + epsilon) * log((p + epsilon) / (p_hat + epsilon)))
-  )) |>
-  mutate(Value = round(Value, 4))
-
-distances |>
-  mutate(Metric = case_when(
-    Metric == "Frobenius" ~ "Frobenius Distance",
-    Metric == "Manhattan" ~ "Manhattan Distance",
-    Metric == "Max" ~ "Max Difference",
-    Metric == "MeanAbs" ~ "Mean Absolute Difference",
-    Metric == "RMSE" ~ "Root Mean Square Error",
-    Metric == "Correlation" ~ "Correlation Distance",
-    Metric == "KL" ~ "Kullback-Leibler Divergence"
-  )) |>
-  knitr::kable(caption = "Distance based metrics", align = "c") |>
-  kableExtra::kable_styling(
-    bootstrap_options = c("striped", "hover"),
-    full_width = FALSE)
-
-```
-
-## Visualisation
-
-Let's visualize the distance metrics between the observed and estimated transition matrices (@fig-metrics).
-
-```{r matrix-distance-visualise}
-#| code-fold: true
-#| code-summary: "Check out my code"
-#| fig-width: 12
-#| label: fig-metrics
-#| fig-cap: "Distance metrics between observed and estimated transition matrices."
-
-caption = stringr::str_glue(
-  "**RMSE:** Root Mean Squared Error\n
-   **KL:** Kullback-Leibler Divergence\n
-   **Correlation:** 1 - Pearson Correlation Coefficient")
-
-distances |>
-  ggplot(aes(x = reorder(Metric, -Value), y = Value, fill = Metric)) +
-  geom_col(colour = "black") +
-  geom_text(aes(label = Value), vjust = -0.5) +
-  ggokabeito::scale_fill_okabe_ito() +
-  scale_y_continuous(expand = expansion(mult = c(0.075, 0.075))) +
-  labs(
-    title = "Distance Between Transition Matrices",
-    x = "Distance Metric",
-    y = "Value",
-    caption = caption) +
-  theme_minimal() +
-  theme(
-    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
-    axis.title = element_text(face = "bold", size = 12),
-    plot.caption = ggtext::element_markdown(size = 10),
-    legend.position = "none")
-```

### 2025-03-28 index.qmd
First site push

---
title: "Procrastination and Dementia"
author: 
  - name: "Cormac Monaghan"
    url: "https://c-monaghan.github.io/"
    orcid: "https://orcid.org/0000-0001-9012-3060"
    affiliation:
      - "Hamilton Institute, Maynooth University, Ireland"
      - "Department of Psychology, Maynooth University, Ireland"
  - name: "Rafael de Andrade Moral"
    url: "https://rafamoral.github.io/"
    orcid: "https://orcid.org/0000-0002-0875-3563"
    affiliation:
      - "Department of Mathematics and Statistics, Maynooth University, Ireland"
  - name: "Joanna McHugh Power"
    orcid: "https://orcid.org/0000-0002-7387-3107"
    affiliation: 
      - "Department of Psychology, Maynooth University, Ireland"
format: 
  html:
    toc: false
sidebar: false
---

### Abstract

**This will come eventually**

### 2025-03-28 R/conversion.R

@@ -1,0 +1,8 @@
+
+data <- readxl::read_xlsx(here::here("data/lw_data_full.xlsx"))
+
+
+
+write.csv(x = data, file = here::here("data/data.csv"))
+saveRDS(object = data, file = here::here("data/data.rds"))
+haven::write_dta(data = data, path = here::here("data/data.dta"))

### 2025-03-28 analysis/continous-markov.qmd
Updating site

---
title: "Continous Time Markov Model"
format: html
---

# Setting up

## Loading packages

```{r packages}
#| code-fold: true
#| code-summary: "Check our my packages"
#| warning: false
#| message: false

library(dplyr)
library(tidyr)
library(ggplot2)
```

## Loading functions

```{r functions}
#| code-fold: true
#| code-summary: "Check our my functions"

extract_years <- function(data, years, impute = TRUE, cog_total = FALSE, absorbing = TRUE) {
  # Extracts cognitive function data for specified years from a dataset.
  # Converts numeric cognitive status codes (1, 2, 3) into descriptive labels
  # ("Normal Cognition", "MCI", "Dementia") for easier interpretation.
  # Arguments:
  #   - data: The input dataset containing cognitive function data.
  #   - years: A vector of years for which data should be extracted.
  # Returns:
  #   - A dataset with ID and cognitive status columns for the specified years.

  # Create dynamic column names based on the years provided
  cogfunction_cols <- paste0("cogfunction", years)
  cogtotal_cols    <- paste0("cogtot27_imp", years)

  if(cog_total == FALSE) {
  data <- data |>
    # Select only the ID column and cognitive function columns for the specified years
    select(ID, any_of(cogfunction_cols)) |>
    mutate(across(!c(ID), ~ case_when(
      .x == 1 ~ "Normal Cognition",
      .x == 2 ~ "MCI",
      .x == 3 ~ "Dementia",
      TRUE ~ NA_character_  # To handle missing/other cases
    )))
  } else {
    data <- data |>
      # Select only the ID column and cognitive function columns for the specified years
      select(ID, any_of(cogtotal_cols)) |>
      rename_with( .cols = !ID, .fn = ~ stringr::str_replace(
        string = .x,
        pattern = "cogtot27_imp", 
        replacement = "cog_score_"))
  }

  if(impute == TRUE){
    data <- data |>
      tidyr::pivot_longer(
        cols = !ID,
        names_to = "Wave",
        values_to = "Status") |>
      group_by(ID) |>
      tidyr::fill(Status, .direction = "down") |>
      ungroup() |>
      tidyr::pivot_wider(names_from = "Wave", values_from = "Status")
  }

  if(absorbing == TRUE) {
    data <- data |>
      tidyr::pivot_longer(
        cols = !ID,
        names_to = "Wave",
        values_to = "Status") |>
      group_by(ID) |>
      mutate(Status = ifelse(cumany(Status == "Dementia"), "Dementia", Status)) |>
      ungroup() |>
      tidyr::pivot_wider(names_from = "Wave", values_from = "Status")
  }

  return(data)
}

create_transitions <- function(data, absorbing = FALSE){
  # Reshapes data from wide to long format to track cognitive status transitions over time.
  # Calculates the next wave's cognitive status for each individual and creates a transition column.
  # Optionally treats "Dementia" as an absorbing state, meaning once an individual is classified
  # with dementia, their status cannot change in subsequent waves.
  # Arguments:
  #   - data: The dataset containing cognitive status data.
  #   - absorbing: A logical flag indicating whether "Dementia" should be treated 
  #   as an absorbing state.
  # Returns:
  #   - A dataset with transition information, including current and next wave statuses.

  # Reshape the data from wide to long format to track cognitive status over waves
  data <- data |>
    select(ID, starts_with("cogfunction")) |>
    tidyr::pivot_longer(cols = !ID,
                        names_to = "Wave",
                        values_to = "Status") |>
    mutate(Wave = as.factor(stringr::str_replace(Wave, "cogfunction", ""))) |>
    # Arrange by ID and Wave to prepare for transition calculation
    arrange(ID, Wave) |>
    group_by(ID) |>
    # Get the next wave's cognitive status for each person
    mutate(next_wave_status = lead(Status)) |>
    ungroup()

  # We can optionally specify dementia as an absorbing state
  # Once an individual is classified with dementia they cannot be classified
  # with anything else 
  if(absorbing == TRUE) {
    data <- data |>
      group_by(ID) |>
      mutate(
        Status = ifelse(cumany(Status == "Dementia"), "Dementia", Status),
        next_wave_status = ifelse(cumany(Status == "Dementia"), "Dementia", next_wave_status),
        transition = paste(Status, next_wave_status, sep = " to ")
      ) |>
      filter(Wave %in% c(2016, 2018, 2020))
  }

  # Filter out rows where either the current or next status is missing
  data <- data |>
    group_by(ID) |>
    filter(!is.na(Status), !is.na(next_wave_status)) |>
    ungroup() |>
    # Create a new column representing the transition from one status to the next
    mutate(
      transition = paste(Status, next_wave_status, sep = " to "),
      transition = factor(
        transition, 
        levels = c("Normal Cognition to Normal Cognition", "Normal Cognition to MCI",
                   "Normal Cognition to Dementia", "MCI to Normal Cognition", "MCI to MCI",
                   "MCI to Dementia", "Dementia to Dementia") 
        ))

  return(data)
}

calculate_probabilties <- function(data) {
  # Calculates the proportion of each cognitive status transition in the dataset.
  # Counts the occurrences of each transition and computes their relative probabilities.
  # Arguments:
  #   - data: The dataset containing transition information.
  # Returns:
  #   - A dataset with transition probabilities, split into "from" and "to" states.

  # Calculate the proportion of each transition by dividing by the total count
  data |>
    count(transition) |>
    mutate(prop = n / sum(n)) |>
    tidyr::separate(transition, into = c("from", "to"), sep = " to ")
}

transition_matrix <- function(data, longitudinal = FALSE) {
  # Constructs a transition matrix from the calculated transition probabilities.
  # The matrix represents the probability of moving from one cognitive state to another.
  # Arguments:
  #   - data: The dataset containing transition probabilities.
  #   - longitudinal: A TRUE/FALSE statement indicating if the probabilities are from multiple waves
  # Returns:
  #   - A transition matrix with rows representing "from" states and columns 
  #   representing "to" states.

  if(longitudinal == FALSE) {
    # Defining empty matrix matrix
    states <- c("Normal Cognition", "MCI", "Dementia")

    transition_matrix <- matrix(
      0,
      nrow = length(states),
      ncol = length(states),
      dimnames = list(from_state = states, to_state = states))

    # Fill the transition matrix with probabilities
    for (i in 1:nrow(data)) {
      from <- data$from[i]
      to <- data$to[i]
      prob <- data$prop[i]

      transition_matrix[from, to] <- prob
    }
  } else if(longitudinal == TRUE){

    # Aggregate by 'from' and 'to' states to sum transition probabilities
    data_agg <- data |>
      group_by(from, to) |>
      summarise(transition_prob = sum(prop), .groups = "drop")

    transition_matrix <- data_agg |>
      tidyr::pivot_wider(names_from = to, values_from = transition_prob, values_fill = 0) |>
      tibble::column_to_rownames("from") |>
      as.matrix()
  }

  return(transition_matrix)
}

plot_matrix <- function(data, year, ts = FALSE) {
  # Visualizes the transition matrix as a heatmap.
  # Uses color gradients to represent transition probabilities and includes labels for clarity.
  # Arguments:
  #   - data: The transition matrix or dataset to plot.
  #   - year: The year or time period associated with the data.
  #   - ts: A logical flag indicating whether the title should include a time series label.
  # Returns:
  #   - A ggplot object representing the transition probabilities as a heatmap.

  if(ts == TRUE) {
    title <- paste0("Transition Probabilities Across Cognitive States - ", year)
  } else {
    title <- paste0("Transition Probabilities Across Cognitive States (", year, "-2022)")
  }

  # Reshaping data for plotting
  data <- data |>
    as.data.frame(row.names = FALSE) |>
    mutate(from_state = c("Normal Cognition", "MCI", "Dementia")) |>
    reshape2::melt(id.vars = "from_state", variable.name = "to_state", value.name = "probability")

  # Plotting probabilities
  data |>
    ggplot(aes(x = from_state, y = to_state, fill = probability)) +
    geom_tile() +
    scale_fill_viridis_c(alpha = 0.8) +
    labs(title = title,
         x = "To", y = "From", fill = "Prob") +
    geom_text(aes(label = round(probability, digits = 3)), size = 5) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
          axis.title = element_text(size = 10),
          axis.text = element_text(size = 8),
          panel.grid = element_blank()
    )
}

normalise <- function(x) {
  # Normalizes a matrix so that each row sums to 1.
  # Arguments:
  #   - x: The input matrix to normalize.
  # Returns:
  #   - A normalized matrix where each row sums to 1.
  x / rowSums(x)
}

reshape_matrix <- function(matrix, wave) {
  # Reshapes a transition matrix into a long format for easier analysis and visualization.
  # Adds a "Wave" column to identify the time period associated with the matrix.
  # Arguments:
  #   - matrix: The transition matrix to reshape.
  #   - wave: The wave or time period associated with the matrix.
  # Returns:
  #   - A long-format dataset with "from", "to", and "transition_prob" columns.

  matrix |>
    as.data.frame() |>
    tibble::rownames_to_column(var = "from") |>
    tidyr::pivot_longer(cols = -from, names_to = "to", values_to = "transition_prob") |>
    mutate(Wave = wave)
}

plot_transition_wave <- function(data, wave) {
  # Plots transition probabilities for a specific wave as a heatmap.
  # Filters the data to include only the main transitions and adds labels for clarity.
  # Arguments:
  #   - data: The dataset containing transition probabilities.
  #   - wave: The wave or time period to plot.
  # Returns:
  #   - A ggplot object representing the transition probabilities for the 
  #   specified wave.

  # Main transitions per wave
  filter_data <- data.frame(
    from = c("Normal Cognition", "Normal Cognition", "Normal Cognition",  "MCI", "MCI", "MCI", "Dementia", "Dementia", "Dementia"),
    to = c("Normal Cognition", "MCI", "Dementia", "Normal Cognition", "MCI", "Dementia", "Normal Cognition", "MCI", "Dementia")
  )

  data |>
    semi_join(filter_data, by = c("from", "to")) |>
    filter(Wave == wave) |>
    ggplot(aes(x = to, y = from, fill = transition_prob)) +
    geom_tile() +
    geom_text(aes(label = transition_prob), size = 5) +
    scale_fill_viridis_c(alpha = 0.8)+
    labs(title = paste0("Transition Probabilities Across Cognitive States (", wave, "-", wave + 2, ")"),
         x = "To", y = "From", fill = "Prob") +
    theme_light() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
      axis.title = element_text(size = 10),
      axis.text = element_text(size = 8),
      panel.grid = element_blank(),
      panel.border = element_blank()
    )
}

backtrack_age <- function(data, base_year, current_year) {
  # Back calculates an individual's age by 2-year intervals
  # Arguments:
  #   - data: The dataset containing the age variable and a 'Wave' column indicating the year of the wave
  #   - current_year: The year for which the age is measured (e.g., 2022)
  # Returns:
  #   - A dataset where the age is subtracted by intervals of 2 for each respective year

  # Map Wave values (1-10) to their corresponding years
  wave_to_year <- function(wave) {
    base_year + (as.numeric(wave) - 1) * 2
  }

  data |>
    group_by(ID) |>
    mutate(Age = Age - (current_year - wave_to_year(Wave))) |>
    ungroup()
}

pivot_data <- function(data, cog_total = FALSE) {
  # Reshapes the input dataset by pivoting cardiovascular, depression, and cognitive data
  # into long format for analysis. Handles two scenarios: one for cognitive status and
  # another for total cognitive scores. Combines the pivoted data into a final dataset
  # for further analysis.
  # Arguments:
  #   - data: The input dataset containing cardiovascular, depression, and cognitive data.
  #   - cog_total: A logical flag indicating whether to pivot total cognitive scores (TRUE)
  #               or cognitive status (FALSE).
  # Returns:
  #   - A reshaped dataset in long format with combined cardiovascular, depression, and
  #     cognitive data, ready for analysis.

  # Pivot cardio dataset
  data_cardio <- data |>
    tidyr::pivot_longer(
      cols = starts_with("Cardio_risk"),
      names_to = "Cardio_Year",
      values_to = "Cardio_risk"
    ) |>
    select(!any_of(c(
      "Cardio_Year", 
      paste0("cogfunction", 2016:2022), 
      paste0("cog_score_", 2016:2022)
      )))

  # Pivot dementia dataset
  data_dep <- data |>
    tidyr::pivot_longer(
      cols = starts_with("Total_dep"),
      names_to = "Dep_Year",
      values_to = "Depression_score"
    ) |>
    select(ID, Depression_score)

  if(cog_total == FALSE) {

    # Pivot cognition dataset
    data_cog <- data |>
      tidyr::pivot_longer(
        cols = starts_with("cogfunction"),
        names_to = "Wave",
        values_to = "Status"
      ) |>
      select(!any_of(c(
        paste0("Cardio_risk_", 16:22), 
        paste0("Total_dep_20", 16:22))
        ))

  # Adding cardio data to cognition data
  combine_data <- data_cog |>
    mutate(Cardio_risk = data_cardio$Cardio_risk, .after = Education_tri) |>
    mutate(Depression = data_dep$Depression_score, .after = Cardio_risk)

  # Producing final dataset
  final_data <- combine_data |>
    mutate(
      Wave = stringr::str_replace(Wave, "cogfunction", "HRS_"),
      Wave = factor(
        Wave,
        levels = c("HRS_2016", "HRS_2018", "HRS_2020", "HRS_2022"),
        labels = c(1, 2, 3, 4)
      )) |>
    group_by(ID) |>
    mutate(
      Status = ifelse(cumany(Status == "Dementia"), "Dementia", Status),
      Status = factor(
        Status,
        levels = c("Normal Cognition", "MCI", "Dementia"),
        labels = c(1, 2, 3)
      )) |>
    ungroup() |>
    mutate(
      Gender = factor(Gender, levels = c(0, 1)),
      Education_tri = factor(Education_tri, levels = c(0, 1, 2)),
      Wave = as.numeric(Wave),
      Status = as.numeric(Status)
    )
  } else {

    # Pivot cog total data
    data_score <- data |>
      tidyr::pivot_longer(
        cols = starts_with("cog_score"),
        names_to = "Wave",
        values_to = "Score"
      ) |>
      mutate(
        Wave = stringr::str_replace(Wave, "cog_score_", ""),
        Wave = factor(
          Wave, 
          levels = c("2016", "2018", "2020", "2022"),
          labels = c(1, 2, 3, 4)),

        Gender = as.factor(Gender),
        Education_tri = as.factor(Education_tri)
      ) |>
      select(!any_of(c(
        paste0("Cardio_risk_", 16:22), 
        paste0("Total_dep_20", 16:22))
      ))

    # Combine cardio and cog total data
    final_data <- data_score |>
      mutate(Cardio_risk = data_cardio$Cardio_risk, .after = Education_tri) |>
      mutate(Depression = data_dep$Depression_score, .after = Cardio_risk)
    }

  return(final_data)
}

gentleman_test <- function(data, model) {
  # Compares observed and expected prevalence of cognitive states (Normal Cognition, MCI, Dementia)
  # over time. Reshapes the observed and expected data into a long format and generates a
  # faceted line plot to visualize the comparison.
  # Arguments:
  #   - data: A list containing observed and expected prevalence data.
  #   - model: A string representing the model name, used in the plot title.
  # Returns:
  #   - A ggplot object displaying the observed vs. expected prevalence of cognitive states
  #     over time.

  # Reshaping observed prevalence
  do1 <- as_tibble(row.names(data$Observed)) |>
    rename(time = value) |>
    mutate(time = as.numeric(time))

  do2 <- as_tibble(data$Observed) |> 
    mutate(type = "observed")

  do <- cbind(do1, do2) |>
    select(-Total) |>
    tidyr::gather(state, number, -c(time, type)) |>
    mutate(state = case_when(state == "State 1" ~ "Normal Cognition",
                             state == "State 2" ~ "MCI",
                             state == "State 3" ~ "Dementia"))

  # Reshaping expected prevalence
  de1 <- as_tibble(row.names(data$Expected)) |>
    rename(time = value) |>
    mutate(time = as.numeric(time))

  de2 <- as_tibble(data$Expected) |>
    mutate(type = "expected")

  de <- cbind(de1,de2) |>
    select(-Total) |>
    tidyr::gather(state, number, -c(time, type))

  # Turn into single data frame and plotting
  rbind(do, de) |>
    mutate(type = factor(type),
           state = factor(state, levels = c("Normal Cognition", "MCI", "Dementia")),
           time = round(time,3)) |>
    group_by(state) |>
    ggplot(aes(x = time, y = number, colour = type)) +
    geom_line(linewidth = 1) +
    labs(title = paste0("Observed vs Fitted Values: ", model), x = "Time", y = "") +
    facet_wrap(~ state) +
    theme_bw() +
    ggeasy::easy_remove_legend_title()
}

process_hr <- function(data, fix_upper = TRUE) {
  # Processes hazard ratio (HR) data by cleaning and reshaping it for visualization.
  # Converts covariate names into descriptive labels and ensures proper formatting
  # of transitions and confidence intervals. Optionally fixes upper confidence intervals
  # that exceed a specified threshold.
  # Arguments:
  #   - data: The input dataset containing hazard ratio data.
  #   - fix_upper: A logical flag indicating whether to fix upper confidence intervals
  #                that exceed a threshold (default: TRUE).
  # Returns:
  #   - A cleaned and reshaped dataset ready for hazard ratio visualization.

  data <- do.call(rbind.data.frame, data) |>
    tibble::rownames_to_column(var = "Covariate") |>
    tidyr::separate_wider_delim(cols = Covariate, 
                                delim = ".", 
                                names = c("Covariate", "Transition")) |>
    mutate(Covariate = case_when(
      Covariate == "Gender1" ~ "Female",
      Covariate == "Education_tri1" ~ "High School",
      Covariate == "Education_tri2" ~ "College",
      Covariate == "Cardio_risk" ~ "Cardiovascular Risk",
      Covariate == "Depression" ~ "Depression",
      Covariate == "Total_p" ~ "Total Procrastination",
      Covariate == "Depression:Total_p" ~ "Depression x Procrastination",
      TRUE ~ Covariate),

      Covariate = factor(
        Covariate, 
        levels = c("Female", "Age", "High School", "College", 
                   "Cardiovascular Risk", "Depression", 
                   "Total Procrastination", "Depression x Procrastination"))
      ) |>
    mutate(Transition = factor(
      Transition, 
      levels = c("Normal Cognition - MCI", 
                 "Normal Cognition - Dementia",
                 "MCI - Normal Cognition",
                 "MCI - Dementia")))

  if(fix_upper == TRUE){
    data <- data |>
      mutate(U = case_when(
        U >= 5 & Transition == "Normal Cognition - Dementia" ~ HR,
        # U >= 5 & HR < 1 ~ 1,
        TRUE ~ U
      ))
  }

  return(data)
}

calculate_p_value <- function(data) {
  # Calculates p-values for hazard ratios (HR) based on log-transformed HR values and
  # their confidence intervals. Adds the calculated p-values to the input dataset.
  # Arguments:
  #   - data: The input dataset containing hazard ratios and confidence intervals.
  # Returns:
  #   - The input dataset with an additional column for p-values

  log_HR <- log(data$HR)
  log_lower_CI <- log(data$L)
  log_upper_CI <- log(data$U)

  SE <- (log_upper_CI - log_lower_CI) / (2 * 1.96)
  Z <- log_HR / SE
  p_value <- 2 * (1 - pnorm(abs(Z)))

  data$pval <- p_value |> round(digits = 3)

  return(data)
}

plot_hr <- function(data) {
  # Generates a forest plot of hazard ratios (HR) for cognitive transitions, including
  # confidence intervals and significance stars. Customizes the plot based on covariate
  # labels and ensures proper formatting for visualization.
  # Arguments:
  #   - data: The input dataset containing hazard ratios, confidence intervals, and p-values.
  # Returns:
  #   - A ggplot object displaying hazard ratios for cognitive transitions, faceted by covariate.

  # Making a plot customization
  labeller <- c(
    Female = "Being female vs. Male",
    `High School` = "Having a highshool education vs. No education",
    College = "Having a further education vs. No education",
    `Cardiovascular Risk` = "Total number of cardiovascular risk factors",
    Depression = "Depression",
    Total_p = "Procrastination",
    `Depression x Procrastination` = "Depression x Procrastination"
  )

  # Add a column for significance stars
  data <- data |>
    mutate(
      stars = case_when(
        pval < 0.001 ~ "***",
        pval < 0.01 ~ "**",
        pval < 0.05 ~ "*",
        pval < 0.10 ~ "+",
        TRUE ~ "" 
      ))

  # Plotting  
  haz_plot <- data |>
    ggplot(aes(x = HR, y = forcats::fct_rev(Transition), colour = HR > 1)) +
    geom_point(size = 4) +
    geom_linerange(aes(xmin = L, xmax = U), linewidth = 1.5, alpha = 0.5) +
    geom_vline(xintercept = 1, colour = "gray75", linewidth = 0.75, linetype = "dashed") +
    geom_text(aes(label = stars), vjust = -0.5, hjust = 0.5, size = 5, color = "black") +
    labs(title = "Hazard Ratio for dementia transitions",
         x = "Hazard Ratio (95% Confidence Interval)", 
         y = "") +
    scale_color_manual(values = c("#E69F00", "#56B4E9"), guide = "none") +
    guides(colour = "none") +
    theme_bw() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
      axis.title.x = element_text(size = 12),
      strip.text = element_text(size = 10)
      )

  if(nrow(data) == 32){
  # Centering the bottom facet because I am pedantic
    haz_plot <- haz_plot + ggh4x::facet_manual(
    ~ Covariate,
    scales = "free_x",
    labeller = labeller(Covariate = labeller),
    design = c(
    "
    AABBCC
    DDEEFF
    #GGHH#
    "
    ))
  } else {
    # Centering the bottom facet because I am pedantic
    haz_plot <- haz_plot + ggh4x::facet_manual(
      ~ Covariate,
      scales = "free_x",
      labeller = labeller(Covariate = labeller),
      design = c(
      "
      AABBCC
      DDEEFF
      ##GG##
      "
      ))
  }

  return(haz_plot)
}
```

## Reading in data

```{r data}
set.seed(34521)

data <- read.csv(here::here("analysis/data/data.csv")) 
```

# Markov Modelling

In this section, we apply Markov modelling to analyze the transitions between cognitive states (Normal Cognition, Mild Cognitive Impairment [MCI], and Dementia) using longitudinal data from the Health and Retirement Study (HRS) from 2016 to 2022. By modeling these transitions, we aim to understand the probabilities of moving between different cognitive states over time.

## Data preparation

The first step is to prepare the transition dataset. We start by transforming the data to create a record of transitions from one cognitive state to another between consecutive waves.

Once the transitions are identified, we calculate the probabilities of each state-to-state transition. These probabilities are computed by counting the occurrences of each transition and dividing by the total number of transitions.

```{r transition-probability}
tran_prob <- data |>
  extract_years(years = seq(2016, 2022, by = 2)) |>
  create_transitions(absorbing = TRUE) |>
  calculate_probabilties()

tran_prob |>
  mutate(prop = round(prop, digits = 3)) |>
  head(n = 9) |>
  gt::gt() |>
  gtExtras::gt_theme_538() |>
  gt::cols_align(align = "center") |>
  gt::tab_header(title = "Overall Transition Probabilties (2016 - 2022)") |>
  gt::tab_options(table.width = gt::pct(100)) |>
  gt::opt_align_table_header(align = "center")
```

## Creating transition matrix

The probability distribution of transitions from one state to another can be represented into a transition matrix $P = (p_{ij})_{i,j}$ where each element of position $(i, j)$ represents the transition probability $p_{ij}$.

$$
P = \begin{bmatrix} 
p_{11} & p_{12} & p_{13} \\
p_{21} & p_{22} & p_{23} \\
p_{31} & p_{32} & p_{33} \\
\end{bmatrix}
$$

```{r transition-matrix}
tran_matrix <- tran_prob |>
  transition_matrix() |>
  normalise() |>
  round(digits = 3)

tran_matrix
```

This code creates a $3 \times 3$ matrix (for the three states: Normal Cognition, MCI, and Dementia) and populates it with the transition probabilities.

## Visualising the matrix

To make the transition matrix more interpretable we visualize the probabilities using a heat map (@fig-transition-matrix)

```{r transition-matrix-plot}
#| fig-width: 8
#| label: fig-transition-matrix
#| fig-cap: Heat map of transition probabilities (2016 - 2022)

plot_matrix(data = t(tran_matrix), year = 2016)
```

We can also create and plot transition matrices for each wave separately allowing us to see how these transition probabilities change across multiple waves of data collection

```{r transition-matrix-longitudinal}
tran_prob_long <- data |>
  extract_years(years = seq(2016, 2022, by = 2)) |>
  create_transitions(absorbing = TRUE) |>
  group_by(Wave) |>
  calculate_probabilties() |>
  group_map(~ transition_matrix(.x, longitudinal = TRUE), .keep = TRUE) |>
  lapply(function(mat) {
    mat[c("Normal Cognition", "MCI", "Dementia"), c("Normal Cognition", "MCI", "Dementia")] |>
        normalise()
    }) |>
  purrr::map2_df(.y = c(seq(2016, 2020, by = 2)), .f = reshape_matrix) |>
  mutate(
    transition_prob = round(transition_prob, digits = 3),
    from = factor(from, levels = c("Normal Cognition", "MCI", "Dementia"))
  )

DT::datatable(
  tran_prob_long,
  filter = "top",
  colnames = c("Probability" = "transition_prob"),
  options = list(pageLength = 6, autoWidth = TRUE)
)
```

```{r transition-matrix-longitudinal-plot, fig.width=8, fig.show="animate", ffmpeg.format="gif", dev="png"}
for(wave in c(seq(2016, 2020, by = 2))){
  print(plot_transition_wave(data = tran_prob_long, wave = wave))
}
```

# Multi-State Markov Model

Now that we have an idea of the disease progression across time we are interested in the effect of different covariates on these transitions. For this, we can fit a multi-state Markov model. Multi-state models (MSMs) provide a comprehensive view of the disease process and make risk-factors analysis and long-term predictions more easily. The conditional distribution of the status of an individual participant at an arbitrary examination given their status at previous examinations was assumed to have the **Markov property**. That is the distribution of the forthcoming state $X_{n+1}$ depends only on the current state $X_n$ and doesn't depend on the previous ones $X_{n-1}, X_{n-2}, \dots, X_1$.

$$Pr(X_{n+1} = x_{n+1} \; \vert \; X_1 = x_1, X_2 = x_2, \dots, X_n = x_n) = Pr(X_{n+1} = x_{n+1} \; \vert \; X_n = x_n)$$

## Data preperation

For our model we are going to look at several different covariates:

- **Gender**: $0 = \text{Male;} \; 1 = \text{Female}$ 
- **Age**: $\text{Range} \; = 50 - 91$ 
- **Education**: $0 = \text{No Degree;} \; 1 = \text{High School Degree;} \; 2 = \text{Further Education}$ 
- **Number of cardiovascular risk factors**: $\text{Range} \; = 0 - 4$
- **Procrastination**: $\text{Range} \; = 4 - 60$

```{r markov-prep}
cols <- c(
  "ID", "Gender", "Age", "Education_tri", 
  paste0("Cardio_risk_", seq(16, 22, by = 2)), 
  paste0("Total_dep_20", seq(16, 22, by = 2)), "Total_p")

markov_data <- data |>
  extract_years(seq(2016, 2022, by = 2)) |>
  na.omit() |>
  inner_join(data[, cols], by = "ID") |>
  pivot_data() |>
  backtrack_age(base_year = 2016, current_year = 2022) |>
  group_by(ID) |>
  mutate(
    Age = min(Age), # Baseline age
    Cardio_risk = Cardio_risk[Wave == 1],
    Depression = Depression[Wave == 1]
    ) |>
  ungroup() |>
  filter(Age >= 50) # Only looking at older adults

DT::datatable(
  markov_data,
  filter = "top",
  options = list(pageLength = 6, autoWidth = TRUE)
)
```

Let's look at a state table to visualize the amount of transitions over time

```{r state-table}
markov_data |>
  select(ID, Wave, Status) |>
  mutate(Status = case_when(
    Status == 1 ~ "Normal Cognition", 
    Status == 2 ~ "MCI", 
    Status == 3 ~ "Dementia")) |>
  msm::statetable.msm(state = Status, subject = ID)
```

## Modelling

### Defining a q-matrix

The Q matrix (also known as the transition intensity matrix or generator matrix) defines the rates at which transitions occur between states in the model.

It is a square matrix where:

- Each row corresponds to a **from** state in the model.
- Each column corresponds to a **to** state in the model.
- The off-diagonal elements $q_{ij}$ represent the transition rate from state $i$ to state $j$.

For a model with $n$ states, the Q matrix has the following structure

$$
Q = \begin{bmatrix} 
q_{11} & q_{12} & \cdots & q_{1n} \\
q_{21} & q_{22} & \cdots & q_{2n} \\
\vdots & \vdots & \ddots &\vdots  \\
q_{n1} & q_{n2} & \cdots & q_{nn}
\end{bmatrix}
$$

```{r q-matrix}
# Creating transition matrix
# Define possible transitions
states <- c("Normal Cognition", "MCI", "Dementia")

# Theoretical transition probabilities
q_matrix <- rbind(
  c(0.75, 0.20, 0.05),  # From Normal, can transition to MCI or Dementia
  c(0.25, 0.50, 0.25),  # From MCI, can transition to Normal or Dementia
  c(0.00, 0.00, 1.00)         # From Dementia, no transitions (absorbing state)
)

rownames(q_matrix) <- states
colnames(q_matrix) <- states

q_matrix
```

## Fitting

We will fit three different multi-state models

1. A model with no covariates
2. A multivariate model with covariates acting on all transitions
3. A multivariate model with covariates (and interactions) acting on all transitions
4. A multivariate model with covariates acting on only NC <-> MCI & MCI - Dementia
5. A multivariate model with covariates (and interactions) acting on only NC <-> MCI & MCI - Dementia

```{r model-fitting}
#| warning: false

# Fitting model with no covariates
model_1 <- msm::msm(
  formula = Status ~ Wave,
  subject = ID,
  data = markov_data,
  qmatrix = q_matrix,
  obstype = 1,
  method = "BFGS", 
  gen.inits = TRUE,
  control = list(
    fnscale = 5000, 
    maxit = 2000 # To reach convergence
  ))

# Multivariate model: covariates acting on all transitions
model_2 <- msm::msm(
  formula = Status ~ Wave,
  subject = ID,
  data = markov_data,
  qmatrix = q_matrix,
  covariates = ~ Gender + Age + Education_tri + Cardio_risk + Depression + Total_p,
  obstype = 1,
  gen.inits = TRUE,
  method = "BFGS",
  control = list(
    fnscale = 6000, 
    maxit = 2000 # To reach convergence
  ))

# Multivariate model: covariates acting on all transitions (with interaction)
model_3 <- msm::msm(
  formula = Status ~ Wave,
  subject = ID,
  data = markov_data,
  qmatrix = q_matrix,
  covariates = ~ Gender + Age + Education_tri + Cardio_risk + (Depression * Total_p),
  obstype = 1,
  gen.inits = TRUE,
  method = "BFGS",
  control = list(
    fnscale = 5000, 
    maxit = 2000 # To reach convergence
  ))

# Multivariate model: covariates acting on NC <-> MCI and MCI -> D
model_4 <- msm::msm(
  formula = Status ~ Wave,
  subject = ID,
  data = markov_data,
  qmatrix = q_matrix,
  covariates = list(
    "1-2" = ~ Gender + Age + Education_tri + Cardio_risk + Depression + Total_p,
    "2-1" = ~ Gender + Age + Education_tri + Cardio_risk + Depression + Total_p,
    "2-3" = ~ Gender + Age + Education_tri + Cardio_risk + Depression + Total_p
  ),
  obstype = 1,
  gen.inits = TRUE,
  method = "BFGS",
  control = list(
    fnscale = 5000, 
    maxit = 2000 # To reach convergence
  ))

# Multivariate model: covariates acting on NC <-> MCI and MCI -> D (with interaction)
model_5 <- msm::msm(
  formula = Status ~ Wave,
  subject = ID,
  data = markov_data,
  qmatrix = q_matrix,
  covariates = list(
    "1-2" = ~ Gender + Age + Education_tri + Cardio_risk + (Depression * Total_p),
    "2-1" = ~ Gender + Age + Education_tri + Cardio_risk + (Depression * Total_p),
    "2-3" = ~ Gender + Age + Education_tri + Cardio_risk + (Depression * Total_p)
  ),
  obstype = 1,
  gen.inits = TRUE,
  method = "BFGS",
  control = list(
    fnscale = 5000, 
    maxit = 2000 # To reach convergence
  ))

```

### Model Checking

After fitting each of our models we can extract some goodness of fit criteria

```{r model-goodness}
log_l <- c(logLik(model_1), 
           logLik(model_2), 
           logLik(model_3), 
           logLik(model_4), 
           logLik(model_5))

aic <- AIC(model_1, model_2, model_3, model_4, model_5)

# Very small difference between model 2 and 3 (makes sense)
cbind(log_l, aic) |>
  tibble::rownames_to_column("Model") |>
  mutate(Model = stringr::str_remove(Model, "model_"))

# Likelihood ratio test
lmtest::lrtest(model_1, model_2, model_3, model_4, model_5)
```

We can also look at a measure of how well the model fits the data using a test proposed by [@gentleman1994]. The function plots the observed numbers of individuals occupying a state at a series of times against forecasts from the fitted model, for each state (@fig-gentleman).

```{r gentleman_test}
#| fig-width: 12
#| fig-height: 12
#| label: fig-gentleman
#| fig-cap: Fitted values versus observed values

# Plotting fitted vs. observed values
g1 <- gentleman_test(data = msm::prevalence.msm(model_1), 
                     model = "Model 1")
g2 <- gentleman_test(data = msm::prevalence.msm(model_2), 
                     model = "Model 2")
g3 <- gentleman_test(data = msm::prevalence.msm(model_3), 
                     model = "Model 3")
g4 <- gentleman_test(data = msm::prevalence.msm(model_4), 
                     model = "Model 4")
g5 <- gentleman_test(data = msm::prevalence.msm(model_5), 
                     model = "Model 5")

ggpubr::ggarrange(g1, g2, g3, g4, g5, 
                  common.legend = TRUE, 
                  nrow = 5,
                  heights = c(2, 2, 2, 2, 2), 
                  legend = "bottom")
```

### Hazard ratios

We now present hazard ratios for each of our fitted models. These ratios represent how the instantaneous risk of making a particular transition is modified by the covariate.

::: {.panel-tabset}

## Model 2

@fig-hazard-ratio-1 presents hazard ratios from model 2 (our multivariate model with covariates acting on all transitions) for each covariate.

```{r}
#| fig-width: 12
#| fig-height: 12
#| label: fig-hazard-ratio-1
#| fig-cap: Ouputted hazard ratios from Markov model. Green bars represent a decreased risk of transition while red bars indicate an increased risk of transition.

# Plotting hazard ratios
model_2 |>
  msm::hazard.msm() |>
  process_hr() |>
  calculate_p_value() |>
  plot_hr()
```

## Model 3

@fig-hazard-ratio-2 presents hazard ratios from model 3 (our multivariate model with covariates **(and interactions)** acting on all transitions) for each covariate.

```{r}
#| fig-width: 12
#| fig-height: 12
#| label: fig-hazard-ratio-2
#| fig-cap: Ouputted hazard ratios from Markov model. Green bars represent a decreased risk of transition while red bars indicate an increased risk of transition.

# Plotting hazard ratios
model_3 |>
  msm::hazard.msm() |>
  process_hr() |>
  calculate_p_value() |>
  plot_hr()
```

## Model 4

@fig-hazard-ratio-3 presents hazard ratios for model 4 (our multivariate model with covariates acting on only NC <-> MCI & MCI -> Dementia) for each covariate

```{r}
#| fig-width: 12
#| fig-height: 12
#| label: fig-hazard-ratio-3
#| fig-cap: Ouputted hazard ratios from Markov model. Green bars represent a decreased risk of transition while red bars indicate an increased risk of transition.

# Plotting hazard ratios
model_4 |>
  msm::hazard.msm() |>
  process_hr() |>
  calculate_p_value() |>
  plot_hr()
```

## Model 5

@fig-hazard-ratio-4 presents hazard ratios for model 5 (our multivariate model with covariates **(and interactions)** acting on only NC <-> MCI & MCI -> Dementia) for each covariate

```{r}
#| fig-width: 12
#| fig-height: 12
#| label: fig-hazard-ratio-4
#| fig-cap: Ouputted hazard ratios from Markov model. Green bars represent a decreased risk of transition while red bars indicate an increased risk of transition.

# Plotting hazard ratios
model_5 |>
  msm::hazard.msm() |>
  process_hr() |>
  calculate_p_value() |>
  plot_hr()
```

:::

### 2025-03-28 analysis/data-exploration.qmd
Updating site

---
title: "Exploratory Data Analysis"
---

# Setting up {.unnumbered}

## Loading packages {.unnumbered}

```{r packages}
#| code-fold: true
#| code-summary: "Check our my packages"
#| warning: false
#| message: false

library(dplyr)
library(tidyr)
library(ggplot2)
library(ggalluvial)
```

## Loading functions {.unnumbered}

```{r functions}
#| code-fold: true
#| code-summary: "Check our my functions"

count_transitions <- function(data, years, absorbing = FALSE) {
  # Counts the occurrences of cognitive status transitions across specified 
  # years.
  # Converts numeric cognitive status codes (1, 2, 3) into descriptive labels
  # ("Normal Cognition", "MCI", "Dementia") and handles missing or unexpected values.
  # Aggregates the data to count transitions and reshapes it into a long format for analysis.
  # Arguments:
  #    - data: The input dataset containing cognitive function data.
  #    - years: A vector of years for which transitions should be counted.
  # Returns:
  #   - A dataset with counts of cognitive status transitions, including
  #   "Missing" and "Other" categories.

  # Create dynamic column names based on the years provided
  cogfunction_cols <- paste0("cogfunction", years)

  data <- data |>
    select(ID, all_of(cogfunction_cols)) |>
    na.omit() |>
    tidyr::pivot_longer(
      cols = !ID,
      names_to = "Wave",
      values_to = "Classification"
    ) |>
    mutate(
      Wave = stringr::str_replace(Wave, "cogfunction", "HRS "),
      Wave = factor(Wave, levels = c(paste0("HRS ", years))),

      Classification = case_when(
        Classification == 1 ~ "Normal Cognition",
        Classification == 2 ~ "MCI",
        Classification == 3 ~ "Dementia")
    )

  if(absorbing == TRUE){
    data <- data |>
      group_by(ID) |>
      mutate(
        Classification = ifelse(
          cumany(Classification == "Dementia"), "Dementia", Classification)) |>
      ungroup()
  }

  data <- data |>
    mutate(Classification = factor(
      Classification,
      levels = c("Normal Cognition", "MCI", "Dementia")))

  return(data)
}

plot_cognitive_scores <- function(data, year) {
  # Plots the distribution of cognitive scores over a range of years.
  # Converts the data from wide to long format and generates histograms for each wave.
  # Arguments:
  #   - data: The dataset containing cognitive scores.
  #   - year: The starting year for the range of cognitive scores to plot.
  # Returns:
  #   - A ggplot object showing the distribution of cognitive scores across waves.
  # Plotting dementia transitions
  data |>
    select(ID, any_of(paste0("cogtot27_imp", year:2022))) |>
    na.omit() |>
    tidyr::pivot_longer(cols = !ID,
                        names_to = "Wave",
                        values_to = "Score") |>
    mutate(Wave = as.double(stringr::str_replace(Wave, "cogtot27_imp", ""))) |>
    ggplot(aes(x = Score)) +
    geom_histogram(fill = "#56B4E9",
                   colour = "black",
                   alpha = 0.5) +
    scale_x_continuous(breaks = seq(0, 27, by = 3)) +
    labs(
      title = paste0("Distribution of cognitive scores ", year, "- 2022"),
      x = "Cognitive Score", y = "") +
    facet_wrap( ~ Wave)
}

plot_transitions <- function(data, size) {
  # Visualizes cognitive status transitions over time using an alluvial plot.
  # Represents the flow of individuals between cognitive states across waves.
  # Arguments:
  #   - data: The dataset containing transition data.
  #   - size: The font size for labels in the plot.
  # Returns:
  #   - A ggplot object showing transitions between cognitive states over time.

  require(ggalluvial)

  data |>
    ggplot(aes(x = Wave, y = n, stratum = Classification, 
               fill = Classification, alluvium = ID
    )) +
    geom_stratum(alpha = 0.5, width = 0.5) +
    geom_flow(width = 0.5) +
    geom_text(
      stat = "stratum",
      aes(label = stringr::str_wrap(Classification, width = 10)),
      hjust = 0.5,
      vjust = 0.5,
      size = size
    ) +
    labs(title = "Dementia transitions across time", 
         x = "", y = "Frequency") +
    scale_fill_viridis_d(direction = -1) +
    theme(axis.text = element_text(size = 10), 
          text = element_text(size = 10)) +
    ggeasy::easy_remove_legend()
}
```

## Setting theme {.unnumbered}

```{r theme}
#| code-fold: true
#| code-summary: "Check our my theme"

colour = "#212427"

theme_set(
theme_minimal() +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold", colour = colour),
    plot.subtitle = element_text(hjust = 0.5, size = 10, colour = colour),
    axis.title = element_text(size = 10, face = "bold", colour = colour),
    strip.text = element_text(size = 10, face = "bold", colour = colour)
    ))
```

## Reading in data {.unnumbered}

```{r data}
data <- read.csv(here::here("analysis/data/data.csv")) 
```

# Dementia Dataset

Our updated dataset contains 13 columns of cognitive classifications spanning from 1996 to 2022. These columns reflect dementia status across each wave, using the same criteria and structure as the original Langa-Weir dataset, now extended to include the most recent 2022 wave.

```{r count-transitions}
dementia_data <- data |>
  count_transitions(years = seq(1996, 2022, by = 2))

glimpse(dementia_data)
```

## Missing data

Since the participant pool is drawn from the 2020 wave of the HRS data, any analysis that looks back at previous waves will inevitably encounter missing data. Not all participants were included in earlier waves, leading to gaps in the data. @fig-missing-data displays the frequency of missing cognitive test data across each HRS wave.

```{r missing-data}
#| warning: false
#| fig-width: 8
#| label: fig-missing-data
#| fig-cap: Occurance of missing data across HRS waves (participants were gathered from the 2020 HRS wave [in red])

# Dynamic colours
colours <- c(rep("#2e2e2e", times = 12), "red", "#2e2e2e")

data |>
  select(any_of(paste0("cogtot27_imp", 1996:2022))) |>
  rename_with( ~ stringr::str_replace(., "cogtot27_imp", ""), starts_with("cogtot27")) |>
  visdat::vis_dat() +
  labs(title = "Missing data across HRS Waves (1996 - 2022)",
       subtitle = "Total Cognitive Scores") +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold", colour = colour),
    plot.subtitle = element_text(hjust = 0.5, size = 12, face = "bold", colour = colour),
    panel.grid = element_blank(),
    axis.text.x = element_text(colour = colours)) +
  ggeasy::easy_remove_legend() +
  ggeasy::easy_x_axis_labels_size(size = 9)

```

The HRS had its most recent participant intake in 2016, which explains the notable decline in missing data occurrences from that point onward. As new participants were not added after 2016, the dataset becomes more complete in subsequent waves, with fewer missing values.

Given this shift, we will conduct our analysis **the reduced dataset (2016 - 2022)** (@fig-complete-data) along with removing the missing values from 2016 and 2018.

```{r complete-data}
#| fig-width: 8
#| label: fig-complete-data
#| fig-cap: HRS dataset after processing

data |>
  select(any_of(paste0("cogtot27_imp", 2016:2022))) |>
  rename_with( ~ stringr::str_replace(., "cogtot27_imp", ""), starts_with("cogtot27")) |>
  na.omit() |>
  visdat::vis_dat() +
  labs(
    title = "Complete dataset across HRS Waves (2016 - 2022)",
    subtitle = "Total Cognitive Scores") +
    theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold", colour = colour),
    plot.subtitle = element_text(hjust = 0.5, size = 12, face = "bold", colour = colour),
    panel.grid = element_blank()) +
  ggeasy::easy_remove_legend() +
  ggeasy::easy_x_axis_labels_size(size = 9)
```

## Exploratory Data Analysis

### Cognitive Test Scores

Initially, we will plot a distribution of the cognitive test scores (ranging from 0 - 27) across time for all HRS participants.

```{r cog-scores}
#| message: false
#| fig-width: 12
#| label: fig-cognitive-test-scores
#| fig-cap: Cognitive test scores (2016 - 2022)

plot_cognitive_scores(data = data, year = 2016)
```

### Classification proportion per year

@fig-dementia-proportions shows the proportion of dementia classifications from HRS 2016 - 2022. The variable **Missing** represents the procrastination HRS respondents who were not yet interviewed by the HRS. Since the analysis focuses on participants included in the 2020 HRS wave, any retroactive analysis of prior waves may result in missing data for certain individuals

```{r cog-class}
#| fig-width: 12
#| label: fig-dementia-proportions
#| fig-cap: Proportion of dementia classifications per year

# Plotting proportions --------------------------------------------------------
data |>
  count_transitions(years = seq(2016, 2022, by = 2), absorbing = TRUE) |>
  ggplot(aes(x = Wave, fill = Classification)) +
  geom_bar(position = "fill", alpha = 0.5, colour = "black") +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Proportion of Cognitive Status Classifications (2016 - 2022)", 
       x = "", y = "Percentage") +
  scale_fill_viridis_d(direction = -1) +
  theme(
    axis.text = element_text(size = 10),
    text = element_text(size = 10)) +
  ggeasy::easy_move_legend("bottom") +
  ggeasy::easy_remove_legend_title()
```

### Dementia transitions per year

@fig-dementia-transitions is an alluvial graph illustrating the transitions in cognitive classifications from one HRS wave to the next.

```{r cog-transition}
#| fig-width: 12
#| label: fig-dementia-transitions
#| fig-cap: Dementia transitions (2016 - 2022)

data |>
  count_transitions(years = seq(2016, 2022, by = 2)) |>
  group_by(ID, Wave) |>
  reframe(plyr::count(Classification)) |>
  rename(Classification = x, n = freq) |>
  plot_transitions(size = 2.5)
```

From this figure we can see that some individuals classified with dementia in one wave transition out of dementia in a subsequent wave. Let's fix this by adding an additional parameter to our `count_transitions()` function to turn dementia into an absorbing state (@fig-dementia-transitions-absorbing). We can do this by using the `cumany()` function from the `dplyr` [@Wickham2023] package.

```{r cog-transition-absorbing}
#| fig-width: 12
#| label: fig-dementia-transitions-absorbing
#| fig-cap: Dementia transitions with dementia as an absorbing state (2016 - 2022)

data |>
  count_transitions(years = seq(2016, 2022, by = 2), absorbing = TRUE) |>
  group_by(ID, Wave) |>
  reframe(plyr::count(Classification)) |>
  rename(Classification = x, n = freq) |>
  plot_transitions(size = 2.5)
```

### 2025-03-28 analysis/data-overview.qmd
Updating site

--- 2025-03-28 _drafts/Markov.qmd
@@
 ---
-title: "Procrastination & Dementia: A Markov Model"
-author: 
-  - name: "Cormac Monaghan"
-    url: "https://c-monaghan.github.io/"
-    orcid: "https://orcid.org/0000-0001-9012-3060"
-    affiliation:
-      - "Hamilton Institute, Maynooth University, Ireland"
-      - "Department of Psychology, Maynooth University, Ireland"
-  - name: "Rafael de Andrade Moral"
-    url: "https://rafamoral.github.io/"
-    orcid: "https://orcid.org/0000-0002-0875-3563"
-    affiliation:
-      - "Department of Mathematics and Statistics, Maynooth University, Ireland"
-  - name: "Joanna McHugh Power"
-    orcid: "https://orcid.org/0000-0002-7387-3107"
-    affiliation: 
-      - "Department of Psychology, Maynooth University, Ireland"
-toc: true
-toc-depth: 4
-toc-title: On this page
-bibliography: references.bib
-csl: apa.csl
+title: "Data Overview"
 format: 
   html:
-    theme: 
-      light: united
-    page-layout: full
-    html-math-method: katex
-    code-tools: true
-    self-contained: true
-    code-fold: true
-    code-summary: "Show the code"
-    include-in-header: 
-      - file: ./html/github-corner.html
-include-after-body: ./html/footer.html
-execute: 
-  warning: false
-editor: source
+    code-tools: false
 ---
-
-```{r setting-up}
-rm(list = ls())
-
-set.seed(43561)
-
-# Packages
-suppressMessages(library(dplyr))
-library(tidyr)
-library(ggplot2)
-library(ggalluvial)
-
-source(here::here("./02__Models/00__Functions.R"))
-
-# Paths -----------------------------------------------------------------------
-path_data <- "./01__Data/02__Processed/"
-```
 
@@
 
-![LWC Workflow](./images/LWC_Workflow.png){#fig-lwc}
-
-# Dementia Dataset
-
-Our updated dataset contains 13 columns of cognitive classifications spanning from 1996 to 2022.
-These columns reflect dementia status across each wave, using the same criteria and structure as the original Langa-Weir dataset, now extended to include the most recent 2022 wave.
-
-```{r data-import}
-data <- readxl::read_xlsx(here::here(path_data, "data.xlsx"))
-
-dementia_data <- data |>
-  count_transitions(years = seq(1996, 2022, by = 2))
-
-glimpse(dementia_data)
-```
-
-## Missing data
-
-Since the participant pool is drawn from the 2020 wave of the HRS data, any analysis that looks back at previous waves will inevitably encounter missing data.
-Not all participants were included in earlier waves, leading to gaps in the data.
-@fig-missing-data displays the frequency of missing cognitive test data across each HRS wave.
-
-```{r missing-data}
-#| fig-width: 8
-#| label: fig-missing-data
-#| fig-cap: Occurance of missing data across HRS waves (participants were gathered from the 2020 HRS wave [in red])
-
-# Dynamic colours
-colours <- c(rep("#2e2e2e", times = 12), "red", "#2e2e2e")
-
-data |>
-  select(any_of(paste0("cogtot27_imp", 1996:2022))) |>
-  rename_with( ~ stringr::str_replace(., "cogtot27_imp", ""), starts_with("cogtot27")) |>
-  visdat::vis_dat(palette = "cb_safe") +
-  labs(title = "Missing data across HRS Waves (1996 - 2022)",
-       subtitle = "Total Cognitive Scores") +
-  theme(
-    plot.title = element_text(hjust = 0.5, size = 12, face = "bold", colour = "#2e2e2e"),
-    plot.subtitle = element_text(hjust = 0.5, size = 10, colour = "#2e2e2e"),
-    axis.text.x = element_text(colour = colours)) +
-  ggeasy::easy_remove_legend() +
-  ggeasy::easy_x_axis_labels_size(size = 9)
-
-```
-
-The HRS had its most recent participant intake in 2016, which explains the notable decline in missing data occurrences from that point onward.
-As new participants were not added after 2016, the dataset becomes more complete in subsequent waves, with fewer missing values.
-
-Given this shift, we will conduct our analysis **the reduced dataset (2016 - 2022)** (@fig-complete-data) along with removing the missing values from 2016 and 2018.
-
-```{r complete-data}
-#| fig-width: 8
-#| label: fig-complete-data
-#| fig-cap: HRS dataset after processing
-
-data |>
-  select(any_of(paste0("cogtot27_imp", 2016:2022))) |>
-  rename_with( ~ stringr::str_replace(., "cogtot27_imp", ""), starts_with("cogtot27")) |>
-  na.omit() |>
-  visdat::vis_dat(palette = "cb_safe") +
-  labs(title = "Complete dataset across HRS Waves (2016 - 2022)",
-       subtitle = "Total Cognitive Scores") +
-  theme(
-    plot.title = element_text(hjust = 0.5, size = 12, face = "bold", colour = "#2e2e2e"),
-    plot.subtitle = element_text(hjust = 0.5, size = 10, colour = "#2e2e2e")) +
-  ggeasy::easy_remove_legend() +
-  ggeasy::easy_x_axis_labels_size(size = 9)
-```
-
-## Exploratory Data Analysis
-
-### Cognitive Test Scores
-
-Initially, we will plot a distribution of the cognitive test scores (ranging from 0 - 27) across time for all HRS participants.
-
-```{r cog-scores}
-#| fig-width: 12
-#| label: fig-cognitive-test-scores
-#| fig-cap: Cognitive test scores (2016 - 2022)
-
-plot_cognitive_scores(data = data, year = 2016)
-```
-
-### Classification proportion per year
-
-@fig-dementia-proportions shows the proportion of dementia classifications from HRS 2016 - 2022.
-The variable **Missing** represents the procrastination HRS respondents who were not yet interviewed by the HRS.
-Since the analysis focuses on participants included in the 2020 HRS wave, any retroactive analysis of prior waves may result in missing data for certain individuals
-
-```{r cog-class}
-#| fig-width: 12
-#| label: fig-dementia-proportions
-#| fig-cap: Proportion of dementia classifications per year
-
-# Plotting proportions --------------------------------------------------------
-data |>
-  count_transitions(years = seq(2016, 2022, by = 2), absorbing = TRUE) |>
-  ggplot(aes(x = Wave, fill = Classification)) +
-  geom_bar(position = "fill", alpha = 0.5, colour = "black") +
-  scale_y_continuous(labels = scales::percent) +
-  labs(title = "Proportion of Cognitive Status Classifications (2016 - 2022)", 
-       x = "", y = "Percentage") +
-  scale_fill_viridis_d(direction = -1) +
-  theme_minimal() +
-  theme(panel.grid = element_blank(),
-        plot.title = element_text(size = 14, hjust = 0.5, face = "bold"),
-        axis.text = element_text(size = 10),
-        axis.title = element_text(size = 10),
-        text = element_text(size = 10)) +
-  ggeasy::easy_move_legend("bottom") +
-  ggeasy::easy_remove_legend_title()
-```
-
-### Dementia transitions per year
-
-@fig-dementia-transitions is an alluvial graph illustrating the transitions in cognitive classifications from one HRS wave to the next.
-
-```{r cog-transition}
-#| fig-width: 12
-#| label: fig-dementia-transitions
-#| fig-cap: Dementia transitions (2016 - 2022)
-
-data |>
-  count_transitions(years = seq(2016, 2022, by = 2)) |>
-  group_by(ID, Wave) |>
-  reframe(plyr::count(Classification)) |>
-  rename(Classification = x, n = freq) |>
-  plot_transitions(size = 2.5)
-```
-
-From this figure we can see that some individuals classified with dementia in one wave transition out of dementia in a subsequent wave.
-Let's fix this by adding an additional parameter to our `count_transitions()` function to turn dementia into an absorbing state (@fig-dementia-transitions-absorbing).
-We can do this by using the `cumany()` function from the `dplyr` [@Wickham2023] package.
-
-```{r cog-transition-absorbing}
-#| fig-width: 12
-#| label: fig-dementia-transitions-absorbing
-#| fig-cap: Dementia transitions with dementia as an absorbing state (2016 - 2022)
-
-data |>
-  count_transitions(years = seq(2016, 2022, by = 2), absorbing = TRUE) |>
-  group_by(ID, Wave) |>
-  reframe(plyr::count(Classification)) |>
-  rename(Classification = x, n = freq) |>
-  plot_transitions(size = 2.5)
-```
-
-# Markov Modelling
-
-In this section, we apply Markov modelling to analyze the transitions between cognitive states (Normal Cognition, Mild Cognitive Impairment [MCI], and Dementia) using longitudinal data from the Health and Retirement Study (HRS) from 2016 to 2022.
-By modeling these transitions, we aim to understand the probabilities of moving between different cognitive states over time.
-
-## Data preparation
-
-The first step is to prepare the transition dataset. We start by transforming the data to create a record of transitions from one cognitive state to another between consecutive waves.
-
-Once the transitions are identified, we calculate the probabilities of each state-to-state transition.
-These probabilities are computed by counting the occurrences of each transition and dividing by the total number of transitions.
-
-```{r transition-probability}
-tran_prob <- data |>
-  extract_years(years = seq(2016, 2022, by = 2)) |>
-  create_transitions(absorbing = TRUE) |>
-  calculate_probabilties()
-
-tran_prob |>
-  mutate(prop = round(prop, digits = 3)) |>
-  head(n = 9) |>
-  gt::gt() |>
-  gtExtras::gt_theme_538() |>
-  gt::cols_align(align = "center") |>
-  gt::tab_header(title = "Overall Transition Probabilties (2016 - 2022)") |>
-  gt::tab_options(table.width = gt::pct(100)) |>
-  gt::opt_align_table_header(align = "center")
-```
-
-## Creating transition matrix
-
-The probability distribution of transitions from one state to another can be represented into a transition matrix $P = (p_{ij})_{i,j}$ where each element of position $(i, j)$ represents the transition probability $p_{ij}$.
-
-$$
-P = \begin{bmatrix} 
-p_{11} & p_{12} & p_{13} \\
-p_{21} & p_{22} & p_{23} \\
-p_{31} & p_{32} & p_{33} \\
-\end{bmatrix}
-$$
-
-```{r transition-matrix}
-tran_matrix <- tran_prob |>
-  transition_matrix() |>
-  normalise() |>
-  round(digits = 3)
-
-tran_matrix
-```
-
-This code creates a $3 \times 3$ matrix (for the three states: Normal Cognition, MCI, and Dementia) and populates it with the transition probabilities.
-
-## Visualising the matrix
-
-To make the transition matrix more interpretable we visualize the probabilities using a heat map (@fig-transition-matrix)
-
-```{r transition-matrix-plot}
-#| fig-width: 8
-#| label: fig-transition-matrix
-#| fig-cap: Heat map of transition probabilities (2016 - 2022)
-
-plot_matrix(data = t(tran_matrix), year = 2016)
-```
-
-We can also create and plot transition matrices for each wave separately allowing us to see how these transition probabilities change across multiple waves of data collection
-
-```{r transition-matrix-longitudinal}
-tran_prob_long <- data |>
-  extract_years(years = seq(2016, 2022, by = 2)) |>
-  create_transitions(absorbing = TRUE) |>
-  group_by(Wave) |>
-  calculate_probabilties() |>
-  group_map(~ transition_matrix(.x, longitudinal = TRUE), .keep = TRUE) |>
-  lapply(function(mat) {
-    mat[c("Normal Cognition", "MCI", "Dementia"), c("Normal Cognition", "MCI", "Dementia")] |>
-        normalise()
-    }) |>
-  purrr::map2_df(.y = c(seq(2016, 2020, by = 2)), .f = reshape_matrix) |>
-  mutate(
-    transition_prob = round(transition_prob, digits = 3),
-    from = factor(from, levels = c("Normal Cognition", "MCI", "Dementia"))
-  )
-
-DT::datatable(
-  tran_prob_long,
-  filter = "top",
-  colnames = c("Probability" = "transition_prob"),
-  options = list(pageLength = 6, autoWidth = TRUE)
-)
-```
-
-```{r transition-matrix-longitudinal-plot, fig.width=8, fig.show="animate", ffmpeg.format="gif", dev="png"}
-for(wave in c(seq(2016, 2020, by = 2))){
-  print(plot_transition_wave(data = tran_prob_long, wave = wave))
-}
-```
-
-# Multi-State Markov Model
-
-Now that we have an idea of the disease progression across time we are interested in the effect of different covariates on these transitions.
-For this, we can fit a multi-state Markov model.
-Multi-state models (MSMs) provide a comprehensive view of the disease process and make risk-factors analysis and long-term predictions more easily.
-The conditional distribution of the status of an individual participant at an arbitrary examination given their status at previous examinations was assumed to have the **Markov property**.
-That is the distribution of the forthcoming state $X_{n+1}$ depends only on the current state $X_n$ and doesn't depend on the previous ones $X_{n-1}, X_{n-2}, \dots, X_1$.
-
-$$Pr(X_{n+1} = x_{n+1} \; \vert \; X_1 = x_1, X_2 = x_2, \dots, X_n = x_n) = Pr(X_{n+1} = x_{n+1} \; \vert \; X_n = x_n)$$
-
-## Data preperation
-
-For our model we are going to look at several different covariates:
-
-- **Gender**: $0 = \text{Male;} \; 1 = \text{Female}$ 
-- **Age**: $\text{Range} \; = 50 - 91$ 
-- **Education**: $0 = \text{No Degree;} \; 1 = \text{High School Degree;} \; 2 = \text{Further Education}$ 
-- **Number of cardiovascular risk factors**: $\text{Range} \; = 0 - 4$
-- **Procrastination**: $\text{Range} \; = 4 - 60$
-
-```{r markov-prep}
-cols <- c(
-  "ID", "Gender", "Age", "Education_tri", 
-  paste0("Cardio_risk_", seq(16, 22, by = 2)), 
-  paste0("Total_dep_20", seq(16, 22, by = 2)), "Total_p")
-
-markov_data <- data |>
-  extract_years(seq(2016, 2022, by = 2)) |>
-  na.omit() |>
-  inner_join(data[, cols], by = "ID") |>
-  pivot_data() |>
-  backtrack_age(base_year = 2016, current_year = 2022) |>
-  group_by(ID) |>
-  mutate(
-    Age = min(Age), # Baseline age
-    Cardio_risk = Cardio_risk[Wave == 1],
-    Depression = Depression[Wave == 1]
-    ) |>
-  ungroup() |>
-  filter(Age >= 50) # Only looking at older adults
-
-DT::datatable(
-  markov_data,
-  filter = "top",
-  options = list(pageLength = 6, autoWidth = TRUE)
-)
-```
-
-Let's look at a state table to visualize the amount of transitions over time
-
-```{r state-table}
-markov_data |>
-  select(ID, Wave, Status) |>
-  mutate(Status = case_when(
-    Status == 1 ~ "Normal Cognition", 
-    Status == 2 ~ "MCI", 
-    Status == 3 ~ "Dementia")) |>
-  msm::statetable.msm(state = Status, subject = ID)
-```
-
-## Modelling
-
-### Defining a q-matrix
-
-The Q matrix (also known as the transition intensity matrix or generator matrix) defines the rates at which transitions occur between states in the model.
-
-It is a square matrix where:
-
-- Each row corresponds to a **from** state in the model.
-- Each column corresponds to a **to** state in the model.
-- The off-diagonal elements $q_{ij}$ represent the transition rate from state $i$ to state $j$.
-
-For a model with $n$ states, the Q matrix has the following structure
-
-$$
-Q = \begin{bmatrix} 
-q_{11} & q_{12} & \cdots & q_{1n} \\
-q_{21} & q_{22} & \cdots & q_{2n} \\
-\vdots & \vdots & \ddots &\vdots  \\
-q_{n1} & q_{n2} & \cdots & q_{nn}
-\end{bmatrix}
-$$
-
-```{r q-matrix}
-# Creating transition matrix
-# Define possible transitions
-states <- c("Normal Cognition", "MCI", "Dementia")
-
-# Theoretical transition probabilities
-q_matrix <- rbind(
-  c(0.75, 0.20, 0.05),  # From Normal, can transition to MCI or Dementia
-  c(0.25, 0.50, 0.25),  # From MCI, can transition to Normal or Dementia
-  c(0.00, 0.00, 1.00)         # From Dementia, no transitions (absorbing state)
-)
-
-rownames(q_matrix) <- states
-colnames(q_matrix) <- states
-
-q_matrix
-```
-
-## Fitting
-
-We will fit three different multi-state models
-
-1. A model with no covariates
-2. A multivariate model with covariates acting on all transitions
-3. A multivariate model with covariates (and interactions) acting on all transitions
-4. A multivariate model with covariates acting on only NC <-> MCI & MCI - Dementia
-5. A multivariate model with covariates (and interactions) acting on only NC <-> MCI & MCI - Dementia
-
-```{r model-fitting}
-# Fitting model with no covariates
-model_1 <- msm::msm(
-  formula = Status ~ Wave,
-  subject = ID,
-  data = markov_data,
-  qmatrix = q_matrix,
-  obstype = 1,
-  method = "BFGS", 
-  gen.inits = TRUE,
-  control = list(
-    fnscale = 5000, 
-    maxit = 2000 # To reach convergence
-  ))
-
-# Multivariate model: covariates acting on all transitions
-model_2 <- msm::msm(
-  formula = Status ~ Wave,
-  subject = ID,
-  data = markov_data,
-  qmatrix = q_matrix,
-  covariates = ~ Gender + Age + Education_tri + Cardio_risk + Depression + Total_p,
-  obstype = 1,
-  gen.inits = TRUE,
-  method = "BFGS",
-  control = list(
-    fnscale = 5000, 
-    maxit = 2000 # To reach convergence
-  ))
-
-# Multivariate model: covariates acting on all transitions (with interaction)
-model_3 <- msm::msm(
-  formula = Status ~ Wave,
-  subject = ID,
-  data = markov_data,
-  qmatrix = q_matrix,
-  covariates = ~ Gender + Age + Education_tri + Cardio_risk + (Depression * Total_p),
-  obstype = 1,
-  gen.inits = TRUE,
-  method = "BFGS",
-  control = list(
-    fnscale = 5000, 
-    maxit = 2000 # To reach convergence
-  ))
-
-# Multivariate model: covariates acting on NC <-> MCI and MCI -> D
-model_4 <- msm::msm(
-  formula = Status ~ Wave,
-  subject = ID,
-  data = markov_data,
-  qmatrix = q_matrix,
-  covariates = list(
-    "1-2" = ~ Gender + Age + Education_tri + Cardio_risk + Depression + Total_p,
-    "2-1" = ~ Gender + Age + Education_tri + Cardio_risk + Depression + Total_p,
-    "2-3" = ~ Gender + Age + Education_tri + Cardio_risk + Depression + Total_p
-  ),
-  obstype = 1,
-  gen.inits = TRUE,
-  method = "BFGS",
-  control = list(
-    fnscale = 5000, 
-    maxit = 2000 # To reach convergence
-  ))
-
-# Multivariate model: covariates acting on NC <-> MCI and MCI -> D (with interaction)
-model_5 <- msm::msm(
-  formula = Status ~ Wave,
-  subject = ID,
-  data = markov_data,
-  qmatrix = q_matrix,
-  covariates = list(
-    "1-2" = ~ Gender + Age + Education_tri + Cardio_risk + (Depression * Total_p),
-    "2-1" = ~ Gender + Age + Education_tri + Cardio_risk + (Depression * Total_p),
-    "2-3" = ~ Gender + Age + Education_tri + Cardio_risk + (Depression * Total_p)
-  ),
-  obstype = 1,
-  gen.inits = TRUE,
-  method = "BFGS",
-  control = list(
-    fnscale = 5000, 
-    maxit = 20002 # To reach convergence
-  ))
-
-```
-
-### Model Checking
-
-After fitting each of our models we can extract some goodness of fit criteria
-
-```{r model-goodness}
-log_l <- c(logLik(model_1), 
-           logLik(model_2), 
-           logLik(model_3), 
-           logLik(model_4), 
-           logLik(model_5))
-
-aic <- AIC(model_1, model_2, model_3, model_4, model_5)
-
-# Very small difference between model 2 and 3 (makes sense)
-cbind(log_l, aic) |>
-  tibble::rownames_to_column("Model") |>
-  mutate(Model = stringr::str_remove(Model, "model_"))
-
-# Likelihood ratio test
-lmtest::lrtest(model_1, model_2, model_3, model_4, model_5)
-```
-
-We can also look at a measure of how well the model fits the data using a test proposed by [@gentleman1994].
-The function plots the observed numbers of individuals occupying a state at a series of times against forecasts from the fitted model, for each state (@fig-gentleman).
-
-```{r gentleman_test}
-#| fig-width: 12
-#| fig-height: 12
-#| label: fig-gentleman
-#| fig-cap: Fitted values versus observed values
-
-# Plotting fitted vs. observed values
-g1 <- gentleman_test(data = msm::prevalence.msm(model_1), 
-                     model = "Model 1")
-g2 <- gentleman_test(data = msm::prevalence.msm(model_2), 
-                     model = "Model 2")
-g3 <- gentleman_test(data = msm::prevalence.msm(model_3), 
-                     model = "Model 3")
-g4 <- gentleman_test(data = msm::prevalence.msm(model_4), 
-                     model = "Model 4")
-g5 <- gentleman_test(data = msm::prevalence.msm(model_5), 
-                     model = "Model 5")
-
-ggpubr::ggarrange(g1, g2, g3, g4, g5, 
-                  common.legend = TRUE, 
-                  nrow = 5,
-                  heights = c(2, 2, 2, 2, 2), 
-                  legend = "bottom")
-```
-
-### Hazard ratios
-
-We now present hazard ratios for each of our fitted models. These ratios represent how the instantaneous risk of making a particular transition is modified by the covariate.
-
-::: {.panel-tabset}
-
-## Model 2
-
-@fig-hazard-ratio-1 presents hazard ratios from model 2 (our multivariate model with covariates acting on all transitions) for each covariate.
-
-```{r}
-#| fig-width: 12
-#| fig-height: 12
-#| label: fig-hazard-ratio-1
-#| fig-cap: Ouputted hazard ratios from Markov model. Green bars represent a decreased risk of transition while red bars indicate an increased risk of transition.
-
-# Plotting hazard ratios
-model_2 |>
-  msm::hazard.msm() |>
-  process_hr() |>
-  calculate_p_value() |>
-  plot_hr()
-```
-
-## Model 3
-
-@fig-hazard-ratio-2 presents hazard ratios from model 3 (our multivariate model with covariates **(and interactions)** acting on all transitions) for each covariate.
-
-```{r}
-#| fig-width: 12
-#| fig-height: 12
-#| label: fig-hazard-ratio-2
-#| fig-cap: Ouputted hazard ratios from Markov model. Green bars represent a decreased risk of transition while red bars indicate an increased risk of transition.
-
-# Plotting hazard ratios
-model_3 |>
-  msm::hazard.msm() |>
-  process_hr() |>
-  calculate_p_value() |>
-  plot_hr()
-```
-
-## Model 4
-
-@fig-hazard-ratio-3 presents hazard ratios for model 4 (our multivariate model with covariates acting on only NC <-> MCI & MCI -> Dementia) for each covariate
-
-```{r}
-#| fig-width: 12
-#| fig-height: 12
-#| label: fig-hazard-ratio-3
-#| fig-cap: Ouputted hazard ratios from Markov model. Green bars represent a decreased risk of transition while red bars indicate an increased risk of transition.
-
-# Plotting hazard ratios
-model_4 |>
-  msm::hazard.msm() |>
-  process_hr() |>
-  calculate_p_value() |>
-  plot_hr()
-```
-
-## Model 5
-
-@fig-hazard-ratio-4 presents hazard ratios for model 5 (our multivariate model with covariates **(and interactions)** acting on only NC <-> MCI & MCI -> Dementia) for each covariate
-
-```{r}
-#| fig-width: 12
-#| fig-height: 12
-#| label: fig-hazard-ratio-4
-#| fig-cap: Ouputted hazard ratios from Markov model. Green bars represent a decreased risk of transition while red bars indicate an increased risk of transition.
-
-# Plotting hazard ratios
-model_5 |>
-  msm::hazard.msm() |>
-  process_hr() |>
-  calculate_p_value() |>
-  plot_hr()
-```
-
-:::
+![LWC Workflow](../files/workflow.svg){#fig-lwc}

### 2025-03-28 analysis/discrete-markov.qmd
Updating site

@@
 ---
-title: "Discrete Time Markov Model Analysis of Cognitive Decline"
+title: "Discrete Time Markov Model"
 subtitle: "Examining the Association Between Procrastination and Cognitive Transitions"
-author: 
-  - name: "Cormac Monaghan"
-    url: "https://c-monaghan.github.io/"
-    orcid: "https://orcid.org/0000-0001-9012-3060"
-    affiliation:
-      - "Hamilton Institute, Maynooth University, Ireland"
-      - "Department of Psychology, Maynooth University, Ireland"
-abstract: |
-  This analysis examines transitions between cognitive states (Normal Cognition, MCI, Dementia) using discrete-time Markov models, with particular focus on procrastination as a risk factor.
-toc: true
-toc-depth: 3
-toc-title: On this page
-number-sections: true
-format: 
-  html:
-    theme: 
-      light: cosmo
-    page-layout: full
-    shift-heading-level-by: 1
-    html-math-method: katex
-    code-tools: true
-    code-copy: true
-    code-block-bg: "#333A3F"
-    highlight-style: monokai
-    fig-align: center
-    self-contained: true
 ---

### 2025-03-28 analysis/data-final.qmd
Launching site

@@
 - CSV file for any program
-- `.rds` file for R <span class="small">(load with `df <- readRDS("weekly_panel.rds")`)</span>
-- `.dta` file for Stata <span class="small">(load with `use "weekly_panel.dta"`)</span>
+- `.rds` file for R <span class="small">(load with `df <- readRDS("data.rds")`)</span>
+- `.dta` file for Stata <span class="small">(load with `use "data.dta"`)</span>
 

### 2025-03-28 README.qmd
Adding hmtl

---
title: "Procrastination and Dementia"
format: 
  gfm: 
    wrap: preserve
    toc: true
  typst:
    toc: true
    toc-title: " "
    author:
    - name: Audrey L. Comstock
      affiliation: Arizona State University
      orcid: 0000-0003-1772-5350
      url: https://audreylcomstock.weebly.com/
      email: audrey.comstock@asu.edu
    - name: Andrew Heiss
      affiliation: Georgia State University
      orcid: 0000-0002-3948-3914
      url: https://www.andrewheiss.com
      email: aheiss@gsu.edu
    - name: Suparna Chaudhry
      affiliation: Lewis and Clark College
      orcid: 0000-0003-3506-7590
      url: https://www.suparnachaudhry.com/
      email: schaudhry@lclark.edu
---

<!-- README.md is generated from README.qmd. Please edit that file -->

```{r setup, include=TRUE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
  fig.retina = 3
)
```

::: {.content-visible when-format="markdown"}

---

[Cormac Monaghan](https://c-monaghan.github.io/)

---

:::

## Note on "crisp clam" project name

Because project titles change all the time we used [{codename}](http://svmiller.com/codename/) to generate an [Ubuntu-style](https://wiki.ubuntu.com/DevelopmentCodeNames)
internal-to-us project name that won’t change.

```{r code-name, eval=FALSE}
library(codename)

codename_message()
#> code name generated by {codename} v.0.5.0. R version 4.4.2 (2024-10-31 ucrt).

codename(seed = "dementia", type = "ubuntu")
#> [1] "crisp clam"
```

### 2025-03-28 analysis/paper.qmd
Adding hmtl

---
title: "Paper"
toc: false
page-layout: full
---

```{=html}
<p class="text-center">
  <a class="btn btn-success btn-sm mb-1" href="/manuscript/output/manuscript.html" target="manuscript">
  <i class="fa-solid fa-file-lines"></i> &ensp;HTML
  </a>

  <a class="btn btn-success btn-sm mb-1" href="/manuscript/output/manuscript.html" target="_blank">
  <i class="fa-solid fa-arrow-up-right-from-square"></i>&ensp;HTML <small>(new window)</small>
  </a>

  <a class="btn btn-danger btn-sm mb-1" href="/manuscript/output/manuscript.pdf" target="manuscript">
  <i class="fa-solid fa-file-pdf"></i>&ensp;PDF
  </a>

  <a class="btn btn-danger btn-sm mb-1" target="_blank" href="/manuscript/output/manuscript.pdf">
  <i class="fa-solid fa-file-arrow-down"></i>&ensp;PDF <small>(download)</small>
  </a>

  <a class="btn btn-warning btn-sm mb-1" href="/manuscript/output/manuscript+ms.pdf" target="manuscript">
  <i class="fa-solid fa-file-pdf"></i>&ensp;Manuscripty PDF
  </a>

  <a class="btn btn-warning btn-sm mb-1" target="_blank" href="/manuscript/output/manuscript+ms.pdf">
  <i class="fa-solid fa-file-arrow-down"></i>&ensp;Manuscripty PDF <small>(download)</small>
  </a>
</p>

<iframe src="/manuscript/output/manuscript.html" frameborder="0" style="overflow:hidden;height:90%;width:100%;border:2px solid #ccc;" height="90%" width="100%" name="manuscript"></iframe>
```

### 2025-03-28 manuscript/manuscript.qmd
Adding hmtl

---
title: "Procrastination and Dementia"
author: 
  - name: "Cormac Monaghan"
    url: "https://c-monaghan.github.io/"
    orcid: "https://orcid.org/0000-0001-9012-3060"
    affiliation:
      - "Hamilton Institute, Maynooth University, Ireland"
      - "Department of Psychology, Maynooth University, Ireland"
    email: cormacmonaghan@protonmail.com
    attributes:
      corresponding: true
  - name: "Rafael de Andrade Moral"
    url: "https://rafamoral.github.io/"
    orcid: "https://orcid.org/0000-0002-0875-3563"
    affiliation:
      - "Department of Mathematics and Statistics, Maynooth University, Ireland"
  - name: "Joanna McHugh Power"
    orcid: "https://orcid.org/0000-0002-7387-3107"
    affiliation: 
      - "Department of Psychology, Maynooth University, Ireland"
correspondence-prefix: "Correspondence concerning this article should be addressed to"
code-repo: "Access the code, data, and analysis at <https://github.com/C-Monaghan/crisp-clam>"
bibliography: references.bib
csl: csl/apa.csl
link-citations: true
---

::: {.content-visible when-format="html"}
# Introduction {.hide}
:::

Dementia is a syndrome characterized by the progressive and typically irreversible decline of cognitive function, leading to memory loss, impaired reasoning, and difficulties with daily activities [@Prince2013; @Sanz-Blasco2022]. It encompasses a range of conditions, including Alzheimer's disease, vascular dementia, and Lewy body dementia [@Cao2020]. The global burden of dementia is substantial, with projections suggesting that the number of affected individuals will rise from 57.4 million in 2019 to 152.8 million by 2050 [@Nichols2022]. Given this projection, identifying and addressing modifiable risk factors is crucial to mitigate the growing prevalence of dementia worldwide.

Given the progressive nature of dementia, early identification of pre-dementia conditions such as mild cognitive impairment (MCI) is essential for timely intervention. MCI is a condition characterized by cognitive changes - such as memory lapses or difficulty making decisions - that exceed typical age-related decline [@Abner2012; @Cooper2015; @Fresnais2023; @Salem2023; @Yu2013]. Studies indicate that approximately 46% of people with MCI transition to dementia within three years and approximately 80% within six years [@Cooper2015; @Shigemizu2020; @Tschanz2006]. As such, identifying factors that influence this progression is critically important for early intervention and personalized care.

Among the behavioural symptoms in MCI and dementia, apathy is one of the most prevalent [@Fresnais2023; @Richard2012; @Salem2023; @vanDalen2018]. Defined as a lack of motivation [@Fresnais2023], apathy is also a multidimensional construct that encompasses deficits in executive and emotional functioning, initiation, and increased functional impairment [@Okura2010; @Radakovic2018]. Individuals with apathy exhibit reduced goal-directed behavior and a diminished desire to pursue rewards or pleasure [@Fahed2021]. Importantly, apathy has been identified as a significant risk factor for the transition from MCI to dementia [@Palmer2010; @Ruthirakuhan2019; @vanDalen2018]. For instance, a meta-analysis by @vanDalen2018 found that apathy almost doubles the risk of transitioning to dementia. Additionally, apathy has been correlated with higher levels of neurofibrillary tangles in individuals with dementia, suggesting a potential connection to underlying neuropathology [@Skogseth2008].

Procrastination, although traditionally viewed as a distinct behavioural issue, may share key characteristics with apathy, suggesting potential common underlying mechanisms.Chronic procrastination, characterized by persistent delays in decision-making and task completion [@Abbasi2015; @Ferrari2010], has been associated with dysfunction in the brain's reward and decision-making systems, particularly the dorsolateral and ventromedial prefrontal cortices [@Friden2020; @Zhang2019]. These brain regions are critical for both initiating and sustaining goal-directed action and are areas where both apathy and procrastination show deficits [@Fahed2021; @Zhang2019].

While apathy primarily reflects a lack of motivation, procrastination involves a delay in action despite an intention to complete such action [@Steel2007]. Both behaviours suggest impaired executive function, particularly in goal-oriented behaviour and decision-making, which are hallmark deficits in MCI and dementia [@Kirova2015; @Stopford2012]. In this context, procrastination could reflect broader motivational and cognitive impairments akin to those seen in apathy.

Given these parallels, it is worth exploring whether chronic procrastination could serve as an early behavioural marker for cognitive impairment, or even a risk factor for dementia, especially in older adults. Procrastination may exacerbate existing cognitive decline by reinforcing patterns of inaction and passivity. Individuals who chronically delay tasks may inadvertently engage in fewer cognitively stimulating activities, such as physical activity, problem-solving, decision-making, and goal-setting---activities that are known to build cognitive resilience and reduce dementia risk [@Chowdhary2022]. By limiting engagement in such activities, procrastination could contribute to the acceleration of cognitive decline. Therefore, while apathy has already been established as a significant risk factor for dementia, the role of procrastination, especially when chronic, may represent an overlooked behavioural trait that warrants similar attention.

Although current research on procrastination in relation to dementia is non-existent, this possible association warrants exploration. Identifying procrastination as a potential risk factor could expand the scope of early interventions aimed at preventing or slowing the progression of dementia. As such, the purpose of this study was to test the hypothesis that higher levels of procrastination would be associated with an increased probability of transitioning from normal cognitive function or MCI to dementia.

# References

### 2025-04-01 analysis/paper.qmd
Update :bug:

@@
   </a>
-
-  <a class="btn btn-warning btn-sm mb-1" href="/manuscript/output/manuscript+ms.pdf" target="manuscript">
-  <i class="fa-solid fa-file-pdf"></i>&ensp;Manuscripty PDF
-  </a>
-
-  <a class="btn btn-warning btn-sm mb-1" target="_blank" href="/manuscript/output/manuscript+ms.pdf">
-  <i class="fa-solid fa-file-arrow-down"></i>&ensp;Manuscripty PDF <small>(download)</small>
-  </a>
 </p>

### 2025-04-01 manuscript/manuscript.qmd
Update :bug:

@@
 
+# Method {#method .unnumbered}
+
+## Data and study population {#data-and-study-population .unnumbered}
+
+Analyses were conducted using a secondary data source; a multi-wave prospective cohort study called the Health and Retirement Study (HRS; [@Fisher2018]), which tracks the health, economic, and social well-being of over $18,000$ American adults primarily aged $>50$.
+The HRS is managed by the Institute for Social Research at the University of Michigan, with data collected every two years.
+Initial data collection of a participant is conducted through a face-to-face interview, with follow-up biennial interviews conducted either by phone or face-to-face.
+The average re-interview response rate ranges from $68.8\%$ to $92.3\%$ [@HRS2017].
+At the time of writing, fifteen years of HRS data are currently archived.
+
+For this study, we focused on four waves of HRS data (from 2016 to 2022).
+Specifically our study sample consisted of respondents who participated in an experimental module assessing procrastination during the 2020 wave.
+These experimental modules, administered at the end of the core HRS interview, consist of concise questionnaires designed to explore new topics or supplement existing core survey data.
+Each respondent receives only one experimental module, with sample sizes for each module constituting approximately 10% of the core sample.
+As a result, while the core HRS sample includes approximately $18,000$ respondents, our initial sample of interest consisted of $1,368$ respondents.
+We excluded respondents with missing cognitive assessment data for any wave $(n = 445)$ and those under 50 years of age $(n = 39)$.
+This resulted in a final analytic sample of $884$ respondents.
+
+## Measures {#measures .unnumbered}
+
+### Outcome: Cognitive Function and Cognitive Category {#outcome-cognitive-function-and-cognitive-category .unnumbered}
+
+Cognitive function in the HRS is assessed using a series of tests adapted from the Telephone Interview for Cognitive Status (TICS; [@Brandt1988; @Fong2009]).
+These tests include an immediate and delayed $10$-noun free recall test, a serial seven subtraction test, and a backward count from $20$ test.
+Based on these assessments, @Crimmins2011 developed both a $27$-point cognitive scale and validated cut-off points to assess and classify cognitive status.
+Using these points, respondents who scored $12 - 27$ were classified as having normal cognition, $7 - 11$ as having MCI, and $0 - 6$ as having dementia.
+
+### Predictor: Procrastination {#predictor-procrastination .unnumbered}
+
+Procrastination was measured using the Pure Procrastination Scale [@steel2010], which consists of 12 items rated on a Likert scale ranging from $1$ (strongly disagree) to $5$ (strongly agree).
+The total procrastination score ranges from $12$ to $60$, with higher scores indicating greater procrastination.
+The Pure Procrastination Scale conducted in 2020 (wave 3) had a Cronbach's $\alpha$ score of $0.91$ in this sample, indicating high internal consistency.
+An example of a question from the scale includes "I delay making decisions until it's too late".
+
+### Confounders {#confounders .unnumbered}
+
+To account for potential confounding, we controlled for variables with established associations with cognitive function and procrastination.
+These included baseline measures conducted in 2016 (wave 1) of age, sex, educational attainment, number of cardiovascular risk factors (history of hypertension, stroke, or heart disease, and a classification on "overweight\" on a body mass index (BMI) scale $(\text{BMI} \ge 30)$), and depressive symptoms [@Abner2012; @Freedman2024; @monaghan2024a; @Yu2013].
+Educational attainment was classified into three categories: no formal education, GED (General Educational Development)/high school diploma, and college/further education.
+Depressive symptoms were measured using an eight-item version of the Center for Epidemiological Studies Depression (CES-D) scale [@Briggs2018], with scores ranging from 0 to 8, demonstrating good internal consistency $(\alpha = 0.82)$.
+
 # References

### 2025-04-01 R/00__Functions.R

@@ -1,810 +1,0 @@
-# A list of functions used throughout analysis
-# Used in 02__Aggregation.R ----------------------------------------------------
-
-process_health_data <- function(health_data, prefix) {
-  health_data |>
-    select(HHID, PN, 
-           Weight = paste0(prefix, "139"),
-           H_feet = paste0(prefix, "141"),
-           H_inches = paste0(prefix, "142"),
-           Hypertension = paste0(prefix, "005"),
-           Stroke = paste0(prefix, "053"),
-           H_problem = paste0(prefix, "036")) |>
-    mutate(across(Hypertension:H_problem, 
-                  ~ ifelse(. %in% c(4, 5, 6), 0, 1))) 
-}
-
-height_weight <- function(data) {
-  data |>
-    mutate(H_inches = round(H_inches, digits = 0)) |>
-    transform(Height = interaction(H_feet, H_inches)) |>
-    mutate(
-      Height = as.numeric(as.character(Height)),
-      Height = round((Height * 0.3048), digits = 2),
-      Weight = round((Weight * 0.453592), digits = 2),
-    ) |>
-    select(!c(H_feet, H_inches)) |>
-    relocate(Height, .before = Weight)
-}
-
-impute_height <- function(data) {
-  data |>
-    select(HHID, PN, starts_with("Height")) |>
-    mutate(ID = row_number()) |>
-    pivot_longer(cols = starts_with("Height"), names_to = "Year", values_to = "Height") |>
-    group_by(ID) |>
-    fill(Height, .direction = "downup") |>
-    ungroup() |>
-    pivot_wider(names_from = "Year", values_from = "Height") |>
-    select(-ID)
-}
-
-process_cognition_data <- function(cog_data, prefix, focus_total = FALSE){
-  cog_data <- cog_data |>
-    select(HHID, PN, any_of(paste0(prefix, 110:117))) |>
-    mutate(
-      across(starts_with(prefix), ~ case_when(
-        .x == 1 ~ 1,
-        .x == 5 ~ 0,
-        TRUE ~ NA_real_
-      )),
-      
-      # Invert values for specific columns
-      across(all_of(paste0(prefix, c(113, 115))), ~ ifelse(.x == 1, 0, 1)),
-      
-      # Calculate the total
-      Total_dep = rowSums(across(starts_with(prefix)), na.rm = TRUE)
-    )
-  
-  if(focus_total == TRUE) {
[…]

### 2025-04-01 R/01__LWC.R

@@ -1,43 +1,0 @@
-# Creating the LW classifications for the 2022 HRS dataset
-
-rm(list = ls())
-
-# Custom made package
-# devtools::install_github("C-Monaghan/lwc2022")
-
-path_data <- "data-raw"
-
-# Reading in data --------------------------------------------------------------
-data <- haven::read_sav(here::here(path_data, "2022/Cognition_2022.sav"))
-proc <- haven::read_sav(here::here(path_data, "2020/Procrastination.sav"))
-
-# Filter only to procrastination participants 
-proc <- proc |>
-  dplyr::filter(!is.na(RV155))
-
-lw_2022_data <- data |>
-  # Getting procrastination participants
-  dplyr::semi_join(proc, by = c("HHID", "PN")) |>
-  # Removing (potential) web respondents
-  dplyr::filter(!is.na(SD182M1), is.na(SD182WM1)) |>
-  lwc2022::extract() |>
-  lwc2022::score() |>
-  lwc2022::classify()
-
-# Joining classifications to original LW datafile -------------------------------
-lw_data <-  haven::read_dta(here::here(path_data, "03__LW_Dementia.dta"))
-
-# Joining 2022 classifications
-lw_data_full <- lw_data |>
-  dplyr::rename(HHID = hhid, PN = pn) |>
-  dplyr::semi_join(lw_2022_data, by = c("HHID", "PN")) |>
-  cbind(lw_2022_data) |>
-  dplyr::distinct() |> # HHID & PN duplicate so this removes the duplication
-  dplyr::select(!c(HHID.1, PN.1))
-
-# Exporting --------------------------------------------------------------------
-export_path <- "data"
-
-writexl::write_xlsx(
-  path = here::here(export_path, "lw_data_full.xlsx"),
-  x = lw_data_full)

### 2025-04-01 R/02__Aggregation.R

@@ -1,178 +1,0 @@
-# Aggregating procrastination data with LWC data
-
-rm(list = ls())
-
-library(dplyr)
-library(tidyr)
-
-source(here::here("R/00__Functions.R"))
-
-path_data <- "data-raw"
-
-# Reading in data --------------------------------------------------------------
-# General data
-tracker    <- haven::read_sav(here::here(path_data, "02__Tracker.sav"))
-lw_data    <- readxl::read_xlsx(here::here(path_data, "../02__Processed/lw_data_full.xlsx"))
-
-# 2016 data
-health_16  <- haven::read_sav(here::here(path_data, "2016/Health_2016.sav"))
-cog_16     <- haven::read_sav(here::here(path_data, "2016/Cognition_2016.sav"))
-
-# 2018 data
-health_18  <- haven::read_sav(here::here(path_data, "2018/Health_2018.sav"))
-cog_18     <- haven::read_sav(here::here(path_data, "2018/Cognition_2018.sav"))
-
-
-# 2020 data
-health_20  <- haven::read_sav(here::here(path_data, "2020/Health_2020.sav"))
-proc       <- haven::read_sav(here::here(path_data, "2020/Procrastination.sav"))
-cog_20     <- haven::read_sav(here::here(path_data, "2020/Cognition_2020.sav"))
-
-# 2022 data
-health_22  <- haven::read_sav(here::here(path_data, "2022/Health_2022.sav"))
-cog_22     <- haven::read_sav(here::here(path_data, "2022/Cognition_2022.sav"))
-
-
-# Filter to procrastination participants ---------------------------------------
-proc_red <- proc |>
-  filter(!is.na(RV155)) |>
-  select(HHID, PN, RV156:RV167) |>
-  rename(
-    P_1 = "RV156",
-    P_2 = "RV157",
-    P_3 = "RV158",
-    P_4 = "RV159",
-    P_5 = "RV160",
-    P_6 = "RV161",
-    P_7 = "RV162",
-    P_8 = "RV163",
-    P_9 = "RV164",
-    P_10 = "RV165",
-    P_11 = "RV166",
-    P_12 = "RV167"
-  ) |>
-  # Recoding missing data values
-  mutate(
-    across(starts_with("P_"), ~ ifelse(. %in% c(-8, 8, 9), NA, .))
-    ) |>
-  # Removing participants not in the LW data file
[…]

### 2025-04-01 R/03__EDA.R

@@ -1,146 +1,0 @@
-# Data exploration -------------------------------------------------------------
-
-rm(list = ls())
-
-library(dplyr)
-library(ggplot2)
-library(ggalluvial)
-
-# source(here::here("02__Models/00__Functions.R"))
-
-count_transitions <- function(data, years) {
-  
-  # Create dynamic column names based on the years provided
-  cogfunction_cols <- paste0("cogfunction", years)
-  
-  data |>
-    select(ID, Total_p, all_of(cogfunction_cols)) |>
-    mutate(across(all_of(cogfunction_cols), ~ case_when(
-      .x == 1 ~ "Normal Cognition",
-      .x == 2 ~ "MCI",
-      .x == 3 ~ "Dementia",
-      is.na(.x) ~ "Missing",  # Handle missing values
-      TRUE ~ "Other"          # Handle other unexpected cases
-    ))) |>
-    count(across(all_of(cogfunction_cols))) |>
-    mutate(ID = row_number()) |>
-    tidyr::pivot_longer(cols = all_of(cogfunction_cols),
-                        names_to = "Wave",
-                        values_to = "Status") |>
-    mutate(Status = factor(Status,
-                           levels = c("Normal Cognition", "MCI", "Dementia", "Missing"))) |>
-    mutate(Wave = stringr::str_replace(Wave, "cogfunction", "HRS "))
-}
-
-# Plotting dementia transitions ------------------------------------------------
-plot_transitions <- function(data, size) {
-  
-  data |>
-    ggplot(aes(x = Wave, y = n, stratum = Status, fill = Status, alluvium = ID)) +
-    geom_stratum(alpha = 0.5, width = 0.5) +
-    geom_flow(width = 0.5) +
-    geom_text(stat = "stratum", aes(label = stringr::str_wrap(Status, width = 10)), 
-              hjust = 0.5, vjust = 0.5, size = size) +
-    labs(title = "Dementia transitions across time", x = "", y = "Frequency") +
-    scale_fill_viridis_d(direction = -1) +
-    theme_minimal() +
-    theme(panel.grid = element_blank(),
-          plot.title = element_text(size = 14, hjust = 0.5, face = "bold"),
-          axis.text = element_text(size = 10),
-          axis.title = element_text(size = 10),
-          text = element_text(size = 10)) +
-    ggeasy::easy_remove_legend()
-}
-
-path_data <- "data/"
-
[…]

### 2025-04-01 R/conversion.R

@@ -1,8 +1,0 @@
-
-data <- readxl::read_xlsx(here::here("data/lw_data_full.xlsx"))
-
-
-
-write.csv(x = data, file = here::here("data/data.csv"))
-saveRDS(object = data, file = here::here("data/data.rds"))
-haven::write_dta(data = data, path = here::here("data/data.dta"))

### 2025-04-01 README.qmd
Update :bug:
Update readme

@@
 ---
-title: "Procrastination and Dementia"
-format: 
-  gfm: 
-    wrap: preserve
-    toc: true
-  typst:
-    toc: true
-    toc-title: " "
-    author:
-    - name: Audrey L. Comstock
-      affiliation: Arizona State University
-      orcid: 0000-0003-1772-5350
-      url: https://audreylcomstock.weebly.com/
-      email: audrey.comstock@asu.edu
-    - name: Andrew Heiss
-      affiliation: Georgia State University
-      orcid: 0000-0002-3948-3914
-      url: https://www.andrewheiss.com
-      email: aheiss@gsu.edu
-    - name: Suparna Chaudhry
-      affiliation: Lewis and Clark College
-      orcid: 0000-0003-3506-7590
-      url: https://www.suparnachaudhry.com/
-      email: schaudhry@lclark.edu
+format: gfm
 ---
@@
 
-```{r setup, include=TRUE}
-knitr::opts_chunk$set(
-  collapse = TRUE,
-  comment = "#>",
-  fig.path = "README-",
-  fig.retina = 3
-)
+```{r setup, include=FALSE}
+library(emoji)
 ```
 
-::: {.content-visible when-format="markdown"}
-
----
-
-[Cormac Monaghan](https://c-monaghan.github.io/)
-
----
-
-:::
-
-## Note on "crisp clam" project name
+##  `r emoji_glue(":potato::spiral_shell:")`: Note on "crisp clam" project name
 

### 2025-04-03 analysis/discrete-markov.qmd
Updating DT model

@@
 title: "Discrete Time Markov Model"
-subtitle: "Examining the Association Between Procrastination and Cognitive Transitions"
 ---
 
-# Setting up {.unnumbered}
-
-## Loading packages {.unnumbered}
+# Setting up
+
+## Loading packages
 
@@
 
-## Loading functions {.unnumbered}
+## Loading functions
 
@@
 
+reshape_matrix <- function(matrix) {
+  # Converts a transition probability matrix into a tidy format suitable for visualization.
+  # Transforms the matrix into long format with explicit factor levels for cognitive states,
+  # preserving the original ordering while preparing for ggplot compatibility.
+  # Arguments:
+  #   - matrix: A square transition probability matrix with states as row/column names
+  # Returns:
+  #   - Tidy data frame with columns:
+  #     * from_state: Factor indicating origin cognitive state
+  #     * to_state: Factor indicating destination cognitive state (reversed for plotting)
+  #     * probability: Numeric transition probability values
+
+  matrix |>
+    as.data.frame() |>
+    mutate(
+      from_state = factor(c(
+        "Normal cognition", "MCI", "Dementia"),
+        levels = c("Normal cognition", "MCI", "Dementia"))
+    ) |>
+    reshape2::melt(
+      id.vars = "from_state", 
+      variable.name = "to_state", 
+      value.name = "probability") |>
+    mutate(to_state = factor(to_state, levels = rev(levels(to_state))))
+}
+
+plot_transition_matrix <- function(matrix, observed = TRUE) {
+  # Creates a heatmap visualization of cognitive state transition probabilities.
+  # Generates either observed or estimated transition plots with consistent formatting,
+  # including labeled probability values and a diverging color scale for emphasis.
+  # Arguments:
+  #   - matrix: Tidy transition matrix from reshape_matrix()
+  #   - observed: Logical flag indicating whether data represents observed (TRUE) 
+  #               or estimated (FALSE) transitions
+  # Returns:
+  #   - ggplot heatmap object with:
+  #     * State transitions as cells
+  #     * Probability values displayed numerically
+  #     * Custom color scale and axis formatting
+
+  if(observed == TRUE) {
+    subtitle <- "Observed state transitions between assessment waves"
+  } else {
+    subtitle <- "Estimated state transitions between assessment waves"
+  }
+
+  matrix |>
+    ggplot(aes(x = from_state, y = to_state, fill = probability)) +
+    geom_tile(color = "white", linewidth = 0.5) +
+    geom_text(
+      aes(label = format(round(probability, 3), nsmall = 3)),
+      size = 4.5, 
+      color = "#212427",
+      fontface = "bold") +
+    colorspace::scale_fill_continuous_diverging(
+      palette = "Blue-Red 3", mid = 0.50, alpha = 0.5, 
+      limits = c(0, 1), name = "Transition \nProbability") +
+    labs(title = "Transition Probabilities Across Cognitive States",
+         subtitle = subtitle,
+         x = "Previous State (t - 1)", 
+         y = "Current State (t)") +
+    theme(
+      axis.text = element_text(size = 10),
+      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
+      legend.position = "right",
+      legend.text = element_text(size = 9),
+      panel.grid = element_blank()
+    )
+}
+
 tran_stack_graph <- function(data) {
@@
     ggeasy::easy_move_legend("bottom")
+}
+
+tidy_output <- function(fit) {
+  output <- broom::tidy(fit, conf.int = TRUE) |>
+    filter(term != "(Intercept)") |>
+    mutate(
+      term = case_when(
+      term == "Gender1" ~ "Being female",
+      term == "Education_tri1" ~ "High school degree vs. No education",
+      term == "Education_tri2" ~ "Further education vs. No education",
+      term == "Total_dep_2016" ~ "Depression Scores (2016)",
+      term == "Total_p" ~ "Procrastination",
+      term == "status_prev2" ~ "Previous state: MCI",
+      term == "status_prev3" ~ "Previous state: Dementia",
+      term == "wave" ~ "Time",
+      TRUE ~ term)) |>
+  mutate(across(c(estimate:p.value), ~ round(x = ., digits = 3)))
+
+  if(any(output$y.level == "1")) {
+    output |>
+      mutate(y.level = case_when(
+      y.level == "1" ~ "MCI - NC",
+      y.level == "3" ~ "MCI - Dementia"))
+  } else {
+    output |>
+      mutate(y.level = case_when(
+      y.level == "2" ~ "NC - MCI",
+      y.level == "3" ~ "NC - Dementia"))
+  }
 }
@@
 
-plot_predictions_stationary <- function(predictions) {
+plot_predictions_stationary <- function(predictions, variable, x_axis) {
   # Visualizes predicted transition probabilities from stationary multi-state models.
@@
   predictions |>
-    ggplot(aes(x = Total_p, y = prob, colour = status_prev)) +
+    ggplot(aes(x = {{variable}}, y = prob, colour = status_prev)) +
     geom_line(linewidth = 1) +
@@
       title = "Predicted transition probabilities (stationary model)",
-      # subtitle = "According to previous cognitive state",
-      x = "Total Procrastination", y = "Probability", colour = "Previous State") +
+      x = x_axis, y = "Probability", colour = "Previous State") +
     facet_wrap(~ status, labeller = labeller(
@@
     Gender = factor(0),
+    Age = mean(data_stack$Age),
+    Education_tri = factor(0),
+    Total_dep_2016 = mean(data_stack$Total_dep_2016),
     Total_p = mean(data_stack$Total_p),
@@
 
-## Setting theme {.unnumbered}
+## Setting theme
 
@@
   ungroup() |> 
-  filter(wave != 2016, !is.na(Total_p))
+  filter(wave != 2016, !is.na(Total_p), Age >= 50)
 
@@
 
-tran_df <- transition_matrix_observed |>
-  as.data.frame() |>
-  mutate(
-    from_state = factor(c("Normal cognition", "MCI", "Dementia"),
-                        levels = c("Normal cognition", "MCI", "Dementia"))
-    ) |>
-  reshape2::melt(
-    id.vars = "from_state", variable.name = "to_state", value.name = "probability") |>
-  mutate(
-    to_state = factor(to_state, levels = rev(levels(to_state)))
-    )
-
-fig_2 <- tran_df |>
-  ggplot(aes(x = from_state, y = to_state, fill = probability)) +
-  geom_tile(color = "white", linewidth = 0.5) +
-  geom_text(
-    aes(label = format(round(probability, 3), nsmall = 3)),
-    size = 4.5, 
-    color = "#212427",
-    fontface = "bold") +
-  colorspace::scale_fill_continuous_diverging(
-    palette = "Blue-Red 3", mid = 0.50, alpha = 0.5, 
-    limits = c(0, 1), name = "Transition \nProbability") +
-  labs(title = "Transition Probabilities Across Cognitive States",
-       subtitle = "Observed state transitions between assessment waves",
-       x = "Previous State (t - 1)", 
-       y = "Current State (t)") +
-  theme(
-    axis.text = element_text(size = 10),
-    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
-    legend.position = "right",
-    legend.text = element_text(size = 9),
-    panel.grid = element_blank()
-  )
-
-fig_2
+fig_3 <- transition_matrix_observed |>
+  reshape_matrix() |>
+  plot_transition_matrix()
+
+fig_3
 ```
@@
 #| 
-# Baseline model (only gender)
-fit_1 <- nnet::multinom(
-  status ~ Gender, family = multinomial, 
+# Baseline model (only gender; reference: NC)
+fit_1a <- nnet::multinom(
+  status ~ Gender + Age + Education_tri + Total_dep_2016, family = multinomial, 
   data = data_stack, trace = FALSE)
 
-# With procrastination
-fit_2 <- nnet::multinom(
-  status ~ Gender + Total_p, family = multinomial, 
+# Baseline model (only gender; reference: MCI)
+fit_1b <- nnet::multinom(
+  status ~ Gender + Age + Education_tri + Total_dep_2016, family = multinomial, 
+  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
+  trace = FALSE)
+
+# With procrastination (reference: NC)
+fit_2a <- nnet::multinom(
+  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p, family = multinomial, 
   data = data_stack, trace = FALSE)
 
-# Full model with previous state
-fit_3 <- nnet::multinom(
-  status ~ Gender + Total_p + status_prev, family = multinomial, 
+# With procrastination (reference: MCI)
+fit_2b <- nnet::multinom(
+  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p, family = multinomial, 
+  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
+  trace = FALSE)
+
+# Full model with previous state (reference NC)
+fit_3a <- nnet::multinom(
+  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev, family = multinomial, 
   data = data_stack, trace = FALSE)
+
+# Full model with previous state (reference MCI)
+fit_3b <- nnet::multinom(
+  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev, family = multinomial, 
+  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
+  trace = FALSE)
 ```
@@
 ```{r model-comparison}
-anova(fit_1, fit_2, fit_3) |>
+anova(fit_1a, fit_2a, fit_3a) |>
   knitr::kable(caption = "Nested Model Comparison") |>
@@
 
+Below we present the estimated parameters for the stationary models. We will transform the estimates to odds ratios and code them based on significance.
+
 ```{r model-output}
@@
 #| 
-broom::tidy(fit_3) |>
-  filter(term != "(Intercept)") |>
+stationary_results <- rbind(tidy_output(fit_3a), tidy_output(fit_3b)) |>
   mutate(
-    y.level = case_when(
-      y.level == "2" ~ "NC - MCI",
-      y.level == "3" ~ "NC - Dementia"),
-    term = case_when(
-      term == "Gender1" ~ "Being female",
-      term == "Total_p" ~ "Procrastination",
-      term == "status_prev2" ~ "Previous state: MCI",
-      term == "status_prev3" ~ "Previous state: Dementia",
-      TRUE ~ term)
-    ) |>
-  mutate(across(c(estimate:p.value), ~ round(x = ., digits = 3)))
+    # Transforming to odds ratios
+    estimate = exp(estimate),
+    conf.low = exp(conf.low),
+    conf.high = exp(conf.high),
+
+    # Mutating to factors
+    y.level = factor(
+      y.level, 
+      levels = c("NC - MCI", "MCI - NC", 
+                 "NC - Dementia", "MCI - Dementia")),
+    term = factor(
+      term, 
+      levels = c("Being female", "Age", "High school degree vs. No education",
+                 "Further education vs. No education", "Depression Scores (2016)",
+                 "Procrastination", "Previous state: MCI", "Previous state: Dementia")),
+
+    # Creating a colour code
+    positive = case_when(
+      estimate > 1 & p.value < 0.05 ~ "Positive",
+      estimate < 1 & p.value < 0.05 ~ "Negative",
+      TRUE ~ "NS"),
+
+    positive = factor(positive, levels = c("Positive", "Negative", "NS"))
+    )
+
+head(stationary_results)
 ```
@@
 
-Let's visualize the predicted transition probabilities from the stationary models. We will show probability curves by procrastination level
+Let's visualise the estimated odds ratios (@fig-odds-ratio) and the predicted transition probabilities (@fig-model-predictions) for the stationary model .
+
+```{r}
+#| message: false
+#| warning: false
+#| code-fold: true
+#| code-summary: "Check out my code"
+#| fig-width: 12
+#| label: fig-odds-ratio
+#| fig-cap: "Estimated odds ratios for the stationary model."
+
+fig_4a <- stationary_results |>
+  filter(!term %in% c("Previous state: MCI", "Previous state: Dementia")) |>
+  ggplot(aes(x = estimate, y = y.level, colour = positive)) +
+  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
+  ggstance::geom_pointrangeh(
+    aes(xmin = conf.low, xmax = conf.high),
+    position = ggstance::position_dodgev(height = 0.5),
+    size = 1.25,
+    fatten = 3) +
+  scale_colour_manual(values = c(
+    "Positive" = "#0072B2", 
+    "Negative" = "#E69F00", 
+    "NS"       = "#B2BEB5")) +
+  labs(title = "Odds Ratios from Multinomial Model",
+       x = "Odds Ratio", y = "Predictor") +
+  guides(colour = "none") +
+  facet_wrap(~ term, scales = "free_x") +
+  theme_bw() +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
+    axis.title = element_text(size = 12, face = "bold"),
+    strip.text = element_text(size = 10, face = "bold"),
+    panel.grid.major.y = element_blank(),
+    panel.grid.minor.y = element_blank())
+
+fig_4a
+```
 
@@
   Gender = factor(0),
+  Age = mean(data_stack$Age),
+  Education_tri = factor(0),
+  Total_dep_2016 = mean(data_stack$Total_dep_2016),
   status_prev = levels(data_stack$status_prev),
@@
 ## Plotting predictions
-fig_3 <- pred_data |>
-  modelr::add_predictions(model = fit_3, var = "pred", type = "probs") |>
+fig_4b <- pred_data |>
+  modelr::add_predictions(model = fit_3a, var = "pred", type = "probs") |>
   tidy_predictions() |>
-  plot_predictions_stationary()
-
-fig_3
+  plot_predictions_stationary(variable = Total_p, x_axis = "Total Procrastination")
+
+fig_4b
 ```
@@
   Gender = factor(0),
+  Age = mean(data_stack$Age),
+  Education_tri = factor(0),
+  Total_dep_2016 = mean(data_stack$Total_dep_2016),
   Total_p = mean(data_stack$Total_p),
   status_prev = states) |>
-  modelr::add_predictions(model = fit_3, var = "pred", type = "probs")
+  modelr::add_predictions(model = fit_3a, var = "pred", type = "probs")
 
@@
 
-tran_df_estimate <- transition_matrix_estimated |>
-  as.data.frame() |>
-  mutate(from_state = factor(
-    c("Normal cognition", "MCI", "Dementia"),
-    levels = c("Normal cognition", "MCI", "Dementia")
-  )) |>
-  reshape2::melt(id.vars = "from_state",
-                 variable.name = "to_state",
-                 value.name = "probability") |>
-  mutate(to_state = factor(to_state, levels = rev(levels(to_state))))
-
-fig_4 <- tran_df_estimate |>
-  ggplot(aes(x = from_state, y = to_state, fill = probability)) +
-  geom_tile(color = "white", linewidth = 0.5) +
-  geom_text(
-    aes(label = format(round(probability, 3), nsmall = 3)),
-    size = 4.5, 
-    color = "#212427",
-    fontface = "bold") +
-  colorspace::scale_fill_continuous_diverging(
-    palette = "Blue-Red 3", mid = 0.50, alpha = 0.5, 
-    limits = c(0, 1), name = "Transition \nProbability") +
-  labs(title = "Transition Probabilities Across Cognitive States",
-       subtitle = "Estimated state transitions between assessment waves",
-       x = "Previous State (t - 1)", 
-       y = "Current State (t)") +
-  theme(
-    axis.text = element_text(size = 10),
-    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
-    legend.position = "right",
-    legend.text = element_text(size = 9),
-    panel.grid = element_blank()
-  )
-
-fig_2 + fig_4 + plot_layout(axis_titles = "collect", guides = "collect")
+fig_5 <- transition_matrix_estimated |>
+  reshape_matrix() |>
+  plot_transition_matrix(observed = FALSE)
+
+fig_3 + fig_5 + plot_layout(axis_titles = "collect", guides = "collect")
 ```
@@
 # Model 4: Additive time effects
-fit_4 <- nnet::multinom(
-  status ~ Gender + Total_p + status_prev + wave, family = multinomial, 
+fit_4a <- nnet::multinom(
+  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev + wave, family = multinomial, 
   data = data_stack, trace = FALSE)
 
+fit_4b <- nnet::multinom(
+  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev + wave, family = multinomial, 
+  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
+  trace = FALSE)
+
 # Model 5: State-specific time effects
-fit_5 <- nnet::multinom(
-  status ~ Gender + Total_p + (status_prev * wave), family = multinomial, 
+fit_5a <- nnet::multinom(
+  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + (status_prev * wave), family = multinomial, 
   data = data_stack, trace = FALSE)
 
+fit_5b <- nnet::multinom(
+  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + (status_prev * wave), family = multinomial, 
+  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
+  trace = FALSE)
+
 # Model 6: Full time interactions
-fit_6 <- nnet::multinom(
-  status ~ (Gender + Total_p + status_prev) * wave, family = multinomial, 
+fit_6a <- nnet::multinom(
+  status ~ (Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev) * wave, family = multinomial, 
   data = data_stack, trace = FALSE)
+
+fit_6b <- nnet::multinom(
+  status ~ (Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev) * wave, family = multinomial, 
+  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
+  trace = FALSE)
 ```
@@
 
-```{r}
+```{r non-stationary-goodness}
 #| code-fold: true
@@
 models <- list(
-  "Stationary" = fit_3,
-  "Additive Time" = fit_4,
-  "State-Time Interaction" = fit_5,
-  "Full Interactions" = fit_6
+  "Stationary" = fit_3a,
+  "Additive Time" = fit_4a,
+  "State-Time Interaction" = fit_5a,
+  "Full Interactions" = fit_6a
 )
@@
 comparison_table <- purrr::map_dfr(models, broom::glance, .id = "Model") |>
-  select(Model:AIC) |>
+  select(edf:AIC) |>
   mutate(across(where(is.numeric), \(x) round(x, digits = 1)))
 
-# Likelihood ratio tests
-lr_tests <- anova(fit_3, fit_4, fit_5, fit_6)
-
-# Outputting tables
-knitr::kable(lr_tests, caption = "Likelihood Ratio Tests") |>
-  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
-
-knitr::kable(comparison_table, caption = "Model Fit Statistics") |>
-  kableExtra::kable_styling(bootstrap_options = c("striped", "hover"))
+# Outputting as table
+anova(fit_3a, fit_4a, fit_5a, fit_6a) |>
+  mutate(
+    Model = c("Stationary", "Additive", "State-Time Interactions", "Full Interactions"),
+    `Resid. Dev` = round(`Resid. Dev`, digits = 3),
+    `LR stat.` = round(`LR stat.`, digits = 3),
+    `Pr(Chi)` = round(`Pr(Chi)`, digits = 3)
+    ) |>
+  cbind(comparison_table) |>
+  DT::datatable(
+    options = list(
+      columnDefs = list(list(className = 'dt-center', targets = "_all"))
+    ),
+    rownames = FALSE,
+    caption = "Likelihood Ratio Tests"
+  )
 ```
@@
 ```{r non-stationary-output}
-#| code-fold: true
-#| code-summary: "Check out my code"
-
-broom::tidy(fit_4) |>
-  filter(term != "(Intercept)") |>
-  mutate(
-    y.level = case_when(
-      y.level == "2" ~ "NC - MCI",
-      y.level == "3" ~ "NC - Dementia"),
-    term = case_when(
-      term == "Gender1" ~ "Being female",
-      term == "Total_p" ~ "Procrastination",
-      term == "status_prev2" ~ "Previous state: MCI",
-      term == "status_prev3" ~ "Previous state: Dementia",
-      term == "wave" ~ "Time",
-      TRUE ~ term)
-    ) |>
-  mutate(across(c(estimate:p.value), ~ round(x = ., digits = 3)))
+non_stationary_results <- rbind(tidy_output(fit_4a), tidy_output(fit_4b))
+
+head(non_stationary_results)
 ```
@@
   wave = unique(data_stack$wave)) |>
-  modelr::add_predictions(model = fit_4, var = "pred", type = "probs") |>
+  modelr::add_predictions(model = fit_4a, var = "pred", type = "probs") |>
   tidy_predictions() |>
@@
   setNames(time_points, paste(time_points, "Years")),
-  ~ get_time_varying_matrix(fit_4, .x)
+  ~ get_time_varying_matrix(fit_4a, .x)
 )

### 2025-04-09 analysis/discrete-markov.qmd
Site re-render :collision:

@@
   #
+
+  # Pivoting depression variables
+  # depression <- data |>
+  #   select(starts_with("Total_dep")) |>
+  #   tidyr::pivot_longer(
+  #     cols = everything(), names_to = "Remove", values_to = "Depression") |>
+  #   select(Depression)
+
+  # Pivoting cardio variables
+  # cardio <- data |>
+  #   select(starts_with("Cardio_risk")) |>
+  #   tidyr::pivot_longer(
+  #     cols = everything(), names_to = "Remove", values_to = "Cardio") |>
+  #   select(Cardio)
+
   data |>
+    # select(!starts_with(c("Total_dep", "Cardio_risk"))) |>
     tidyr::pivot_longer(
@@
     relocate(status, .after = wave)
+    #cbind(depression, cardio)
 }
@@
 
-tidy_output <- function(fit) {
+tidy_output <- function(fit, multiplicative = FALSE) {
   output <- broom::tidy(fit, conf.int = TRUE) |>
     filter(term != "(Intercept)") |>
+  mutate(across(c(estimate:p.value), ~ round(x = ., digits = 3)))
+
+  if(any(output$y.level == "1")) {
+    output <- output |>
+      mutate(y.level = case_when(
+      y.level == "1" ~ "MCI - NC",
+      y.level == "3" ~ "MCI - Dementia"))
+  } else {
+    output <- output |>
+      mutate(y.level = case_when(
+      y.level == "2" ~ "NC - MCI",
+      y.level == "3" ~ "NC - Dementia"))
+  }
+
+  if(multiplicative == FALSE) {
+    output <- output |>
       mutate(
@@
       term == "Total_dep_2016" ~ "Depression Scores (2016)",
-      term == "Total_p" ~ "Procrastination",
+      term == "Total_p" ~ "Procrastination (2020)",
       term == "status_prev2" ~ "Previous state: MCI",
@@
       term == "wave" ~ "Time",
-      TRUE ~ term)) |>
-  mutate(across(c(estimate:p.value), ~ round(x = ., digits = 3)))
-
-  if(any(output$y.level == "1")) {
-    output |>
-      mutate(y.level = case_when(
-      y.level == "1" ~ "MCI - NC",
-      y.level == "3" ~ "MCI - Dementia"))
+      TRUE ~ term))
   } else {
-    output |>
-      mutate(y.level = case_when(
-      y.level == "2" ~ "NC - MCI",
-      y.level == "3" ~ "NC - Dementia"))
-  }
+    output <- output |>
+      mutate(
+      term = case_when(
+      term == "Gender1" ~ "Being female",
+      term == "Education_tri1" ~ "High school degree vs. No education",
+      term == "Education_tri2" ~ "Further education vs. No education",
+      term == "Depression" ~ "Depression Scores (2016)",
+      term == "Total_p" ~ "Procrastination (2020)",
+      term == "status_prev2" ~ "Previous state: MCI",
+      term == "status_prev3" ~ "Previous state: Dementia",
+      term == "wave" ~ "Time",
+      term == "Gender1:wave" ~ "Gender & Time",
+      term == "Age:wave" ~ "Age & Time",
+      term == "Education_tri1:wave" ~ "Education (0 - 1) & Time",
+      term == "Education_tri2:wave" ~ "Education (0 - 2) & Time",
+      term == "Depression:wave" ~ "Depression & Time",
+      term == "Total_p:wave" ~ "Procrastination & Time",
+      term == "status_prev2:wave" ~ "Previous state: MCI & Time",
+      term == "status_prev3:wave" ~ "Previous state: Dementia & Time",
+      TRUE ~ term))
+  }
+
+  return(output)
 }
@@
 
+create_additive_predictions <- function(
+    data, model, var, var_seq, hold_constant = list(Gender = factor(0), Education_tri = factor(0))) {
+
+  # Set default prediction sequence based on variable type
+  if (is.null(var_seq)) {
+    if (is.numeric(data[[var]])) {
+      predict_seq <- seq(min(data[[var]], max(data[[var]]), length = 200))
+    } else if (is.factor(data[[var]])) {
+      predict_seq <- levels(data[[var]])
+    } else {
+      predict_seq <- unique(data[[var]])
+    }
+  }
+
+  # Create list of all variables needed for prediction
+  all_vars <- all.vars(formula(model)[-2])  # Get RHS variables from model formula
+
+  # Create base grid with all variables EXCEPT the prediction variable
+  base_grid <- expand.grid(
+    lapply(all_vars, function(v) {
+      if (v == var) return(NULL)  # Skip prediction variable
+
+      # Special cases - keep all levels
+      if (v %in% c("status_prev", "wave")) {
+        return(unique(data[[v]]))
+      }
+
+      if (v %in% names(hold_constant)) {
+        # Use user-specified constant value
+        hold_constant[[v]]
+      } else if (is.numeric(data[[v]])) {
+        # Use mean for numeric variables
+        mean(data[[v]], na.rm = TRUE)
+      } else if (is.factor(data[[v]])) {
+        # Use first level for factors
+        factor(levels(data[[v]])[1], levels = levels(data[[v]]))
+      } else {
+        # Default to most common value
+        names(sort(table(data[[v]]), decreasing = TRUE))[1]
+      }
+    }) |> purrr::compact() |> setNames(all_vars[all_vars != var]),
+    KEEP.OUT.ATTRS = FALSE
+  )
+
+  # Create full prediction grid
+  pred_grid <- base_grid |> 
+    tidyr::crossing(!!var := var_seq) |>
+    modelr::add_predictions(model = model, var = "pred", type = "probs") |>
+    tidy_predictions() |>
+    mutate(wave = factor(wave))
+
+  return(pred_grid)
+}
+
+plot_additive_predictions <- function(data, var, x_label, subtitle) {
+
+  # Generating predictions
+  data |>
+    ggplot(aes(x = as.numeric(.data[[var]]), y = prob, color = status_prev)) +
+    geom_line(linewidth = 1) +
+    ggokabeito::scale_color_okabe_ito() +
+    labs(
+      title = "Transition Probabilities (Additive Model)",
+      subtitle = subtitle,
+      x = x_label, 
+      y = "Predicted Probability", 
+      color = "Previous State"
+    ) +
+    facet_grid(wave ~ status, labeller = labeller(
+      wave = function(x) case_when(
+        x == 2 ~ "Year = 2018",
+        x == 3 ~ "Year = 2020",
+        x == 4 ~ "Year = 2022",
+      ),
+      status = function(x) paste("Transition to:", x)
+    )) +
+    theme(
+      plot.subtitle = element_text(hjust = 0.5, size = 12, face = "bold"),
+      panel.spacing = unit(1, "lines")) 
+}
+
 get_time_varying_matrix <- function(model, time_point) {
@@
 ```{r covariates}
-cols <- c("ID", "Gender", "Age", "Education_tri", "Cardio_risk_16", "Total_dep_2016", "Total_p")
+cols <- c(
+  "ID", "Gender", "Age", "Education_tri",
+  paste0("Cardio_risk_", seq(16, 22, by = 2)), 
+  paste0("Total_dep_", seq(2016, 2022, by = 2)),
+  "Total_p")
 ```
@@
 ```{r extract-years}
-data_stack <- data |> extract_years(years = seq(2016, 2022, by = 2), impute = TRUE, absorbing = TRUE) |> na.omit()
+data_stack <- data |> 
+  extract_years(years = seq(2016, 2022, by = 2), impute = TRUE, absorbing = TRUE) |>
+  na.omit()
 
@@
   mutate(status_prev = lag(status), .after = status) |>
+  filter(!is.na(Total_p), Age >= 50) |>
   ungroup() |> 
-  filter(wave != 2016, !is.na(Total_p), Age >= 50)
+  filter(wave != 2016)
 
@@
 fit_1a <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Total_dep_2016, family = multinomial, 
+  status ~ Gender + Age + Education_tri + Total_dep_2016, 
+  family = multinomial, 
   data = data_stack, trace = FALSE)
@@
 fit_1b <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Total_dep_2016, family = multinomial, 
+  status ~ Gender + Age + Education_tri + Total_dep_2016, 
+  family = multinomial, 
   data = data_stack |> mutate(status = relevel(status, ref = 2)), 
@@
 fit_2a <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p, family = multinomial, 
+  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p, 
+  family = multinomial, 
   data = data_stack, trace = FALSE)
@@
 fit_2b <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p, family = multinomial, 
+  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p, 
+  family = multinomial, 
   data = data_stack |> mutate(status = relevel(status, ref = 2)), 
@@
 fit_3a <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev, family = multinomial, 
+  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev, 
+  family = multinomial, 
   data = data_stack, trace = FALSE)
@@
 fit_3b <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev, family = multinomial, 
+  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev, 
+  family = multinomial, 
   data = data_stack |> mutate(status = relevel(status, ref = 2)), 
@@
 
-We evaluate model improvement using likelihood ratio tests:
+We evaluate model improvement (@tbl-stationary-comparison) using likelihood ratio tests:
 
@@
 ```{r model-comparison}
+#| label: tbl-stationary-comparison
+#| tbl-cap: "Likelihood ratio test for stationary models"
+#| tbl-cap-location: bottom
+
 anova(fit_1a, fit_2a, fit_3a) |>
-  knitr::kable(caption = "Nested Model Comparison") |>
-  kableExtra::kable_styling(
-    bootstrap_options = c("striped", "hover"),
-    full_width = FALSE)
+  mutate(
+    Model = c("Baseline", "Procrastination", "Procrastination + Previous Status"),
+    `Resid. Dev`= round(`Resid. Dev`, digits = 3),
+    `LR stat.` = round(`LR stat.`, digits = 3),
+    `Pr(Chi)` = round(`Pr(Chi)`, digits = 3)
+    ) |>
+  DT::datatable(
+    options = list(
+      pageLength = 3,
+      dom = "t",
+      ordering = FALSE,
+      columnDefs = list(list(className = "dt-center", targets = "_all"))
+    ),
+    rownames = FALSE
+  )
 ```
@@
 
-Below we present the estimated parameters for the stationary models. We will transform the estimates to odds ratios and code them based on significance.
+Below we present the estimated parameters for the stationary models (@tbl-stationary-output). We will transform the estimates to odds ratios and colour them based on significance.
 
@@
 #| code-summary: "Check out my code"
-#| 
+#| label: tbl-stationary-output
+#| tbl-cap: "Estimated parameters for the stationary model"
+#| tbl-cap-location: bottom
+
 stationary_results <- rbind(tidy_output(fit_3a), tidy_output(fit_3b)) |>
+  rename(transition = y.level) |>
   mutate(
@@
     # Mutating to factors
-    y.level = factor(
-      y.level, 
+    transition = factor(
+      transition, 
       levels = c("NC - MCI", "MCI - NC", 
@@
                  "Further education vs. No education", "Depression Scores (2016)",
-                 "Procrastination", "Previous state: MCI", "Previous state: Dementia")),
+                 "Procrastination (2020)", "Previous state: MCI", 
+                 "Previous state: Dementia")),
 
     # Creating a colour code
-    positive = case_when(
+    colour = case_when(
       estimate > 1 & p.value < 0.05 ~ "Positive",
@@
 
-    positive = factor(positive, levels = c("Positive", "Negative", "NS"))
-    )
-
-head(stationary_results)
+    colour = factor(colour, levels = c("Positive", "Negative", "NS"))
+    )
+
+stationary_results |>
+  mutate(across(c(estimate, conf.low, conf.high), ~ round(x = ., digits = 3))) |>
+  DT::datatable(
+    options = list(
+      pageLength = 6,
+      dom = "tip",
+      ordering = FALSE,
+      columnDefs = list(list(className = "dt-center", targets = "_all"))
+    ),
+    rownames = FALSE
+  )
 ```
@@
   filter(!term %in% c("Previous state: MCI", "Previous state: Dementia")) |>
-  ggplot(aes(x = estimate, y = y.level, colour = positive)) +
+  ggplot(aes(x = estimate, y = transition, colour = colour)) +
   geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
@@
     "NS"       = "#B2BEB5")) +
-  labs(title = "Odds Ratios from Multinomial Model",
+  labs(title = "Odds ratios (stationary model)",
        x = "Odds Ratio", y = "Predictor") +
@@
 
-Again, we will evaluate model improvement using likelihood ratio tests (@eq-liklihood). However, this time, we will also compare the best fitted non-stationary models `fit_3`.
+Again, we will evaluate model improvement using likelihood ratio tests (@eq-liklihood; @tbl-non-stationary-comparison).
+However, this time, we will also compare the best fitted non-stationary models `fit_3`.
 
@@
 #| code-summary: "Check out my code"
-#| layout-nrow: 2
+#| label: tbl-non-stationary-comparison
+#| tbl-cap: "Likelihood ratio test for non-stationary models"
+#| tbl-cap-location: bottom
 
@@
     options = list(
+      pagelength = 4,
+      dom = "t",
+      ordering = FALSE,
       columnDefs = list(list(className = 'dt-center', targets = "_all"))
     ),
+    rownames = FALSE
+  )
+```
+
+**This is interesting!! Looks like adding in time as an additive term improved things. However, having full time interactions is better than having state-time interactions**
+
+::: {.panel-tabset .nav-pills}
+## Additive Model
+
+### Estimated parameters
+
+Key coefficients from the selected non-stationary model `fit_4`
+
+```{r non-stationary-output}
+additive_results <- rbind(tidy_output(fit_4a), tidy_output(fit_4b))
+
+additive_results |>
+  rename(transiton = y.level) |>
+  mutate(across(c(estimate, conf.low, conf.high), ~ round(x = ., digits = 3))) |>
+  DT::datatable(
+    options = list(
+      pageLength = 6,
+      ordering = FALSE,
+      columnDefs = list(list(className = "dt-center", targets = "_all"))
+    ),
     rownames = FALSE,
-    caption = "Likelihood Ratio Tests"
-  )
-```
-
-**Looks like adding in time as an additive term improved things**
-
-### Estimated parameters
-
-Key coefficients from the selected non-stationary model `fit_4`
-
-```{r non-stationary-output}
-non_stationary_results <- rbind(tidy_output(fit_4a), tidy_output(fit_4b))
-
-head(non_stationary_results)
+    colnames = c("Transition", "Predictor", "Estimate", "Standard Error",
+                 "Statistic", "p-val", "Lower CI", "Upper CI")
+  )
 ```
@@
 
-Let's visualize the predicted transition probabilities from the non-stationary model @fig-non-stationary. We will show probability curves by procrastination level
-
-```{r non-stationary-visualise}
+Let's visualize the predicted transition probabilities from the non-stationary model.
+
+- @fig-non-stationary-age shows the predicted transition probabilities for age
+- @fig-non-stationary-depression shows the predicted transition probabilities for depression
+- @fig-non-stationary-education shows the predicted transition probabilities for per education level
+- @fig-non-stationary-procrastination shows the predicted transition probabilities for procrastination
+
+::: {.panel-tabset .nav-pills}
+
+## Age
+
+```{r}
 #| code-fold: true
 #| code-summary: "Check our my code"
-#| fig-width: 12
-#| fig-height: 10
-#| label: fig-non-stationary
-#| fig-cap: "Predicted transition probabilities from the non-stationary model."
+#| fig-width: 10
+#| fig-height: 8
+#| label: fig-non-stationary-age
+#| fig-cap: "Predicted transition probabilities from the non-stationary model (for age)."
 #|
-expand.grid(
-  Gender = factor(0),
-  Age = mean(data_stack$Age),
-  Education_tri = factor(0),
-  Total_dep_2016 = mean(data_stack$Total_dep_2016),
-  status_prev = levels(data_stack$status_prev),
-  Total_p = seq(0, 60, length = 200),
-  wave = unique(data_stack$wave)) |>
-  modelr::add_predictions(model = fit_4a, var = "pred", type = "probs") |>
-  tidy_predictions() |>
-  mutate(wave = factor(wave)) |>
-  ggplot(aes(x = Total_p, y = prob, color = status_prev)) +
-  geom_line(linewidth = 1) +
-  ggokabeito::scale_color_okabe_ito() +
-  labs(
-    title = "Time-Varying Transition Probabilities",
-    x = "Total Procrastination", 
-    y = "Predicted Probability", color = "Previous State") +
-  facet_grid(wave ~ status, labeller = labeller(
-    wave = function(x) paste("Years since baseline:", x),
-    status = function(x) paste("Transition to:", x)
-  )) +
-  theme(panel.spacing = unit(1, "lines"))
-
-```
+
+create_additive_predictions(data = data_stack, model = fit_4a, var = "Age", var_seq = seq(50, 97, by = 5)) |>
+  plot_additive_predictions(var = "Age", x_label = "Age", subtitle = "By Age")
+
+```
+
+## Depression
+
+```{r}
+#| code-fold: true
+#| code-summary: "Check our my code"
+#| fig-width: 10
+#| fig-height: 8
+#| label: fig-non-stationary-depression
+#| fig-cap: "Predicted transition probabilities from the non-stationary model (for depression)."
+#|
+
+create_additive_predictions(data = data_stack, model = fit_4a, var = "Total_dep_2016", var_seq = seq(0, 8, by = 1)) |>
+  plot_additive_predictions(var = "Total_dep_2016", x_label = "Depression", subtitle = "By depression")
+
+```
+
+## Education Level
+
+```{r}
+#| code-fold: true
+#| code-summary: "Check our my code"
+#| fig-width: 10
+#| fig-height: 8
+#| label: fig-non-stationary-education
+#| fig-cap: "Predicted transition probabilities from the non-stationary model (for having a high school degree)."
+
+create_additive_predictions(data = data_stack, model = fit_4a, var = "Education_tri", var_seq = factor(c(0:2)), hold_constant = list(Gender = factor(0))) |>
+  plot_additive_predictions(var = "Education_tri", x_label = "Education", subtitle = "By education level") +
+  geom_point(size = 2) +
+  scale_x_continuous(breaks = c(0, 1, 2), labels = c("No Education", "High School", "Further Education")) +
+  theme(axis.text.x = element_text(size = 8, angle = 60, vjust = 0.90, hjust = 1))
+```
+
+## Procrastination
+
+```{r non-stationary-visualise-procrastination}
+#| code-fold: true
+#| code-summary: "Check our my code"
+#| fig-width: 10
+#| fig-height: 8
+#| label: fig-non-stationary-procrastination
+#| fig-cap: "Predicted transition probabilities from the non-stationary model (for procrastination)."
+#|
+
+create_additive_predictions(data = data_stack, model = fit_4a, var = "Total_p", var_seq = seq(0, 60, by = 1)) |>
+  plot_additive_predictions(var = "Total_p", x_label = "Procrastination", subtitle = "By procrastination")
+
+```
+:::
 
@@
 
-Let's visualize the estimated time-varying transition matrices @fig-non-stationary-matrix-plot
+@fig-non-stationary-matrix-plot shows the estimated time-varying transition matrices
 
@@
 
+## Multiplicative Model
+
+### Estimated parameters
+
+Key coefficients from the selected non-stationary model `fit_6`
+
+```{r multiplicative-output}
+multiplicative_output <- rbind(
+  tidy_output(fit_6a, multiplicative = TRUE),
+  tidy_output(fit_6b, multiplicative = TRUE))
+
+multiplicative_output |>
+  mutate(across(c(estimate, conf.low, conf.high), ~ round(x = ., digits = 3))) |>
+  DT::datatable(
+    options = list(
+      pageLength = 6,
+      ordering = FALSE,
+      columnDefs = list(list(className = "dt-center", targets = "_all"))
+    ),
+    rownames = FALSE,
+    colnames = c("Transition", "Predictor", "Estimate", "Standard Error",
+                 "Statistic", "p-val", "Lower CI", "Upper CI")
+  )
+
+```
+:::
+
 # Matrix distance metrics
@@
 ```
+
+```{r test-code}
+#| include: false
+#| eval: false
+preds <- create_additive_predictions(data = data_stack, model = fit_4a, var = "Age", var_seq = seq(50, 97, by = 5)) |>
+  # Add color column based on transition significance
+  mutate(
+    transition_type = paste(status_prev, status, sep = " → "),
+    colour = case_when(
+      transition_type == "Normal Cognition → Normal Cognition" ~ "#E69F00",
+      transition_type == "Normal Cognition → MCI" ~ "#E69F00",
+      transition_type == "Normal Cognition → Dementia" ~ "#B2BEB5",
+      transition_type == "MCI → Normal Cognition" ~ "#56B4E9",
+      transition_type == "MCI → MCI" ~ "#56B4E9",
+      transition_type == "MCI → Dementia" ~ "darkgrey",
+      transition_type == "Dementia → Normal Cognition" ~ "#009E73",
+      transition_type == "Dementia → MCI" ~ "#009E73",
+      transition_type == "Dementia → Dementia" ~ "#009E73"
+    ))
+
+ggplot(preds, aes(Age, prob, color = colour, group = interaction(wave, transition_type))) +
+  geom_line(linewidth = 1) +
+  scale_color_identity("Previous State", 
+                      breaks = c("#E69F00", "#56B4E9", "#009E73"),
+                      labels = c("Normal Cognition", "MCI", "Dementia"),
+                      guide = "legend") +
+  labs(
+    title = "Transition Probabilities (Additive Model)",
+    subtitle = "By Age",
+    x = "Age", 
+    y = "Predicted Probability",
+    caption = "*Note:* Grey lines indicate non-significant transitions",
+    ) +
+    facet_grid(wave ~ status, labeller = labeller(
+      wave = function(x) case_when(
+        x == 2 ~ "Year = 2018",
+        x == 3 ~ "Year = 2020",
+        x == 4 ~ "Year = 2022",
+      ),
+      status = function(x) paste("Transition to:", x)
+    )) +
+    theme(
+      plot.subtitle = element_text(hjust = 0.5, size = 12, face = "bold"),
+      plot.caption = ggtext::element_markdown(),
+      panel.spacing = unit(1, "lines"))
+
+```

### 2025-04-09 manuscript/manuscript.qmd
Site re-render :collision:

@@
 
-# Method {#method .unnumbered}
+# Method
 
@@
 
-For this study, we focused on four waves of HRS data (from 2016 to 2022).
+For this study, we focused on four biennial waves of HRS data (from 2016 to 2022).
 Specifically our study sample consisted of respondents who participated in an experimental module assessing procrastination during the 2020 wave.
@@
 As a result, while the core HRS sample includes approximately $18,000$ respondents, our initial sample of interest consisted of $1,368$ respondents.
-We excluded respondents with missing cognitive assessment data for any wave $(n = 445)$ and those under 50 years of age $(n = 39)$.
-This resulted in a final analytic sample of $884$ respondents.
+We excluded respondents with missing cognitive assessment data for any wave $(n = 445)$, missing procrastination data $(n = 4)$, and those under 50 years of age $(n = 16)$.
+This resulted in a final analytic sample of $903$ respondents.
 
-## Measures {#measures .unnumbered}
+## Measures
 
-### Outcome: Cognitive Function and Cognitive Category {#outcome-cognitive-function-and-cognitive-category .unnumbered}
+### Outcome: Cognitive Function and Cognitive Category
 
@@
 
-### Predictor: Procrastination {#predictor-procrastination .unnumbered}
+### Predictor: Procrastination
 
@@
 
-### Confounders {#confounders .unnumbered}
+### Confounders
 
@@
 
+## Data analysis
+
+All data analysis was carried out in R [@R2024].
+Within the final dataset, $0.69\%$ of the data was missing, with all all missing values occurring within the educational attainment variable.
+To examine longitudinal patterns of cognitive transitions, we implemented a discrete time, first-order Markov model [@braun2020].
+This framework estimates probabilities of transitioning between three discrete states: normal cognition (State 1), mild cognitive impairment (MCI; State 2), and dementia (State 3).
+The model relies on the Markov property, which assumes that the probability of transitioning to a future state depends only on the current state, not the full history of prior states.
+
+### Model specification
+Formally, let $X_t$ denote an individual’s cognitive state at time $t$, where $X_t \in S = \{1, 2, 3\}$.
+Under the Markov property, the probability of transitioning to state $j$ at time $t + 1$ depends only on state $i$ at time $t$, independent of all earlier states.
+We can write this conditional probability as:
+
+$$
+P(X_{t+1} = j \vert X_t = i, X_{t-1} = i_{t-1}, \dots X_0 = i_0) = P(X_{t+1} = j \vert X_t = i)
+$$
+
+These conditional probabilities, denoted $p_{ij} = P(X_{t+1} = j \vert X_t = i$ are called transition probabilities and can be organised into a $3 \times 3$ transition matrix $P$:
+
+$$
+P = 
+\begin{bmatrix} 
+p_{11} & p_{12} & p_{13} \\
+p_{21} & p_{22} & p_{23} \\
+p_{31} & p_{32} & p_{33} \\
+\end{bmatrix}
+$$
+
+wherein, each row represents the cognitive state at time $t$, each column represents the cognitive state at time $t + 1$, and each entry $p_{ij}$ represents the probability of transitioning from state $i$ to state $j$.
+
+### Non-stationary transition probabilities
+
+In a stationary Markov model, transition probabilities $p_{ij}$ are assumed constant over time.
+However, cognitive transitions may vary with age, time since diagnosis, or other covariates.
+To account for this, we generalize to a non-stationary Markov model, where $p_{ij}^{(t)}$ depends on covariates $\boldsymbol{x}^{(t)}$
+
+$$
+p_{ij}^{(t)} = P(X_{t+1} = j \vert X_t = i, \boldsymbol{x}^{(t)})
+$$
+
+#### Multinomial Logistic Regression
+
+We model these time-varying transition probabilities using multinomial logistic regression.
+For a system with $K = 3$ states (using state $K$ as reference), the log-odds of transitioning to state $j$ relative to state $K$ are linear functions of covariates $\boldsymbol{x}(t)$:
+
+$$
+log \left( \frac{p_{ij}^{(t)}}{p_{ik}^{(t)}} \right) = \alpha_j^{(t)} + \beta_j^{T \;(t)} \boldsymbol{x}^{(t)} \qquad \text{for } j = 1, \dots K-1
+$$
+where, $\alpha_j^{(t)}$ is the intercept for state $j$ at time $t$, $\beta_j^{(t)}$ is the vector of regression coefficients for state $j$ at time $t$, and $\boldsymbol{x}^{(t)}$ is the vector of covariates at time $t$.
+The model estimates the transition probabilities as a function of covariates, allowing us to assess how procrastination and other factors influence cognitive transitions.
+
+# Results
+
+In our sample of $903$ older American adults aged 50 years or older, $63.01\%$ were female $(n = 569)$, with a mean sample age of $67.73$ years $(\pm 9.84)$.
+At baseline, $84.81\%$ of the sample was classified as having normal cognition, $14.96\%$ as having MCI, and $1.35\%$ as having dementia.
+
+The mean procrastination score was $36.73$ $(\pm 8.23)$, with a range of $12 - 60$. The sample characteristics are presented in Table \ref{tab:sample-characteristics}.
+
 # References

### 2025-04-09 analysis/discrete-markov.qmd
Updating DS model

@@
   group_by(ID) |>
-  mutate(Age = Age - (2022 - as.numeric(as.character(wave))))
+  mutate(Age = Age - (2022 - as.numeric(as.character(wave)))) |>
+  select(!starts_with("Cardio") & !starts_with(paste0("Total_dep_20", 18:22)))
 

### 2025-04-29 R/anderson-goodman-test.R
Adding seperate functions

@@ -1,0 +1,41 @@
+anderson_goodman_test <- function(data, stationary_probs, non_stationary_models) {
+  
+  # Get observed vs. expected counts
+  obs_counts <- data |>
+    group_by(wave, status_prev, status) |>  
+    summarise(n = n(), .groups = "drop")|>
+    complete(wave, status_prev, status, fill = list(n = 0))
+  
+  exp_counts <- purrr::map_dfr(
+    names(non_stationary_models), function(w) {
+      data_w <- data_list[[w]]
+      pred_w <- predict(non_stationary_models[[w]], type = "probs", newdata = pred_grid)
+      
+      pred_w |>
+        as.data.frame() |>
+        mutate(wave = as.numeric(w), status_prev = pred_grid$status_prev) |>
+        pivot_longer(cols = !c(wave, status_prev), 
+                     names_to = "status", 
+                     values_to = "prob_nonstat")
+  })
+  
+  # Calculate test statistic components
+  ag_test <- obs_counts |>
+    left_join(stationary_probs, by = c("status", "status_prev")) |>
+    left_join(exp_counts, by = c("wave", "status_prev", "status")) |>
+    mutate(
+      component = ifelse(prob_stat > 0, n * (prob_nonstat - prob_stat)^2 / prob_stat, 0)
+    ) |>
+    summarise(ag_stat = sum(component)) |>
+    pull(ag_stat)
+  
+  ag_df <- (nrow(exp_counts_nonstat) - nrow(stationary_probs))
+  ag_p_val <- pchisq(ag_test, df = ag_df, lower.tail = FALSE)
+  
+  return(list(
+    ag_test = ag_test,
+    ag_df = ag_df,
+    ag_p_val = ag_p_val
+  ))
+}
+

### 2025-04-29 R/create-additive-predictions.R

@@ -1,0 +1,57 @@
+create_additive_predictions <- function(
+    data, 
+    model, 
+    var, 
+    var_seq, 
+    hold_constant = list(Gender = factor(0), Education_tri = factor(0))) {
+  
+  # Set default prediction sequence based on variable type
+  if (is.null(var_seq)) {
+    if (is.numeric(data[[var]])) {
+      predict_seq <- seq(min(data[[var]], max(data[[var]]), length = 200))
+    } else if (is.factor(data[[var]])) {
+      predict_seq <- levels(data[[var]])
+    } else {
+      predict_seq <- unique(data[[var]])
+    }
+  }
+  
+  # Create list of all variables needed for prediction
+  all_vars <- all.vars(formula(model)[-2])  # Get RHS variables from model formula
+  
+  # Create base grid with all variables EXCEPT the prediction variable
+  base_grid <- expand.grid(
+    lapply(all_vars, function(v) {
+      if (v == var) return(NULL)  # Skip prediction variable
+      
+      # Special cases - keep all levels
+      if (v %in% c("status_prev", "wave")) {
+        return(unique(data[[v]]))
+      }
+      
+      if (v %in% names(hold_constant)) {
+        # Use user-specified constant value
+        hold_constant[[v]]
+      } else if (is.numeric(data[[v]])) {
+        # Use mean for numeric variables
+        mean(data[[v]], na.rm = TRUE)
+      } else if (is.factor(data[[v]])) {
+        # Use first level for factors
+        factor(levels(data[[v]])[1], levels = levels(data[[v]]))
+      } else {
+        # Default to most common value
+        names(sort(table(data[[v]]), decreasing = TRUE))[1]
+      }
+    }) |> purrr::compact() |> setNames(all_vars[all_vars != var]),
+    KEEP.OUT.ATTRS = FALSE
+  )
+  
+  # Create full prediction grid
+  pred_grid <- base_grid |> 
+    tidyr::crossing(!!var := var_seq) |>
+    modelr::add_predictions(model = model, var = "pred", type = "probs") |>
+    tidy_predictions() |>
+    mutate(wave = factor(wave))
+  
+  return(pred_grid)
+}

### 2025-04-29 R/create-transition-dataset.R

@@ -1,0 +1,25 @@
+create_transition_dataset <- function(data, transition_results) {
+  # Combines multiple transition tables into a single analysis-ready dataset.
+  # Formats period labels, ensures consistent factor levels, and structures data for visualization.
+  # Arguments:
+  #   - data: List of year pairs to process (e.g., list(c(2016,2018)))
+  #   - transition_results: List containing transition tables from create_transition_table()
+  # Returns:
+  #   - Tidy dataset with transition frequencies between all specified periods
+  # 
+  data |>
+    purrr::map_dfr(~ {
+      period_name <- paste(.x[2], .x[1], sep = " - ")
+      tbl <- transition_results[[paste(.x[2], .x[1], sep = "-")]]$table
+      
+      as.data.frame(tbl) |>
+        rename(t_minus_1 = 1, t = 2) %>%  # Positional renaming
+        mutate(Period = period_name, .before = t_minus_1)
+    }) |>
+    mutate(
+      Period = stringr::str_replace(Period, "(\\d+) - (\\d+)", "\\2 - \\1"),
+      Period = factor(Period, levels = c("2016 - 2018", "2018 - 2020", "2020 - 2022")),
+      t_minus_1 = factor(t_minus_1, levels = c("Normal Cognition", "MCI", "Dementia")),
+      t = factor(t, levels = c("Normal Cognition", "MCI", "Dementia"))
+    )
+}

### 2025-04-29 R/create-transition-table.R

@@ -1,0 +1,29 @@
+create_transition_table <- function(year_to, year_from) {
+  # Creates a transition frequency table between two specified years of cognitive status data.
+  # Calculates row sums and total observations for transition analysis.
+  # Arguments:
+  #   - year_to: Target year for transitions (character or numeric)
+  #   - year_from: Origin year for transitions (character or numeric)
+  # Returns:
+  #   - List containing: 
+  #     - transition frequency table
+  #     - row sums
+  #     - total observations
+  # 
+  tbl <- table(
+    table_data[[paste0("HRS_", year_to)]],
+    table_data[[paste0("HRS_", year_from)]],
+    dnn = c(year_to, year_from)
+  )
+  
+  # Calculate row sums and total
+  row_sums <- rowSums(tbl)
+  total <- sum(row_sums)
+  
+  # Return as a list with both the table and summary stats
+  list(
+    table = tbl,
+    row_sums = row_sums,
+    total = total
+  )
+}

### 2025-04-29 R/create-transitions.R

@@ -1,0 +1,57 @@
+create_transitions <- function(data, absorbing = FALSE){
+  # Reshapes data from wide to long format to track cognitive status transitions over time.
+  # Calculates the next wave's cognitive status for each individual and creates a transition column.
+  # Optionally treats "Dementia" as an absorbing state, meaning once an individual is classified
+  # with dementia, their status cannot change in subsequent waves.
+  # Arguments:
+  #   - data: The dataset containing cognitive status data.
+  #   - absorbing: A logical flag indicating whether "Dementia" should be treated 
+  #   as an absorbing state.
+  # Returns:
+  #   - A dataset with transition information, including current and next wave statuses.
+  
+  # Reshape the data from wide to long format to track cognitive status over waves
+  data <- data |>
+    select(ID, starts_with("cogfunction")) |>
+    tidyr::pivot_longer(cols = !ID,
+                        names_to = "Wave",
+                        values_to = "Status") |>
+    mutate(Wave = as.factor(stringr::str_replace(Wave, "cogfunction", ""))) |>
+    # Arrange by ID and Wave to prepare for transition calculation
+    arrange(ID, Wave) |>
+    group_by(ID) |>
+    # Get the next wave's cognitive status for each person
+    mutate(next_wave_status = lead(Status)) |>
+    ungroup()
+  
+  # We can optionally specify dementia as an absorbing state
+  # Once an individual is classified with dementia they cannot be classified
+  # with anything else 
+  if(absorbing == TRUE) {
+    data <- data |>
+      group_by(ID) |>
+      mutate(
+        Status = ifelse(cumany(Status == "Dementia"), "Dementia", Status),
+        next_wave_status = ifelse(cumany(Status == "Dementia"), "Dementia", next_wave_status),
+        transition = paste(Status, next_wave_status, sep = " to ")
+      ) |>
+      filter(Wave %in% c(2016, 2018, 2020))
+  }
+  
+  # Filter out rows where either the current or next status is missing
+  data <- data |>
+    group_by(ID) |>
+    filter(!is.na(Status), !is.na(next_wave_status)) |>
+    ungroup() |>
+    # Create a new column representing the transition from one status to the next
+    mutate(
+      transition = paste(Status, next_wave_status, sep = " to "),
+      transition = factor(
+        transition, 
+        levels = c("Normal Cognition to Normal Cognition", "Normal Cognition to MCI",
[…]

### 2025-04-29 R/extract-years.R

@@ -1,0 +1,60 @@
+extract_years <- function(data, years, impute = TRUE, cog_total = FALSE, absorbing = TRUE) {
+  # Extracts cognitive function data for specified years from a dataset.
+  # Converts numeric cognitive status codes (1, 2, 3) into descriptive labels
+  # ("Normal Cognition", "MCI", "Dementia") for easier interpretation.
+  # Arguments:
+  #   - data: The input dataset containing cognitive function data.
+  #   - years: A vector of years for which data should be extracted.
+  # Returns:
+  #   - A dataset with ID and cognitive status columns for the specified years.
+  
+  # Create dynamic column names based on the years provided
+  cogfunction_cols <- paste0("cogfunction", years)
+  cogtotal_cols    <- paste0("cogtot27_imp", years)
+  
+  if(cog_total == FALSE) {
+    data <- data |>
+      # Select only the ID column and cognitive function columns for the specified years
+      select(ID, any_of(cogfunction_cols)) |>
+      mutate(across(!c(ID), ~ case_when(
+        .x == 1 ~ "Normal Cognition",
+        .x == 2 ~ "MCI",
+        .x == 3 ~ "Dementia",
+        TRUE ~ NA_character_  # To handle missing/other cases
+      )))
+  } else {
+    data <- data |>
+      # Select only the ID column and cognitive function columns for the specified years
+      select(ID, any_of(cogtotal_cols)) |>
+      rename_with( .cols = !ID, .fn = ~ stringr::str_replace(
+        string = .x,
+        pattern = "cogtot27_imp", 
+        replacement = "cog_score_"))
+  }
+  
+  if(impute == TRUE){
+    data <- data |>
+      tidyr::pivot_longer(
+        cols = !ID,
+        names_to = "Wave",
+        values_to = "Status") |>
+      group_by(ID) |>
+      tidyr::fill(Status, .direction = "down") |>
+      ungroup() |>
+      tidyr::pivot_wider(names_from = "Wave", values_from = "Status")
+  }
+  
+  if(absorbing == TRUE) {
+    data <- data |>
+      tidyr::pivot_longer(
+        cols = !ID,
+        names_to = "Wave",
+        values_to = "Status") |>
+      group_by(ID) |>
+      mutate(Status = ifelse(cumany(Status == "Dementia" & !is.na(Status)), "Dementia", Status)) |>
+      ungroup() |>
+      tidyr::pivot_wider(names_from = "Wave", values_from = "Status")
+  }
+  
+  return(data)
+}

### 2025-04-29 R/get-time-varying-matrix.R

@@ -1,0 +1,17 @@
+get_time_varying_matrix <- function(model, time_point) {
+  expand.grid(
+    Gender = factor(0),
+    Age = mean(data_stack$Age),
+    Education_tri = factor(0),
+    Depression = mean(data_stack$Depression),
+    Total_p = mean(data_stack$Total_p),
+    status_prev = factor(1:3),
+    wave = time_point) |> 
+    modelr::add_predictions(model, var = "prob", type = "probs") |> 
+    select(status_prev, starts_with("prob")) |> 
+    mutate(
+      status_prev = factor(status_prev, labels = state_names),
+      across(starts_with("prob"), ~round(., 3))) |>
+    tibble::column_to_rownames("status_prev") |>
+    as.matrix() |> `colnames<-`(state_names)
+}

### 2025-04-29 R/normalise.R

@@ -1,0 +1,8 @@
+normalise <- function(x) {
+  # Normalizes a matrix so that each row sums to 1.
+  # Arguments:
+  #   - x: The input matrix to normalize.
+  # Returns:
+  #   - A normalized matrix where each row sums to 1.
+  x / rowSums(x)
+}

### 2025-04-29 R/observed-transition-matrix.R

@@ -1,0 +1,25 @@
+observed_transition_matrix <- function(data) {
+  # Converts transition data into a properly formatted probability transition matrix.
+  # Ensures consistent state ordering and converts proportions to matrix format suitable
+  # for multi-state modeling and visualization.
+  # Arguments:
+  #   - data: Transition dataset from create_transitions()
+  # Returns:
+  #   - Square transition probability matrix with states as row/column names
+  # 
+  # The order I want my matrix in
+  state_order <- c("Normal Cognition", "MCI", "Dementia")
+  
+  data |>
+    group_by(Status, next_wave_status) |>
+    summarise(freq = n(),  .groups = "drop") |>
+    group_by(Status) |>
+    mutate(freq = round(proportions(freq), 3)) |>
+    ungroup() |>
+    complete(Status, next_wave_status, fill = list(freq = 0)) |>
+    tidyr::pivot_wider(names_from = next_wave_status, values_from = freq, names_sort = TRUE) |>
+    tibble::column_to_rownames("Status") |>
+    select(all_of(state_order)) |>
+    _[state_order, ] |>
+    as.matrix()
+}

### 2025-04-29 R/pivot-and-factorise.R

@@ -1,0 +1,59 @@
+pivot_and_factorise <- function(data, time_vary = FALSE) {
+  # Converts cognitive function data from wide to long format and factorizes key variables.
+  # Pivots multiple cogfunction columns into wave/status pairs and converts categorical
+  # variables (Gender, Education, status) to factors with meaningful labels.
+  # Arguments:
+  #   - data: Dataset containing cognitive function variables in wide format
+  # Returns:
+  #   - Long-format dataset with factorized variables, ordered by ID and wave
+  # 
+  
+  if(time_vary == TRUE) {
+    # Pivoting depression variable
+    dep_scores <- data |>
+      select(ID, starts_with("Total_dep")) |>
+      pivot_longer(
+        cols = !ID,
+        names_to = "remove",
+        values_to = "Depression") |>
+      select(Depression)
+    
+    # Pivoting whole dataset
+    data <- data |>
+      select(!starts_with("Total_dep")) |>
+      pivot_longer(
+        cols = starts_with("cogfunction"),
+        names_to = "wave",
+        names_prefix = "cogfunction",
+        values_to = "status")
+    
+    # Adding in depression scores
+    data <- data |> cbind(dep_scores)
+  } else{
+    # No pivoting of depression
+    data <- data |>
+      tidyr::pivot_longer(
+        cols = starts_with("cogfunction"), 
+        names_to = "wave",
+        names_prefix = "cogfunction", 
+        values_to = "status") |>
+      rename(Depression = Total_dep_2016) |>
+      select(!starts_with("Total_dep_"))
+  }
+  
+  # Factorizing
+  data <- data |>
+    mutate(
+      Gender = factor(Gender, levels = c(0, 1)),
+      Education_tri = factor(Education_tri, levels = c(0, 1, 2)),
+      wave = factor(wave),
+      status = factor(
+        status,
+        levels = c("Normal Cognition", "MCI", "Dementia"),
+        labels = c(1, 2, 3))) |>
+    relocate(wave, .after = ID) |>
+    relocate(status, .after = wave)
+    
+  
+  return(data)
+}

### 2025-04-29 R/plot-additive-predictions.R

@@ -1,0 +1,26 @@
+plot_additive_predictions <- function(data, var, x_label, subtitle) {
+  
+  # Generating predictions
+  data |>
+    ggplot(aes(x = as.numeric(.data[[var]]), y = prob, color = status_prev)) +
+    geom_line(linewidth = 1) +
+    ggokabeito::scale_color_okabe_ito() +
+    labs(
+      title = "Transition Probabilities (Additive Model)",
+      subtitle = subtitle,
+      x = x_label, 
+      y = "Predicted Probability", 
+      color = "Previous State"
+    ) +
+    facet_grid(wave ~ status, labeller = labeller(
+      wave = function(x) case_when(
+        x == 2 ~ "Year = 2018",
+        x == 3 ~ "Year = 2020",
+        x == 4 ~ "Year = 2022",
+      ),
+      status = function(x) paste("Transition to:", x)
+    )) +
+    theme(
+      plot.subtitle = element_text(hjust = 0.5, size = 12, face = "bold"),
+      panel.spacing = unit(1, "lines")) 
+}

### 2025-04-29 R/plot-predictions-stationary.R

@@ -1,0 +1,19 @@
+plot_predictions_stationary <- function(predictions, variable, x_axis) {
+  # Visualizes predicted transition probabilities from stationary multi-state models.
+  # Shows probability curves by procrastination level, stratified by baseline state.
+  # Arguments:
+  #   - predictions: Tidy predictions from tidy_predictions()
+  # Returns:
+  #   - Faceted ggplot showing predicted probability curves
+  # 
+  predictions |>
+    ggplot(aes(x = {{variable}}, y = prob, colour = status_prev)) +
+    geom_line(linewidth = 1) +
+    ggokabeito::scale_colour_okabe_ito() +
+    labs(
+      title = "Predicted transition probabilities (stationary model)",
+      x = x_axis, y = "Probability", colour = "Previous State") +
+    facet_wrap(~ status, labeller = labeller(
+      status = function(x){paste0("Transition to: ", x)}
+    ))
+}

### 2025-04-29 R/plot-transition-matrix.R

@@ -1,0 +1,43 @@
+plot_transition_matrix <- function(matrix, observed = TRUE) {
+  # Creates a heatmap visualization of cognitive state transition probabilities.
+  # Generates either observed or estimated transition plots with consistent formatting,
+  # including labeled probability values and a diverging color scale for emphasis.
+  # Arguments:
+  #   - matrix: Tidy transition matrix from reshape_matrix()
+  #   - observed: Logical flag indicating whether data represents observed (TRUE) 
+  #               or estimated (FALSE) transitions
+  # Returns:
+  #   - ggplot heatmap object with:
+  #     * State transitions as cells
+  #     * Probability values displayed numerically
+  #     * Custom color scale and axis formatting
+  
+  if(observed == TRUE) {
+    subtitle <- "Observed state transitions between assessment waves"
+  } else {
+    subtitle <- "Estimated state transitions between assessment waves"
+  }
+  
+  matrix |>
+    ggplot(aes(x = from_state, y = to_state, fill = probability)) +
+    geom_tile(color = "white", linewidth = 0.5) +
+    geom_text(
+      aes(label = format(round(probability, 3), nsmall = 3)),
+      size = 4.5, 
+      color = "#212427",
+      fontface = "bold") +
+    colorspace::scale_fill_continuous_diverging(
+      palette = "Blue-Red 3", mid = 0.50, alpha = 0.5, 
+      limits = c(0, 1), name = "Transition \nProbability") +
+    labs(title = "Transition Probabilities Across Cognitive States",
+         subtitle = subtitle,
+         x = "Previous State (t - 1)", 
+         y = "Current State (t)") +
+    theme(
+      axis.text = element_text(size = 10),
+      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
+      legend.position = "right",
+      legend.text = element_text(size = 9),
+      panel.grid = element_blank()
+    )
+}

### 2025-04-29 R/reshape-matrix.R

@@ -1,0 +1,25 @@
+reshape_matrix <- function(matrix) {
+  # Converts a transition probability matrix into a tidy format suitable for visualization.
+  # Transforms the matrix into long format with explicit factor levels for cognitive states,
+  # preserving the original ordering while preparing for ggplot compatibility.
+  # Arguments:
+  #   - matrix: A square transition probability matrix with states as row/column names
+  # Returns:
+  #   - Tidy data frame with columns:
+  #     * from_state: Factor indicating origin cognitive state
+  #     * to_state: Factor indicating destination cognitive state (reversed for plotting)
+  #     * probability: Numeric transition probability values
+  
+  matrix |>
+    as.data.frame() |>
+    mutate(
+      from_state = factor(c(
+        "Normal cognition", "MCI", "Dementia"),
+        levels = c("Normal cognition", "MCI", "Dementia"))
+    ) |>
+    reshape2::melt(
+      id.vars = "from_state", 
+      variable.name = "to_state", 
+      value.name = "probability") |>
+    mutate(to_state = factor(to_state, levels = rev(levels(to_state))))
+}

### 2025-04-29 R/test-stationarity.R

@@ -1,0 +1,54 @@
+test_stationarity <- function(stationary_model, non_stationary_models) {
+  
+  # Extract log-likelihoods from both models -----------------------------------
+  ll_stat <- as.numeric(logLik(stationary_model))
+  
+  ll_nonstat <- sum(sapply(non_stationary_models, logLik))
+  
+  # Extract coefficients from both models --------------------------------------
+  coef_stat <- c(t(coef(stationary_model)))
+  
+  coef_nonstat <- unlist(lapply(
+    non_stationary_models, function(m) c(t(coef(m)))
+    ))
+  
+  # Calculate difference -------------------------------------------------------
+  # Ensure coefs are of the same length
+  min_length <- min(length(coefs_stat), length(coefs_nonstat))
+  coef_diff <- coefs_nonstat[1:min_length] - coefs_stat[1:min_length]
+  
+  # Create a covariance matrix -------------------------------------------------
+  vcov_nonstat <- Matrix::bdiag(lapply(non_stationary_models, vcov))
+  
+  # Add small ridge regularization to make matrix invertible
+  vcov_reg <- as.matrix(vcov_nonstat) + 1e-6*diag(nrow(vcov_nonstat))
+  
+  # Subset matrix
+  vcov_subset <- vcov_reg[1:min_length, 1:min_length]
+  
+  # Calculate test statistics --------------------------------------------------
+  # Likelihood ratio test
+  LR_test <- -2 * (ll_stat - ll_nonstat)
+  
+  # Wald test
+  wald_stat <- tryCatch({
+    
+    t(coef_diff) %*% solve(vcov_subset) %*% coef_diff
+  }, error = function(e) {
+    
+    MASS::ginv(vcov_subset) %*% coef_diff
+    
+    t(coef_diff) %*% MASS::ginv(vcov_subset) %*% coef_diff
+  })
+  
+  # Degrees of freedom
+  df_LR <- length(coef_nonstat) - length(coef_stat)
+  df_Wald <- min_length
+  
+  list(
+    LR = LR_test,
+    df_LR = length(coef_nonstat) - length(coef_stat),
+    Wald = wald_stat,
+    df_Wald = length(coef_nonstat)
+  )
+}

### 2025-04-29 R/tidy-output.R

@@ -1,0 +1,55 @@
+tidy_output <- function(fit, multiplicative = FALSE) {
+  output <- broom::tidy(fit, conf.int = TRUE) |>
+    filter(term != "(Intercept)") |>
+    mutate(across(c(estimate:p.value), ~ round(x = ., digits = 3)))
+  
+  if(any(output$y.level == "1")) {
+    output <- output |>
+      mutate(y.level = case_when(
+        y.level == "1" ~ "MCI - NC",
+        y.level == "3" ~ "MCI - Dementia"))
+  } else {
+    output <- output |>
+      mutate(y.level = case_when(
+        y.level == "2" ~ "NC - MCI",
+        y.level == "3" ~ "NC - Dementia"))
+  }
+  
+  if(multiplicative == FALSE) {
+    output <- output |>
+      mutate(
+        term = case_when(
+          term == "Gender1" ~ "Being female",
+          term == "Education_tri1" ~ "High school degree vs. No education",
+          term == "Education_tri2" ~ "Further education vs. No education",
+          term == "Total_dep_2016" ~ "Depression Scores (2016)",
+          term == "Total_p" ~ "Procrastination (2020)",
+          term == "status_prev2" ~ "Previous state: MCI",
+          term == "status_prev3" ~ "Previous state: Dementia",
+          term == "wave" ~ "Time",
+          TRUE ~ term))
+  } else {
+    output <- output |>
+      mutate(
+        term = case_when(
+          term == "Gender1" ~ "Being female",
+          term == "Education_tri1" ~ "High school degree vs. No education",
+          term == "Education_tri2" ~ "Further education vs. No education",
+          term == "Depression" ~ "Depression Scores (2016)",
+          term == "Total_p" ~ "Procrastination (2020)",
+          term == "status_prev2" ~ "Previous state: MCI",
+          term == "status_prev3" ~ "Previous state: Dementia",
+          term == "wave" ~ "Time",
+          term == "Gender1:wave" ~ "Gender & Time",
+          term == "Age:wave" ~ "Age & Time",
+          term == "Education_tri1:wave" ~ "Education (0 - 1) & Time",
+          term == "Education_tri2:wave" ~ "Education (0 - 2) & Time",
+          term == "Depression:wave" ~ "Depression & Time",
+          term == "Total_p:wave" ~ "Procrastination & Time",
+          term == "status_prev2:wave" ~ "Previous state: MCI & Time",
[…]

### 2025-04-29 R/tidy-predictions.R

@@ -1,0 +1,31 @@
+tidy_predictions <- function(predictions) {
+  # Restructures model prediction matrices into tidy format for visualization.
+  # Converts numeric codes to factor labels and reshapes multiple prediction columns
+  # into key-value pairs suitable for ggplot.
+  # Arguments:
+  #   - predictions: Raw prediction matrix from model output
+  # Returns:
+  #   - Long-format dataset with probabilities for each cognitive state
+  #   
+  predictions |>
+    as.matrix() |>
+    as_tibble() |>
+    mutate(
+      Gender = factor(ifelse(Gender == 0, "Male", "Female"), levels = c("Male", "Female")),
+      status_prev = case_when(
+        status_prev == 1 ~ "Normal Cognition",
+        status_prev == 2 ~ "MCI",
+        status_prev == 3 ~ "Dementia",
+      ),
+      status_prev = factor(status_prev, levels = c("Normal Cognition", "MCI", "Dementia")),
+      Total_p = as.numeric(Total_p),
+      across(c(pred.1:pred.3), as.numeric)) |>
+    tidyr::pivot_longer(cols = c(pred.1:pred.3), names_to = "status", values_to = "prob") |>
+    mutate(status = case_when(
+      status == "pred.1" ~ "Normal Cognition",
+      status == "pred.2" ~ "MCI",
+      status == "pred.3" ~ "Dementia"
+    ),
+    status = factor(status, levels = c("Normal Cognition", "MCI", "Dementia")),
+    status_prev = factor(status_prev, levels = c("Normal Cognition", "MCI", "Dementia")))
+}

### 2025-04-29 R/tran-heat-map.R

@@ -1,0 +1,23 @@
+tran_heat_map <- function(data) {
+  # Generates heatmap visualization of transition frequencies between cognitive states.
+  # Uses color intensity and labeled values to show transition patterns across time periods.
+  # Arguments:
+  #   - data: Transition dataset from create_transition_dataset()
+  # Returns:
+  #   - ggplot heatmap with state transitions as cells
+  # 
+  data |>
+    ggplot(aes(x = t_minus_1, y = t, fill = Freq)) +
+    geom_tile(color = "white") +
+    geom_text(aes(label = Freq), color = "black", size = 3.5) +  # Add counts
+    scale_fill_gradient(low = "white", high = "steelblue") +
+    facet_wrap(~ Period, ncol = 3) +  # Split by time period
+    labs(
+      x = "Previous State (t-1)", 
+      y = "Current State (t)",
+      title = "Cognitive State Transitions Between Time Periods",
+      fill = "Frequency"
+    ) +
+    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+    ggeasy::easy_move_legend("bottom")
+}

### 2025-04-29 R/tran-stack-graph.R

@@ -1,0 +1,21 @@
+tran_stack_graph <- function(data) {
+  # Creates stacked bar charts visualizing cognitive state transitions across time periods.
+  # Shows composition of current states by previous state, faceted by observation period.
+  # Arguments:
+  #   - data: Transition dataset from create_transition_dataset()
+  # Returns:
+  #   - ggplot object showing stacked transition proportions
+  # 
+  data |>
+    ggplot(aes(x = t_minus_1, y = Freq, fill = t)) +
+    geom_col(position = "stack", colour = "black") +
+    facet_wrap(~ Period, ncol = 3) +
+    labs(
+      x = "Previous State (t-1)", 
+      y = "Count",
+      title = "Outcomes by Prior Cognitive State",
+      fill = "Current State (t)"
+    ) +
+    ggokabeito::scale_fill_okabe_ito() +
+    ggeasy::easy_move_legend("bottom")
+}

### 2025-04-29 analysis/LGCM.qmd
Re rendering site :fire:

---
title: "Latent Growth Curve Model"
---

# Setting up

## Loading packages

```{r packages}
#| code-fold: true
#| code-summary: "Check out my packages"
#| message: false
#| warning: false

library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(ggplot2)
library(patchwork)

library(lavaan)
```

## Loading functions

```{r functions}
#| code-fold: true
#| code-summary: "Check out my functions"

extract_years <- function(data, years, impute = FALSE, cog_total = FALSE, absorbing = FALSE) {
  # Extracts cognitive function data for specified years from a dataset.
  # Converts numeric cognitive status codes (1, 2, 3) into descriptive labels
  # ("Normal Cognition", "MCI", "Dementia") for easier interpretation.
  # Arguments:
  #   - data: The input dataset containing cognitive function data.
  #   - years: A vector of years for which data should be extracted.
  # Returns:
  #   - A dataset with ID and cognitive status columns for the specified years.

  # Create dynamic column names based on the years provided
  cogfunction_cols <- paste0("cogfunction", years)
  cogtotal_cols    <- paste0("cogtot27_imp", years)

  if(cog_total == FALSE) {
    data <- data |>
      # Select only the ID column and cognitive function columns for the specified years
      select(ID, any_of(cogfunction_cols)) |>
      mutate(across(!c(ID), ~ case_when(
        .x == 1 ~ "Normal Cognition",
        .x == 2 ~ "MCI",
        .x == 3 ~ "Dementia",
        TRUE ~ NA_character_  # To handle missing/other cases
      )))
  } else {
    data <- data |>
      # Select only the ID column and cognitive function columns for the specified years
      select(ID, any_of(cogtotal_cols)) |>
      rename_with( .cols = !ID, .fn = ~ stringr::str_replace(
        string = .x,
        pattern = "cogtot27_imp", 
        replacement = "cog_score_"))
  }

  if(impute == TRUE){
    data <- data |>
      tidyr::pivot_longer(
        cols = !ID,
        names_to = "Wave",
        values_to = "Status") |>
      group_by(ID) |>
      tidyr::fill(Status, .direction = "down") |>
      ungroup() |>
      tidyr::pivot_wider(names_from = "Wave", values_from = "Status")
  }

  if(cog_total == FALSE & absorbing == TRUE) {
    data <- data |>
      tidyr::pivot_longer(
        cols = !ID,
        names_to = "Wave",
        values_to = "Status") |>
      group_by(ID) |>
      mutate(Status = ifelse(cumany(Status == "Dementia" & !is.na(Status)), "Dementia", Status)) |>
      ungroup() |>
      tidyr::pivot_wider(names_from = "Wave", values_from = "Status")
  } else if(cog_total == TRUE & absorbing == TRUE) {
    data <- data |>
      tidyr::pivot_longer(cols = !ID, names_to = "Wave", values_to = "Score") |>
      group_by(ID) |>
      mutate(Score = ifelse(cumany(Score <= 6 & !is.na(Score)), pmin(Score, 6), Score)) |>
      pivot_wider(names_from = Wave, values_from = Score)
  }

  return(data)
}
```

## Loading data

```{r}
data <- read.csv(here::here("analysis/data/data.csv"))
```

# Data processing

```{r}
cog_totals <- data |> 
  extract_years(
    years = seq(2016, 2022, by = 2),
    cog_total = TRUE,
    impute = TRUE,
    absorbing = TRUE) |>
  na.omit()

head(cog_totals)
```

```{r}
cols <- c(
  "ID", "Gender", "Age", "Education_tri",
  paste0("Total_dep_", seq(2016, 2022, by = 2)),
  "Total_p")

data_final <- cog_totals |>
  inner_join(data[, cols], by = "ID") |>
  pivot_longer(
    cols = starts_with("cog_score"),
    names_to = "Wave",
    values_to = "Score",
    names_prefix = "cog_score_") |>
  mutate(
    Wave = factor(Wave, levels = c(2016, 2018, 2020, 2022)),
    Gender = factor(Gender, levels = c(0, 1)),
    GED = factor(ifelse(Education_tri == 1, 1, 0)),
    FE = factor(ifelse(Education_tri == 2, 1, 0)),
    Education_tri = factor(Education_tri, levels = c(0, 1, 2)),
    ) |>
  relocate(Wave, .after = ID) |>
  relocate(Score, .after = Wave) |>
  select(!starts_with(paste0("Total_dep_20", 18:22))) |>
  filter(!is.na(Total_p)) |>
  ungroup()

head(data_final)
```

```{r}
observed_trajectories <- data_final |>
  ggplot(aes(x = Wave, y = Score, group = ID)) +
  geom_line(alpha = 0.25) +
  stat_summary(
    aes(group = 1),
    fun = mean,
    geom = "line",
    linewidth = 1.5,
    color = "red"
  ) +
  labs(title = "Observed trajectories of cognitive function",
       subtitle = "Cognitive function scores across waves",
       x = "Year", y = "Cognitive Score") +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, colour = "grey11", face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 12, colour = "grey11", face = "bold"),
    axis.title = element_text(size = 10, colour = "grey11", face = "bold")
  )

observed_trajectories
```

### 2025-04-29 analysis/discrete-markov.qmd
Re rendering site :fire:

--- 2025-04-09 analysis/discrete-markov.qmd
@@
 
-## Loading packages
-
-```{r packages}
-#| code-fold: true
-#| code-summary: "Check out my packages"
+```{r}
+#| label: setup
 #| warning: false
+#| message: false
+#| output: false
+#| code-fold: true
+#| code-summary: "Check out my code"
 
@@
 library(patchwork)
-```
-
-## Loading functions
-
-```{r functions}
-#| code-fold: true
-#| code-summary: "Check out my functions"
-#| warning: false
 
 # Functions --------------------------------------------------------------------
-extract_years <- function(data, years, impute = TRUE, cog_total = FALSE, absorbing = TRUE) {
-  # Extracts cognitive function data for specified years from a dataset.
-  # Converts numeric cognitive status codes (1, 2, 3) into descriptive labels
-  # ("Normal Cognition", "MCI", "Dementia") for easier interpretation.
-  # Arguments:
-  #   - data: The input dataset containing cognitive function data.
-  #   - years: A vector of years for which data should be extracted.
-  # Returns:
-  #   - A dataset with ID and cognitive status columns for the specified years.
-
-  # Create dynamic column names based on the years provided
-  cogfunction_cols <- paste0("cogfunction", years)
-  cogtotal_cols    <- paste0("cogtot27_imp", years)
-
-  if(cog_total == FALSE) {
-    data <- data |>
-      # Select only the ID column and cognitive function columns for the specified years
-      select(ID, any_of(cogfunction_cols)) |>
-      mutate(across(!c(ID), ~ case_when(
-        .x == 1 ~ "Normal Cognition",
-        .x == 2 ~ "MCI",
-        .x == 3 ~ "Dementia",
-        TRUE ~ NA_character_  # To handle missing/other cases
-      )))
-  } else {
-    data <- data |>
-      # Select only the ID column and cognitive function columns for the specified years
-      select(ID, any_of(cogtotal_cols)) |>
-      rename_with( .cols = !ID, .fn = ~ stringr::str_replace(
-        string = .x,
-        pattern = "cogtot27_imp", 
-        replacement = "cog_score_"))
-  }
-
-  if(impute == TRUE){
-    data <- data |>
-      tidyr::pivot_longer(
-        cols = !ID,
-        names_to = "Wave",
-        values_to = "Status") |>
-      group_by(ID) |>
-      tidyr::fill(Status, .direction = "down") |>
-      ungroup() |>
-      tidyr::pivot_wider(names_from = "Wave", values_from = "Status")
-  }
-
-  if(absorbing == TRUE) {
-    data <- data |>
-      tidyr::pivot_longer(
-        cols = !ID,
-        names_to = "Wave",
-        values_to = "Status") |>
-      group_by(ID) |>
-      mutate(Status = ifelse(cumany(Status == "Dementia"), "Dementia", Status)) |>
-      ungroup() |>
-      tidyr::pivot_wider(names_from = "Wave", values_from = "Status")
-  }
-
-  return(data)
-}
-
-pivot_and_factorise <- function(data) {
-  # Converts cognitive function data from wide to long format and factorizes key variables.
-  # Pivots multiple cogfunction columns into wave/status pairs and converts categorical
-  # variables (Gender, Education, status) to factors with meaningful labels.
-  # Arguments:
-  #   - data: Dataset containing cognitive function variables in wide format
-  # Returns:
-  #   - Long-format dataset with factorized variables, ordered by ID and wave
-  #
-
-  # Pivoting depression variables
-  # depression <- data |>
-  #   select(starts_with("Total_dep")) |>
-  #   tidyr::pivot_longer(
-  #     cols = everything(), names_to = "Remove", values_to = "Depression") |>
-  #   select(Depression)
-
-  # Pivoting cardio variables
-  # cardio <- data |>
-  #   select(starts_with("Cardio_risk")) |>
-  #   tidyr::pivot_longer(
-  #     cols = everything(), names_to = "Remove", values_to = "Cardio") |>
-  #   select(Cardio)
-
-  data |>
-    # select(!starts_with(c("Total_dep", "Cardio_risk"))) |>
-    tidyr::pivot_longer(
-      cols = starts_with("cogfunction"), names_to = "wave",
-      names_prefix = "cogfunction", values_to = "status") |>
-    mutate(
-      Gender = factor(Gender, levels = c(0, 1)),
-      Education_tri = factor(Education_tri, levels = c(0, 1, 2)),
-      wave = factor(wave),
-      status = factor(status, 
-                      levels = c("Normal Cognition", "MCI", "Dementia"),
-                      labels = c(1, 2, 3))) |>
-    relocate(wave, .after = ID) |>
-    relocate(status, .after = wave)
-    #cbind(depression, cardio)
-}
-
-create_transition_table <- function(year_to, year_from) {
-  # Creates a transition frequency table between two specified years of cognitive status data.
-  # Calculates row sums and total observations for transition analysis.
-  # Arguments:
-  #   - year_to: Target year for transitions (character or numeric)
-  #   - year_from: Origin year for transitions (character or numeric)
-  # Returns:
-  #   - List containing: 
-  #     - transition frequency table
-  #     - row sums
-  #     - total observations
-  # 
-  tbl <- table(
-    table_data[[paste0("HRS_", year_to)]],
-    table_data[[paste0("HRS_", year_from)]],
-    dnn = c(year_to, year_from)
-  )
-
-  # Calculate row sums and total
-  row_sums <- rowSums(tbl)
-  total <- sum(row_sums)
-
-  # Return as a list with both the table and summary stats
-  list(
-    table = tbl,
-    row_sums = row_sums,
-    total = total
-  )
-}
-
-create_transition_dataset <- function(data, transition_results) {
-  # Combines multiple transition tables into a single analysis-ready dataset.
-  # Formats period labels, ensures consistent factor levels, and structures data for visualization.
-  # Arguments:
-  #   - data: List of year pairs to process (e.g., list(c(2016,2018)))
-  #   - transition_results: List containing transition tables from create_transition_table()
-  # Returns:
-  #   - Tidy dataset with transition frequencies between all specified periods
-  # 
-  data |>
-    purrr::map_dfr(~ {
-      period_name <- paste(.x[2], .x[1], sep = " - ")
-      tbl <- transition_results[[paste(.x[2], .x[1], sep = "-")]]$table
-
-      as.data.frame(tbl) |>
-        rename(t_minus_1 = 1, t = 2) %>%  # Positional renaming
-        mutate(Period = period_name, .before = t_minus_1)
-    }) |>
-    mutate(
-      Period = stringr::str_replace(Period, "(\\d+) - (\\d+)", "\\2 - \\1"),
-      Period = factor(Period, levels = c("2016 - 2018", "2018 - 2020", "2020 - 2022")),
-      t_minus_1 = factor(t_minus_1, levels = c("Normal Cognition", "MCI", "Dementia")),
-      t = factor(t, levels = c("Normal Cognition", "MCI", "Dementia"))
-    )
-}
-
-create_transitions <- function(data, absorbing = FALSE){
-  # Reshapes data from wide to long format to track cognitive status transitions over time.
-  # Calculates the next wave's cognitive status for each individual and creates a transition column.
-  # Optionally treats "Dementia" as an absorbing state, meaning once an individual is classified
-  # with dementia, their status cannot change in subsequent waves.
-  # Arguments:
-  #   - data: The dataset containing cognitive status data.
-  #   - absorbing: A logical flag indicating whether "Dementia" should be treated 
-  #   as an absorbing state.
-  # Returns:
-  #   - A dataset with transition information, including current and next wave statuses.
-
-  # Reshape the data from wide to long format to track cognitive status over waves
-  data <- data |>
-    select(ID, starts_with("cogfunction")) |>
-    tidyr::pivot_longer(cols = !ID,
-                        names_to = "Wave",
-                        values_to = "Status") |>
-    mutate(Wave = as.factor(stringr::str_replace(Wave, "cogfunction", ""))) |>
-    # Arrange by ID and Wave to prepare for transition calculation
-    arrange(ID, Wave) |>
-    group_by(ID) |>
-    # Get the next wave's cognitive status for each person
-    mutate(next_wave_status = lead(Status)) |>
-    ungroup()
-
-  # We can optionally specify dementia as an absorbing state
-  # Once an individual is classified with dementia they cannot be classified
-  # with anything else 
-  if(absorbing == TRUE) {
-    data <- data |>
-      group_by(ID) |>
-      mutate(
-        Status = ifelse(cumany(Status == "Dementia"), "Dementia", Status),
-        next_wave_status = ifelse(cumany(Status == "Dementia"), "Dementia", next_wave_status),
-        transition = paste(Status, next_wave_status, sep = " to ")
-      ) |>
-      filter(Wave %in% c(2016, 2018, 2020))
-  }
-
-  # Filter out rows where either the current or next status is missing
-  data <- data |>
-    group_by(ID) |>
-    filter(!is.na(Status), !is.na(next_wave_status)) |>
-    ungroup() |>
-    # Create a new column representing the transition from one status to the next
-    mutate(
-      transition = paste(Status, next_wave_status, sep = " to "),
-      transition = factor(
-        transition, 
-        levels = c("Normal Cognition to Normal Cognition", "Normal Cognition to MCI",
-                   "Normal Cognition to Dementia", "MCI to Normal Cognition", "MCI to MCI",
-                   "MCI to Dementia", "Dementia to Dementia") 
-      ))
-
-  return(data)
-}
-
-observed_transition_matrix <- function(data) {
-  # Converts transition data into a properly formatted probability transition matrix.
-  # Ensures consistent state ordering and converts proportions to matrix format suitable
-  # for multi-state modeling and visualization.
-  # Arguments:
-  #   - data: Transition dataset from create_transitions()
-  # Returns:
-  #   - Square transition probability matrix with states as row/column names
-  # 
-  # The order I want my matrix in
-  state_order <- c("Normal Cognition", "MCI", "Dementia")
-
-  data |>
-    group_by(Status, next_wave_status) |>
-    summarise(freq = n(),  .groups = "drop") |>
-    group_by(Status) |>
-    mutate(freq = round(proportions(freq), 3)) |>
-    ungroup() |>
-    complete(Status, next_wave_status, fill = list(freq = 0)) |>
-    tidyr::pivot_wider(names_from = next_wave_status, values_from = freq, names_sort = TRUE) |>
-    tibble::column_to_rownames("Status") |>
-    select(all_of(state_order)) |>
-    _[state_order, ] |>
-    as.matrix()
-}
-
-normalise <- function(x) {
-  # Normalizes a matrix so that each row sums to 1.
-  # Arguments:
-  #   - x: The input matrix to normalize.
-  # Returns:
-  #   - A normalized matrix where each row sums to 1.
-  x / rowSums(x)
-}
-
-reshape_matrix <- function(matrix) {
-  # Converts a transition probability matrix into a tidy format suitable for visualization.
-  # Transforms the matrix into long format with explicit factor levels for cognitive states,
-  # preserving the original ordering while preparing for ggplot compatibility.
-  # Arguments:
-  #   - matrix: A square transition probability matrix with states as row/column names
-  # Returns:
-  #   - Tidy data frame with columns:
-  #     * from_state: Factor indicating origin cognitive state
-  #     * to_state: Factor indicating destination cognitive state (reversed for plotting)
-  #     * probability: Numeric transition probability values
-
-  matrix |>
-    as.data.frame() |>
-    mutate(
-      from_state = factor(c(
-        "Normal cognition", "MCI", "Dementia"),
-        levels = c("Normal cognition", "MCI", "Dementia"))
-    ) |>
-    reshape2::melt(
-      id.vars = "from_state", 
-      variable.name = "to_state", 
-      value.name = "probability") |>
-    mutate(to_state = factor(to_state, levels = rev(levels(to_state))))
-}
-
-plot_transition_matrix <- function(matrix, observed = TRUE) {
-  # Creates a heatmap visualization of cognitive state transition probabilities.
-  # Generates either observed or estimated transition plots with consistent formatting,
-  # including labeled probability values and a diverging color scale for emphasis.
-  # Arguments:
-  #   - matrix: Tidy transition matrix from reshape_matrix()
-  #   - observed: Logical flag indicating whether data represents observed (TRUE) 
-  #               or estimated (FALSE) transitions
-  # Returns:
-  #   - ggplot heatmap object with:
-  #     * State transitions as cells
-  #     * Probability values displayed numerically
-  #     * Custom color scale and axis formatting
-
-  if(observed == TRUE) {
-    subtitle <- "Observed state transitions between assessment waves"
-  } else {
-    subtitle <- "Estimated state transitions between assessment waves"
-  }
-
-  matrix |>
-    ggplot(aes(x = from_state, y = to_state, fill = probability)) +
-    geom_tile(color = "white", linewidth = 0.5) +
-    geom_text(
-      aes(label = format(round(probability, 3), nsmall = 3)),
-      size = 4.5, 
-      color = "#212427",
-      fontface = "bold") +
-    colorspace::scale_fill_continuous_diverging(
-      palette = "Blue-Red 3", mid = 0.50, alpha = 0.5, 
-      limits = c(0, 1), name = "Transition \nProbability") +
-    labs(title = "Transition Probabilities Across Cognitive States",
-         subtitle = subtitle,
-         x = "Previous State (t - 1)", 
-         y = "Current State (t)") +
-    theme(
-      axis.text = element_text(size = 10),
-      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
-      legend.position = "right",
-      legend.text = element_text(size = 9),
-      panel.grid = element_blank()
-    )
-}
-
-tran_stack_graph <- function(data) {
-  # Creates stacked bar charts visualizing cognitive state transitions across time periods.
-  # Shows composition of current states by previous state, faceted by observation period.
-  # Arguments:
-  #   - data: Transition dataset from create_transition_dataset()
-  # Returns:
-  #   - ggplot object showing stacked transition proportions
-  # 
-  data |>
-    ggplot(aes(x = t_minus_1, y = Freq, fill = t)) +
-    geom_col(position = "stack", colour = "black") +
-    facet_wrap(~ Period, ncol = 3) +
-    labs(
-      x = "Previous State (t-1)", 
-      y = "Count",
-      title = "Outcomes by Prior Cognitive State",
-      fill = "Current State (t)"
-    ) +
-    ggokabeito::scale_fill_okabe_ito() +
-    ggeasy::easy_move_legend("bottom")
-}
-
-tran_heat_map <- function(data) {
-  # Generates heatmap visualization of transition frequencies between cognitive states.
-  # Uses color intensity and labeled values to show transition patterns across time periods.
-  # Arguments:
-  #   - data: Transition dataset from create_transition_dataset()
-  # Returns:
-  #   - ggplot heatmap with state transitions as cells
-  # 
-  data |>
-    ggplot(aes(x = t_minus_1, y = t, fill = Freq)) +
-    geom_tile(color = "white") +
-    geom_text(aes(label = Freq), color = "black", size = 3.5) +  # Add counts
-    scale_fill_gradient(low = "white", high = "steelblue") +
-    facet_wrap(~ Period, ncol = 3) +  # Split by time period
-    labs(
-      x = "Previous State (t-1)", 
-      y = "Current State (t)",
-      title = "Cognitive State Transitions Between Time Periods",
-      fill = "Frequency"
-    ) +
-    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
-    ggeasy::easy_move_legend("bottom")
-}
-
-tidy_output <- function(fit, multiplicative = FALSE) {
-  output <- broom::tidy(fit, conf.int = TRUE) |>
-    filter(term != "(Intercept)") |>
-  mutate(across(c(estimate:p.value), ~ round(x = ., digits = 3)))
-
-  if(any(output$y.level == "1")) {
-    output <- output |>
-      mutate(y.level = case_when(
-      y.level == "1" ~ "MCI - NC",
-      y.level == "3" ~ "MCI - Dementia"))
-  } else {
-    output <- output |>
-      mutate(y.level = case_when(
-      y.level == "2" ~ "NC - MCI",
-      y.level == "3" ~ "NC - Dementia"))
-  }
-
-  if(multiplicative == FALSE) {
-    output <- output |>
-      mutate(
-      term = case_when(
-      term == "Gender1" ~ "Being female",
-      term == "Education_tri1" ~ "High school degree vs. No education",
-      term == "Education_tri2" ~ "Further education vs. No education",
-      term == "Total_dep_2016" ~ "Depression Scores (2016)",
-      term == "Total_p" ~ "Procrastination (2020)",
-      term == "status_prev2" ~ "Previous state: MCI",
-      term == "status_prev3" ~ "Previous state: Dementia",
-      term == "wave" ~ "Time",
-      TRUE ~ term))
-  } else {
-    output <- output |>
-      mutate(
-      term = case_when(
-      term == "Gender1" ~ "Being female",
-      term == "Education_tri1" ~ "High school degree vs. No education",
-      term == "Education_tri2" ~ "Further education vs. No education",
-      term == "Depression" ~ "Depression Scores (2016)",
-      term == "Total_p" ~ "Procrastination (2020)",
-      term == "status_prev2" ~ "Previous state: MCI",
-      term == "status_prev3" ~ "Previous state: Dementia",
-      term == "wave" ~ "Time",
-      term == "Gender1:wave" ~ "Gender & Time",
-      term == "Age:wave" ~ "Age & Time",
-      term == "Education_tri1:wave" ~ "Education (0 - 1) & Time",
-      term == "Education_tri2:wave" ~ "Education (0 - 2) & Time",
-      term == "Depression:wave" ~ "Depression & Time",
-      term == "Total_p:wave" ~ "Procrastination & Time",
-      term == "status_prev2:wave" ~ "Previous state: MCI & Time",
-      term == "status_prev3:wave" ~ "Previous state: Dementia & Time",
-      TRUE ~ term))
-  }
-
-  return(output)
-}
-
-tidy_predictions <- function(predictions) {
-  # Restructures model prediction matrices into tidy format for visualization.
-  # Converts numeric codes to factor labels and reshapes multiple prediction columns
-  # into key-value pairs suitable for ggplot.
-  # Arguments:
-  #   - predictions: Raw prediction matrix from model output
-  # Returns:
-  #   - Long-format dataset with probabilities for each cognitive state
-  #   
-  predictions |>
-    as.matrix() |>
-    as_tibble() |>
-    mutate(
-      Gender = factor(ifelse(Gender == 0, "Male", "Female"), levels = c("Male", "Female")),
-      status_prev = case_when(
-        status_prev == 1 ~ "Normal Cognition",
-        status_prev == 2 ~ "MCI",
-        status_prev == 3 ~ "Dementia",
-      ),
-      status_prev = factor(status_prev, levels = c("Normal Cognition", "MCI", "Dementia")),
-      Total_p = as.numeric(Total_p),
-      across(c(pred.1:pred.3), as.numeric)) |>
-    tidyr::pivot_longer(cols = c(pred.1:pred.3), names_to = "status", values_to = "prob") |>
-    mutate(status = case_when(
-      status == "pred.1" ~ "Normal Cognition",
-      status == "pred.2" ~ "MCI",
-      status == "pred.3" ~ "Dementia"
-    ),
-    status = factor(status, levels = c("Normal Cognition", "MCI", "Dementia")),
-    status_prev = factor(status_prev, levels = c("Normal Cognition", "MCI", "Dementia")))
-}
-
-plot_predictions_stationary <- function(predictions, variable, x_axis) {
-  # Visualizes predicted transition probabilities from stationary multi-state models.
-  # Shows probability curves by procrastination level, stratified by baseline state.
-  # Arguments:
-  #   - predictions: Tidy predictions from tidy_predictions()
-  # Returns:
-  #   - Faceted ggplot showing predicted probability curves
-  # 
-  predictions |>
-    ggplot(aes(x = {{variable}}, y = prob, colour = status_prev)) +
-    geom_line(linewidth = 1) +
-    ggokabeito::scale_colour_okabe_ito() +
-    labs(
-      title = "Predicted transition probabilities (stationary model)",
-      x = x_axis, y = "Probability", colour = "Previous State") +
-    facet_wrap(~ status, labeller = labeller(
-      status = function(x){paste0("Transition to: ", x)}
-    ))
-}
-
-create_additive_predictions <- function(
-    data, model, var, var_seq, hold_constant = list(Gender = factor(0), Education_tri = factor(0))) {
-
-  # Set default prediction sequence based on variable type
-  if (is.null(var_seq)) {
-    if (is.numeric(data[[var]])) {
-      predict_seq <- seq(min(data[[var]], max(data[[var]]), length = 200))
-    } else if (is.factor(data[[var]])) {
-      predict_seq <- levels(data[[var]])
-    } else {
-      predict_seq <- unique(data[[var]])
-    }
-  }
-
-  # Create list of all variables needed for prediction
-  all_vars <- all.vars(formula(model)[-2])  # Get RHS variables from model formula
-
-  # Create base grid with all variables EXCEPT the prediction variable
-  base_grid <- expand.grid(
-    lapply(all_vars, function(v) {
-      if (v == var) return(NULL)  # Skip prediction variable
-
-      # Special cases - keep all levels
-      if (v %in% c("status_prev", "wave")) {
-        return(unique(data[[v]]))
-      }
-
-      if (v %in% names(hold_constant)) {
-        # Use user-specified constant value
-        hold_constant[[v]]
-      } else if (is.numeric(data[[v]])) {
-        # Use mean for numeric variables
-        mean(data[[v]], na.rm = TRUE)
-      } else if (is.factor(data[[v]])) {
-        # Use first level for factors
-        factor(levels(data[[v]])[1], levels = levels(data[[v]]))
-      } else {
-        # Default to most common value
-        names(sort(table(data[[v]]), decreasing = TRUE))[1]
-      }
-    }) |> purrr::compact() |> setNames(all_vars[all_vars != var]),
-    KEEP.OUT.ATTRS = FALSE
-  )
-
-  # Create full prediction grid
-  pred_grid <- base_grid |> 
-    tidyr::crossing(!!var := var_seq) |>
-    modelr::add_predictions(model = model, var = "pred", type = "probs") |>
-    tidy_predictions() |>
-    mutate(wave = factor(wave))
-
-  return(pred_grid)
-}
-
-plot_additive_predictions <- function(data, var, x_label, subtitle) {
-
-  # Generating predictions
-  data |>
-    ggplot(aes(x = as.numeric(.data[[var]]), y = prob, color = status_prev)) +
-    geom_line(linewidth = 1) +
-    ggokabeito::scale_color_okabe_ito() +
-    labs(
-      title = "Transition Probabilities (Additive Model)",
-      subtitle = subtitle,
-      x = x_label, 
-      y = "Predicted Probability", 
-      color = "Previous State"
-    ) +
-    facet_grid(wave ~ status, labeller = labeller(
-      wave = function(x) case_when(
-        x == 2 ~ "Year = 2018",
-        x == 3 ~ "Year = 2020",
-        x == 4 ~ "Year = 2022",
-      ),
-      status = function(x) paste("Transition to:", x)
-    )) +
-    theme(
-      plot.subtitle = element_text(hjust = 0.5, size = 12, face = "bold"),
-      panel.spacing = unit(1, "lines")) 
-}
-
-get_time_varying_matrix <- function(model, time_point) {
-  expand.grid(
-    Gender = factor(0),
-    Age = mean(data_stack$Age),
-    Education_tri = factor(0),
-    Total_dep_2016 = mean(data_stack$Total_dep_2016),
-    Total_p = mean(data_stack$Total_p),
-    status_prev = factor(1:3),
-    wave = time_point) |> 
-    modelr::add_predictions(model, var = "prob", type = "probs") |> 
-    select(status_prev, starts_with("prob")) |> 
-    mutate(
-      status_prev = factor(status_prev, labels = state_names),
-      across(starts_with("prob"), ~round(., 3))) |>
-    tibble::column_to_rownames("status_prev") |>
-    as.matrix() |> `colnames<-`(state_names)
-}
-```
-
-## Setting theme
-
-```{r set-theme}
-#| code-fold: true
-#| code-summary: "Check out my theme"
-
+files <- list.files(here::here("R/"), full.names = TRUE)
+
+sapply(files, source)
+
+# Theme ------------------------------------------------------------------------
 colour <- "#212427"
@@
       ))
-```
-
-## Reading in data {.unnumbered}
-
-```{r data}
+
+# Data -------------------------------------------------------------------------
 data <- read.csv(here::here("analysis/data/data.csv"))
@@
   "ID", "Gender", "Age", "Education_tri",
-  paste0("Cardio_risk_", seq(16, 22, by = 2)), 
   paste0("Total_dep_", seq(2016, 2022, by = 2)),
@@
   inner_join(data[, cols], by = "ID") |>
-  pivot_and_factorise() |>
+  # mutate(Age = Age - (2022 - 2016)) |>  # Age of person in 2016
+  # filter(Age >= 50) |>                  # Remove those with were not initially 50+ in 2016
+  pivot_and_factorise(time_vary = FALSE) |>
   group_by(ID) |>
-  mutate(Age = Age - (2022 - as.numeric(as.character(wave))))
+  mutate(Age = Age - (2022 - as.numeric(as.character(wave)))) # Make age time varying
 
@@
 fit_1a <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Total_dep_2016, 
+  status ~ Gender + Age + Education_tri + Depression, 
   family = multinomial, 
@@
 fit_1b <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Total_dep_2016, 
+  status ~ Gender + Age + Education_tri + Depression, 
   family = multinomial, 
@@
 fit_2a <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p, 
+  status ~ Gender + Age + Education_tri + Depression + Total_p, 
   family = multinomial, 
@@
 fit_2b <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p, 
+  status ~ Gender + Age + Education_tri + Depression + Total_p, 
   family = multinomial, 
@@
 fit_3a <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev, 
+  status ~ Gender + Age + Education_tri + Depression + Total_p + status_prev, 
   family = multinomial, 
@@
 fit_3b <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev, 
+  status ~ Gender + Age + Education_tri + Depression + Total_p + status_prev, 
   family = multinomial, 
@@
       levels = c("Being female", "Age", "High school degree vs. No education",
-                 "Further education vs. No education", "Depression Scores (2016)",
+                 "Further education vs. No education", "Depression",
                  "Procrastination (2020)", "Previous state: MCI", 
@@
 
-Let's visualise the estimated odds ratios (@fig-odds-ratio) and the predicted transition probabilities (@fig-model-predictions) for the stationary model .
+Let's visualize the estimated odds ratios (@fig-odds-ratio) and the predicted transition probabilities (@fig-model-predictions) for the stationary model .
 
@@
   Education_tri = factor(0),
-  Total_dep_2016 = mean(data_stack$Total_dep_2016),
+  Depression = mean(data_stack$Depression),
   status_prev = levels(data_stack$status_prev),
@@
   Education_tri = factor(0),
-  Total_dep_2016 = mean(data_stack$Total_dep_2016),
+  Depression = mean(data_stack$Depression),
   Total_p = mean(data_stack$Total_p),
@@
 fit_4a <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev + wave, family = multinomial, 
+  status ~ Gender + Age + Education_tri + Depression + Total_p + status_prev + wave, family = multinomial, 
   data = data_stack, trace = FALSE)
@@
 fit_4b <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev + wave, family = multinomial, 
+  status ~ Gender + Age + Education_tri + Depression + Total_p + status_prev + wave, family = multinomial, 
   data = data_stack |> mutate(status = relevel(status, ref = 2)), 
@@
 fit_5a <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + (status_prev * wave), family = multinomial, 
+  status ~ Gender + Age + Education_tri + Depression + Total_p + (status_prev * wave), family = multinomial, 
   data = data_stack, trace = FALSE)
@@
 fit_5b <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Total_dep_2016 + Total_p + (status_prev * wave), family = multinomial, 
+  status ~ Gender + Age + Education_tri + Depression + Total_p + (status_prev * wave), family = multinomial, 
   data = data_stack |> mutate(status = relevel(status, ref = 2)), 
@@
 fit_6a <- nnet::multinom(
-  status ~ (Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev) * wave, family = multinomial, 
+  status ~ (Gender + Age + Education_tri + Depression + Total_p + status_prev) * wave, family = multinomial, 
   data = data_stack, trace = FALSE)
@@
 fit_6b <- nnet::multinom(
-  status ~ (Gender + Age + Education_tri + Total_dep_2016 + Total_p + status_prev) * wave, family = multinomial, 
+  status ~ (Gender + Age + Education_tri + Depression + Total_p + status_prev) * wave, family = multinomial, 
   data = data_stack |> mutate(status = relevel(status, ref = 2)), 

### 2025-04-29 manuscript/manuscript.qmd
Re rendering site :fire:

@@
 However, cognitive transitions may vary with age, time since diagnosis, or other covariates.
-To account for this, we generalize to a non-stationary Markov model, where $p_{ij}^{(t)}$ depends on covariates $\boldsymbol{x}^{(t)}$
+To account for this, we generalize to a non-stationary Markov model, where $p_{ij}^{(t)}$ depends on covariates $\boldsymbol{X}^{(t)}$
 
 $$
-p_{ij}^{(t)} = P(X_{t+1} = j \vert X_t = i, \boldsymbol{x}^{(t)})
+p_{ij}^{(t)} = P(X_{t+1} = j \vert X_t = i, \boldsymbol{X}^{(t)})
 $$
@@
 We model these time-varying transition probabilities using multinomial logistic regression.
-For a system with $K = 3$ states (using state $K$ as reference), the log-odds of transitioning to state $j$ relative to state $K$ are linear functions of covariates $\boldsymbol{x}(t)$:
+For a system with $K = 3$ states (using state $K$ as reference), the log-odds of transitioning to state $j$ relative to state $K$ are linear functions of covariates $\boldsymbol{X}(t)$:
 
 $$
-log \left( \frac{p_{ij}^{(t)}}{p_{ik}^{(t)}} \right) = \alpha_j^{(t)} + \beta_j^{T \;(t)} \boldsymbol{x}^{(t)} \qquad \text{for } j = 1, \dots K-1
+log \left( \frac{p_{ij}^{(t)}}{p_{ik}^{(t)}} \right) = \alpha_j^{(t)} + \beta_j^{T \;(t)} \boldsymbol{X}^{(t)} \qquad \text{for } j = 1, \dots K-1
 $$
-where, $\alpha_j^{(t)}$ is the intercept for state $j$ at time $t$, $\beta_j^{(t)}$ is the vector of regression coefficients for state $j$ at time $t$, and $\boldsymbol{x}^{(t)}$ is the vector of covariates at time $t$.
+where, $\alpha_j^{(t)}$ is the intercept for state $j$ at time $t$, $\beta_j^{(t)}$ is the vector of regression coefficients for state $j$ at time $t$, and $\boldsymbol{X}^{(t)}$ is the vector of covariates at time $t$.
 The model estimates the transition probabilities as a function of covariates, allowing us to assess how procrastination and other factors influence cognitive transitions.
@@
 At baseline, $84.81\%$ of the sample was classified as having normal cognition, $14.96\%$ as having MCI, and $1.35\%$ as having dementia.
-
-The mean procrastination score was $36.73$ $(\pm 8.23)$, with a range of $12 - 60$. The sample characteristics are presented in Table \ref{tab:sample-characteristics}.
+A
 

### 2025-05-21 manuscript/manuscript.qmd
Update manuscript

--- 2025-04-01 manuscript/manuscript.qmd
@@
 
-Dementia is a syndrome characterized by the progressive and typically irreversible decline of cognitive function, leading to memory loss, impaired reasoning, and difficulties with daily activities [@Prince2013; @Sanz-Blasco2022].
-It encompasses a range of conditions, including Alzheimer's disease, vascular dementia, and Lewy body dementia [@Cao2020].
+Dementia is a syndrome characterized initially by the progressive and typically irreversible decline of cognitive function, leading to memory loss, impaired reasoning, and difficulties with activities of daily living [@Prince2013; @Sanz-Blasco2022].
+It arises from various conditions, including Alzheimer's disease, vascular dementia, and Lewy body dementia [@Cao2020].
 The global burden of dementia is substantial, with projections suggesting that the number of affected individuals will rise from 57.4 million in 2019 to 152.8 million by 2050 [@Nichols2022].
-Given this projection, identifying and addressing modifiable risk factors is crucial to mitigate the growing prevalence of dementia worldwide.
+Given this projection, identifying and addressing modifiable risk factors is crucial to mitigate the growing prevalence difficulties with activities of daily living of dementia worldwide.
+In fact, targeting these factors could prevent up to 50% of dementia cases worldwide [@Livingston2024].
+The *Lancet* Commission highlights 14 key modifiable risks, emphasizing that dementia risk can be reduced at any age.
+Nevertheless, age remains the strongest non-modifiable risk factor, accounting for the largest proportion of dementia cases.
 
-Given the progressive nature of dementia, early identification of pre-dementia conditions such as mild cognitive impairment (MCI) is essential for timely intervention.
-MCI is a condition characterized by cognitive changes - such as memory lapses or difficulty making decisions - that exceed typical age-related decline [@Abner2012; @Cooper2015; @Fresnais2023; @Salem2023; @Yu2013].
-Studies indicate that approximately 46% of people with MCI transition to dementia within three years and approximately 80% within six years [@Cooper2015; @Shigemizu2020; @Tschanz2006].
-As such, identifying factors that influence this progression is critically important for early intervention and personalized care.
+Mild cognitive impairment (MCI) is a condition characterized by cognitive changes --- such as memory lapses or difficulty making decisions --- that exceed typical age-related decline [@Gauthier2006].
+Prior studies suggest that 46% of people with MCI transition to a dementia state within three years and 80% within six years [@Cooper2015; @Shigemizu2020; @Tschanz2006].
+However, progression is neither uniform nor inevitable.
+Heterogeneity in outcomes underscores the importance of distinguishing between two classes of predictors: causal risk factors (e.g.
+smoking, hypertension, or diabetes) which may accelerate dementia onset and are potential targets for intervention [@Livingston2024], and prodromal markers (e.g.
+memory loss, apathy, or struggles in daily living) that reflect early disease processes and may signal limited opportunities for prevention.
+While causal factors offer avenues for mitigating risk, prodromal features often emerge closer to irreversible pathological changes, highlighting the need for early detection and tailored interventions.
 
-Among the behavioural symptoms in MCI and dementia, apathy is one of the most prevalent [@Fresnais2023; @Richard2012; @Salem2023; @vanDalen2018].
-Defined as a lack of motivation [@Fresnais2023], apathy is also a multidimensional construct that encompasses deficits in executive and emotional functioning, initiation, and increased functional impairment [@Okura2010; @Radakovic2018].
-Individuals with apathy exhibit reduced goal-directed behavior and a diminished desire to pursue rewards or pleasure [@Fahed2021].
-Importantly, apathy has been identified as a significant risk factor for the transition from MCI to dementia [@Palmer2010; @Ruthirakuhan2019; @vanDalen2018].
+Apathy, defined as a significant loss of motivation, distinct from depression and cognitive impairment [@Fresnais2023; @Richard2012] is among the most prevalent neuropsychiatric symptoms in both MCI and dementia [@Donovan2014; @Fresnais2023; @Salem2023; @vanDalen2018].
+Individuals with apathy exhibit reduced goal-directed behavior and a diminished desire to pursue rewards or pleasure [@Fresnais2023].
+Importantly, apathy has been identified as a significant predictor for the transition from MCI to dementia [@Ruthirakuhan2019; @Salem2023; @vanDalen2018].
 For instance, a meta-analysis by @vanDalen2018 found that apathy almost doubles the risk of transitioning to dementia.
-Additionally, apathy has been correlated with higher levels of neurofibrillary tangles in individuals with dementia, suggesting a potential connection to underlying neuropathology [@Skogseth2008].
-
-Procrastination, although traditionally viewed as a distinct behavioural issue, may share key characteristics with apathy, suggesting potential common underlying mechanisms.Chronic procrastination, characterized by persistent delays in decision-making and task completion [@Abbasi2015; @Ferrari2010], has been associated with dysfunction in the brain's reward and decision-making systems, particularly the dorsolateral and ventromedial prefrontal cortices [@Friden2020; @Zhang2019].
-These brain regions are critical for both initiating and sustaining goal-directed action and are areas where both apathy and procrastination show deficits [@Fahed2021; @Zhang2019].
+Additionally, levels of apathy measured using both the *"Neuropsychiatric Inventory Questionnaire\"* [@Kaufer2000] and *"Apathy Evaluation Scale\"* [@Marin1991] have shown to be correlated with higher levels of amyloid burden and neurofibrillary tangles in individuals with dementia, suggesting a potential connection to underlying neuropathology [@Donovan2014; @Marshall2013].
 
 While apathy primarily reflects a lack of motivation, procrastination involves a delay in action despite an intention to complete such action [@Steel2007].
-Both behaviours suggest impaired executive function, particularly in goal-oriented behaviour and decision-making, which are hallmark deficits in MCI and dementia [@Kirova2015; @Stopford2012].
-In this context, procrastination could reflect broader motivational and cognitive impairments akin to those seen in apathy.
+Although both behaviours indicate disruptions in goal-directed activity, they differ in their phenomenology: apathy is marked by diminished drive or emotional engagement, whereas procrastination typically involves intact intention but impaired follow-through.
+These differences suggest that procrastination may capture aspects of executive dysfunction and emotional regulation not fully accounted for by apathy alone.
 
-Given these parallels, it is worth exploring whether chronic procrastination could serve as an early behavioural marker for cognitive impairment, or even a risk factor for dementia, especially in older adults.
-Procrastination may exacerbate existing cognitive decline by reinforcing patterns of inaction and passivity.
-Individuals who chronically delay tasks may inadvertently engage in fewer cognitively stimulating activities, such as physical activity, problem-solving, decision-making, and goal-setting---activities that are known to build cognitive resilience and reduce dementia risk [@Chowdhary2022].
-By limiting engagement in such activities, procrastination could contribute to the acceleration of cognitive decline.
-Therefore, while apathy has already been established as a significant risk factor for dementia, the role of procrastination, especially when chronic, may represent an overlooked behavioural trait that warrants similar attention.
+Both behaviours have been linked with impaired executive function, particularly in goal-oriented behaviours, decision-making, and self-regulation [@Gustavson2015; @Kawagoe2017; @steel2010], domains which are frequently compromised in individuals with MCI and dementia [@Kirova2015; @Stopford2012].
+While apathy has been well-established as a prodromal marker and risk factor for dementia progression [@Ruthirakuhan2019; @Salem2023; @vanDalen2018], emerging evidence also suggests a potential link between procrastination and early cognitive decline.
+For instance, @Friden2020 and @Zhang2019 highlight that chronic procrastination is associated with dysfunction in both the dorsolateral and ventromedial prefrontal cortices, regions of the brain known to not only show early degeneration in dementia [@Joseph2021], but also be linked to higher levels of apathy [@Fahed2021; @Zhang2019].
+
+Furthermore, procrastination may exacerbate existing cognitive decline by reinforcing patterns of inaction and passivity.
+Individuals who chronically delay tasks may inadvertently engage in fewer cognitively stimulating activities, such as physical activity, problem-solving, decision-making, and goal-setting.
+Such activities are known to build cognitive resilience and reduce dementia risk [@Chowdhary2022; @Livingston2024].
+By limiting engagement in such activities, procrastination may contribute to a self-reinforcing cycle of cognitive disengagement, potentially accelerating decline over time.
+
+Given these behavioural and neuro-cognitive parallels with apathy, it is worth exploring whether chronic procrastination could serve as an early behavioural marker for cognitive impairment, or even a risk factor for dementia, especially in older adults.
+By examining procrastination alongside apathy, we may uncover a broader behavioural phenotype indicative of early neuropathological changes.
+Crucially, procrastination may identify individuals with motivational or executive dysfunction who do not yet meet the threshold for apathy, offering a potentially valuable and earlier behavioural marker for neurodegenerative risk.
+
+Moreover, since age remains the strongest predictor of MCI and dementia, it is important to consider how behavioural predictors such as procrastination operate across the lifespan.
+Many established modifiable risk factors such as, hypertension, hearing loss, smoking, and social isolation, exert age-dependent effects, with certain factors carrying more weight at midlife than in late life [@livingston2020; @Livingston2024].
+In light of this, we sought to explore whether procrastination also demonstrates an age-dependent risk profile.
+Understanding how procrastination interacts with age may help clarify its role in the aetiology of cognitive decline and identify windows of opportunity for targeted intervention.
 
@@
 Identifying procrastination as a potential risk factor could expand the scope of early interventions aimed at preventing or slowing the progression of dementia.
-As such, the purpose of this study was to test the hypothesis that higher levels of procrastination would be associated with an increased probability of transitioning from normal cognitive function or MCI to dementia.
+As such, the purpose of this study was to (a) assess group differences in procrastination levels across the normal cognition, MCI, and dementia groups, and b) test the hypotheses that higher levels of procrastination would be associated with an increased probability of transitioning from normal cognitive function or MCI to dementia.
 
-# Method {#method .unnumbered}
+# Method
 
@@
 
-Analyses were conducted using a secondary data source; a multi-wave prospective cohort study called the Health and Retirement Study (HRS; [@Fisher2018]), which tracks the health, economic, and social well-being of over $18,000$ American adults primarily aged $>50$.
+Analyses were conducted using a secondary data source; a multi-wave prospective cohort study called the Health and Retirement Study (HRS; [@Fisher2018], which tracks the health, economic, and social well-being of over $18,000$ American adults primarily aged $>50$.
 The HRS is managed by the Institute for Social Research at the University of Michigan, with data collected every two years.
@@
 As a result, while the core HRS sample includes approximately $18,000$ respondents, our initial sample of interest consisted of $1,368$ respondents.
-We excluded respondents with missing cognitive assessment data for any wave $(n = 445)$ and those under 50 years of age $(n = 39)$.
-This resulted in a final analytic sample of $884$ respondents.
+We excluded respondents with missing cognitive assessment data for any wave $(n = 445)$, missing procrastination data $(n = 4)$, and those under 50 years of age  $(n = 16)$.
+This resulted in a final analytic sample of $903$ respondents.
 
-## Measures {#measures .unnumbered}
+## Measures
 
-### Outcome: Cognitive Function and Cognitive Category {#outcome-cognitive-function-and-cognitive-category .unnumbered}
+### Outcome: Cognitive Function and Cognitive Category
 
-Cognitive function in the HRS is assessed using a series of tests adapted from the Telephone Interview for Cognitive Status (TICS; [@Brandt1988; @Fong2009]).
+Cognitive function in the HRS is assessed using a series of tests adapted from the Telephone Interview for Cognitive Status (TICS; [@Brandt1988; @Fong2009].
 These tests include an immediate and delayed $10$-noun free recall test, a serial seven subtraction test, and a backward count from $20$ test.

### 2025-05-21 analysis/discrete-markov.qmd
Doing interactions :zap:

@@
 #| code-summary: "Check out my code"
+
+rm(list = ls())
 
@@
   inner_join(data[, cols], by = "ID") |>
-  # mutate(Age = Age - (2022 - 2016)) |>  # Age of person in 2016
-  # filter(Age >= 50) |>                  # Remove those with were not initially 50+ in 2016
+  # Removing baseline age below 60
+  filter(Age - (2022 - 2016) >= 60) |>
   pivot_and_factorise(time_vary = FALSE) |>
   group_by(ID) |>
-  mutate(Age = Age - (2022 - as.numeric(as.character(wave)))) # Make age time varying
+  # ^ Make age time varying
+  mutate(Age = Age - (2022 - as.numeric(as.character(wave))))
 
@@
   mutate(status_prev = lag(status), .after = status) |>
-  filter(!is.na(Total_p), Age >= 50) |>
+  filter(!is.na(Total_p)) |>
   ungroup() |> 
@@
 
-```{r visualise-transition}
+```{r}
 #| code-fold: true
@@
 
-```{r visualise-matrix}
+```{r}
 #| code-fold: true
@@
   trace = FALSE)
+
+# Full model with interaction (reference NC)
+fit_4a <- nnet::multinom(
+  status ~ Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + status_prev, 
+  family = multinomial, 
+  data = data_stack, trace = FALSE)
+
+# Full model with interaction (reference MCI)
+fit_4b <- nnet::multinom(
+  status ~ Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + status_prev, 
+  family = multinomial, 
+  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
+  trace = FALSE)
 ```
@@
 
-```{r model-comparison}
+```{r}
 #| label: tbl-stationary-comparison
@@
 
-anova(fit_1a, fit_2a, fit_3a) |>
+anova(fit_1a, fit_2a, fit_3a, fit_4a) |>
   mutate(
-    Model = c("Baseline", "Procrastination", "Procrastination + Previous Status"),
+    Model = c("Baseline", 
+              "Procrastination", 
+              "Procrastination + Previous Status",
+              "Interaction Model"),
     `Resid. Dev`= round(`Resid. Dev`, digits = 3),
@@
     options = list(
-      pageLength = 3,
+      pageLength = 4,
       dom = "t",
@@
 
+::: {.panel-tabset .nav-pills}
+## Non-interaction model
+
 ### Estimated parameters
 
-Below we present the estimated parameters for the stationary models (@tbl-stationary-output). We will transform the estimates to odds ratios and colour them based on significance.
-
-```{r model-output}
+Below we present the estimated parameters for the stationary model (fit 3) (@tbl-stationary-output). We will transform the estimates to odds ratios and colour them based on significance.
+
+```{r}
 #| code-fold: true
@@
 
-```{r compare-matrix}
+```{r}
 #| code-fold: true
@@
 ```
+
+## Interaction model
+
+### Estimated parameters
+
+Below we present the estimated parameters for the interaction model (fit 4; @tbl-interaction-output). We will transform the estimates to odds ratios and colour them based on significance.
+
+```{r}
+#| code-fold: true
+#| code-summary: "Check out my code"
+#| label: tbl-interaction-output
+#| tbl-cap: "Estimated parameters for the interaction model"
+#| tbl-cap-location: bottom
+
+interaction_results <- rbind(tidy_output(fit_4a), tidy_output(fit_4b)) |>
+  rename(transition = y.level) |>
+  mutate(
+    # Transforming to odds ratios
+    estimate = exp(estimate),
+    conf.low = exp(conf.low),
+    conf.high = exp(conf.high),
+
+    # Mutating to factors
+    transition = factor(
+      transition, 
+      levels = c("NC - MCI", "MCI - NC", 
+                 "NC - Dementia", "MCI - Dementia")),
+    term = factor(
+      term, 
+      levels = c("Being female", "Age", 
+                 "High school degree vs. No education",
+                 "Further education vs. No education", "Depression",
+                 "Procrastination (2020)",  
+                 "Age x Procrastination", "Previous state: MCI", 
+                 "Previous state: Dementia")),
+
+    # Creating a colour code
+    colour = case_when(
+      estimate > 1 & p.value < 0.05 ~ "Positive",
+      estimate < 1 & p.value < 0.05 ~ "Negative",
+      TRUE ~ "NS"),
+
+    colour = factor(colour, levels = c("Positive", "Negative", "NS"))
+    )
+
+interaction_results |>
+  mutate(across(c(estimate, conf.low, conf.high), ~ round(x = ., digits = 3))) |>
+  DT::datatable(
+    options = list(
+      pageLength = 6,
+      dom = "tip",
+      ordering = FALSE,
+      columnDefs = list(list(className = "dt-center", targets = "_all"))
+    ),
+    rownames = FALSE
+  )
+```
+
+### Model visualisation
+
+Let’s visualize the estimated odds ratios (@fig-odds-ratio-interaction) and predicted transition probabilities (@fig-interaction-predictions) for our interaction model.
+
+```{r}
+#| message: false
+#| warning: false
+#| code-fold: true
+#| code-summary: "Check out my code"
+#| fig-width: 12
+#| fig-height: 10
+#| label: fig-odds-ratio-interaction
+#| fig-cap: "Estimated odds ratios for the interaction model."
+
+# Design matrix for faceting
+design <- "
+  AABBCC 
+  DDEEFF
+  ##GG##
+"
+
+interaction_results |>
+  filter(!term %in% c("Previous state: MCI", "Previous state: Dementia")) |>
+  ggplot(aes(x = estimate, y = transition, colour = colour)) +
+  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
+  ggstance::geom_pointrangeh(
+    aes(xmin = conf.low, xmax = conf.high),
+    position = ggstance::position_dodgev(height = 0.5),
+    size = 1.25,
+    fatten = 3) +
+  scale_colour_manual(values = c(
+    "Positive" = "#0072B2", 
+    "Negative" = "#E69F00", 
+    "NS"       = "#B2BEB5")) +
+  labs(title = "Odds ratios (stationary model)",
+       x = "Odds Ratio", y = "Predictor") +
+  guides(colour = "none") +
+  ggh4x::facet_manual(~ term, scales = "free_x", design = design) +
+  theme_bw() +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
+    axis.title = element_text(size = 12, face = "bold"),
+    strip.text = element_text(size = 10, face = "bold"),
+    panel.grid.major.y = element_blank(),
+    panel.grid.minor.y = element_blank())
+
+```
+
+```{r}
+#| code-fold: true
+#| code-summary: "Check out my code"
+#| warning: false
+#| fig-width: 12
+#| fig-height: 10
+#| label: fig-interaction-predictions
+#| fig-cap: "Predicted transition probabilities from the interaction model."
+
+# Making a new predicted dataset
+pred_data <- expand.grid(
+  Gender = factor(0),
+  Age = seq(60, 97, length = 200),
+  Education_tri = factor(0),
+  Depression = mean(data_stack$Depression),
+  status_prev = levels(data_stack$status_prev),
+  Total_p = seq(0, 60, length = 200))
+
+# Reveling original data
+data_plot <- data_stack |>
+  mutate(
+    # Relevel
+    status = case_when(
+      status == 1 ~ "Normal Cognition",
+      status == 2 ~ "MCI",
+      status == 3 ~ "Dementia"),
+    status_prev = case_when(
+      status_prev == 1 ~ "Normal Cognition",
+      status_prev == 2 ~ "MCI",
+      status_prev == 3 ~ "Dementia")
+    ) |>
+  mutate(
+    status = factor(
+      status, levels = c("Normal Cognition", "MCI", "Dementia")),
+    status_prev = factor(
+      status_prev, levels = c("Normal Cognition", "MCI", "Dementia"))) |>
+  filter(status_prev %in% c("Normal Cognition", "MCI") & 
+         status %in%  c("Normal Cognition", "MCI"))
+
+# Plotting interaction effect ----------------------------------------------
+pred_data |>
+  # Adding predictions
+  modelr::add_predictions(model = fit_4a, var = "pred", type = "probs") |>
+  tidy_predictions() |>
+  filter(status_prev %in% c("Normal Cognition", "MCI") & 
+         status %in%  c("Normal Cognition", "MCI")) |>
+  ggplot(aes(x = Total_p, y = as.numeric(Age))) +
+  # Adding heat mat with contour lines
+  geom_raster(aes(fill = prob), alpha = 0.7) +
+  geom_contour(aes(z = prob), colour = "black", size = 0.3) +
+  # Adding orginal data points
+  geom_point(
+    data = data_plot, aes(x = Total_p, y = Age), size = 1.75, alpha = 0.3) +
+  # Customising plot
+  scale_x_continuous(
+    breaks = seq(0, 60, by = 10),
+    sec.axis = sec_axis(~ ., name = "Previous State",
+                        breaks = NULL, labels = NULL)) +
+  scale_y_continuous(
+    breaks = seq(50, 100, by = 5),
+    sec.axis = sec_axis(~ ., name = "Current State",
+                        breaks = NULL, labels = NULL)) +
+  scale_fill_viridis_c(option = "plasma") +
+  labs(
+    title = "Predicted transition probabilities from the interaction model",
+    x = "Total Procrastination", y = "Age", 
+    fill = expression(hat(p))) +
+  facet_grid(status ~ status_prev) +
+  theme_classic() +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
+    axis.title = element_text(size = 12, face = "bold"),
+    strip.text = element_text(size = 10, face = "bold"),
+    panel.grid.major.y = element_blank(),
+    panel.grid.minor.y = element_blank(),
+    legend.title = element_text(size = 10, face = "bold"))
+
+```
+
+Finally, (@fig-interaction-age-predictions) shows the the predicted transition probabilities for different age cohorts.
+
+```{r}
+#| code-fold: true
+#| code-summary: "Check out my code"
+#| warning: false
+#| fig-width: 12
+#| fig-height: 8
+#| label: fig-interaction-age-predictions
+#| fig-cap: "Predicted transition probabilities for different age cohorts."
+
+# Create predictions with age info
+age_cohots <- expand.grid(
+  Gender = factor(0),
+  Age = c(60, 70, 80, 90),
+  Education_tri = factor(0),
+  Depression = mean(data_stack$Depression),
+  status_prev = levels(data_stack$status_prev),
+  Total_p = seq(0, 60, length = 200)) |>
+  modelr::add_predictions(model = fit_4a, var = "pred", type = "probs") |>
+  tidy_predictions() |>
+  filter(status_prev %in% c("Normal Cognition", "MCI") & 
+         status %in%  c("Normal Cognition", "MCI"))
+
+# Identifying right most point for each label point
+label_data <- age_cohots |>
+  group_by(status_prev, Age) |>
+  filter(Total_p == max(Total_p)) |>
+  ungroup()
+
+# Plotting ---------------------------------------------------------------------
+ggplot(data = age_cohots, aes(x = Total_p, y = prob, colour = Age)) +
+  geom_line(linewidth = 1) + 
+  ggrepel::geom_text_repel(
+    data = label_data,
+    aes(label = paste0("Age ", Age), colour = factor(Age)),
+    direction = "y", nudge_x = 4, hjust = 0,
+    segment.size = 0.2, segment.colour = "grey50",
+    size = 3.5, show.legend = FALSE) +
+  scale_x_continuous(
+    breaks = seq(0, 60, by = 10),
+    sec.axis = sec_axis(
+      ~., name = "Previous State",
+      breaks = NULL, labels = NULL)) +
+  scale_y_continuous(
+    sec.axis = sec_axis(
+      ~ ., name = "Current State",
+      breaks = NULL, labels = NULL)) +
+  ggokabeito::scale_color_okabe_ito() +
+  labs(
+    title = "Predicted transition probabilities for different age cohorts",
+    x = "Procrastination", y = "Probability", colour = NULL) +
+  facet_grid(status ~ status_prev) +
+  theme_bw() +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
+    axis.title = element_text(size = 12, face = "bold"),
+    strip.text = element_text(size = 10, face = "bold"),
+    panel.grid.major.y = element_blank(),
+    panel.grid.minor.y = element_blank(),
+    legend.title = element_text(size = 10, face = "bold"),
+    legend.position = "none")
+
+```
+
+:::
 
@@
 
-```{r non-stationary-goodness}
+```{r}
 #| code-fold: true
@@
 
-```{r non-stationary-visualise-procrastination}
+```{r}
 #| code-fold: true
@@
 
-```{r non-stationary-matrix-visualise}
+```{r}
 #| code-fold: true
@@
 
-```{r matrix-distance-visualise}
+```{r}
 #| code-fold: true

### 2025-05-21 R/tidy-output.R
Update

@@ -26,6 +26,7 @@ tidy_output <- function(fit, multiplicative = FALSE) {
           term == "Total_p" ~ "Procrastination (2020)",
           term == "status_prev2" ~ "Previous state: MCI",
           term == "status_prev3" ~ "Previous state: Dementia",
+          term == "Age:Total_p" ~ "Age x Procrastination",
           term == "wave" ~ "Time",
           TRUE ~ term))
   } else {
@@ -39,6 +40,7 @@ tidy_output <- function(fit, multiplicative = FALSE) {
           term == "Total_p" ~ "Procrastination (2020)",
           term == "status_prev2" ~ "Previous state: MCI",
           term == "status_prev3" ~ "Previous state: Dementia",
+          # term == "Age:Total_p" ~ "Age x Procrastination",
           term == "wave" ~ "Time",
           term == "Gender1:wave" ~ "Gender & Time",
           term == "Age:wave" ~ "Age & Time",

### 2025-06-24 analysis/discrete-markov.qmd
Update

@@
 
-sapply(files, source)
+purrr::walk(files, source)
 
@@
   filter(Age - (2022 - 2016) >= 60) |>
-  pivot_and_factorise(time_vary = FALSE) |>
+  pivot_and_factorise(time_vary = TRUE) |>
   group_by(ID) |>
@@
 stationary_results <- rbind(tidy_output(fit_3a), tidy_output(fit_3b)) |>
-  rename(transition = y.level) |>
   mutate(
-    # Transforming to odds ratios
-    estimate = exp(estimate),
-    conf.low = exp(conf.low),
-    conf.high = exp(conf.high),
-
     # Mutating to factors
@@
                  "Procrastination (2020)", "Previous state: MCI", 
-                 "Previous state: Dementia")),
-
-    # Creating a colour code
-    colour = case_when(
-      estimate > 1 & p.value < 0.05 ~ "Positive",
-      estimate < 1 & p.value < 0.05 ~ "Negative",
-      TRUE ~ "NS"),
-
-    colour = factor(colour, levels = c("Positive", "Negative", "NS"))
+                 "Previous state: Dementia"))
     )
@@
   Education_tri = factor(0),
-  Depression = mean(data_stack$Depression),
+  Depression = mean(data_stack$Depression, na.rm = TRUE),
   status_prev = levels(data_stack$status_prev),
@@
   Education_tri = factor(0),
-  Depression = mean(data_stack$Depression),
+  Depression = mean(data_stack$Depression, na.rm = TRUE),
   Total_p = mean(data_stack$Total_p),
@@
 interaction_results <- rbind(tidy_output(fit_4a), tidy_output(fit_4b)) |>
-  rename(transition = y.level) |>
   mutate(
-    # Transforming to odds ratios
-    estimate = exp(estimate),
-    conf.low = exp(conf.low),
-    conf.high = exp(conf.high),
-
     # Mutating to factors
@@
     term = factor(
-      term, 
+      term, =
       levels = c("Being female", "Age", 
@@
                  "Previous state: Dementia")),
-
-    # Creating a colour code
-    colour = case_when(
-      estimate > 1 & p.value < 0.05 ~ "Positive",
-      estimate < 1 & p.value < 0.05 ~ "Negative",
-      TRUE ~ "NS"),
-
-    colour = factor(colour, levels = c("Positive", "Negative", "NS"))
     )
@@
 interaction_results |>
-  mutate(across(c(estimate, conf.low, conf.high), ~ round(x = ., digits = 3))) |>
+  mutate(across(c(estimate, conf.low, conf.high), ~ round(x = ., digits = 4))) |>
   DT::datatable(
     options = list(
-      pageLength = 6,
+      pageLength = 10,
       dom = "tip",
@@
   Gender = factor(0),
-  Age = c(60, 70, 80, 90),
+  Age = c(62:97),
   Education_tri = factor(0),
-  Depression = mean(data_stack$Depression),
+  Depression = mean(data_stack$Depression, na.rm = TRUE),
   status_prev = levels(data_stack$status_prev),
@@
   filter(status_prev %in% c("Normal Cognition", "MCI") &
-         status %in%  c("Normal Cognition", "MCI"))
+         status %in%  c("Normal Cognition", "MCI")) |>
+  mutate(label = case_when(
+    Age %in% c(70, 80, 90) ~ as.character(Age), 
+    TRUE ~ "Other"
+  ))
 
@@
 label_data <- age_cohots |>
+  filter(Age %in% c(70, 80, 90)) |>
   group_by(status_prev, Age) |>
@@
 # Plotting ---------------------------------------------------------------------
-ggplot(data = age_cohots, aes(x = Total_p, y = prob, colour = Age)) +
-  geom_line(linewidth = 1) + 
+ggplot(data = age_cohots, aes(x = Total_p, y = prob, 
+                              group = interaction(Age, status_prev, status),
+                              colour = label)) +
+  geom_line(aes(linewidth = label)) + 
   ggrepel::geom_text_repel(
     data = label_data,
-    aes(label = paste0("Age ", Age), colour = factor(Age)),
+    aes(label = paste0("Age ", Age), colour = as.character(Age)),
     direction = "y", nudge_x = 4, hjust = 0,
-    segment.size = 0.2, segment.colour = "grey50",
-    size = 3.5, show.legend = FALSE) +
+    segment.size = 0.5, segment.colour = "grey50",
+    size = 3, show.legend = FALSE) +
   scale_x_continuous(
     breaks = seq(0, 60, by = 10),
-    sec.axis = sec_axis(
-      ~., name = "Previous State",
-      breaks = NULL, labels = NULL)) +
-  scale_y_continuous(
-    sec.axis = sec_axis(
-      ~ ., name = "Current State",
-      breaks = NULL, labels = NULL)) +
-  ggokabeito::scale_color_okabe_ito() +
+    sec.axis = sec_axis( ~., name = "Previous State", breaks = NULL, labels = NULL)) +
+  scale_y_continuous(sec.axis = sec_axis( ~ ., name = "Current State", breaks = NULL, labels = NULL)) +
+  scale_colour_manual(values = c(
+    "70"    = "#E69F00", 
+    "80"    = "#56B4E9", 
+    "90"    = "#009E73",
+    "Other" = "grey80")) +
+  scale_linewidth_manual(values = c(
+    "70"    = 1.5, 
+    "80"    = 1.5, 
+    "90"    = 1.5,
+    "Other" = 0.5)) +
   labs(
@@
     legend.position = "none")
-
 ```
@@
 
-# Model 4: Additive time effects
-fit_4a <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Depression + Total_p + status_prev + wave, family = multinomial, 
+# Model 5: Additive time effects
+fit_5a <- nnet::multinom(
+  status ~ Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) +  status_prev + wave, family = multinomial, 
   data = data_stack, trace = FALSE)
 
-fit_4b <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Depression + Total_p + status_prev + wave, family = multinomial, 
+fit_5b <- nnet::multinom(
+  status ~ Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + status_prev + wave, family = multinomial, 
   data = data_stack |> mutate(status = relevel(status, ref = 2)), 
@@
 
-# Model 5: State-specific time effects
-fit_5a <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Depression + Total_p + (status_prev * wave), family = multinomial, 
+# Model 6: State-specific time effects
+fit_6a <- nnet::multinom(
+  status ~ Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + (status_prev * wave), family = multinomial, 
   data = data_stack, trace = FALSE)
 
-fit_5b <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Depression + Total_p + (status_prev * wave), family = multinomial, 
+fit_6b <- nnet::multinom(
+  status ~ Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + (status_prev * wave), family = multinomial, 
   data = data_stack |> mutate(status = relevel(status, ref = 2)), 
@@
 
-# Model 6: Full time interactions
-fit_6a <- nnet::multinom(
-  status ~ (Gender + Age + Education_tri + Depression + Total_p + status_prev) * wave, family = multinomial, 
+# Model 7: Full time interactions
+fit_7a <- nnet::multinom(
+  status ~ (Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + status_prev) * wave, family = multinomial, 
   data = data_stack, trace = FALSE)
 
-fit_6b <- nnet::multinom(
-  status ~ (Gender + Age + Education_tri + Depression + Total_p + status_prev) * wave, family = multinomial, 
+fit_7b <- nnet::multinom(
+  status ~ (Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + status_prev) * wave, family = multinomial, 
   data = data_stack |> mutate(status = relevel(status, ref = 2)), 
@@
 models <- list(
-  "Stationary" = fit_3a,
-  "Additive Time" = fit_4a,
-  "State-Time Interaction" = fit_5a,
-  "Full Interactions" = fit_6a
+  "Interaction" = fit_4a,
+  "Additive Time" = fit_5a,
+  "State-Time Interaction" = fit_6a,
+  "Full Interactions" = fit_7a
 )
@@
 # Outputting as table
-anova(fit_3a, fit_4a, fit_5a, fit_6a) |>
+anova(fit_4a, fit_5a, fit_6a, fit_7a) |>
   mutate(
-    Model = c("Stationary", "Additive", "State-Time Interactions", "Full Interactions"),
+    Model = c("Interaction", "Additive", "State-Time Interactions", "Full Interactions"),
     `Resid. Dev` = round(`Resid. Dev`, digits = 3),
@@
 ```{r non-stationary-output}
-additive_results <- rbind(tidy_output(fit_4a), tidy_output(fit_4b))
+additive_results <- rbind(tidy_output(fit_5a), tidy_output(fit_5b))
 
 additive_results |>
-  rename(transiton = y.level) |>
   mutate(across(c(estimate, conf.low, conf.high), ~ round(x = ., digits = 3))) |>
@@
 
-create_additive_predictions(data = data_stack, model = fit_4a, var = "Age", var_seq = seq(50, 97, by = 5)) |>
+create_additive_predictions(data = data_stack, model = fit_5a, var = "Age", var_seq = seq(50, 97, by = 5)) |>
   plot_additive_predictions(var = "Age", x_label = "Age", subtitle = "By Age")
@@
 
-create_additive_predictions(data = data_stack, model = fit_4a, var = "Total_dep_2016", var_seq = seq(0, 8, by = 1)) |>
-  plot_additive_predictions(var = "Total_dep_2016", x_label = "Depression", subtitle = "By depression")
+create_additive_predictions(data = data_stack, model = fit_5a, var = "Depression", var_seq = seq(0, 8, by = 1)) |>
+  plot_additive_predictions(var = "Depression", x_label = "Depression", subtitle = "By depression")
 
@@
 
-create_additive_predictions(data = data_stack, model = fit_4a, var = "Education_tri", var_seq = factor(c(0:2)), hold_constant = list(Gender = factor(0))) |>
+create_additive_predictions(data = data_stack, model = fit_5a, var = "Education_tri", var_seq = factor(c(0:2)), hold_constant = list(Gender = factor(0))) |>
   plot_additive_predictions(var = "Education_tri", x_label = "Education", subtitle = "By education level") +
@@
 
-create_additive_predictions(data = data_stack, model = fit_4a, var = "Total_p", var_seq = seq(0, 60, by = 1)) |>
+create_additive_predictions(data = data_stack, model = fit_5a, var = "Total_p", var_seq = seq(0, 60, by = 1)) |>
   plot_additive_predictions(var = "Total_p", x_label = "Procrastination", subtitle = "By procrastination")
@@
 ```
+
+## Interaction
+
+```{r}
+#| code-fold: true
+#| code-summary: "Check our my code"
+#| fig-width: 12
+#| fig-height: 10
+#| label: fig-non-stationary-interaction
+#| fig-cap: "Predicted transition probabilities from the non-stationary model (interaction effect)."
+
+# Making prediction dataset ----------------------------------------------------
+pred_data <- expand.grid(
+  Gender = factor(0),
+  Age = c(62:97),
+  Education_tri = factor(0),
+  Depression = mean(data_stack$Depression),
+  status_prev = levels(data_stack$status_prev),
+  Total_p = seq(0, 60, length = 200),
+  wave = c(2:4)
+) |>
+  modelr::add_predictions(model = fit_5a, var = "pred", type = "prob") |>
+  tidy_predictions() |>
+  filter(status_prev != "Dementia", status != "Dementia") |>
+  mutate(label = case_when(
+    Age %in% c(70, 80, 90) ~ as.character(Age), 
+    TRUE ~ "Other"
+  ))
+
+# For adding age labels --------------------------------------------------------
+label_data <- pred_data |>
+  filter(Age %in% c(70, 80, 90)) |>
+  group_by(status_prev, Age) |>
+  filter(Total_p == max(Total_p)) |>
+  ungroup()
+
+pred_data |>
+  ggplot(aes(x = Total_p, y = prob, 
+             group = interaction(Age, status_prev, status),
+             colour = label)) +
+  geom_line(aes(linewidth = label)) +
+  ggrepel::geom_text_repel(
+    data = label_data,
+    aes(label = paste0("Age ", Age), colour = as.character(Age)),
+    direction = "y", nudge_x = 4, hjust = 0,
+    segment.size = 1, segment.colour = "grey50",
+    size = 2.5, show.legend = FALSE) +
+  scale_x_continuous(
+    breaks = seq(0, 60, by = 10), expand = expansion(mult = c(0, 0.2)),
+    sec.axis = sec_axis( ~., name = "Previous State", breaks = NULL, labels = NULL)) +
+  scale_y_continuous(sec.axis = sec_axis( ~ ., name = "Current State", breaks = NULL, labels = NULL)) +
+  scale_colour_manual(values = c(
+    "70"    = "#E69F00", 
+    "80"    = "#56B4E9", 
+    "90"    = "#009E73",
+    "Other" = "grey80")) +
+  scale_linewidth_manual(values = c(
+    "70"    = 1.5, 
+    "80"    = 1.5, 
+    "90"    = 1.5,
+    "Other" = 0.5)) +
+  labs(
+    title = "Predicted transition probabilities for different age cohorts",
+    subtitle = "Across time",
+    x = "Procrastination", y = "Probability", colour = NULL) +
+  facet_grid(status ~ status_prev + wave, labeller = labeller(
+    wave = c("2" = "2016 - 2018", "3" = "2018 - 2020", "4" = "2020 - 2022"),
+  )) +
+  theme_bw() +
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
+    plot.subtitle = element_text(hjust = 0.5, size = 14, face = "bold"),
+    axis.title = element_text(size = 12, face = "bold"),
+    strip.text = element_text(size = 9, face = "bold"),
+    panel.grid.major.y = element_blank(),
+    panel.grid.minor.y = element_blank(),
+    legend.title = element_text(size = 10, face = "bold"),
+    legend.position = "none")
+
+```
+
 :::
@@
 multiplicative_output <- rbind(
-  tidy_output(fit_6a, multiplicative = TRUE),
-  tidy_output(fit_6b, multiplicative = TRUE))
+  tidy_output(fit_7a, multiplicative = TRUE),
+  tidy_output(fit_7b, multiplicative = TRUE))
 
@@
     colnames = c("Transition", "Predictor", "Estimate", "Standard Error",
-                 "Statistic", "p-val", "Lower CI", "Upper CI")
-  )
-
-```
+                 "Statistic", "p-val", "Lower CI", "Upper CI", "colour")
+  )
+```
+
+### Model visualisation
+
+```{r}
+multiplicative_output |>
+  filter(!term %in% c("Previous state: MCI", "Previous state: Dementia", "Previous state: MCI & Time", "Previous state: Dementia & Time")) |>
+  ggplot(aes(x = estimate, y = transition, colour = colour)) +
+  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
+    ggstance::geom_pointrangeh(
+    aes(xmin = conf.low, xmax = conf.high),
+    position = ggstance::position_dodgev(height = 0.5),
+    size = 1.25,
+    fatten = 3) +
+  scale_colour_manual(values = c(
+    "Positive" = "#0072B2", 
+    "Negative" = "#E69F00", 
+    "NS"       = "#B2BEB5")) +
+  facet_wrap(~term, scales = "free_x")
+```
+
 :::

### 2025-06-24 manuscript/manuscript.qmd
Update

@@
 
-## Data and study population {#data-and-study-population .unnumbered}
+## Data and study population
 
@@
 
-For this study, we focused on four biennial waves of HRS data (from 2016 to 2022).
+For this study, we focused on four waves of HRS data (from 2016 to 2022).
 Specifically our study sample consisted of respondents who participated in an experimental module assessing procrastination during the 2020 wave.
@@
 As a result, while the core HRS sample includes approximately $18,000$ respondents, our initial sample of interest consisted of $1,368$ respondents.
-We excluded respondents with missing cognitive assessment data for any wave $(n = 445)$, missing procrastination data $(n = 4)$, and those under 50 years of age  $(n = 16)$.
-This resulted in a final analytic sample of $903$ respondents.
+We excluded respondents with missing cognitive assessment data for any wave $(n = 419)$, those under 60 years of age $(n = 398)$, and those with complete missing values across the procrastination measure $(n = 2)$.
+This resulted in a final analytic sample of $549$ respondents.
 
@@
 The total procrastination score ranges from $12$ to $60$, with higher scores indicating greater procrastination.
-The Pure Procrastination Scale conducted in 2020 (wave 3) had a Cronbach's $\alpha$ score of $0.91$ in this sample, indicating high internal consistency.
+The Pure Procrastination Scale conducted in 2020 (wave 3) had a Cronbach's $\alpha$ score of $0.92$ in this sample, indicating high internal consistency.
 An example of a question from the scale includes "I delay making decisions until it's too late".
@@
 
-To account for potential confounding, we controlled for variables with established associations with cognitive function and procrastination.
-These included baseline measures conducted in 2016 (wave 1) of age, sex, educational attainment, number of cardiovascular risk factors (history of hypertension, stroke, or heart disease, and a classification on "overweight\" on a body mass index (BMI) scale $(\text{BMI} \ge 30)$), and depressive symptoms [@Abner2012; @Freedman2024; @monaghan2024a; @Yu2013].
-Educational attainment was classified into three categories: no formal education, GED (General Educational Development)/high school diploma, and college/further education.
-Depressive symptoms were measured using an eight-item version of the Center for Epidemiological Studies Depression (CES-D) scale [@Briggs2018], with scores ranging from 0 to 8, demonstrating good internal consistency $(\alpha = 0.82)$.
+To account for potential confounding, we controlled for variables withestablished associations with cognitive function and procrastination.
+These included measures of of age, sex, educational attainment, and depressive symptoms [@Abner2012; @Freedman2024; @Monaghan2024a; @Yu2013].
+Educationalattainment was classified into three categories: no formal education, GED (General Educational Development)/high school diploma, and college/further education.
+Depressive symptoms were measured using an eight-item version of the Center for Epidemiological Studies Depression (CES-D) scale [@Briggs2018], with scores ranging from 0 to 8, demonstrating good internal consistency $(\alpha = 0.60)$.
+
+## Data Analysis
+
+All data analysis was carried out in R [@R2025].
+Within the final dataset, missing data accounted for only 0.3%, all of which was contained within the educational attainment variable.
+To model transitions in cognitive states over time, we employed a first order discrete-time Markov model, a class of stochastic processes that satisfy the Markov property [@Zhang2010], which can be formally expressed as:
+
+$$P(X_{t+1} = j \vert X_t = i, X_{t-1} = i_{t-1}, \dots X_0 = i_0) = P(X_{t+1} = j \vert X_t = i)$$
+
+This property asserts that the probability of transitioning to a future state $X_{t+1}$ depends only on the current state $X_t$, and not on the full history of preceding states.
+
+Unlike continuous-time models, discrete-time Markov models are not readily supported by a dedicated R package for deriving transition probabilities.
+Therefore, we implemented the model manually using multinomial logistic regression from the `nnet` package [@Venables2022].
+This approach estimates the log-odds of transitioning to each non-reference state as a linear function of covariates, relative to a chosen reference category.
+For a system with $K$ cognitive states (with state $K$ as the reference), the model takes the form:
+
+$$log \left( \frac{P(Y = j \vert x)}{P(Y = k \vert x)} \right) = \alpha + \beta_j^Tx \qquad \text{for } j = 1, \dots K-1$$
+
+From these equations, the predicted transition probabilities for the non-reference states $j = 1, \dots, K - 1$ are derived as:
+
+$$P(Y = j \vert x) = \frac{e^{\alpha_j + \beta_j^Tx}}{1 + \sum^{k - 1}_{k = 1} e^{\alpha_{k} + \beta_k^Tx}}$$
+
+and for the reference state $K$ as:
+
+$$P(Y = k \vert x) = \frac{1}{1 + \sum^{k - 1}_{k = 1} e^{\alpha_{k} + \beta_j^Tx }}$$
+
+# Results
+
+As mentioned, our initial sample of interest consisted of $1,368$ respondents.
+However, after excluding $819$ respondents our final analytic sample comprised $549$ respondents with the following age distribution: 60 - 70 $(n = 186)$, 71 - 80 $(n = 203)$, 81 - 90 $(n = 142)$, and 90+ $(n = 18)$.
+Descriptive statistics for the full sample, as well as stratified by cognitive status (normal cognition, MCI, and dementia), are presented in Table [\[table:1\]](#table:1){reference-type="ref" reference="table:1"}.
+Additionally, Table [1](#table:2){reference-type="ref" reference="table:2"} presents both the frequencies and estimated transition probabilities of cognitive status changes across waves.
+Specifically, it captures unconditional transitions between wave one and two (first transition), wave two and three (second transition), and wave three and four (third transition), yielding a total of 1,647 observed transitions over time.
+
+*Note.* Descriptives for continuous and categorical variables are
+represented using means $\pm$ standard\
+\* deviations and percentages and frequencies respectively. NC = Normal
+cognition; MCI = Mild\
+\* cognitive impairment; GED = General Educational Development.
+[]{#table:1 label="table:1"}
+
+::: {#table:2}
+  **2016**
+  ---------- -------- -------- ----------
+             NC       MCI      Dementia
+  NC         (0.75)   (0.07)   (0.001)
+  MCI        (0.09)   (0.06)   (0.01)
+  Dementia   \(0\)    \(0\)    (0.02)
+  **2018**
+             NC       MCI      Dementia
+  NC         (0.75)   (0.09)   \(0\)
+  MCI        (0.07)   (0.03)   (0.02)
+  Dementia   \(0\)    \(0\)    (0.03)
+  **2020**
+             NC       MCI      Dementia
+  NC         (0.68)   (0.13)   (0.02)
+  MCI        (0.05)   (0.05)   (0.01)
+  Dementia   \(0\)    \(0\)    (0.05)
+
+  : Frequencies (and transition probabilities) of cognitive status
+  transitions from HRS waves.
+:::
+
+*Note.* NC = Normal cognition; MCI = Mild cognitive impairment
+[]{#table:2 label="table:2"}
+
+Initially, a Kruskal-Wallis test was conducted to examine differences in
+procrastination scores (measured in 2020) across three cognitive status
+groups after Levene's test indicated violation of homogeneity of
+variance $(p = 0.039)$. The analysis revealed a statistically
+significant effect of cognitive status, $(\chi^2(2) = 17.54, p < .001)$,
+indicating that procrastination levels differed significantly between at
+least two of the groups. Post-hoc analysis using a pairwise Wilcoxon
+rank-sum test with a Benjamini-Hochberg correction showed that
+participants with normal cognition $(M = 27.7, SD = 11.7)$ reported
+significantly lower levels of procrastination than those with both MCI
+$(M = 32.1; SD = 11.3; p = 0.004)$ and dementia
+$(M = 36.2; SD = 14.8; p = 0.005)$. No significance difference was found
+between those with MCI and dementia $(p = 0.334)$. Figure
+[1](#fig:1){reference-type="ref" reference="fig:1"} displays the
+distribution of procrastination scores across groups, with both boxplots
+and dotplots showing median values and individual data points.
+Significance bars indicate the pairwise differences described above.
+
+<figure id="fig:1">
+<p><embed src="Figures/fig_1.pdf" /> <span id="fig:1"
+data-label="fig:1"></span></p>
+<figcaption>Group differences in 2020 procrastination scores according
+to cognitive status.</figcaption>
+</figure>
+
+## Markov analysis
+
+Results from the discrete-time Markov analysis, showed that all covariates significantly influenced the likelihood of transitioning
+between cognitive states (see Table [34](#table:3){reference-type="ref"
+reference="table:3"}). Notably, procrastination interacted significantly
+with age to affect two key transitions: increasing the likelihood of
+transitioning from normal cognition to MCI
+$(\text{OR} = 1.001; p < 0.001)$ and decreasing the likelihood of
+reverting from MCI to normal cognition $(\text{OR} = 0.999; p < 0.001)$.
+
+In addition women were significantly less likely than men to transition
+from both normal cognition to dementia $(\text{OR} = 0.068; p < 0.001)$
+and from MCI to dementia $(\text{OR} = 0.70; p < 0.001$). Educational
+attainment was another strong predictor of cognitive transitions.
+Individuals with a GED were less likely to transition from normal
+cognition to either MCI $(\text{OR} = 0.49; p < 0.001)$ or dementia
+$(\text{OR} = 0.29; p < 0.001)$ and from MCI to dementia
+$(\text{OR} = 0.64; p < 0.001)$. They were also more likely to back
+transition from MCI to normal cognition $(\text{OR} = 2.18; p < 0.001)$.
+
+Furthermore, those with a college level education or higher demonstrated
+a significantly reduced likelihood of transitioning from normal
+cognition to either MCI $(\text{OR} = 0.32; p < 0.001)$ or dementia
+$(\text{OR} = 0.07; p < 0.001)$ and from MCI to dementia
+$(\text{OR} = 0.22; p < 0.001)$. They were also more likely to back
+transition from MCI to normal cognition $(\text{OR} = 3.13; p < 0.001)$.
+
+Finally, higher levels of depressive symptomatology were associated with
+an increased likelihood of transitioning from normal cognition to both
+MCI $(\text{OR} = 1.13; p = 0.002)$ and dementia
+$(\text{OR} = 1.18; p = 0.046)$ and and a decreased likelihood of
+transitioning from MCI back to normal cognition
+$(\text{OR} = 0.89; p = 0.001)$.
+
+|                |         Intercept         |           Gender          |            Age            |            GED            |
+|----------------|:-------------------------:|:-------------------------:|:-------------------------:|:-------------------------:|
+| NC - MCI       | 0.114***  (0.113 - 0.114) |   0.941  (0.690 - 1.284)  |  1.008**  (1.002 - 1.014) | 0.446***  (0.360 - 0.551) |
+| NC - Dementia  | 0.009***  (0.009 - 0.009) | 0.669***  (0.662 - 0.687) |  1.023*** (1.010 - 1.035) | 0.295***  (0.289 - 0.302) |
+| MCI - NC       | 9.008***  (8.986 - 9.030) |   1.061 (0.789 - 1.482)   |  0.992**  (0.986 - 0.997) |  2.46***  (1.829 - 2.759) |
+| MCI - Dementia |  0.13***  (0.129 - 0.130) | 0.708***  (0.678 - 0.740) |   1.008  (0.995 - 1.021)  | 0.646***  (0.630 - 0.662) |
+|                |     Further Education     |         Depression        |      Procrastination      |   Procrastination x Age   |
+| NC - MCI       | 0.310***  (0.264 - 0.363) | 1.147***  (1.067 - 1.233) | 0.921***  (0.882 - 0.962) | 1.001***  (1.001 - 1.002) |
+| NC - Dementia  | 0.068***  (0.068 - 0.068) |   1.150  (0.979 - 1.353)  |   0.981  (0.888 - 1.084)  |   1.000  (0.888 - 1.084)  |
+| MCI - NC       | 3.240***  (2.789 - 3.762) | 0.872***  (0.811 - 0.937) | 1.085***  (1.039 - 1.134) | 0.999***  (0.998 - 0.999) |
+| MCI - Dementia | 0.214***  (0.212 - 0.216) |   1.004  (0.852 - 1.184)  |   1.005  (0.952 - 1.168)  |   0.999  (0.998 - 1.001)  |
+
+Finally, Figure┬á[2](#fig:2){reference-type="ref" reference="fig:2"}
+presents the predicted transition probabilities across varying levels of
+age and procrastination. These estimates illustrate how the interaction
+between age and procrastination influences the likelihood of progressing
+between cognitive states over time. Notably, while transition
+probabilities remain relatively stable at very low levels of
+procrastination, substantial shifts emerge as both age and
+procrastination increase. In particular, older individuals with higher
+procrastination scores show an elevated probability of cognitive decline
+and a reduced likelihood of cognitive recovery, highlighting the
+compounded risk posed by these two variables in later life.
+
+<figure id="fig:2">
+<p><embed src="Figures/fig_3.pdf" /> <span><em>Note.</em> Each curve
+represents a different age profile, ranging from the minimum (62 years)
+to the maximum (97 years) of the dataset. Specific ages of interest (70,
+80, and 90) are highlighted.</span> <span id="fig:2"
+data-label="fig:2"></span></p>
+<figcaption>Predicted transition probabilities by procrastination and
+age.</figcaption>
+</figure>
 

### 2025-06-24 R/count_transitions.R

@@ -1,0 +1,50 @@
+count_transitions <- function(data, years, absorbing = FALSE) {
+  # Counts the occurrences of cognitive status transitions across specified 
+  # years.
+  # Converts numeric cognitive status codes (1, 2, 3) into descriptive labels
+  # ("Normal Cognition", "MCI", "Dementia") and handles missing or unexpected values.
+  # Aggregates the data to count transitions and reshapes it into a long format for analysis.
+  # Arguments:
+  #    - data: The input dataset containing cognitive function data.
+  #    - years: A vector of years for which transitions should be counted.
+  # Returns:
+  #   - A dataset with counts of cognitive status transitions, including
+  #   "Missing" and "Other" categories.
+  
+  # Create dynamic column names based on the years provided
+  cogfunction_cols <- paste0("cogfunction", years)
+  
+  data <- data |>
+    select(ID, all_of(cogfunction_cols)) |>
+    na.omit() |>
+    tidyr::pivot_longer(
+      cols = !ID,
+      names_to = "Wave",
+      values_to = "Classification"
+    ) |>
+    mutate(
+      Wave = stringr::str_replace(Wave, "cogfunction", "HRS "),
+      Wave = factor(Wave, levels = c(paste0("HRS ", years))),
+      
+      Classification = case_when(
+        Classification == 1 ~ "Normal Cognition",
+        Classification == 2 ~ "MCI",
+        Classification == 3 ~ "Dementia")
+    )
+  
+  if(absorbing == TRUE){
+    data <- data |>
+      group_by(ID) |>
+      mutate(
+        Classification = ifelse(
+          cumany(Classification == "Dementia"), "Dementia", Classification)) |>
+      ungroup()
+  }
+  
+  data <- data |>
+    mutate(Classification = factor(
+      Classification,
+      levels = c("Normal Cognition", "MCI", "Dementia")))
+  
+  return(data)
+}

### 2025-06-24 R/plot_cognitive_scores.R

@@ -1,0 +1,26 @@
+plot_cognitive_scores <- function(data, year) {
+  # Plots the distribution of cognitive scores over a range of years.
+  # Converts the data from wide to long format and generates histograms for each wave.
+  # Arguments:
+  #   - data: The dataset containing cognitive scores.
+  #   - year: The starting year for the range of cognitive scores to plot.
+  # Returns:
+  #   - A ggplot object showing the distribution of cognitive scores across waves.
+  # Plotting dementia transitions
+  data |>
+    select(ID, any_of(paste0("cogtot27_imp", year:2022))) |>
+    na.omit() |>
+    tidyr::pivot_longer(cols = !ID,
+                        names_to = "Wave",
+                        values_to = "Score") |>
+    mutate(Wave = as.double(stringr::str_replace(Wave, "cogtot27_imp", ""))) |>
+    ggplot(aes(x = Score)) +
+    geom_histogram(fill = "#56B4E9",
+                   colour = "black",
+                   alpha = 0.5) +
+    scale_x_continuous(breaks = seq(0, 27, by = 3)) +
+    labs(
+      title = paste0("Distribution of cognitive scores ", year, "- 2022"),
+      x = "Cognitive Score", y = "") +
+    facet_wrap( ~ Wave)
+}

### 2025-06-24 R/plot_transitions.R

@@ -1,0 +1,31 @@
+plot_transitions <- function(data, size) {
+  # Visualizes cognitive status transitions over time using an alluvial plot.
+  # Represents the flow of individuals between cognitive states across waves.
+  # Arguments:
+  #   - data: The dataset containing transition data.
+  #   - size: The font size for labels in the plot.
+  # Returns:
+  #   - A ggplot object showing transitions between cognitive states over time.
+  
+  require(ggalluvial)
+  
+  data |>
+    ggplot(aes(x = Wave, y = n, stratum = Classification, 
+               fill = Classification, alluvium = ID
+    )) +
+    geom_stratum(alpha = 0.5, width = 0.5) +
+    geom_flow(width = 0.5) +
+    geom_text(
+      stat = "stratum",
+      aes(label = stringr::str_wrap(Classification, width = 10)),
+      hjust = 0.5,
+      vjust = 0.5,
+      size = size
+    ) +
+    labs(title = "Dementia transitions across time", 
+         x = "", y = "Frequency") +
+    scale_fill_viridis_d(direction = -1) +
+    theme(axis.text = element_text(size = 10), 
+          text = element_text(size = 10)) +
+    ggeasy::easy_remove_legend()
+}

### 2025-06-24 R/tidy-output.R

@@ -1,18 +1,28 @@
 tidy_output <- function(fit, multiplicative = FALSE) {
   output <- broom::tidy(fit, conf.int = TRUE) |>
+    rename(transition = y.level) |>
     filter(term != "(Intercept)") |>
+    mutate(
+      estimate = exp(estimate), conf.low = exp(conf.low), conf.high = exp(conf.high),
+      colour = case_when(
+        estimate > 1 & p.value < 0.05 ~ "Positive",
+        estimate < 1 & p.value < 0.05 ~ "Negative",
+        TRUE ~ "NS"),
+      colour = factor(colour, levels = c("Positive", "Negative", "NS"))
+      ) |>
     mutate(across(c(estimate:p.value), ~ round(x = ., digits = 3)))
+    
   
-  if(any(output$y.level == "1")) {
+  if(any(output$transition == "1")) {
     output <- output |>
-      mutate(y.level = case_when(
-        y.level == "1" ~ "MCI - NC",
-        y.level == "3" ~ "MCI - Dementia"))
+      mutate(transition = case_when(
+        transition == "1" ~ "MCI - NC",
+        transition == "3" ~ "MCI - Dementia"))
   } else {
     output <- output |>
-      mutate(y.level = case_when(
-        y.level == "2" ~ "NC - MCI",
-        y.level == "3" ~ "NC - Dementia"))
+      mutate(transition = case_when(
+        transition == "2" ~ "NC - MCI",
+        transition == "3" ~ "NC - Dementia"))
   }
   
   if(multiplicative == FALSE) {
@@ -40,7 +50,7 @@ tidy_output <- function(fit, multiplicative = FALSE) {
           term == "Total_p" ~ "Procrastination (2020)",
           term == "status_prev2" ~ "Previous state: MCI",
           term == "status_prev3" ~ "Previous state: Dementia",
-          # term == "Age:Total_p" ~ "Age x Procrastination",
+          term == "Age:Total_p" ~ "Age x Procrastination",
           term == "wave" ~ "Time",
           term == "Gender1:wave" ~ "Gender & Time",
           term == "Age:wave" ~ "Age & Time",
@@ -50,6 +60,7 @@ tidy_output <- function(fit, multiplicative = FALSE) {
           term == "Total_p:wave" ~ "Procrastination & Time",
           term == "status_prev2:wave" ~ "Previous state: MCI & Time",
[…]

### 2025-06-24 analysis/discrete-markov.qmd
Adding apathy example

@@
 # Data -------------------------------------------------------------------------
-data <- read.csv(here::here("analysis/data/data.csv"))
+data <- read.csv(here::here("analysis/data/data_apathy.csv"))
 ```
@@
   "ID", "Gender", "Age", "Education_tri",
-  paste0("Total_dep_", seq(2016, 2022, by = 2)),
+  paste0("apathy_", seq(2016, 2022, by = 2)),
   "Total_p")
@@
     term = factor(
-      term, =
+      term,
       levels = c("Being female", "Age", 

### 2025-06-24 R/pivot-and-factorise.R

@@ -11,7 +11,7 @@ pivot_and_factorise <- function(data, time_vary = FALSE) {
   if(time_vary == TRUE) {
     # Pivoting depression variable
     dep_scores <- data |>
-      select(ID, starts_with("Total_dep")) |>
+      select(ID, starts_with("apathy")) |>
       pivot_longer(
         cols = !ID,
         names_to = "remove",
@@ -20,7 +20,7 @@ pivot_and_factorise <- function(data, time_vary = FALSE) {
     
     # Pivoting whole dataset
     data <- data |>
-      select(!starts_with("Total_dep")) |>
+      select(!starts_with("apathy")) |>
       pivot_longer(
         cols = starts_with("cogfunction"),
         names_to = "wave",

### 2025-07-02 analysis/discrete-markov.qmd
Update

@@
   filter(!term %in% c("Previous state: MCI", "Previous state: Dementia")) |>
+  # Adding text labels
+  mutate(
+    label_pos = case_when(
+      stringr::str_detect(term, "Further") & 
+        stringr::str_detect(transition, "MCI - NC") ~ estimate - 0.5,
+      stringr::str_detect(term, "Further") ~ estimate - 0.5,
+      TRUE ~ estimate
+    ),
+    label_text = sprintf(
+    "<span style='line-height:3.5'>OR = %.3f<br><br>[%.3f, %.3f]</span>",
+    estimate, conf.low, conf.high)) |>
+  # Plotting
   ggplot(aes(x = estimate, y = transition, colour = colour)) +
   geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
+  # Odds ratios
   ggstance::geom_pointrangeh(
     aes(xmin = conf.low, xmax = conf.high),
-    position = ggstance::position_dodgev(height = 0.5),
-    size = 1.25,
-    fatten = 3) +
+    position = ggstance::position_dodgev(height = 0.6),
+    size = 1,
+    fatten = 2) +
+  # Estimates and CIS
+  ggtext::geom_richtext(
+    aes(x = label_pos, label = label_text),
+    position = ggstance::position_dodge2v(height = 0.5),
+    hjust = 0,
+    size = 3,
+    fill = NA, label.colour = NA,
+    show.legend = FALSE) +
+  # Customization
   scale_colour_manual(values = c(
@@
     "NS"       = "#B2BEB5")) +
-  labs(title = "Odds ratios (stationary model)",
+  # scale_x_continuous(expand = expansion(mult = 0.025, add = 0.025)) +
+  labs(# title = "Odds ratios (stationary model)",
        x = "Odds Ratio", y = "Predictor") +
   guides(colour = "none") +
-  ggh4x::facet_manual(~ term, scales = "free_x", design = design) +
+  ggh4x::facet_manual(
+    ~ term, scales = "free_x", 
+    design = design, labeller = labeller(
+     term = c("Depression" = "Apathy")
+    )) +
+  coord_cartesian(clip = "off") +
   theme_bw() +
@@
     x = "Procrastination", y = "Probability", colour = NULL) +
-  facet_grid(status ~ status_prev) +
+  facet_grid(status ~ status_prev, labeller = labeller(
+    status = c("Normal Cognition" = "Normal Cognitive Function"),
+    status_prev = c("Normal Cognition" = "Normal Cognitive Function")
+  )) +
   theme_bw() +

### 2025-07-02 analysis/data-exploration.qmd
Update
Updating analysis scripts

@@
 
-# Setting up {.unnumbered}
+# Setting up
 
-## Loading packages {.unnumbered}
-
-```{r packages}
+```{r}
+#| label: set-up
 #| code-fold: true
-#| code-summary: "Check our my packages"
+#| code-summary: "Check out my code"
 #| warning: false
@@
 
-library(dplyr)
-library(tidyr)
-library(ggplot2)
-library(ggalluvial)
-```
-
-## Loading functions {.unnumbered}
-
-```{r functions}
-#| code-fold: true
-#| code-summary: "Check our my functions"
-
-count_transitions <- function(data, years, absorbing = FALSE) {
-  # Counts the occurrences of cognitive status transitions across specified 
-  # years.
-  # Converts numeric cognitive status codes (1, 2, 3) into descriptive labels
-  # ("Normal Cognition", "MCI", "Dementia") and handles missing or unexpected values.
-  # Aggregates the data to count transitions and reshapes it into a long format for analysis.
-  # Arguments:
-  #    - data: The input dataset containing cognitive function data.
-  #    - years: A vector of years for which transitions should be counted.
-  # Returns:
-  #   - A dataset with counts of cognitive status transitions, including
-  #   "Missing" and "Other" categories.
-
-  # Create dynamic column names based on the years provided
-  cogfunction_cols <- paste0("cogfunction", years)
-
-  data <- data |>
-    select(ID, all_of(cogfunction_cols)) |>
-    na.omit() |>
-    tidyr::pivot_longer(
-      cols = !ID,
-      names_to = "Wave",
-      values_to = "Classification"
-    ) |>
-    mutate(
-      Wave = stringr::str_replace(Wave, "cogfunction", "HRS "),
-      Wave = factor(Wave, levels = c(paste0("HRS ", years))),
-
-      Classification = case_when(
-        Classification == 1 ~ "Normal Cognition",
-        Classification == 2 ~ "MCI",
-        Classification == 3 ~ "Dementia")
+# Packages ---------------------------------------------------------------------
+pacman::p_load(
+  dplyr,         # For data manipulation
+  tidyr,         # For even more data manipulation
+  stringr,       # String manipulation
+  ggplot2,       # For cool plots
+  ggalluvial     # For alluvial diagrams
 )
 
-  if(absorbing == TRUE){
-    data <- data |>
-      group_by(ID) |>
-      mutate(
-        Classification = ifelse(
-          cumany(Classification == "Dementia"), "Dementia", Classification)) |>
-      ungroup()
-  }
+# Functions --------------------------------------------------------------------
+func_files <- c(
+  here::here("R/count_transitions.R"),
+  here::here("R/extract-years.R"),
+  here::here("R/plot_cognitive_scores.R"),
+  here::here("R/plot_transitions.R")
+)
 
-  data <- data |>
-    mutate(Classification = factor(
-      Classification,
-      levels = c("Normal Cognition", "MCI", "Dementia")))
+purrr::walk(func_files, source)
 
-  return(data)
-}
-
-plot_cognitive_scores <- function(data, year) {
-  # Plots the distribution of cognitive scores over a range of years.
-  # Converts the data from wide to long format and generates histograms for each wave.
-  # Arguments:
-  #   - data: The dataset containing cognitive scores.
-  #   - year: The starting year for the range of cognitive scores to plot.
-  # Returns:
-  #   - A ggplot object showing the distribution of cognitive scores across waves.
-  # Plotting dementia transitions
-  data |>
-    select(ID, any_of(paste0("cogtot27_imp", year:2022))) |>
-    na.omit() |>
-    tidyr::pivot_longer(cols = !ID,
-                        names_to = "Wave",
-                        values_to = "Score") |>
-    mutate(Wave = as.double(stringr::str_replace(Wave, "cogtot27_imp", ""))) |>
-    ggplot(aes(x = Score)) +
-    geom_histogram(fill = "#56B4E9",
-                   colour = "black",
-                   alpha = 0.5) +
-    scale_x_continuous(breaks = seq(0, 27, by = 3)) +
-    labs(
-      title = paste0("Distribution of cognitive scores ", year, "- 2022"),
-      x = "Cognitive Score", y = "") +
-    facet_wrap( ~ Wave)
-}
-
-plot_transitions <- function(data, size) {
-  # Visualizes cognitive status transitions over time using an alluvial plot.
-  # Represents the flow of individuals between cognitive states across waves.
-  # Arguments:
-  #   - data: The dataset containing transition data.
-  #   - size: The font size for labels in the plot.
-  # Returns:
-  #   - A ggplot object showing transitions between cognitive states over time.
-
-  require(ggalluvial)
-
-  data |>
-    ggplot(aes(x = Wave, y = n, stratum = Classification, 
-               fill = Classification, alluvium = ID
-    )) +
-    geom_stratum(alpha = 0.5, width = 0.5) +
-    geom_flow(width = 0.5) +
-    geom_text(
-      stat = "stratum",
-      aes(label = stringr::str_wrap(Classification, width = 10)),
-      hjust = 0.5,
-      vjust = 0.5,
-      size = size
-    ) +
-    labs(title = "Dementia transitions across time", 
-         x = "", y = "Frequency") +
-    scale_fill_viridis_d(direction = -1) +
-    theme(axis.text = element_text(size = 10), 
-          text = element_text(size = 10)) +
-    ggeasy::easy_remove_legend()
-}
-```
-
-## Setting theme {.unnumbered}
-
-```{r theme}
-#| code-fold: true
-#| code-summary: "Check our my theme"
-
+# Theme ------------------------------------------------------------------------
 colour = "#212427"
@@
     theme(
+      plot.title = element_text(
+        hjust = 0.5,
+        size = 14,
+        face = "bold",
+        colour = colour
+      ),
+      plot.subtitle = element_text(
+        hjust = 0.5,
+        size = 10,
+        colour = colour
+      ),
+      axis.title = element_text(
+        size = 10,
+        face = "bold",
+        colour = colour
+      ),
+      strip.text = element_text(
+        size = 10,
+        face = "bold",
+        colour = colour
+      ),
       panel.grid = element_blank(),
-    plot.title = element_text(hjust = 0.5, size = 14, face = "bold", colour = colour),
-    plot.subtitle = element_text(hjust = 0.5, size = 10, colour = colour),
-    axis.title = element_text(size = 10, face = "bold", colour = colour),
-    strip.text = element_text(size = 10, face = "bold", colour = colour)
-    ))
+    )
+)
 ```
 
-## Reading in data {.unnumbered}
+## Reading in data
 
-```{r data}
+```{r}
+#| label: data
+
 data <- read.csv(here::here("analysis/data/data.csv")) 
@@
 
-```{r count-transitions}
+```{r}
+#| label: count-transitions
+
 dementia_data <- data |>
@@
 
-```{r missing-data}
-#| warning: false
+```{r}
 #| fig-width: 8
+#| code-fold: true
+#| code-summary: "Check out my code"
 #| label: fig-missing-data
 #| fig-cap: Occurance of missing data across HRS waves (participants were gathered from the 2020 HRS wave [in red])
+#| warning: false
 
@@
   select(any_of(paste0("cogtot27_imp", 1996:2022))) |>
-  rename_with( ~ stringr::str_replace(., "cogtot27_imp", ""), starts_with("cogtot27")) |>
+  rename_with(~ str_replace(., "cogtot27_imp", ""), starts_with("cogtot27")) |>
   visdat::vis_dat() +
@@
   theme(
-    plot.title = element_text(hjust = 0.5, size = 14, face = "bold", colour = colour),
-    plot.subtitle = element_text(hjust = 0.5, size = 12, face = "bold", colour = colour),
+    plot.title = element_text(
+      hjust = 0.5, size = 14, face = "bold", colour = colour),
+    plot.subtitle = element_text(
+      hjust = 0.5, size = 12, face = "bold", colour = colour),
     panel.grid = element_blank(),
-    axis.text.x = element_text(colour = colours)) +
-  ggeasy::easy_remove_legend() +
-  ggeasy::easy_x_axis_labels_size(size = 9)
+    axis.text.x = element_text(size = 9, colour = colours),
+    legend.position = "none"
+  )
 
@@
 
-```{r complete-data}
+```{r}
 #| fig-width: 8
+#| code-fold: true
+#| code-summary: "Check out my code"
 #| label: fig-complete-data
 #| fig-cap: HRS dataset after processing
+#| warning: false
 
@@
   select(any_of(paste0("cogtot27_imp", 2016:2022))) |>
-  rename_with( ~ stringr::str_replace(., "cogtot27_imp", ""), starts_with("cogtot27")) |>
+  rename_with(~ str_replace(., "cogtot27_imp", ""), starts_with("cogtot27")) |>
   na.omit() |>
   visdat::vis_dat() +
-  labs(
-    title = "Complete dataset across HRS Waves (2016 - 2022)",
+  labs(title = "Complete dataset across HRS Waves (2016 - 2022)", 
        subtitle = "Total Cognitive Scores") +
   theme(
-    plot.title = element_text(hjust = 0.5, size = 14, face = "bold", colour = colour),
-    plot.subtitle = element_text(hjust = 0.5, size = 12, face = "bold", colour = colour),
-    panel.grid = element_blank()) +
-  ggeasy::easy_remove_legend() +
-  ggeasy::easy_x_axis_labels_size(size = 9)
+    plot.title = element_text(
+      hjust = 0.5, size = 14, face = "bold", colour = colour),
+    plot.subtitle = element_text(
+      hjust = 0.5, size = 12, face = "bold", colour = colour),
+    axis.text.x = element_text(size = 9),
+    panel.grid = element_blank(),
+    legend.position = "none"
+  )
 ```
@@
 
-```{r cog-scores}
+```{r}
 #| message: false
@@
 @fig-dementia-proportions shows the proportion of dementia classifications from HRS 2016 - 2022.
-The variable **Missing** represents the procrastination HRS respondents who were not yet interviewed by the HRS.
-Since the analysis focuses on participants included in the 2020 HRS wave, any retroactive analysis of prior waves may result in missing data for certain individuals
 
-```{r cog-class}
+```{r}
 #| fig-width: 12
@@
 
-```{r cog-transition}
+```{r}
 #| fig-width: 12
@@
 
-```{r cog-transition-absorbing}
+```{r}
 #| fig-width: 12

### 2025-07-02 analysis/data-overview.qmd
Updating analysis scripts

@@
 title: "Data Overview"
-format: 
-  html:
-    code-tools: false
+bibliography: ../manuscript/references.bib
+csl: ../manuscript/csl/apa.csl
 ---
@@
 
-- **Normal Cognition:** Individuals in this category show no significant signs of cognitive impairment and are able to function independently in daily activities.
+- **Normative Cognitive Function:** Individuals in this category show no significant signs of cognitive impairment and are able to function independently in daily activities.
 - **Mild Cognitive Impairment (MCI):** This stage reflects a slight but noticeable decline in cognitive abilities, such as memory or thinking skills, that goes beyond what would be expected with normal aging but does not yet interfere significantly with daily life.
@@
 
-For previous waves of HRS data (1995 - 2020) there is a researcher contributed dataset of dementia classifications [@langa2023].
+For previous waves of HRS data (1995 - 2020) there is a researcher contributed dataset of dementia classifications [@Langa2023].
 Researchers have used this dataset to study the trajectories of cognitive aging, dementia risk, and related health outcomes in older adults.
@@
 
-In order to address this we developed the `LWC2022` package [@Monaghan2024] to replicate the methodology of @langa2023.
+In order to address this we developed the `LWC2022` package [@Monaghan2024b] to replicate the methodology of @Langa2023.
 This package automates the classification process, ensuring consistency with previous waves and providing researchers with an up-to-date resource for studying cognitive decline and dementia across all available HRS waves.
-@fig-lwc illustrates the LWC workflow.
-
-![LWC Workflow](../files/workflow.svg){#fig-lwc}

### 2025-07-02 analysis/hypothesis.qmd
Launching site
Updating analysis scripts

@@
     code-tools: false
+    toc: false
 ---
 
-## Primary Hypothesis
+## Hypothesis 1
 
@@
 
-- **H1:** Higher levels of procrastination are associated with:
-  - Increased risk of transitioning from Normal Cognition to MCI $(\beta = x, p < .05)$
+- **H0:** There are no group differences in procrastination scores between cognitive groups (normative cognitive function; MCI; Dementia).
+- **H1:** There are group differences in procrastination scores between cognitive groups (normative cognitive function; MCI; Dementia)
+
+## Hypothesis 2
+
+We hypothesize that:
+
+- **H0:** There is no association between procrastination and cognitive decline transitions
+- **H1:** Higher scores on a measure of procrastination are associated with:
+  - Increased risk of transitioning from normative cognitive function to MCI / dementia $(\beta = x, p < .05)$
   - Increased risk of transitioning from MCI to Dementia $(\beta = y, p < .05)$
 
-- **H0:** There is no association between procrastination and cognitive decline transitions
-
-In order to test this hypothesis we will be using a **discrete time Markov Model**
+In order to test our hypotheses we will be using both a **non-parametric ANOVA** and a post-hoc test to compare procrastination scores between cognitive groups, and a **discrete time Markov Model** to assess the association between procrastination and cognitive decline transitions.

### 2025-09-29 README.qmd
Woops :bug:

@@
 ```
+
+## Procrastination as a Marker of Cognitive Decline: Evidence from Longitudinal Transitions in the Older Adult Population
+
+- [Cormac Monaghan](https://c-monaghan.github.io/) • Maynooth University
+- [Rafael de Andrade Moral](https://rafamoral.github.io/) • Maynooth University
+- [Michelle Kelly](https://www.ncirl.ie/About/A-Z-Staff-Directory/Staff/346) • National College of Ireland
+- [Joanna McHugh Power](https://www.maynoothuniversity.ie/people/joanna-mchugh-power) • Maynooth University
 

### 2025-09-30 README.qmd
Adding scripts, files, and manuscript :fire:

@@
 ---
+title: "Procrastination as a Marker of Cognitive Decline: Evidence from Longitudinal Transitions in the Older Adult Population"
 format: gfm
@@
 library(emoji)
+
+yaml_metadata <- rmarkdown::yaml_front_matter(here::here("manuscript", "manuscript.qmd"))
+
+title <- yaml_metadata$title
+abstract <- yaml_metadata$abstract
+
+emoji_out <- emoji_glue(":potato::spiral_shell:")
 ```
 
-## Procrastination as a Marker of Cognitive Decline: Evidence from Longitudinal Transitions in the Older Adult Population
+::: {.content-visible when-format="markdown"}
+
+---
 
@@
 
-##  `r emoji_glue(":potato::spiral_shell:")`: Note on "crisp clam" project name
+---
+
+:::
+
+---
+
+> Monaghan, C., de Andrade Moral, R, Kelly, M, & McHugh Power, J. "`r title`".
+
+---
+
+## Abstract
+
+`r abstract`
+
+## `r emoji_out`: Note on "crisp clam" project name
 

### 2025-09-30 analysis/data-analysis.qmd
Adding scripts, files, and manuscript :fire:

---
title: "Analysis Plan"
format: 
  html:
    toc: false
---

## Cross sectional analysis

To examine whether procrastination scores differed by cognitive status, we employed the **Kruskal–Wallis test**, a non-parametric alternative to one-way ANOVA. This test is appropriate when the assumption of homogeneity of variance is violated. The Kruskal–Wallis test evaluates whether the distribution of ranks differs across groups and is formally expressed as:

$$
H = \frac{n - 1}{n} \times \sum^k_{i = 1}\frac{n_i \times (\bar{R} - E_R)^2}{\sigma^2}
$$

where $k$ denotes the number of groups, $n_i$ the number of participants in group $i$, $n$ the total sample size, $\bar{R}$ the mean rank sum of each group, $E_R$ the expected value of the rankings and $\sigma^2$ the rank variance. Under the null hypothesis that all groups have the same distribution, $H$ approximately follows a $\chi^2$ distribution with $k-1$ degrees of freedom.

## Longitudinal analysis

To model transitions in cognitive states over time, we employed a **first order discrete-time Markov model**, a class of stochastic processes that satisfy the Markov property, which can be formally expressed as:

$$
P(X_{t+1} = j \; \vert \; X_t = i, X_{t-1} = i_{t-1}, \dots X_0 = i_0) = P(X_{t+1} = j \; \vert x_t = 1)
$$

This property asserts that the probability of transitioning from state $X_t = i$ to a future state $X_{t+1} = j$ depends only on the current state $X_t$, and not on the full history of preceding states.

Unlike continuous-time models, discrete-time Markov models are not readily supported by a dedicated R package for deriving transition probabilities. Therefore, we implemented the model manually using multinomial logistic regression from the *nnet* package. This approach estimates the log-odds of transitioning to each non-reference state as a linear function of covariates, relative to a chosen reference state. For a system with $K$ cognitive states (with state $K$ as the reference), the model takes the form:

$$
log \left( \frac{P(Y = j \vert x)}{P(Y = k \vert x)} \right) = \beta_0 + \beta_j^Tx \qquad \text{for } j = 1, \dots K-1
$$

For non-reference states $j = 1, \dots, K - 1$

$$
P(Y = j \vert x) = \frac{e^{\beta_{0j} + \beta_j^Tx}}{1 + \sum^{k - 1}_{k = 1} e^{\beta_{0k} + \beta_k^Tx}}
$$

For the reference state $K$:

$$
P(Y = k \vert x) = \frac{1}{1 + \sum^{k - 1}_{k = 1} e^{\beta_{0k} + \beta_j^Tx }}
$$

Additional, to assess model fit and guide model selection for the covariates in the $\beta_j$ vectors, we conducted likelihood-ratio tests comparing full and reduced models. Such tests are defined as:

$$
D = 2 \times (\ell_{reduced} - \ell_{full})
$$

where $\ell_{full}$ and $\ell_{reduced}$ denote the log-likelihoods of the respective models. Under the null hypothesis that the full model does not represent an improved fit when compared to the reduced model, $D$ has an asymptotic $\chi^2$ distribution with degrees of freedom corresponding to the difference between the numbers of parameters estimated by the two models.

### 2025-09-30 analysis/data-exploration.qmd
Adding scripts, files, and manuscript :fire:

@@
 ---
-
-# Setting up
 
@@
 
-theme_set(
+theme_clean <- function() {
   theme_minimal() +
     theme(
-      plot.title = element_text(
-        hjust = 0.5,
-        size = 14,
-        face = "bold",
-        colour = colour
-      ),
-      plot.subtitle = element_text(
-        hjust = 0.5,
-        size = 10,
-        colour = colour
-      ),
-      axis.title = element_text(
-        size = 10,
-        face = "bold",
-        colour = colour
-      ),
-      strip.text = element_text(
-        size = 10,
-        face = "bold",
-        colour = colour
-      ),
+      plot.title = element_text(hjust = 0.5, size = 14, face = "bold", colour = colour ),
+      plot.subtitle = element_text( hjust = 0.5, size = 10, colour = colour),
+      axis.title = element_text(size = 10,face = "bold",colour = colour),
+      strip.text = element_text(size = 10,face = "bold",colour = colour),
       panel.grid = element_blank(),
     )
-)
-```
+  }
 
-## Reading in data
-
-```{r}
-#| label: data
-
+# Read in data -----------------------------------------------------------------
 data <- read.csv(here::here("analysis/data/data.csv")) 
 ```
-
-# Dementia Dataset
 
@@
 ```{r}
-#| fig-width: 8
+#| label: fig-missing-data
 #| code-fold: true
 #| code-summary: "Check out my code"
-#| label: fig-missing-data
+#| fig-width: 8
 #| fig-cap: Occurance of missing data across HRS waves (participants were gathered from the 2020 HRS wave [in red])
@@
        subtitle = "Total Cognitive Scores") +
-  theme(
-    plot.title = element_text(
-      hjust = 0.5, size = 14, face = "bold", colour = colour),
-    plot.subtitle = element_text(
-      hjust = 0.5, size = 12, face = "bold", colour = colour),
-    panel.grid = element_blank(),
-    axis.text.x = element_text(size = 9, colour = colours),
-    legend.position = "none"
-  )
+  theme_clean() +
+  theme(legend.position = "none")
 

### 2025-09-30 analysis/data-final.qmd
Adding scripts, files, and manuscript :fire:

@@
 ---
-title: "Download final data"
+title: "Download data"
 format:
@@
 
-## Data
+## Data (wide format)
 
 <a class="btn btn-primary btn-sm" target="_blank" href="data/data.csv">{{< fa table >}} &ensp;<span class="font-monospace">data.csv</span></a> &emsp; <a class="btn btn-success btn-sm" target="_blank" href="data/data.rds">{{< fa brands r-project >}} &ensp;<span class="font-monospace">data.rds</span></a> &emsp; <a class="btn btn-warning btn-sm" target="_blank" href="data/data.dta">{{< fa table-cells >}} &ensp;<span class="font-monospace">data.dta</span></a>
+
+## Data (long format)
+
+<a class="btn btn-primary btn-sm" target="_blank" href="data/data_long.csv">{{< fa table >}} &ensp;<span class="font-monospace">data.csv</span></a> &emsp; <a class="btn btn-success btn-sm" target="_blank" href="data/data_long.rds">{{< fa brands r-project >}} &ensp;<span class="font-monospace">data.rds</span></a> &emsp; <a class="btn btn-warning btn-sm" target="_blank" href="data/data_long.dta">{{< fa table-cells >}} &ensp;<span class="font-monospace">data.dta</span></a>

### 2025-09-30 analysis/discrete-markov.qmd
Updating analysis scripts
Adding scripts, files, and manuscript :fire:

@@
 # Data -------------------------------------------------------------------------
-data <- read.csv(here::here("analysis/data/data_apathy.csv"))
+data <- read.csv(here::here("analysis/data/data.csv"))
 ```
@@
 
-```{r covariates}
+```{r}
+#| label: covariates
+#| 
 cols <- c(
@@
 
-```{r extract-years}
+```{r}
+#| label: extract-years
+#| collapse: true
 data_stack <- data |> 
@@
 
-```{r long_data}
+```{r}
+#| label: long_data
+#| collapse: true
 data_stack <- data_stack |>
+  # Join covariate data
   inner_join(data[, cols], by = "ID") |>
@@
   group_by(ID) |>
-  # ^ Make age time varying
+  # Make age time varying
   mutate(Age = Age - (2022 - as.numeric(as.character(wave))))
@@
 
-```{r stack-data}
+```{r}
+#| label: stack-data
+#| collapse: true
+#| 
 data_stack <- data_stack |>
@@
 
-# Transition frequencies
-
-We will now calculate the transition frequencies between cognitive states for each time period.
-We will use the `create_transition_table()` function to create a transition table for each time period.
-This will be done with the help of the `map()` function from the `purrr` package.
-Finally, to combine all the transition tables into one dataset, we will use the `create_transition_dataset()` function.
-
-```{r transition-tables}
-# Creating a table dataset
-table_data <- data |>
-  extract_years(seq(2016, 2022, by = 2)) |>
-  rename_with(~ gsub("cogfunction", "HRS_", .)) |>
-  mutate(
-    across(c(HRS_2016:HRS_2022), ~ factor(.x, levels = c("Normal Cognition", "MCI", "Dementia")))
-  )
-
-# Creating transition frequencies ---------------------------------------------
-## These are the time periods we are interested in
-time_periods <- list(
-  c("2016", "2018"),
-  c("2018", "2020"), 
-  c("2020", "2022")
-)
-
-# Applying function
-transition_results <- purrr::map(time_periods, ~ create_transition_table(.x[1], .x[2]))
-names(transition_results) <- purrr::map_chr(time_periods, ~ paste(.x[2], .x[1], sep = "-"))
-
-### Creating one dataset
-transition_frequencies <- time_periods |>
-  create_transition_dataset(transition_results = transition_results)
-
-head(transition_frequencies)
-```
-
-## Observed transition matrix
-
-The probability distribution of transitions from one state to another can be represented into a transition matrix $P = (p_{ij})_{i,j}$ where each element of position $(i, j)$ represents the transition probability $p_{ij}$.
-
-In order to create this matrix we will use both the `create_transitions()` and `observed_transition_matrix()` functions.
-
-```{r observed-matrix}
-## Creating observed transition matrix -----------------------------------------
-transition_matrix_observed <- data |>
-  extract_years(seq(2016, 2022, by = 2)) |>
-  create_transitions() |>
-  observed_transition_matrix()
-
-transition_matrix_observed
-```
-
-## Visualisation
-
-Let's visualize both the transition frequencies (@fig-frequencies) and matrix (@fig-matrix).
-
-```{r}
-#| code-fold: true
-#| code-summary: "Check out my code"
-#| fig-width: 12
-#| fig-height: 10
-#| label: fig-frequencies
-#| fig-cap: "Transition frequencies between cognitive states for each time period."
-
-fig_1 <- transition_frequencies |> tran_stack_graph()
-fig_2 <- transition_frequencies |> tran_heat_map()
-
-fig_1 / fig_2
-```
-
-```{r}
-#| code-fold: true
-#| code-summary: "Check out my code"
-#| fig-width: 12
-#| label: fig-matrix
-#| fig-cap: "Observed transition matrix between cognitive states for the years 2016 - 2022."
-
-fig_3 <- transition_matrix_observed |>
-  reshape_matrix() |>
-  plot_transition_matrix()
-
-fig_3
-```
-
 # Modelling
-
 ## Markov Process Fundamentals
@@
 
-### Transition Probability Matrix
-
-For our three-state system (Normal Cognition, Mild Cognitive Impairment [MCI], Dementia), the transition matrix $P$ captures all possible transition probabilities:
-
-$$
-P = \begin{bmatrix} 
-p_{11} & p_{12} & p_{13} \\
-p_{21} & p_{22} & p_{23} \\
-p_{31} & p_{32} & p_{33} \\
-\end{bmatrix}
-$$ {#eq-t-matrix}
-
-where:
-
-- $p_{ij} = P(X_{t+1} = j \vert X_t = i)$ represents the probability of transitioning from state $i$ to state $j$.
-- Each row sums to 1, $\sum^3_{j=1} p_{ij} = 1 \quad  \forall_i \in \{1, 2, 3\}$
-
 ### Multinomial Logistic Regression Formulation
@@
 $$
-log \left( \frac{P(Y = j \vert x)}{P(Y = k \vert x)} \right) = \beta_{j0} + \beta_j^Tx \qquad \text{for } j = 1, \dots K-1
+log \left( \frac{P(Y = j \vert x)}{P(Y = k \vert x)} \right) = \beta_0 + \beta_j^Tx \qquad \text{for } j = 1, \dots K-1
 $$ {#eq-multinomial}
@@
 
-### Time-Homogeneous Approach
-
-Our implementation assumes time-homogeneous transitions, but the framework can be extended to time-varying probabilities:
-
-$$
-P^{(t)} = \begin{bmatrix} 
-p_{11}^{(t)} & p_{12}^{(t)} & \cdots & p_{1n}^{(t)} \\
-p_{21}^{(t)} & p_{22}^{(t)} & \cdots & p_{2n}^{(t)} \\
-\vdots & \vdots & \ddots & \vdots \\
-p_{n1}^{(t)} & \cdots & \cdots & p_{nn}^{(t)} \\
-\end{bmatrix}
-$$ {#eq-non-stationary}
+### Absorbing State Specification
 
@@
 
-### Absorbing State Specification
-
-We model Dementia as an absorbing state:
+However, we model Dementia as an absorbing state:
 
@@
 
-## Stationary model
+## Our model
 
-We estimate three progressively complex stationary models using `nnet::multinom()`
+As mentioned, we estimate our Markov model using `nnet::multinom()`
 
-```{r model-fitting}
-#| code-fold: true
-#| code-summary: "Check out my models"
-#| 
-# Baseline model (only gender; reference: NC)
-fit_1a <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Depression, 
+```{r}
+#| label: model-fit
+
+fit_a <- nnet::multinom(
+  status ~ Gender + Education_tri + Depression + (Total_p * Age) + status_prev, 
   family = multinomial, 
@@
 
-# Baseline model (only gender; reference: MCI)
-fit_1b <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Depression, 
-  family = multinomial, 
-  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
-  trace = FALSE)
-
-# With procrastination (reference: NC)
-fit_2a <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Depression + Total_p, 
-  family = multinomial, 
-  data = data_stack, trace = FALSE)
-
-# With procrastination (reference: MCI)
-fit_2b <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Depression + Total_p, 
-  family = multinomial, 
-  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
-  trace = FALSE)
-
-# Full model with previous state (reference NC)
-fit_3a <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Depression + Total_p + status_prev, 
-  family = multinomial, 
-  data = data_stack, trace = FALSE)
-
-# Full model with previous state (reference MCI)
-fit_3b <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Depression + Total_p + status_prev, 
-  family = multinomial, 
-  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
-  trace = FALSE)
-
-# Full model with interaction (reference NC)
-fit_4a <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + status_prev, 
-  family = multinomial, 
-  data = data_stack, trace = FALSE)
-
-# Full model with interaction (reference MCI)
-fit_4b <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + status_prev, 
+fit_b <- nnet::multinom(
+  status ~ Gender + Education_tri + Depression + (Total_p * Age) + status_prev, 
   family = multinomial, 
@@
 
-### Model comparison
+### Estimated parameters
 
-We evaluate model improvement (@tbl-stationary-comparison) using likelihood ratio tests:
-
-$$
-D = -2 \times \ell_{\text{reduced}} - \ell_{\text{full}}
-$$ {#eq-liklihood}
+Below we present the estimated parameters for the interaction model (@tbl-interaction-output). We will transform the estimates to odds ratios and colour them based on significance.
 
 ```{r}
-#| label: tbl-stationary-comparison
-#| tbl-cap: "Likelihood ratio test for stationary models"
-#| tbl-cap-location: bottom
-
-anova(fit_1a, fit_2a, fit_3a, fit_4a) |>
-  mutate(
-    Model = c("Baseline", 
-              "Procrastination", 
-              "Procrastination + Previous Status",
-              "Interaction Model"),
-    `Resid. Dev`= round(`Resid. Dev`, digits = 3),
-    `LR stat.` = round(`LR stat.`, digits = 3),
-    `Pr(Chi)` = round(`Pr(Chi)`, digits = 3)
-    ) |>
-  DT::datatable(
-    options = list(
-      pageLength = 4,
-      dom = "t",
-      ordering = FALSE,
-      columnDefs = list(list(className = "dt-center", targets = "_all"))
-    ),
-    rownames = FALSE
-  )
-```
-
-::: {.panel-tabset .nav-pills}
-## Non-interaction model
-
-### Estimated parameters
-
-Below we present the estimated parameters for the stationary model (fit 3) (@tbl-stationary-output). We will transform the estimates to odds ratios and colour them based on significance.
-
-```{r}
-#| code-fold: true
-#| code-summary: "Check out my code"
-#| label: tbl-stationary-output
-#| tbl-cap: "Estimated parameters for the stationary model"
-#| tbl-cap-location: bottom
-
-stationary_results <- rbind(tidy_output(fit_3a), tidy_output(fit_3b)) |>
-  mutate(
-    # Mutating to factors
-    transition = factor(
-      transition, 
-      levels = c("NC - MCI", "MCI - NC", 
-                 "NC - Dementia", "MCI - Dementia")),
-    term = factor(
-      term, 
-      levels = c("Being female", "Age", "High school degree vs. No education",
-                 "Further education vs. No education", "Depression",
-                 "Procrastination (2020)", "Previous state: MCI", 
-                 "Previous state: Dementia"))
-    )
-
-stationary_results |>
-  mutate(across(c(estimate, conf.low, conf.high), ~ round(x = ., digits = 3))) |>
-  DT::datatable(
-    options = list(
-      pageLength = 6,
-      dom = "tip",
-      ordering = FALSE,
-      columnDefs = list(list(className = "dt-center", targets = "_all"))
-    ),
-    rownames = FALSE
-  )
-```
-
-### Model visualisation
-
-Let's visualize the estimated odds ratios (@fig-odds-ratio) and the predicted transition probabilities (@fig-model-predictions) for the stationary model .
-
-```{r}
-#| message: false
-#| warning: false
-#| code-fold: true
-#| code-summary: "Check out my code"
-#| fig-width: 12
-#| label: fig-odds-ratio
-#| fig-cap: "Estimated odds ratios for the stationary model."
-
-fig_4a <- stationary_results |>
-  filter(!term %in% c("Previous state: MCI", "Previous state: Dementia")) |>
-  ggplot(aes(x = estimate, y = transition, colour = colour)) +
-  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
-  ggstance::geom_pointrangeh(
-    aes(xmin = conf.low, xmax = conf.high),
-    position = ggstance::position_dodgev(height = 0.5),
-    size = 1.25,
-    fatten = 3) +
-  scale_colour_manual(values = c(
-    "Positive" = "#0072B2", 
-    "Negative" = "#E69F00", 
-    "NS"       = "#B2BEB5")) +
-  labs(title = "Odds ratios (stationary model)",
-       x = "Odds Ratio", y = "Predictor") +
-  guides(colour = "none") +
-  facet_wrap(~ term, scales = "free_x") +
-  theme_bw() +
-  theme(
-    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
-    axis.title = element_text(size = 12, face = "bold"),
-    strip.text = element_text(size = 10, face = "bold"),
-    panel.grid.major.y = element_blank(),
-    panel.grid.minor.y = element_blank())
-
-fig_4a
-```
-
-```{r}
-#| code-fold: true
-#| code-summary: "Check out my code"
-#| fig-width: 12
-#| label: fig-model-predictions
-#| fig-cap: "Predicted transition probabilities from the stationary model."
-#| 
-## Making a prediction dataset -------------------------------------------------
-pred_data <- expand.grid(
-  Gender = factor(0),
-  Age = mean(data_stack$Age),
-  Education_tri = factor(0),
-  Depression = mean(data_stack$Depression, na.rm = TRUE),
-  status_prev = levels(data_stack$status_prev),
-  Total_p = seq(0, 60, length = 200))
-
-## Plotting predictions
-fig_4b <- pred_data |>
-  modelr::add_predictions(model = fit_3a, var = "pred", type = "probs") |>
-  tidy_predictions() |>
-  plot_predictions_stationary(variable = Total_p, x_axis = "Total Procrastination")
-
-fig_4b
-```
-
-### Estimated transition matrix
-
-Finally, we can compare the observed and predicted transition matrices. We will use the estimated probabilities from the model to fill in the transition matrix.
-
-```{r estimated-matrix}
-# Get all unique states
-names <- c("Normal Cognition", "MCI", "Dementia")
-states <- sort(unique(data_stack$status))
-n_states <- length(states)
-
-# Create empty transition matrix
-transition_matrix_estimated <- matrix(
-  0, nrow = n_states, ncol = n_states,
-  dimnames = list(paste("From", names), paste("To", names))
-  )
-
-# Getting estimated probabilities
-estimated_probs <- expand.grid(
-  Gender = factor(0),
-  Age = mean(data_stack$Age),
-  Education_tri = factor(0),
-  Depression = mean(data_stack$Depression, na.rm = TRUE),
-  Total_p = mean(data_stack$Total_p),
-  status_prev = states) |>
-  modelr::add_predictions(model = fit_3a, var = "pred", type = "probs")
-
-# Filling in matrix
-for(i in 1:n_states) {
-  transition_matrix_estimated[i, ] <- estimated_probs$pred[i, ]
-}
-
-transition_matrix_estimated |> round(digits = 3)
-```
-
-Let's plot this and then compare with our observed matrix
-
-```{r}
-#| code-fold: true
-#| code-summary: "Check out my code"
-#| fig-width: 12
-#| label: fig-matrix-comparison
-#| fig-cap: "Comparison of observed and estimated transition matrices."
-
-fig_5 <- transition_matrix_estimated |>
-  reshape_matrix() |>
-  plot_transition_matrix(observed = FALSE)
-
-fig_3 + fig_5 + plot_layout(axis_titles = "collect", guides = "collect")
-```
-
-## Interaction model
-
-### Estimated parameters
-
-Below we present the estimated parameters for the interaction model (fit 4; @tbl-interaction-output). We will transform the estimates to odds ratios and colour them based on significance.
-
-```{r}
-#| code-fold: true
-#| code-summary: "Check out my code"
 #| label: tbl-interaction-output
@@
 #| tbl-cap-location: bottom
+#| code-fold: true
+#| code-summary: "Check out my code"
 
-interaction_results <- rbind(tidy_output(fit_4a), tidy_output(fit_4b)) |>
+interaction_results <- rbind(tidy_output(fit_a), tidy_output(fit_b)) |>
   mutate(
@@
 ```{r}
+#| label: fig-odds-ratio-interaction
+#| fig-width: 12
+#| fig-height: 10
+#| code-fold: true
+#| code-summary: "Check out my code"
+#| fig-cap: "Estimated odds ratios for the interaction model."
 #| message: false
 #| warning: false
-#| code-fold: true
-#| code-summary: "Check out my code"
-#| fig-width: 12
-#| fig-height: 10
-#| label: fig-odds-ratio-interaction
-#| fig-cap: "Estimated odds ratios for the interaction model."
 
@@
 ```{r}
+#| label: fig-interaction-predictions
+#| fig-width: 12
+#| fig-height: 10
 #| code-fold: true
 #| code-summary: "Check out my code"
+#| fig-cap: "Predicted transition probabilities from the interaction model."
 #| warning: false
-#| fig-width: 12
-#| fig-height: 10
-#| label: fig-interaction-predictions
-#| fig-cap: "Predicted transition probabilities from the interaction model."
 
@@
   # Adding predictions
-  modelr::add_predictions(model = fit_4a, var = "pred", type = "probs") |>
+  modelr::add_predictions(model = fit_a, var = "pred", type = "probs") |>
   tidy_predictions() |>
@@
 ```{r}
+#| label: fig-interaction-age-predictions
+#| fig-width: 12
+#| fig-height: 8
 #| code-fold: true
 #| code-summary: "Check out my code"
+#| fig-cap: "Predicted transition probabilities for different age cohorts."
 #| warning: false
-#| fig-width: 12
-#| fig-height: 8
-#| label: fig-interaction-age-predictions
-#| fig-cap: "Predicted transition probabilities for different age cohorts."
 
@@
   Total_p = seq(0, 60, length = 200)) |>
-  modelr::add_predictions(model = fit_4a, var = "pred", type = "probs") |>
+  modelr::add_predictions(model = fit_a, var = "pred", type = "probs") |>
   tidy_predictions() |>
@@
   labs(
-    title = "Predicted transition probabilities for different age cohorts",
+    # title = "Predicted transition probabilities for different age cohorts",
     x = "Procrastination", y = "Probability", colour = NULL) +
   facet_grid(status ~ status_prev, labeller = labeller(
-    status = c("Normal Cognition" = "Normal Cognitive Function"),
-    status_prev = c("Normal Cognition" = "Normal Cognitive Function")
+    status = c("Normal Cognition" = "Normative Cognitive Function"),
+    status_prev = c("Normal Cognition" = "Normative Cognitive Function")
   )) +
@@
 ```
-
-:::
-
-## Non-stationary model
-
-We can also estimate a non-stationary model (@eq-non-stationary) by incorporating time `(wave)` into our model. This model allows for time-varying transition probabilities $p^{(t)}_{ij}$.
-
-We estimate three progressively complex non-stationary models:
-
-- Additive time effects `(fit_4)`: Baseline covariates + wave
-- State-specific time effects `(fit_5)`: Interaction between previous state and wave
-- Full time interactions `(fit_6)`: All covariates interacting with wave
-
-```{r non-stationary}
-#| code-fold: true
-#| code-summary: "Check out my models"
-
-## Small processing to fix wave column
-data_stack <- data_stack |>
-  mutate(wave = case_when(
-    wave == "2016" ~ 1,
-    wave == "2018" ~ 2,
-    wave == "2020" ~ 3,
-    wave == "2022" ~ 4
-  ))
-
-# Model 5: Additive time effects
-fit_5a <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) +  status_prev + wave, family = multinomial, 
-  data = data_stack, trace = FALSE)
-
-fit_5b <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + status_prev + wave, family = multinomial, 
-  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
-  trace = FALSE)
-
-# Model 6: State-specific time effects
-fit_6a <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + (status_prev * wave), family = multinomial, 
-  data = data_stack, trace = FALSE)
-
-fit_6b <- nnet::multinom(
-  status ~ Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + (status_prev * wave), family = multinomial, 
-  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
-  trace = FALSE)
-
-# Model 7: Full time interactions
-fit_7a <- nnet::multinom(
-  status ~ (Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + status_prev) * wave, family = multinomial, 
-  data = data_stack, trace = FALSE)
-
-fit_7b <- nnet::multinom(
-  status ~ (Gender + Age + Education_tri + Depression + Total_p + (Total_p * Age) + status_prev) * wave, family = multinomial, 
-  data = data_stack |> mutate(status = relevel(status, ref = 2)), 
-  trace = FALSE)
-```
-
-### Model comparison
-
-Again, we will evaluate model improvement using likelihood ratio tests (@eq-liklihood; @tbl-non-stationary-comparison).
-However, this time, we will also compare the best fitted non-stationary models `fit_3`.
-
-```{r}
-#| code-fold: true
-#| code-summary: "Check out my code"
-#| label: tbl-non-stationary-comparison
-#| tbl-cap: "Likelihood ratio test for non-stationary models"
-#| tbl-cap-location: bottom
-
-# Fit statistics
-models <- list(
-  "Interaction" = fit_4a,
-  "Additive Time" = fit_5a,
-  "State-Time Interaction" = fit_6a,
-  "Full Interactions" = fit_7a
-)
-
-# Create comparison table
-comparison_table <- purrr::map_dfr(models, broom::glance, .id = "Model") |>
-  select(edf:AIC) |>
-  mutate(across(where(is.numeric), \(x) round(x, digits = 1)))
-
-# Outputting as table
-anova(fit_4a, fit_5a, fit_6a, fit_7a) |>
-  mutate(
-    Model = c("Interaction", "Additive", "State-Time Interactions", "Full Interactions"),
-    `Resid. Dev` = round(`Resid. Dev`, digits = 3),
-    `LR stat.` = round(`LR stat.`, digits = 3),
-    `Pr(Chi)` = round(`Pr(Chi)`, digits = 3)
-    ) |>
-  cbind(comparison_table) |>
-  DT::datatable(
-    options = list(
-      pagelength = 4,
-      dom = "t",
-      ordering = FALSE,
-      columnDefs = list(list(className = 'dt-center', targets = "_all"))
-    ),
-    rownames = FALSE
-  )
-```
-
-**This is interesting!! Looks like adding in time as an additive term improved things. However, having full time interactions is better than having state-time interactions**
-
-::: {.panel-tabset .nav-pills}
-## Additive Model
-
-### Estimated parameters
-
-Key coefficients from the selected non-stationary model `fit_4`
-
-```{r non-stationary-output}
-additive_results <- rbind(tidy_output(fit_5a), tidy_output(fit_5b))
-
-additive_results |>
-  mutate(across(c(estimate, conf.low, conf.high), ~ round(x = ., digits = 3))) |>
-  DT::datatable(
-    options = list(
-      pageLength = 6,
-      ordering = FALSE,
-      columnDefs = list(list(className = "dt-center", targets = "_all"))
-    ),
-    rownames = FALSE,
-    colnames = c("Transition", "Predictor", "Estimate", "Standard Error",
-                 "Statistic", "p-val", "Lower CI", "Upper CI")
-  )
-```
-
-### Model visualisation
-
-Let's visualize the predicted transition probabilities from the non-stationary model.
-
-- @fig-non-stationary-age shows the predicted transition probabilities for age
-- @fig-non-stationary-depression shows the predicted transition probabilities for depression
-- @fig-non-stationary-education shows the predicted transition probabilities for per education level
-- @fig-non-stationary-procrastination shows the predicted transition probabilities for procrastination
-
-::: {.panel-tabset .nav-pills}
-
-## Age
-
-```{r}
-#| code-fold: true
-#| code-summary: "Check our my code"
-#| fig-width: 10
-#| fig-height: 8
-#| label: fig-non-stationary-age
-#| fig-cap: "Predicted transition probabilities from the non-stationary model (for age)."
-#|
-
-create_additive_predictions(data = data_stack, model = fit_5a, var = "Age", var_seq = seq(50, 97, by = 5)) |>
-  plot_additive_predictions(var = "Age", x_label = "Age", subtitle = "By Age")
-
-```
-
-## Depression
-
-```{r}
-#| code-fold: true
-#| code-summary: "Check our my code"
-#| fig-width: 10
-#| fig-height: 8
-#| label: fig-non-stationary-depression
-#| fig-cap: "Predicted transition probabilities from the non-stationary model (for depression)."
-#|
-
-create_additive_predictions(data = data_stack, model = fit_5a, var = "Depression", var_seq = seq(0, 8, by = 1)) |>
-  plot_additive_predictions(var = "Depression", x_label = "Depression", subtitle = "By depression")
-
-```
-
-## Education Level
-
-```{r}
-#| code-fold: true
-#| code-summary: "Check our my code"
-#| fig-width: 10
-#| fig-height: 8
-#| label: fig-non-stationary-education
-#| fig-cap: "Predicted transition probabilities from the non-stationary model (for having a high school degree)."
-
-create_additive_predictions(data = data_stack, model = fit_5a, var = "Education_tri", var_seq = factor(c(0:2)), hold_constant = list(Gender = factor(0))) |>
-  plot_additive_predictions(var = "Education_tri", x_label = "Education", subtitle = "By education level") +
-  geom_point(size = 2) +
-  scale_x_continuous(breaks = c(0, 1, 2), labels = c("No Education", "High School", "Further Education")) +
-  theme(axis.text.x = element_text(size = 8, angle = 60, vjust = 0.90, hjust = 1))
-```
-
-## Procrastination
-
-```{r}
-#| code-fold: true
-#| code-summary: "Check our my code"
-#| fig-width: 10
-#| fig-height: 8
-#| label: fig-non-stationary-procrastination
-#| fig-cap: "Predicted transition probabilities from the non-stationary model (for procrastination)."
-#|
-
-create_additive_predictions(data = data_stack, model = fit_5a, var = "Total_p", var_seq = seq(0, 60, by = 1)) |>
-  plot_additive_predictions(var = "Total_p", x_label = "Procrastination", subtitle = "By procrastination")
-
-```
-
-## Interaction
-
-```{r}
-#| code-fold: true
-#| code-summary: "Check our my code"
-#| fig-width: 12
-#| fig-height: 10
-#| label: fig-non-stationary-interaction
-#| fig-cap: "Predicted transition probabilities from the non-stationary model (interaction effect)."
-
-# Making prediction dataset ----------------------------------------------------
-pred_data <- expand.grid(
-  Gender = factor(0),
-  Age = c(62:97),
-  Education_tri = factor(0),
-  Depression = mean(data_stack$Depression),
-  status_prev = levels(data_stack$status_prev),
-  Total_p = seq(0, 60, length = 200),
-  wave = c(2:4)
-) |>
-  modelr::add_predictions(model = fit_5a, var = "pred", type = "prob") |>
-  tidy_predictions() |>
-  filter(status_prev != "Dementia", status != "Dementia") |>
-  mutate(label = case_when(
-    Age %in% c(70, 80, 90) ~ as.character(Age), 
-    TRUE ~ "Other"
-  ))
-
-# For adding age labels --------------------------------------------------------
-label_data <- pred_data |>
-  filter(Age %in% c(70, 80, 90)) |>
-  group_by(status_prev, Age) |>
-  filter(Total_p == max(Total_p)) |>
-  ungroup()
-
-pred_data |>
-  ggplot(aes(x = Total_p, y = prob, 
-             group = interaction(Age, status_prev, status),
-             colour = label)) +
-  geom_line(aes(linewidth = label)) +
-  ggrepel::geom_text_repel(
-    data = label_data,
-    aes(label = paste0("Age ", Age), colour = as.character(Age)),
-    direction = "y", nudge_x = 4, hjust = 0,
-    segment.size = 1, segment.colour = "grey50",
-    size = 2.5, show.legend = FALSE) +
-  scale_x_continuous(
-    breaks = seq(0, 60, by = 10), expand = expansion(mult = c(0, 0.2)),
-    sec.axis = sec_axis( ~., name = "Previous State", breaks = NULL, labels = NULL)) +
-  scale_y_continuous(sec.axis = sec_axis( ~ ., name = "Current State", breaks = NULL, labels = NULL)) +
-  scale_colour_manual(values = c(
-    "70"    = "#E69F00", 
-    "80"    = "#56B4E9", 
-    "90"    = "#009E73",
-    "Other" = "grey80")) +
-  scale_linewidth_manual(values = c(
-    "70"    = 1.5, 
-    "80"    = 1.5, 
-    "90"    = 1.5,
-    "Other" = 0.5)) +
-  labs(
-    title = "Predicted transition probabilities for different age cohorts",
-    subtitle = "Across time",
-    x = "Procrastination", y = "Probability", colour = NULL) +
-  facet_grid(status ~ status_prev + wave, labeller = labeller(
-    wave = c("2" = "2016 - 2018", "3" = "2018 - 2020", "4" = "2020 - 2022"),
-  )) +
-  theme_bw() +
-  theme(
-    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
-    plot.subtitle = element_text(hjust = 0.5, size = 14, face = "bold"),
-    axis.title = element_text(size = 12, face = "bold"),
-    strip.text = element_text(size = 9, face = "bold"),
-    panel.grid.major.y = element_blank(),
-    panel.grid.minor.y = element_blank(),
-    legend.title = element_text(size = 10, face = "bold"),
-    legend.position = "none")
-
-```
-
-:::
-
-### Estimated transition matrices
-
-@fig-non-stationary-matrix-plot shows the estimated time-varying transition matrices
-
-```{r}
-#| code-fold: true
-#| code-summary: "Check out my code"
-#| fig-width: 12
-#| label: fig-non-stationary-matrix-plot
-#| fig-cap: "Estimated time-varying transition matrices."
-
-# Create prediction grid for each time point
-time_points <- unique(data_stack$wave)
-state_names <- c("Normal Cognition", "MCI", "Dementia")
-
-# Get matrices for all time points
-time_varying_matrices <- purrr::map(
-  setNames(time_points, paste(time_points, "Years")),
-  ~ get_time_varying_matrix(fit_4a, .x)
-)
-
-# Converting into a tidy data frame
-non_stationary_matrices <- purrr::imap_dfr(
-  time_varying_matrices,
-  ~ as.data.frame(.x) |> 
-    tibble::rownames_to_column("From") |> 
-    pivot_longer(-From, names_to = "To", values_to = "probability") |> 
-    mutate(Time = .y),
-  .id = "Time_point"
-) |> 
-  mutate(
-    From = factor(From, levels = state_names),
-    To = factor(To, levels = rev(state_names)),
-    Time_point = forcats::fct_inorder(Time_point)
-  )
-
-# Plotting 
-non_stationary_matrices |>
-ggplot(aes(x = From, y = To, fill = probability)) +
-  geom_tile(color = "white", linewidth = 0.5) +
-  geom_text(
-    aes(label = format(round(probability, 3), nsmall = 3)),
-    size = 4.5, 
-    color = "#212427",
-    fontface = "bold") +
-    colorspace::scale_fill_continuous_diverging(
-    palette = "Blue-Red 3", mid = 0.50, alpha = 0.5, 
-    limits = c(0, 1), name = "Transition \nProbability") +
-  labs(
-    title = "Estimated Time-Varying Transition Matrices",
-    subtitle = "Showing changes in transition probabilities over time",
-    x = "Previous State (t-1)",
-    y = "Next State (t)",
-    fill = "Probability"
-  ) +
-  facet_wrap(~ Time_point, ncol = 3, labeller = labeller(
-    Time_point = function(x){
-      case_when(x == "2 Years" ~ "2018",
-                x == "3 Years" ~ "2020",
-                x == "4 Years" ~ "2022")
-    })) +
-  theme(
-    axis.text = element_text(size = 10),
-    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
-    legend.position = "right",
-    legend.text = element_text(size = 9),
-    panel.grid = element_blank()
-  )
-```
-
-## Multiplicative Model
-
-### Estimated parameters
-
-Key coefficients from the selected non-stationary model `fit_6`
-
-```{r multiplicative-output}
-multiplicative_output <- rbind(
-  tidy_output(fit_7a, multiplicative = TRUE),
-  tidy_output(fit_7b, multiplicative = TRUE))
-
-multiplicative_output |>
-  mutate(across(c(estimate, conf.low, conf.high), ~ round(x = ., digits = 3))) |>
-  DT::datatable(
-    options = list(
-      pageLength = 6,
-      ordering = FALSE,
-      columnDefs = list(list(className = "dt-center", targets = "_all"))
-    ),
-    rownames = FALSE,
-    colnames = c("Transition", "Predictor", "Estimate", "Standard Error",
-                 "Statistic", "p-val", "Lower CI", "Upper CI", "colour")
-  )
-```
-
-### Model visualisation
-
-```{r}
-multiplicative_output |>
-  filter(!term %in% c("Previous state: MCI", "Previous state: Dementia", "Previous state: MCI & Time", "Previous state: Dementia & Time")) |>
-  ggplot(aes(x = estimate, y = transition, colour = colour)) +
-  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
-    ggstance::geom_pointrangeh(
-    aes(xmin = conf.low, xmax = conf.high),
-    position = ggstance::position_dodgev(height = 0.5),
-    size = 1.25,
-    fatten = 3) +
-  scale_colour_manual(values = c(
-    "Positive" = "#0072B2", 
-    "Negative" = "#E69F00", 
-    "NS"       = "#B2BEB5")) +
-  facet_wrap(~term, scales = "free_x")
-```
-
-:::
-
-# Matrix distance metrics
-
-Given two matrices $P = (p_{ij})$ and $\hat{P} = (\hat{p}_{ij})$ of size $m \times n$, we define the following distance measures
-
-**Frobenius Norm (Matrix Euclidean Distance)**
-
-$$D_{\text{Frobenius}}(P, \hat{P}) = \|P - \hat{P}\|_F = \sqrt{\sum_{i=1}^{m}\sum_{j=1}^{n} |p_{ij} - \hat{p}_{ij}|^2}$$
-
-- Measures the Euclidean distance in matrix space
-- Sensitive to large differences due to squaring
-- Equivalent to the vector 2-norm of the flattened matrix
-
-**Manhattan Distance (L1 Norm)**
-
-$$D_{\text{Manhattan}}(P, \hat{P}) = \sum_{i=1}^{m}\sum_{j=1}^{n} |p_{ij} - \hat{p}_{ij}|$$
-
-- Sum of absolute differences between corresponding elements
-- Less sensitive to outliers than Frobenius norm
-- Useful when sparse differences are important
-
-**Maximum Difference (Infinity Norm)**
-
-$$D_{\text{Max}}(A, B) = \max_{\substack{1 \leq i \leq m \\ 1 \leq j \leq n}} |a_{ij} - b_{ij}|$$
-
-- Captures the single largest difference between elements
-- Useful for worst-case analysis
-- Ignores the distribution of other differences
-
-**Mean Absolute Difference**
-
-$$D_{\text{MeanAbs}}(P, \hat{P}) = \frac{1}{mn}\sum_{i=1}^{m}\sum_{j=1}^{n} |p_{ij} - \hat{p}_{ij}|$$
-
-- Average absolute difference across all elements
-- Scales with matrix size (unlike Manhattan distance)
-- Easy to interpret as average error per element
-
-**Root Mean Square Error (RMSE)**
-
-$$D_{\text{RMSE}}(P, \hat{P}) = \sqrt{\frac{1}{mn}\sum_{i=1}^{m}\sum_{j=1}^{n} (p_{ij} - \hat{p}_{ij})^2}$$
-
-- Similar to Frobenius but normalized by matrix size
--  Sensitive to large errors due to squaring
-- Common in statistical and machine learning applications
-
-**Correlation-Based Distance**
-
-$$D_{\text{Corr}}(P, \hat{P}) = 1 - \frac{\sum_{i=1}^{m}\sum_{j=1}^{n} (p_{ij} - \bar{P})(\hat{p}_{ij} - \bar{\hat{P}})}{\sqrt{\sum_{i=1}^{m}\sum_{j=1}^{n} (p_{ij} - \bar{P})^2 \sum_{i=1}^{m}\sum_{j=1}^{n} (\hat{p}_{ij} - \bar{\hat{p}})^2}}$$
-
-- Measures linear relationship between matrix elements
-- $\bar{P} = \frac{1}{mn}\sum_{i,j}p_{ij}$ (mean of all elements in $P$)
-- $\bar{\hat{P}} = \frac{1}{mn}\sum_{i,j}\hat{p}_{ij}$ (mean of all elements in $\hat{P}$)
-- Range: [0,2] where 0 = perfect positive correlation
-
-**Kullback-Leibler Divergence**
-
-For matrices where each row sums to 1 ($\sum_j p_{ij} = \sum_j \hat{P}_{ij} = 1$) and $p_{ij}, \hat{p}_{ij} > 0$:
-
-$$
-D_{\text{KL}}(P \parallel \hat{P}) = \sum_{i=1}^{m}\sum_{j=1}^{n} p_{ij} \log\left(\frac{p_{ij}}{\hat{p}_{ij}}\right)
-$$
-
-- Measures information loss when $\hat{P}$ approximates $P$
-- Asymmetric: $D_{\text{KL}}(P \parallel \hat{P}) \neq D_{\text{KL}}(\hat{P} \parallel P)$
-- $0\log0 = 0$ by convention
-- In practice, add $\epsilon > 0$ (e.g., $10^{-10}$) to avoid zeros
-
-## Implementation
-
-Let's calculate the distance metrics between the observed $P$ and estimated transition matrices $\hat{P}$.
-
-```{r matrix-distance}
-p       <- transition_matrix_observed
-p_hat   <- transition_matrix_estimated
-epsilon <- 1e-10
-
-# Creating tibble of distance metrics
-distances <- tibble(
-  Metric = c("Frobenius", "Manhattan", "Max", "MeanAbs", "RMSE", "Correlation", "KL"),
-  Value = c(
-    norm(p - p_hat, type = "F"),
-    sum(abs(p - p_hat)),
-    max(abs(p - p_hat)),
-    mean(abs(p - p_hat)),
-    sqrt(mean((p - p_hat)^2)),
-    1 - cor(c(p), c(p_hat)),
-    sum((p + epsilon) * log((p + epsilon) / (p_hat + epsilon)))
-  )) |>
-  mutate(Value = round(Value, 4))
-
-distances |>
-  mutate(Metric = case_when(
-    Metric == "Frobenius" ~ "Frobenius Distance",
-    Metric == "Manhattan" ~ "Manhattan Distance",
-    Metric == "Max" ~ "Max Difference",
-    Metric == "MeanAbs" ~ "Mean Absolute Difference",
-    Metric == "RMSE" ~ "Root Mean Square Error",
-    Metric == "Correlation" ~ "Correlation Distance",
-    Metric == "KL" ~ "Kullback-Leibler Divergence"
-  )) |>
-  knitr::kable(caption = "Distance based metrics", align = "c") |>
-  kableExtra::kable_styling(
-    bootstrap_options = c("striped", "hover"),
-    full_width = FALSE)
-
-```
-
-## Visualisation
-
-Let's visualize the distance metrics between the observed and estimated transition matrices (@fig-metrics).
-
-```{r}
-#| code-fold: true
-#| code-summary: "Check out my code"
-#| fig-width: 12
-#| label: fig-metrics
-#| fig-cap: "Distance metrics between observed and estimated transition matrices."
-
-caption = stringr::str_glue(
-  "**RMSE:** Root Mean Squared Error\n
-   **KL:** Kullback-Leibler Divergence\n
-   **Correlation:** 1 - Pearson Correlation Coefficient")
-
-distances |>
-  ggplot(aes(x = reorder(Metric, -Value), y = Value, fill = Metric)) +
-  geom_col(colour = "black") +
-  geom_text(aes(label = Value), vjust = -0.5) +
-  ggokabeito::scale_fill_okabe_ito() +
-  scale_y_continuous(expand = expansion(mult = c(0.075, 0.075))) +
-  labs(
-    title = "Distance Between Transition Matrices",
-    x = "Distance Metric",
-    y = "Value",
-    caption = caption) +
-  theme_minimal() +
-  theme(
-    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
-    axis.title = element_text(face = "bold", size = 12),
-    plot.caption = ggtext::element_markdown(size = 10),
-    legend.position = "none")
-```
-
-```{r test-code}
-#| include: false
-#| eval: false
-preds <- create_additive_predictions(data = data_stack, model = fit_4a, var = "Age", var_seq = seq(50, 97, by = 5)) |>
-  # Add color column based on transition significance
-  mutate(
-    transition_type = paste(status_prev, status, sep = " → "),
-    colour = case_when(
-      transition_type == "Normal Cognition → Normal Cognition" ~ "#E69F00",
-      transition_type == "Normal Cognition → MCI" ~ "#E69F00",
-      transition_type == "Normal Cognition → Dementia" ~ "#B2BEB5",
-      transition_type == "MCI → Normal Cognition" ~ "#56B4E9",
-      transition_type == "MCI → MCI" ~ "#56B4E9",
-      transition_type == "MCI → Dementia" ~ "darkgrey",
-      transition_type == "Dementia → Normal Cognition" ~ "#009E73",
-      transition_type == "Dementia → MCI" ~ "#009E73",
-      transition_type == "Dementia → Dementia" ~ "#009E73"
-    ))
-
-ggplot(preds, aes(Age, prob, color = colour, group = interaction(wave, transition_type))) +
-  geom_line(linewidth = 1) +
-  scale_color_identity("Previous State", 
-                      breaks = c("#E69F00", "#56B4E9", "#009E73"),
-                      labels = c("Normal Cognition", "MCI", "Dementia"),
-                      guide = "legend") +
-  labs(
-    title = "Transition Probabilities (Additive Model)",
-    subtitle = "By Age",
-    x = "Age", 
-    y = "Predicted Probability",
-    caption = "*Note:* Grey lines indicate non-significant transitions",
-    ) +
-    facet_grid(wave ~ status, labeller = labeller(
-      wave = function(x) case_when(
-        x == 2 ~ "Year = 2018",
-        x == 3 ~ "Year = 2020",
-        x == 4 ~ "Year = 2022",
-      ),
-      status = function(x) paste("Transition to:", x)
-    )) +
-    theme(
-      plot.subtitle = element_text(hjust = 0.5, size = 12, face = "bold"),
-      plot.caption = ggtext::element_markdown(),
-      panel.spacing = unit(1, "lines"))
-
-```

### 2025-09-30 analysis/kruskal-wallis.qmd
Adding scripts, files, and manuscript :fire:

---
title: "Kruskal Wallis Test"
format:
  html: 
    toc: false
---

```{r}
#| label: setup
#| code-fold: true
#| code-summary: "Check out my code"
#| warning: false
#| message: false

# Packages ---------------------------------------------------------------------
pacman::p_load(
  dplyr,
  ggplot2
)

# Data -------------------------------------------------------------------------
data <- readRDS(here::here("analysis/data/data_long.RDS")) |>
  filter(wave == "2020") |>
  select(ID, status, Total_p)
```

To examine whether procrastination scores differed across cognitive status groups in 2020, we first tested the assumption of homogeneity of variances using Levene’s test, which was violated ($p = 0.039$). Consequently, we employed the **Kruskal–Wallis test**, a rank-based non-parametric alternative to one-way ANOVA, to evaluate whether procrastination scores differed significantly among participants with normative cognitive function, mild cognitive impairment (MCI), and dementia.

```{r}
#| label: kw_test
#| collapse: true

# Running test
kw_results <- kruskal.test(Total_p ~ status, data = data)

# Outputting result
broom::tidy(kw_results)
```
The test revealed a significant effect of cognitive status on procrastination scores ($\chi^2(2) = 17.54, \; p < 0.001$), suggesting that at least two groups differ.

## Post-hoc analysis

To identify where these differences lay, we conducted **pairwise Wilcoxon rank-sum tests with Benjamini–Hochberg (BH) correction** for multiple comparisons:

```{r}
#| label: posthoc
#| collapse: true

# Running multiple comparison
pairwise_res <- pairwise.wilcox.test(data$Total_p, data$status, p.adjust.method = "BH")

# Outputting result
broom::tidy(pairwise_res) |>
  mutate(across(c(group1, group2), ~ case_when(
  .x == "1" ~ "Normative Cognitive Function",
  .x == "2" ~ "MCI",
  .x == "3" ~ "Dementia"
  )))
```

The results indicated that:

- Normative cognitive function participants reported significantly lower procrastination scores than those with MCI ($p = 0.004$).
- Normative cognitive function participants also reported significantly lower scores than those with dementia ($p = 0.005$).
- No significant difference was found between the MCI and dementia groups ($p = 0.334$).

## Visualisation

Figure fig-kruskal illustrates the distribution of procrastination scores across cognitive status groups. Individual scores are displayed as dotplots (left side of each box), and boxplots summarize the group distributions (right side). Significance bars represent the results of the pairwise comparisons described above.

```{r}
#| label: fig-kruskal
#| fig-width: 12
#| code-fold: true
#| code-summary: "Check out my code"
#| fig-cap: "Procrastination scores across cognitive status groups (2020)."
#| warning: false
#| message: false

anova_fig <- data |>
  ggplot(aes(x = status, y = Total_p, fill = status, colour = status)) +
  gghalves::geom_half_point(
    side = "l", 
    transformation = ggbeeswarm::position_quasirandom(width = 0.15),
    size = 1.5, alpha = 0.5) +
  gghalves::geom_half_boxplot(side = "r", colour = "black") +
  ggsignif::geom_signif(
    comparisons = list(c("1", "2"), c("2", "3"), c("1", "3")),
    annotations = c("**", "NS", "**"),
    size = 1,
    textsize = 4.5,
    step_increase = 0.05,
    margin_top = 0.05,
    y_position = c(60, 65, 70),
    tip_length = 0.01,
    vjust = 0.01
  ) +
  ggokabeito::scale_fill_okabe_ito() +
  ggokabeito::scale_color_okabe_ito() +
  scale_x_discrete(labels = c(
    "1" = "Normative Cognitive Function", 
    "2" = "MCI", 
    "3" = "Dementia")) +
  scale_y_continuous(breaks = seq(0, 60, by = 10), 
                     expand = expansion(mult = 0.05, add = 1)) +
  labs(x = NULL, y = "Procrastination") +
  guides(fill = "none", colour = "none") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 10, face = "bold"))

anova_fig
```

### 2025-09-30 index.qmd
Adding scripts, files, and manuscript :fire:

@@
 ---
-title: "Procrastination and Dementia"
-author: 
-  - name: "Cormac Monaghan"
-    url: "https://c-monaghan.github.io/"
-    orcid: "https://orcid.org/0000-0001-9012-3060"
-    affiliation:
-      - "Hamilton Institute, Maynooth University, Ireland"
-      - "Department of Psychology, Maynooth University, Ireland"
-  - name: "Rafael de Andrade Moral"
-    url: "https://rafamoral.github.io/"
-    orcid: "https://orcid.org/0000-0002-0875-3563"
-    affiliation:
-      - "Department of Mathematics and Statistics, Maynooth University, Ireland"
-  - name: "Joanna McHugh Power"
-    orcid: "https://orcid.org/0000-0002-7387-3107"
-    affiliation: 
-      - "Department of Psychology, Maynooth University, Ireland"
+title: "Procrastination as a Marker of Cognitive Decline: Evidence from Longitudinal Transitions in the Older Adult Population"
+author: "Cormac Monaghan, Rafael de Andrade Moral, <br>Michelle Kelly, and Joanna McHugh Power"
 format: 
@@
     toc: false
-sidebar: false
+    code-tools: false
+    page-layout: full
 ---
@@
 
-**This will come eventually**
+Cognitive decline is a global health concern, making the identification of early, modifiable risk factors essential.
+While apathy is a recognized prodromal marker, procrastination may also signal early executive dysfunction.
+We used longitudinal secondary data from the United States Health and Retirement Study among adults aged 60+ $(n = 549; \bar{x} =69.70; \sigma = 7.58)$.
+Cognitive function, procrastination, and a proxy measure of apathy were assessed.
+Transitions between normative cognitive function, mild cognitive impairment (MCI), and dementia were modeled using a discrete-time first-order Markov model.
+Procrastination scores were higher among individuals with MCI or dementia than those with normative cognitive function.
+Procrastination also interacted with age, disproportionately increasing the risk of decline in the oldest participants.
+Procrastination was associated with cognitive impairment and predicted transitions to MCI, suggesting it may serve as both an early behavioral marker and compounding risk factor.
+
+### Code
+
+All the raw code and data for this paper is available in a [GitHub repository](https://github.com/C-Monaghan/crisp-clam).

### 2025-09-30 manuscript/manuscript.qmd
Adding scripts, files, and manuscript :fire:

--- 2025-05-21 manuscript/manuscript.qmd
@@
 ---
-title: "Procrastination and Dementia"
+title: "Procrastination as a Marker of Cognitive Decline: Evidence from Longitudinal Transitions in the Older Adult Population"
 author: 
-  - name: "Cormac Monaghan"
-    url: "https://c-monaghan.github.io/"
-    orcid: "https://orcid.org/0000-0001-9012-3060"
+  - name: Cormac Monaghan
+    url: https://c-monaghan.github.io/
+    email: cormacmonaghan@protonmail.com
+    orcid: https://orcid.org/0000-0001-9012-3060
     affiliation:
-      - "Hamilton Institute, Maynooth University, Ireland"
-      - "Department of Psychology, Maynooth University, Ireland"
-    email: cormacmonaghan@protonmail.com
+      - Hamilton Institute, Maynooth University, Ireland
+      - Department of Psychology, Maynooth University, Ireland
     attributes:
       corresponding: true
-  - name: "Rafael de Andrade Moral"
-    url: "https://rafamoral.github.io/"
-    orcid: "https://orcid.org/0000-0002-0875-3563"
-    affiliation:
-      - "Department of Mathematics and Statistics, Maynooth University, Ireland"
-  - name: "Joanna McHugh Power"
-    orcid: "https://orcid.org/0000-0002-7387-3107"
-    affiliation: 
-      - "Department of Psychology, Maynooth University, Ireland"
+  - name: Rafael de Andrade Moral
+    url: https://rafamoral.github.io/
+    orcid: https://orcid.org/0000-0002-0875-3563
+    affiliation: Department of Mathematics and Statistics, Maynooth University, Ireland
+  - name: Michelle Kelly
+    url: https://www.ncirl.ie/About/A-Z-Staff-Directory/Staff/346
+    orcid: https://orcid.org/0000-0001-6159-6707
+    affiliation: Department of Psychology, National College of Ireland
+  - name: Joanna McHugh Power
+    url: https://www.maynoothuniversity.ie/people/joanna-mchugh-power
+    orcid: https://orcid.org/0000-0002-7387-3107
+    affiliation: Department of Psychology, Maynooth University, Ireland
 correspondence-prefix: "Correspondence concerning this article should be addressed to"
+abstract: |
+  Cognitive decline is a global health concern, making the identification of early, modifiable risk factors essential.
+While apathy is a recognized prodromal marker, procrastination may also signal early executive dysfunction.
+We used longitudinal secondary data from the United States Health and Retirement Study among adults aged 60+ $(n = 549; \bar{x} = 69.70;\sigma = 7.58)$.
+Cognitive function, procrastination, and a proxy measure of apathy were assessed.
+Transitions between normative cognitive function, mild cognitive impairment (MCI), and dementia were modeled using a discrete-time first-order Markov model.
+Procrastination scores were higher among individuals with MCI or dementia than those with normative cognitive function.
+Procrastination also interacted with age, disproportionately increasing the risk of decline in the oldest participants.
+Procrastination was associated with cognitive impairment and predicted transitions to MCI, suggesting it may serve as both an early behavioral marker and compounding risk factor.
+additional-info: |
+  **Data Availability:** All data and replication code are publicly available at <https://github.com/C-Monaghan/crisp-clam>
+keywords:
+  - Dementia risk
+  - Mild cognitive impairment
+  - Procrastination
+  - Apathy
+  - Executive dysfunction
+  - Aging;
 code-repo: "Access the code, data, and analysis at <https://github.com/C-Monaghan/crisp-clam>"
@@
 link-citations: true
+editor: 
+  markdown: 
+    wrap: 80
 ---
+
+```{r}
+#| label: setup
+#| include: false
+
+library(ggplot2)
+
+knitr::opts_chunk$set(echo = FALSE, include = FALSE, 
+warning = FALSE, message = FALSE)
+
+knitr::opts_chunk$set(
+  fig.align = "center", fig.retina = 3,
+  fig.width = 6, fig.height = (6 * 0.618),
+  out.width = "100%", collapse = TRUE
+)
+
+theme_set(
+  theme_minimal() +
+    theme(
+      plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
+      plot.subtitle = element_text(hjust = 0.5, size = 10),
+      plot.caption = ggtext::element_markdown(hjust = 1),
+      axis.title = element_text(size = 10, face = "bold"),
+      axis.text = element_text(size = 9),
+      strip.text = element_text(size = 9, face = "bold"),
+      panel.grid = element_blank()
+    )
+  )
+```
 
@@
 
-Dementia is a syndrome characterized initially by the progressive and typically irreversible decline of cognitive function, leading to memory loss, impaired reasoning, and difficulties with activities of daily living [@Prince2013; @Sanz-Blasco2022].
-It arises from various conditions, including Alzheimer's disease, vascular dementia, and Lewy body dementia [@Cao2020].
-The global burden of dementia is substantial, with projections suggesting that the number of affected individuals will rise from 57.4 million in 2019 to 152.8 million by 2050 [@Nichols2022].
-Given this projection, identifying and addressing modifiable risk factors is crucial to mitigate the growing prevalence difficulties with activities of daily living of dementia worldwide.
-In fact, targeting these factors could prevent up to 50% of dementia cases worldwide [@Livingston2024].
-The *Lancet* Commission highlights 14 key modifiable risks, emphasizing that dementia risk can be reduced at any age.
-Nevertheless, age remains the strongest non-modifiable risk factor, accounting for the largest proportion of dementia cases.
+Dementia is characterized by progressive decline of cognitive function, leading
+to memory loss and difficulties with daily living [@prince2013;
+@sanz-blasco2022]. The global burden of dementia is substantial, with cases
+projected to rise to 152.8 million by 2050 [@nichols2022]. Identifying and
+addressing modifiable risk factors is crucial to mitigate this projected rise in
+prevalence. Two particularly vulnerable groups are older adults and those with
+mild cognitive impairment (MCI), of whom up to 80% progress to dementia within
+six years \[@cooper2015; @shigemizu2020 @tschanz2006}.
 
-Mild cognitive impairment (MCI) is a condition characterized by cognitive changes --- such as memory lapses or difficulty making decisions --- that exceed typical age-related decline [@Gauthier2006].
-Prior studies suggest that 46% of people with MCI transition to a dementia state within three years and 80% within six years [@Cooper2015; @Shigemizu2020; @Tschanz2006].
-However, progression is neither uniform nor inevitable.
-Heterogeneity in outcomes underscores the importance of distinguishing between two classes of predictors: causal risk factors (e.g.
-smoking, hypertension, or diabetes) which may accelerate dementia onset and are potential targets for intervention [@Livingston2024], and prodromal markers (e.g.
-memory loss, apathy, or struggles in daily living) that reflect early disease processes and may signal limited opportunities for prevention.
-While causal factors offer avenues for mitigating risk, prodromal features often emerge closer to irreversible pathological changes, highlighting the need for early detection and tailored interventions.
+The Lancet Commission on Dementia identified fourteen causal risk factors
+including smoking, hypertension, and diabetes, which account for approximately
+45% of dementia cases worldwide [@livingston2024]. Prodromal markers, on the
+other hand, reflect early disease processes and may signal limited opportunities
+for prevention [@teipel2025]. One such prodromal marker is apathy, or the
+significant loss of motivation, as distinct from depression and cognitive
+impairment [@fresnais2023; @richard2012]. Apathy is prevalent in both MCI and
+dementia subgroups [@donovan2014; @fresnais2023, @salem2023; @vandalen2018].
+Apathy is a significant predictor of the transition from MCI to dementia
+[@ruthirakuhan2019; @salem2023, @vandalen2018]. While apathy has been
+well-established as a prodromal marker and risk factor for dementia progression,
+emerging evidence suggests that procrastination may also relate to early
+cognitive decline.
 
-Apathy, defined as a significant loss of motivation, distinct from depression and cognitive impairment [@Fresnais2023; @Richard2012] is among the most prevalent neuropsychiatric symptoms in both MCI and dementia [@Donovan2014; @Fresnais2023; @Salem2023; @vanDalen2018].
-Individuals with apathy exhibit reduced goal-directed behavior and a diminished desire to pursue rewards or pleasure [@Fresnais2023].
-Importantly, apathy has been identified as a significant predictor for the transition from MCI to dementia [@Ruthirakuhan2019; @Salem2023; @vanDalen2018].
-For instance, a meta-analysis by @vanDalen2018 found that apathy almost doubles the risk of transitioning to dementia.
-Additionally, levels of apathy measured using both the *"Neuropsychiatric Inventory Questionnaire\"* [@Kaufer2000] and *"Apathy Evaluation Scale\"* [@Marin1991] have shown to be correlated with higher levels of amyloid burden and neurofibrillary tangles in individuals with dementia, suggesting a potential connection to underlying neuropathology [@Donovan2014; @Marshall2013].
+Though superficially similar, apathy and procrastination are behaviorally
+distinct. Apathy reflects reduced internal drive and emotional engagement,
+impairing initiation of action [@fresnais2023], whereas procrastination reflects
+intact intention but delayed execution [@steel2007]. While both reflect
+impairments in executive functioning, procrastination may additionally capture
+unique aspects of emotional and motivational dysregulation . For instance,
+person A may feel no motivation to attend an exercise class and therefore never
+books it (apathy). Person B, however, books the class and intends to go, but
+postpones at the last minute (procrastination). Both apathy and procrastination
+have been linked to dysfunction in prefrontal regions implicated in dementia
+[@friden2020; @zhang2010; @joseph2021; @fahed2021].
 
-While apathy primarily reflects a lack of motivation, procrastination involves a delay in action despite an intention to complete such action [@Steel2007].
-Although both behaviours indicate disruptions in goal-directed activity, they differ in their phenomenology: apathy is marked by diminished drive or emotional engagement, whereas procrastination typically involves intact intention but impaired follow-through.
-These differences suggest that procrastination may capture aspects of executive dysfunction and emotional regulation not fully accounted for by apathy alone.
+These parallels raise the possibility that procrastination could serve as an
+early marker of cognitive impairment or even a modifiable risk factor. Chronic
+procrastination may exacerbate decline by limiting engagement in cognitively
+stimulating activities such as physical exercise, problem-solving, and
+goal-setting, which build cognitive resilience and reduce dementia risk
+[@kelly2021; @mohammadibytamar2020; @chowdhary2022; @livingston2024]. Reduced
+engagement may contribute to a cycle of cognitive disengagement, accelerating
+decline.
 
-Both behaviours have been linked with impaired executive function, particularly in goal-oriented behaviours, decision-making, and self-regulation [@Gustavson2015; @Kawagoe2017; @steel2010], domains which are frequently compromised in individuals with MCI and dementia [@Kirova2015; @Stopford2012].
-While apathy has been well-established as a prodromal marker and risk factor for dementia progression [@Ruthirakuhan2019; @Salem2023; @vanDalen2018], emerging evidence also suggests a potential link between procrastination and early cognitive decline.
-For instance, @Friden2020 and @Zhang2019 highlight that chronic procrastination is associated with dysfunction in both the dorsolateral and ventromedial prefrontal cortices, regions of the brain known to not only show early degeneration in dementia [@Joseph2021], but also be linked to higher levels of apathy [@Fahed2021; @Zhang2019].
+Importantly, procrastination is responsive to intervention through
+cognitive-behavioral strategies and self-regulation training [@rozental2014;
+@vaneerde2018]. Crucially, procrastination may identify individuals with
+motivational or executive dysfunction who do not yet meet the long-term symptom
+threshold for apathy, offering a potentially valuable and earlier marker (and
+possible target) for neurodegenerative risk.
 
-Furthermore, procrastination may exacerbate existing cognitive decline by reinforcing patterns of inaction and passivity.
-Individuals who chronically delay tasks may inadvertently engage in fewer cognitively stimulating activities, such as physical activity, problem-solving, decision-making, and goal-setting.
-Such activities are known to build cognitive resilience and reduce dementia risk [@Chowdhary2022; @Livingston2024].
-By limiting engagement in such activities, procrastination may contribute to a self-reinforcing cycle of cognitive disengagement, potentially accelerating decline over time.
+Since age remains the strongest predictor of MCI and dementia, it is important
+to consider how procrastination operates across the lifespan (i.e. interaction
+with age). Many established modifiable risk factors such as hypertension,
+hearing loss, smoking, and social isolation, exert age-dependent effects, with
+certain factors carrying more weight at midlife than in late life
+[@livingston2020; @livingston2024]. Understanding how procrastination interacts
+with age may help clarify its role in the aetiology of cognitive decline and
+identify windows of opportunity for targeted intervention.
 
-Given these behavioural and neuro-cognitive parallels with apathy, it is worth exploring whether chronic procrastination could serve as an early behavioural marker for cognitive impairment, or even a risk factor for dementia, especially in older adults.
-By examining procrastination alongside apathy, we may uncover a broader behavioural phenotype indicative of early neuropathological changes.
-Crucially, procrastination may identify individuals with motivational or executive dysfunction who do not yet meet the threshold for apathy, offering a potentially valuable and earlier behavioural marker for neurodegenerative risk.
-
-Moreover, since age remains the strongest predictor of MCI and dementia, it is important to consider how behavioural predictors such as procrastination operate across the lifespan.
-Many established modifiable risk factors such as, hypertension, hearing loss, smoking, and social isolation, exert age-dependent effects, with certain factors carrying more weight at midlife than in late life [@livingston2020; @Livingston2024].
-In light of this, we sought to explore whether procrastination also demonstrates an age-dependent risk profile.
-Understanding how procrastination interacts with age may help clarify its role in the aetiology of cognitive decline and identify windows of opportunity for targeted intervention.
-
-Although current research on procrastination in relation to dementia is non-existent, this possible association warrants exploration.
-Identifying procrastination as a potential risk factor could expand the scope of early interventions aimed at preventing or slowing the progression of dementia.
-As such, the purpose of this study was to (a) assess group differences in procrastination levels across the normal cognition, MCI, and dementia groups, and b) test the hypotheses that higher levels of procrastination would be associated with an increased probability of transitioning from normal cognitive function or MCI to dementia.
+To our knowledge, no studies have directly examined procrastination as a
+predictor of cognitive decline or dementia progression. Accordingly, the present
+study aimed to a) assess differences in procrastination levels across three
+groups: individuals with dementia, individuals with MCI, and individuals with
+neither dementia nor MCI, and b) test whether higher procrastination scores
+predict transition from normative cognition function to dementia or MCI to
+dementia.
 
@@
 
-## Data and study population {#data-and-study-population .unnumbered}
+## Data and study population
 
-Analyses were conducted using a secondary data source; a multi-wave prospective cohort study called the Health and Retirement Study (HRS; [@Fisher2018], which tracks the health, economic, and social well-being of over $18,000$ American adults primarily aged $>50$.
-The HRS is managed by the Institute for Social Research at the University of Michigan, with data collected every two years.
-Initial data collection of a participant is conducted through a face-to-face interview, with follow-up biennial interviews conducted either by phone or face-to-face.
-The average re-interview response rate ranges from $68.8\%$ to $92.3\%$ [@HRS2017].
-At the time of writing, fifteen years of HRS data are currently archived.
+Analyses were conducted using a secondary data source; a multi-wave prospective
+cohort study called the Health and Retirement Study (HRS; [@fisher2018], which
+tracks the health, economic, and social well-being of over $18,000$ American
+adults primarily aged $50+$. The HRS is managed by the Institute for Social
+Research at the University of Michigan, with data collected every two years.
+Initial data collection of a participant is conducted through a face-to-face
+interview, with follow-up biennial interviews conducted either by phone or
+face-to-face. The average retention rate ranges from $68.8\%$ to $92.3\%$
+[@HRS2017]. At the time of writing, fifteen years of HRS data are currently
+archived.
 
-For this study, we focused on four biennial waves of HRS data (from 2016 to 2022).
-Specifically our study sample consisted of respondents who participated in an experimental module assessing procrastination during the 2020 wave.
-These experimental modules, administered at the end of the core HRS interview, consist of concise questionnaires designed to explore new topics or supplement existing core survey data.
-Each respondent receives only one experimental module, with sample sizes for each module constituting approximately 10% of the core sample.
-As a result, while the core HRS sample includes approximately $18,000$ respondents, our initial sample of interest consisted of $1,368$ respondents.
-We excluded respondents with missing cognitive assessment data for any wave $(n = 445)$, missing procrastination data $(n = 4)$, and those under 50 years of age  $(n = 16)$.
-This resulted in a final analytic sample of $903$ respondents.
+We focused on four waves of HRS data from 2016 to 2022. Specifically, our study
+sample consisted of respondents who participated in an experimental module
+assessing procrastination during the 2020 wave. These experimental modules,
+administered at the end of the core HRS interview, consist of concise
+questionnaires designed to explore new topics or supplement existing core survey
+data [@juster1995]. Each respondent receives only one experimental module, with
+sample sizes for each module constituting approximately 10% of the core sample.
+As a result, while the core HRS sample includes approximately $18,000$
+respondents, our initial sample of interest consisted of $1,368$ respondents. We
+excluded respondents with missing cognitive assessment data for any wave
+$(n = 419)$, those under 60 years of age, (as cognitive symptoms typically occur
+around this age $n = 398$), and those with complete missing values across the
+procrastination measure $(n = 2)$. This resulted in a final analytic sample of
+$549$ respondents.
 
@@
 
-Cognitive function in the HRS is assessed using a series of tests adapted from the Telephone Interview for Cognitive Status (TICS; [@Brandt1988; @Fong2009].
-These tests include an immediate and delayed $10$-noun free recall test, a serial seven subtraction test, and a backward count from $20$ test.
-Based on these assessments, @Crimmins2011 developed both a $27$-point cognitive scale and validated cut-off points to assess and classify cognitive status.
-Using these points, respondents who scored $12 - 27$ were classified as having normal cognition, $7 - 11$ as having MCI, and $0 - 6$ as having dementia.
+Cognitive function in the HRS is assessed using a series of tests adapted from
+the Telephone Interview for Cognitive Status (TICS; [@brandt1988; @fong2009].
+These tests include an immediate and delayed $10$-noun free recall test (to
+assess episodic memory), a serial seven subtraction test (to assess working
+memory), and a backward count from $20$ test (to assess mental processing).
+Based on these assessments, @crimmins2011 developed both a $27$-point cognitive
+scale and validated cut-off points to assess and classify cognitive status.
+Using these points, respondents who scored $12 - 27$ were classified as having
+normal cognition, $7 - 11$ as having MCI, and $0 - 6$ as having dementia.
 
@@
 
-Procrastination was measured using the Pure Procrastination Scale [@steel2010], which consists of 12 items rated on a Likert scale ranging from $1$ (strongly disagree) to $5$ (strongly agree).
-The total procrastination score ranges from $12$ to $60$, with higher scores indicating greater procrastination.
-The Pure Procrastination Scale conducted in 2020 (wave 3) had a Cronbach's $\alpha$ score of $0.91$ in this sample, indicating high internal consistency.
-An example of a question from the scale includes "I delay making decisions until it's too late".
+Procrastination was measured using the Pure Procrastination Scale [@steel2010],
+a psychometrically validated scale for evaluating procrastination when
+conceptualized as a dysfunctional delay. The PPS consists of 12 items rated on a
+Likert scale ranging from $1$ (strongly disagree) to $5$ (strongly agree). In
+responding to the scale items, participants were asked to reflect on their
+general behavioural tendencies, with no specific time-frame provided. Total
+procrastination scores range from $12$ to $60$, with higher scores indicating
+greater tendency to procrastinate. In the HRS, the PPS was administered only in
+wave 3 (2020) of the respective waves. As such, we use the wave 3 measure as
+both a retrospective and prospective proxy for procrastination scores across all
+waves in the analysis. This approach assumes relative temporal stability in
+procrastination over the study period. In this sample, the PSS had a Cronbach’s
+$\alpha$ score of 0.92, indicating high internal consistency. In responding to
+the scale items, participants were asked to reflect on their general behavioural
+tendencies, with no specific time-frame provided. An example of a question from
+the scale includes \`\`I delay making decisions until it’s too late”.
+
+### Covariate: Apathy
+
+While no direct measure of apathy exists within the HRS, we utilised two
+questions from the eight-item version of the Center for Epidemiological Studies
+Depression (CES-D8) scale [@briggs2018] as proxies for apathy: "You felt that
+everything you did was an effort" and "You could not get going". Both items
+capture core features of apathy (behavioural and motivational disengagement)
+and, while not a comprehensive measure of apathy, provide a valid and pragmatic
+approximation for modelling purposes. Each item was measured on a binary
+“yes/no” scale with total scores ranging from \$0\$ to \$2\$.
 
@@
 
-To account for potential confounding, we controlled for variables with established associations with cognitive function and procrastination.
-These included baseline measures conducted in 2016 (wave 1) of age, sex, educational attainment, number of cardiovascular risk factors (history of hypertension, stroke, or heart disease, and a classification on "overweight\" on a body mass index (BMI) scale $(\text{BMI} \ge 30)$), and depressive symptoms [@Abner2012; @Freedman2024; @monaghan2024a; @Yu2013].
-Educational attainment was classified into three categories: no formal education, GED (General Educational Development)/high school diploma, and college/further education.
-Depressive symptoms were measured using an eight-item version of the Center for Epidemiological Studies Depression (CES-D) scale [@Briggs2018], with scores ranging from 0 to 8, demonstrating good internal consistency $(\alpha = 0.82)$.
+To account for potential confounding, we controlled for demographic variables
+with established associations with both cognitive function and procrastination
+[@vanderweele2019]. These included measures of age, sex, and educational
+attainment [@abner2012; @freedman2024; @yu2013]. Educational attainment was
+classified into three categories: no formal education, GED (General Educational
+Development)/high school diploma, and college/further education.
+
+## Data Analysis
+
+All data analysis was carried out in R [@rcoreteam2025]. To model transitions in
+cognitive states over time, we employed a first order discrete-time Markov
+model, a class of stochastic processes that satisfy the Markov property
+[@zhang2010], which can be formally expressed as:
+
+$$P(X_{t+1} = j \vert X_t = i, X_{t-1} = i_{t-1}, \dots X_0 = i_0) = P(X_{t+1} = j \vert X_t = i)$$
+
+This property asserts that the probability of transitioning from state \$X_t =
+i\$ to a future state \$X\_{t+1} = j\$ depends only on the current state
+\$X_t\$, and not on the full history of preceding states. Here, there are three
+potential states: normative cognitive function, mild cognitive impairment, and
+dementia. We consider dementia to be an absorbing state, i.e. once an individual
+reaches this state, they cannot return to the other states in a future time
+point.
+
+Unlike continuous-time models, discrete-time Markov models are not readily
+supported by a dedicated R package for deriving transition probabilities.
+Therefore, we implemented the model manually using multinomial logistic
+regression from the *nnet* package [@venables2002]. This approach estimates the
+log-odds of transitioning to each non-reference state as a linear function of
+covariates, relative to a chosen reference category. For a system with $K$
+cognitive states (with state $K$ as the reference), the model takes the form:
+
+$$log \frac{P(Y = j \; \vert \; x)}{P(Y = K \vert x)} = \beta_0 + \beta_j^\top x \qquad \text{for } j = 1, \dots K-1$$
+
+From these equations, the predicted transition probabilities for the
+non-reference states $j = 1, \dots, K - 1$ are derived as:
+
+$$P(Y = j \vert x) = \frac{e^{\beta_{0j} + \beta_j^\top x}}{1 + \sum^{K - 1}_{k = 1} e^{\beta_{0k} + \beta_k^ \top x}}$$
+
+and for the reference state $K$ as:
+
+$$ P(Y = K \; \vert \; x) = \frac{1}{1 + \sum^{K - 1}_{k = 1} e^{\beta_{0k} + \beta_j^ \top x }} $$
+
+To assess model fit and guide model selection for covariates in the $\beta_j$
+vectors, we conducted likelihood-ratio tests comparing full and reduced models,
+whose test statistic is defined as:
+
+$$ D = 2 \times (\ell_{\text{reduced}} - \ell_{\text{full}}) $$
+
+where $\ell_{\text{full}}$ and $\ell_{\text{reduced}}$ denote the
+log-likelihoods of the respective models. Under the null hypothesis that the
+full model does not represent an improved fit when compared to the reduced
+model, $D$ has an asymptotic $\chi^2$ distribution with degrees of freedom
+corresponding to the difference between the numbers of parameters estimated by
+the two models.
+
+# Results
+
+Our final analytic sample comprised $549$ respondents with the following age
+distribution: 60 - 70 $(n = 186)$, 71 - 80 $(n = 203)$, 81 - 90 $(n = 142)$, and
+90+ $(n = 18)$. Descriptive statistics for the full sample, as well as data
+stratified by cognitive status (normative cognitive function, MCI, and
+dementia), are presented in @tbl-descriptives. Both @fig-river and supplementary
+figure S1 capture the unconditional transitions and transition probabilities
+(respectively) between wave one and two (first transition), wave two and three
+(second transition), and wave three and four (third transition), yielding a
+total of $1,647$ observed transitions over time.
+
+```{r}
+#| label: tbl-descriptives
+#| fig-width: 12
+#| tbl-cap: "Baseline descriptives for the study sample and stratified by cognitive status category."
+#| tbl-cap-location: top
+#| include: true
+
+# Table information
+tbl_descriptives <- tibble::tribble(
+  ~"", ~"Full sample <br>(n = 549)", ~"NC <br>(n = 452)", ~"MCI <br>(n = 86)", ~"Dementia <br>(n = 11)",
+  "Age (years)", "69.70 ± 7.58", "69.70 ± 7.55",	"69.60 ± 7.55",	"70.00 ± 9.89",
+  "**Sex**", " ", " ", " ", " ",
+  "Male", "38.43% <br>(n = 211)", "39.82% <br>(n =180)", "32.56% <br>(n = 28)", "27.27% <br>(n = 3)",
+  "Female", "61.57% <br>(n = 338)", "60.18%<br>(n = 272)", "67.44% <br>(n = 58)",	"72.73 <br>(n = 8)",
+  "**Education**", " ", " ", " ", " ",
+  "No degree",	"16.24% <br>(n = 98)", "10.56% <br>(n = 47)",	"39.53% <br>(n = 34)",	"63.54% <br>(n = 7)",
+  "GED", "51.58% <br>(n = 279)",	"51.91% <br>(n = 231)",	"52.33% <br>(n = 45)",	"27.27 <br>(n = 3)",
+  "Further <br>education", "32.29% <br>(n = 175)", "37.53% <br>(n = 167)", "8.14% <br>(n = 7)",	"9.09% <br>(n = 1)",
+  "Apathy",	"0.37 ± 0.63", "0.31 ± 0.59",	"0.65 ± 0.76",	"0.36 ± 0.51",
+  "Procrastination",	"28.60 ± 12.00", "27.70 ± 11.30",	"32.10 ± 13.80",	"39.00 ± 16.00"
+)
+
+# Captions and footnotes
+caption <- glue::glue(
+  "*Note:* Descriptives for continuous and categorical variables are represented 
+  using means ± standard deviations and percentages and frequencies respectively. 
+  NC = Normative cognitive function; MCI = Mild cognitive impairment; 
+  GED = General Educational Development.")
+
+footnote <- glue::glue(
+  "Procrastination scores were collected in 2020 (Wave 3) and are presented 
+  here for descriptive comparison, although they were not measured at 
+  baseline.")
+
+# Outputting table 
+tbl_descriptives |>
+  setNames(c(" ", "Full sample <br>(n = 549)", "NC <br>(n = 452)", "MCI <br>(n = 86)", "Dementia <br>(n = 11)")) |>
+  tinytable::tt(notes = list(
+    caption, 
+    a = list(i = 10, j = 1, text = footnote))
+    ) |>
+  tinytable::style_tt(j = 2:5, align = "c") |>
+  tinytable::format_tt("notes", markdown = TRUE) |>
+  tinytable::format_tt(j = 1, markdown = TRUE)
+```
+
+```{r}
+#| label: fig-river
+#| fig-width: 15
+#| fig-height: 10
+#| fig-cap: "Frequency of cognitive status transitions between HRS waves."
+#| include: true
+
+fig_1 <- readRDS(here::here("files/figure_1.RDS"))
+
+fig_1
+```
+
+## Cross sectional differences
+
+Initially, a Kruskal-Wallis test was conducted to examine differences in
+procrastination scores (measured in 2020) across three cognitive status groups
+after Levene's test indicated violation of homogeneity of variance
+$(p = 0.039)$. The analysis revealed a statistically significant effect of
+cognitive status, $(\chi^2(2) = 17.54, p < .001)$, indicating that
+procrastination levels differed significantly between at least two of the
+groups. Post-hoc analysis using a pairwise Wilcoxon rank-sum test with a
+Benjamini-Hochberg correction showed that participants with normal cognition
+$(M = 27.7, SD = 11.7)$ reported significantly lower levels of procrastination
+than those with both MCI $(M = 32.1; SD = 11.3; p = 0.004)$ and dementia
+$(M = 36.2; SD = 14.8; p = 0.005)$. No significance difference was found between
+those with MCI and dementia $(p = 0.334)$. @fig-cross-sectional displays the
+distribution of procrastination scores across groups, with both boxplots and
+dotplots showing median values and individual data points. Significance bars
+indicate the pairwise differences described above.
+
+```{r}
+#| label: fig-cross-sectional
+#| fig-width: 10
+#| fig-cap: "Group differences in 2020 procrastination scores according to cognitive status."
+#| include: true
+
+fig_2 <- readRDS(here::here("files/figure_2.RDS"))
+
+fig_2
+```
+
+## Markov analysis
+
+Results from the discrete-time Markov analysis, showed that all covariates
+significantly influenced the likelihood of transitioning between cognitive
+states (see @fig-markov). Notably, procrastination interacted significantly with
+*age* to affect two key transitions: increasing the likelihood of transitioning
+from normative cognitive function to MCI $(\text{OR} = 1.001; p < 0.001)$ and
+decreasing the likelihood of reverting from MCI to normative cognitive function
+$(\text{OR} = 0.999; p < 0.001)$. *Women* were significantly less likely than
+men to transition from both normative cognitive function to dementia
+$(\text{OR} = 0.068; p < 0.001)$ and from MCI to dementia
+$(\text{OR} = 0.70; p < 0.001$).
+
+For *education attainment*, individuals with a GED were less likely to
+transition from normative cognitive function to either MCI
+$(\text{OR} = 0.49; p < 0.001)$ or dementia $(\text{OR} = 0.29; p < 0.001)$ and
+from MCI to dementia $(\text{OR} = 0.64; p < 0.001)$. They were also more likely
+to back transition from MCI to normative cognitive function
+$(\text{OR} = 2.18; p < 0.001)$. Those with a college level education or higher
+demonstrated a significantly reduced likelihood of transitioning from normative
+cognitive function to either MCI $(\text{OR} = 0.32; p < 0.001)$ or dementia
+$(\text{OR} = 0.07; p < 0.001)$ and from MCI to dementia
+$(\text{OR} = 0.22; p < 0.001)$. They were also more likely to back transition
+from MCI to normative cognitive function $(\text{OR} = 3.13; p < 0.001)$.
+
+Finally, higher levels of *apathy* were associated with an increased likelihood
+of transitioning from normative cognitive function to both MCI
+$(\text{OR} = 1.35; p = 0.005)$ and dementia $(\text{OR} = 1.36; p<0.001)$ and a
+decreased likelihood of transitioning from MCI back to normative cognitive
+function $(\text{OR} = 0.74; p = 0.003)$.
+
+```{r}
+#| label: fig-markov
+#| fig-width: 12
+#| fig-height: 10
+#| fig-cap: "Odds ratios (95% Confidence Intervals) of transitioning from one cognitive state to another calculated using generalised logit models.
+<br>*Note*.
+Odds ratios significantly different from 1 at a 5% significance rate are presented in blue (positive) or orange (negative),  otherwise they are presented in gray.
+NC = Normative cognitive function;  MCI = Mild cognitive impairment."
+#| include: true
+
+fig_3 <- readRDS(here::here("files/figure_3.RDS"))
+
+fig_3
+```
+
+@fig-age-preds presents the predicted transition probabilities across varying
+levels of age and procrastination. These estimates illustrate how the
+interaction between age and procrastination influences the likelihood of
+progressing between cognitive states over time. Notably, while transition
+probabilities remain relatively stable at very low levels of procrastination,
+substantial shifts emerge as both age and procrastination increase. In
+particular, older individuals with higher procrastination scores show an
+elevated probability of cognitive decline transitioning from normative cognitive
+function to mild cognitive impairment (MCI) and a reduced likelihood of
+transitioning back from MCI to normative cognitive function, highlighting the
+compounded risk posed by these two variables in later life.
+
+```{r}
+#| label: fig-age-preds
+#| fig-width: 12
+#| fig-height: 8
+#| fig-cap: "Predicted transition probabilities by procrastination and age.
+<br>*Note.* Each curve represents a different age profile, ranging from the minimum (62 years) to the maximum (97 years) of the dataset.
+Specific ages of interest (70, 80, and 90) are highlighted."
+#| include: true
+
+# Read in plot
+fig_4 <- readRDS(here::here("files/figure_4.RDS"))
+
+# Display plot
+fig_4
+```
+
+# Discussion
+
+Dementia poses a growing global health burden, with modifiable risk factors and
+prodromal markers offering important targets for intervention [@livingston2024;
+@livingston2020]. While apathy has been well established as a prodromal marker
+and predictor of dementia [@donovan2014; @fresnais2023; @salem2023;
+@vandalen2018], emerging evidence suggests that procrastination may also signal
+early cognitive dysfunction [@friden2020; @zhang2019]. This study examined
+whether procrastination, particularly in later life, could serve as a predictor
+of cognitive decline. We assessed both cross-sectional differences in
+procrastination across three cognitive groups (normative cognitive function, MCI
+and dementia) and longitudinal associations with transitions in cognitive
+status, to clarify whether procrastination functions as a novel and
+age-sensitive marker of emerging cognitive impairment.
+
+Our analysis revealed significant group differences in procrastination, with
+individuals experiencing cognitive impairment, both MCI and dementia, reporting
+higher procrastination scores than those with normative cognitive function.
+These findings support the hypothesis that procrastination may be associated
+with cognitive decline and aligns with emerging evidence linking procrastination
+to cognitive dysfunction [@friden2020; @zhang2019]. Interestingly, while both
+MCI and dementia groups exhibited elevated procrastination scores, no
+significant difference emerged between these two groups. This suggests that
+increases in procrastination may occur relatively early in the neurodegenerative
+process, potentially preceding or paralleling the emergence of more overt
+cognitive symptoms.
+
+However, it should be noted that the number of participants classified as having
+MCI $(n = 67)$ or dementia $(n = 27)$ was relatively small compared to those
+with normative cognitive function $(n = 455)$. This imbalance in group sizes may
+have limited the sensitivity of our comparisons and raises the possibility that
+subtle differences between MCI and dementia groups were obscured by statistical
+power constraints. Future research should aim to replicate these findings in
+more evenly distributed samples to better determine whether procrastination
+continues to increase across progressive stages of impairment or plateaus
+following early decline.
+
+Beyond these cross-sectional differences, our discrete-time Markov analysis
+offered insight into how procrastination alongside other demographic and
+psychological factors, influences the likelihood of transitioning between
+cognitive states over time. Notably, procrastination emerged as a dynamic
+predictor of cognitive change, interacting with age to significantly increase
+the odds of transitioning from normative cognitive function to MCI, while
+simultaneously decreasing the odds of moving from MCI back to normative
+cognitive function.
+
+These findings imply that procrastination may serve dual roles, both as a marker
+of early cognitive dysfunction and, potentially, as a behavioral impediment to
+moving from MCI state to the normative cognitive state. One plausible pathway
+involves the reinforcement of maladaptive behaviors such as reduced cognitive
+engagement or reduced participation in cognitively protective activities, such
+as physical exercise, social interaction or medical adherence [@hajek2025;
+@kelly2021; @sirois2007; @stead2010]. In parallel, procrastination has been
+associated with chronic stress and elevated cortisol levels [@sirois2023a],
+which may accelerate hippocampal atrophy, β-amyloid plaque deposition, and brain
+inflammation [@franks2021; @wallensten2023], further undermining cognitive
+resilience. These behavioral and biological mechanisms align with
+self-regulation theories suggesting that procrastination reflects executive
+dysfunction [@rozental2014; @steel2007], an early feature of neurodegenerative
+progression [@clark2012].
+
+The observed interaction with age (see @fig-age-preds) suggests that the
+influence of procrastination on cognitive trajectories may grow stronger with
+advancing age, a period during which neuro-plasticity diminishes, and behavioral
+risk factors exert greater influence [@livingston2020; @livingston2024]. This
+effect was particularly pronounced among the oldest participants. As illustrated
+in @fig-age-preds, the effect of procrastination is relatively modest for
+younger-older adults (those age 70). However, the slope of these transitions
+steepens considerably for individuals aged 80 and especially for those age 90.
+These findings indicate that procrastination is increasingly associated with a
+higher risk of cognitive decline and a lower likelihood of improvement in
+cognitive status in late old age.
+
+This pattern underscores the possibility that procrastination functions as a
+compounding risk factor in older adulthood, particularly among the oldest
+individuals, by both reflecting emerging cognitive difficulties and potentially
+accelerating their progression. In younger-old adults, cognitive reserve and
+compensatory mechanisms [@gooijers2024] may buffer against the impact of poor
+self-regulation. However, as individuals reach more advanced ages, these
+protective systems weaken [@roberts2015]. Consequently, maladaptive behavioral
+tendencies like procrastination can exert a disproportionate toll on cognitive
+function. These findings underscore the potential value of addressing behavioral
+regulation and self-management in older adults as part of broader dementia
+risk-reduction strategies, particularly for those of a much older age.
+
+By revealing a significant interaction between age and procrastination, this
+study highlights the importance of broadening dementia risk models to
+incorporate dynamic lifestyle and psychological variables. Procrastination, as
+both a modifiable and measurable behavioral tendency [@vaneerde2018], represents
+a promising target for low-cost, non-invasive interventions aimed at enhancing
+cognitive resilience in older populations. Tracking self-regulatory behavioral
+tendencies such as procrastination may offer an early warning system for
+emerging cognitive risk, opening avenues for preventative action before
+irreversible decline takes hold.
+
+## Limitations
+
+Despite the insights offered by this study, several limitations should be
+considered. Most notably, although our longitudinal Markov modelling approach
+captured changes in cognitive status, a key constraint was the measurement of
+procrastination at only a single time point (2020). As a result, we assumed
+temporal stability in procrastination scores across time and used this measure
+as both a retrospective and prospective proxy for procrastination scores.
+However, this precluded analysis of within-person changes in this behavior over
+time and whether such changes affect cognitive transitions. It should be noted
+that this limitation is inherent to the HRS dataset rather than a methodological
+oversight (Juster & Suzman, 1995). Future studies should prioritise repeated
+assessments of procrastination to determine whether increases in procrastination
+precedes, accompanies, or follows cognitive decline. Moreover, our use of a
+proxy measure of apathy may not have fully captured the multidimensional nature
+of apathy. Future research should incorporate more validated apathy assessments
+to strengthen the behavioral inferences drawn.
+
+Additionally, as noted earlier, the relatively small number of individuals in
+the MCI and dementia groups may have constrained statistical power for certain
+comparisons and increased the potential for misclassification bias. Finally,
+while the models adjusted for several key demographic and psychological
+covariates, unmeasured confounding (undiagnosed medical conditions, medication
+use, or sleep quality) could also have influenced cognitive outcomes. Future
+research should seek to replicate and extend these findings using more
+clinically diverse samples.
+
+## Conclusions
+
+In summary, this study offers preliminary evidence that procrastination may
+function as an early behavioral tendency marker of cognitive decline,
+particularly in older age. Individuals with MCI and dementia reported higher
+procrastination scores, and longitudinal modelling revealed that
+procrastination, especially when coupled with advancing age, was predictive of
+increased cognitive decline. These findings underscore the importance of
+everyday self-regulatory behaviors in dementia risk and resilience. As a
+modifiable and measurable construct, procrastination holds promise as a target
+for early detection and preventative intervention strategies aimed at sustaining
+cognitive health in ageing populations.
 

### 2025-10-03 manuscript/manuscript.qmd
Update :fire:

@@
     email: cormacmonaghan@protonmail.com
-    orcid: https://orcid.org/0000-0001-9012-3060
-    affiliation:
-      - Hamilton Institute, Maynooth University, Ireland
-      - Department of Psychology, Maynooth University, Ireland
+    orcid: 0000-0001-9012-3060
+    affiliation: Maynooth University
     attributes:
@@
     url: https://rafamoral.github.io/
-    orcid: https://orcid.org/0000-0002-0875-3563
-    affiliation: Department of Mathematics and Statistics, Maynooth University, Ireland
+    orcid: 0000-0002-0875-3563
+    affiliation: Maynooth University
   - name: Michelle Kelly
     url: https://www.ncirl.ie/About/A-Z-Staff-Directory/Staff/346
-    orcid: https://orcid.org/0000-0001-6159-6707
-    affiliation: Department of Psychology, National College of Ireland
+    orcid: 0000-0001-6159-6707
+    affiliation: National College of Ireland
   - name: Joanna McHugh Power
     url: https://www.maynoothuniversity.ie/people/joanna-mchugh-power
-    orcid: https://orcid.org/0000-0002-7387-3107
-    affiliation: Department of Psychology, Maynooth University, Ireland
+    orcid: 0000-0002-7387-3107
+    affiliation: Maynooth University
 correspondence-prefix: "Correspondence concerning this article should be addressed to"
@@
     wrap: 80
+keep-tex: true
 ---
@@
 library(ggplot2)
+library(tinytable)
 
@@
 mild cognitive impairment (MCI), of whom up to 80% progress to dementia within
-six years \[@cooper2015; @shigemizu2020 @tschanz2006}.
+six years [@cooper2015; @shigemizu2020 @tschanz2006].
 
@@
 
+\clearpage
+
 ## Data Analysis
@@
 #| tbl-cap-location: top
+#| tbl-pos: H
 #| include: true
@@
 tbl_descriptives <- tibble::tribble(
-  ~"", ~"Full sample <br>(n = 549)", ~"NC <br>(n = 452)", ~"MCI <br>(n = 86)", ~"Dementia <br>(n = 11)",
+  ~"", ~Full, ~NC, ~MCI, ~Dementia,
   "Age (years)", "69.70 ± 7.58", "69.70 ± 7.55",	"69.60 ± 7.55",	"70.00 ± 9.89",
-  "**Sex**", " ", " ", " ", " ",
-  "Male", "38.43% <br>(n = 211)", "39.82% <br>(n =180)", "32.56% <br>(n = 28)", "27.27% <br>(n = 3)",
-  "Female", "61.57% <br>(n = 338)", "60.18%<br>(n = 272)", "67.44% <br>(n = 58)",	"72.73 <br>(n = 8)",
-  "**Education**", " ", " ", " ", " ",
-  "No degree",	"16.24% <br>(n = 98)", "10.56% <br>(n = 47)",	"39.53% <br>(n = 34)",	"63.54% <br>(n = 7)",
-  "GED", "51.58% <br>(n = 279)",	"51.91% <br>(n = 231)",	"52.33% <br>(n = 45)",	"27.27 <br>(n = 3)",
-  "Further <br>education", "32.29% <br>(n = 175)", "37.53% <br>(n = 167)", "8.14% <br>(n = 7)",	"9.09% <br>(n = 1)",
+  "Male", "38.43% (n = 211)", "39.82% (n =180)", "32.56% (n = 28)", "27.27% (n = 3)",
+  "Female", "61.57% (n = 338)", "60.18% (n = 272)", "67.44% (n = 58)",	"72.73% (n = 8)",
+  "No degree",	"16.24% (n = 98)", "10.56% (n = 47)",	"39.53% (n = 34)",	"63.54% (n = 7)",
+  "GED", "51.58% (n = 279)",	"51.91% (n = 231)",	"52.33% (n = 45)",	"27.27% (n = 3)",
+  "Further education", "32.29% (n = 175)", "37.53% (n = 167)", "8.14% (n = 7)",	"9.09% (n = 1)",
   "Apathy",	"0.37 ± 0.63", "0.31 ± 0.59",	"0.65 ± 0.76",	"0.36 ± 0.51",
@@
 tbl_descriptives |>
-  setNames(c(" ", "Full sample <br>(n = 549)", "NC <br>(n = 452)", "MCI <br>(n = 86)", "Dementia <br>(n = 11)")) |>
-  tinytable::tt(notes = list(
+  setNames(c(" ", "Full sample (n = 549)", "NC (n = 452)", "MCI (n = 86)", "Dementia (n = 11)")) |>
+   tt(notes = list(
     caption, 
@@
   ) |>
-  tinytable::style_tt(j = 2:5, align = "c") |>
-  tinytable::format_tt("notes", markdown = TRUE) |>
-  tinytable::format_tt(j = 1, markdown = TRUE)
+   group_tt(i = list("**Sex**" = 2, "**Education**" = 4)) |>
+   style_tt(j = 2:5, align = "c") |>
+   style_tt(i = c(2, 5), align = "l") |>
+   format_tt(escape = TRUE) |>
+   format_tt("notes", markdown = TRUE) |>
+   format_tt(j = 1, markdown = TRUE) |>
+  style_tt(
+     bootstrap_class = "table table-sm"
+  )
+
 ```
@@
 #| fig-cap: "Frequency of cognitive status transitions between HRS waves."
+#| fig-pos: H
 #| include: true
@@
 #| fig-cap: "Odds ratios (95% Confidence Intervals) of transitioning from one cognitive state to another calculated using generalised logit models.
-<br>*Note*.
+*Note*.
 Odds ratios significantly different from 1 at a 5% significance rate are presented in blue (positive) or orange (negative),  otherwise they are presented in gray.
@@
 #| fig-cap: "Predicted transition probabilities by procrastination and age.
-<br>*Note.* Each curve represents a different age profile, ranging from the minimum (62 years) to the maximum (97 years) of the dataset.
+*Note.* Each curve represents a different age profile, ranging from the minimum (62 years) to the maximum (97 years) of the dataset.
 Specific ages of interest (70, 80, and 90) are highlighted."

### 2025-10-03 manuscript/manuscript.tex
Update :fire:

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{xcolor}
\usepackage[top=10pc,bottom=10pc,left=11pc,right=11pc,heightrounded]{geometry}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
  \setmainfont[Numbers=Proportional,Numbers=OldStyle]{Linux Libertine O}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% -----------------------
% CUSTOM PREAMBLE STUFF
% -----------------------

% -----------------
% Typography tweaks
% -----------------
% Sets the paragraph indent size. 1pc is a pica (12pt). 
% This controls how much the first line of a paragraph is indented.
\setlength{\parindent}{1pc}

% The nowidow package tries to prevent widows and orphans 
% (single lines of a paragraph stranded at top/bottom of a page). 
% defaultlines=2 requests at least two lines to remain on a page when possible.
\usepackage[all,defaultlines=2]{nowidow}

% enumitem gives fine control of lists.

  % \setlist[1]{labelindent=\parindent}: for first-level lists, 
  % set the label indent equal to the paragraph indent 
  % (keeps list labels aligned with normal text).
  % 
  % leftmargin=* on itemize/enumerate: tells enumitem to align the 
  % list body with the surrounding text (remove extra left offset).

\usepackage{enumitem}
\setlist[1]{labelindent=\parindent}
\setlist[itemize]{leftmargin=*}
\setlist[enumerate]{leftmargin=*}

 % description with style=unboxed makes description-list terms wrap 
% nicely (no boxed layout).
\setlist[description]{style=unboxed}

% Improves Table of Contents formatting (the titles option makes it 
% play nicely with section title formatting).
\usepackage[titles]{tocloft}

% Remove left margin in lists inside longtables
% https://tex.stackexchange.com/a/378190/11851
\AtBeginEnvironment{longtable}{\setlist[itemize]{nosep, wide=0pt, leftmargin=*, before=\vspace*{-\baselineskip}, after=\vspace*{-\baselineskip}}}

% Allow for /singlespacing and /doublespacing
\usepackage{setspace}

% -----------------
% Title block stuff
% -----------------

% Abstract
\usepackage[overload]{textcase}
\usepackage[runin]{abstract}
\renewcommand{\abstractnamefont}{\sffamily\footnotesize\bfseries\MakeUppercase}
\renewcommand{\abstracttextfont}{\sffamily\small}
\setlength{\absleftindent}{\parindent * 2}
\setlength{\absrightindent}{\parindent * 2}
\abslabeldelim{\quad}
\setlength{\abstitleskip}{-\parindent}

% Keywords
\newenvironment{keywords}
{\vskip -3em \hspace{\parindent}\small\sffamily{\sffamily\footnotesize\bfseries\MakeUppercase{Keywords}}\quad}
{\vskip 3em}

% Title
\usepackage{titling}
\setlength{\droptitle}{3em}
\pretitle{\par\vskip 5em \begin{flushleft}\LARGE\sffamily\bfseries}
\posttitle{\par\end{flushleft}\vskip 0.75em}

% Authors
%
% PHEW this is complicated for a number of reasons!
%
% When using \and with multiple authors, the article class in LaTeX wraps each 
% author block in a tabluar environment with a hardcoded center alignment. It's 
% possible to use \preauthor{} to start tabulars with a left alignment {l}, but 
% that only applies to the first author because the others all use \and with the 
% hardcoded {c}. But we can override the \and command and add our own {l}
%
% (See https://github.com/rstudio/rmarkdown/issues/1716#issuecomment-560601691 
% for an example of redefining \and to just be \\)
%
% That's all great, except tabulars have some amount of default horizontal 
% padding, which makes left-aligned author blocks not actuall get fully 
% left-aligned on the page. We can set the horizontal padding for the column to 
% 0, but it requires some wonky syntax: {@{\hspace{0em}}l@{}}
\renewcommand{\and}{\end{tabular} \hskip 3em \begin{tabular}[t]{@{\hspace{0em}}l@{}}}
\preauthor{\begin{flushleft}
           \lineskip 1.5em 
           \begin{tabular}[t]{@{\hspace{0em}}l@{}}}
\postauthor{\end{tabular}\par\end{flushleft}}

% Omit the date since the \published command does that
\predate{}
\postdate{}

% Command for a note at the top of the first page describing the publication
% status of the paper.
\newcommand{\published}[1]{%
   \gdef\puB{#1}}
   \newcommand{\puB}{}
   \renewcommand{\maketitlehooka}{%
       \par\noindent\footnotesize\sffamily \puB}

% ------------------
% Section headings
% ------------------
\usepackage{titlesec}
\titleformat*{\section}{\Large\sffamily\bfseries\raggedright}
\titleformat*{\subsection}{\large\sffamily\bfseries\raggedright}
\titleformat*{\subsubsection}{\normalsize\sffamily\bfseries\raggedright}
\titleformat*{\paragraph}{\small\sffamily\bfseries\raggedright}

% \titlespacing{<command>}{<left>}{<before-sep>}{<after-sep>}
% Starred version removes indentation in following paragraph
\titlespacing*{\section}{0em}{2em}{0.1em}
\titlespacing*{\subsection}{0em}{1.25em}{0.1em}
\titlespacing*{\subsubsection}{0em}{0.75em}{0em}

% -----------
% Footnotes
% -----------
% NB: footmisc has to come after setspace and biblatex because of conflicts
\usepackage[bottom, flushmargin]{footmisc}
\renewcommand*{\footnotelayout}{\footnotesize}

\addtolength{\skip\footins}{10pt}    % vertical space between rule and main text
\setlength{\footnotesep}{5pt}  % vertical space between footnotes

% ----------
% Captions
% ----------
\usepackage[font={small,sf}, labelfont={small,sf,bf}]{caption}

% --------
% Macros
% --------
% pandoc will not convert text within \begin{} XXX \end{} to Markdown and will
% treat it as regular TeX. Because of this, it's impossible to do stuff like
% this:

% \begin{landscape}
%
% | One | Two   |
% |-----+-------|
% | my  | table |
% | is  | nice  |
%
% \end{landscape}
%
% Since it'll render like: | One | Two | |—–+——-| | my | table | | is | nice |
% 
% BUT, from this http://stackoverflow.com/a/41945462/120898 we can get around
% this by creating new commands for \begin and \end, like this:
\usepackage{pdflscape}
\newcommand{\blandscape}{\begin{landscape}}
\newcommand{\elandscape}{\end{landscape}}

% \blandscape
%
% | One | Two   |
% |-----+-------|
% | my  | table |
% | is  | nice  |
%
% \elandscape

% Same thing, but for generic groups
% But can't use \bgroup and \egroup because those are built-in TeX things
\newcommand{\stgroup}{\begingroup}
\newcommand{\fingroup}{\endgroup}

% ---------------------------
% END CUSTOM PREAMBLE STUFF
% ---------------------------
\usepackage{float}
\usepackage{tabularray}
\usepackage[normalem]{ulem}
\usepackage{graphicx}
\usepackage{rotating}
\UseTblrLibrary{siunitx}
\NewTableCommand{\tinytableDefineColor}[3]{\definecolor{#1}{#2}{#3}}
\newcommand{\tinytableTabularrayUnderline}[1]{\underline{#1}}
\newcommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
\usepackage{mathtools}
\everydisplay\expandafter{\the\everydisplay\small }

\SetTblrStyle{foot}{font=\footnotesize}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Procrastination as a Marker of Cognitive Decline: Evidence from Longitudinal Transitions in the Older Adult Population},
  pdfauthor={Cormac Monaghan; Rafael de Andrade Moral; Michelle Kelly; Joanna McHugh Power},
  pdfkeywords={Dementia risk, Mild cognitive
impairment, Procrastination, Apathy, Executive dysfunction, Aging;},
  colorlinks=true,
  linkcolor={DarkSlateBlue},
  filecolor={Maroon},
  citecolor={DarkSlateBlue},
  urlcolor={DarkSlateBlue},
  pdfcreator={LaTeX via pandoc}}

% -----------------------
% END-OF-PREAMBLE STUFF
% -----------------------

% ---------------------- 
% Title block elements
% ---------------------- 
\usepackage{orcidlink}  % Create automatic ORCID icons/links

\title{Procrastination as a Marker of Cognitive Decline: Evidence from
Longitudinal Transitions in the Older Adult Population}

\author{
{\large Cormac Monaghan~\orcidlink{0000-0001-9012-3060}}%
\thanks{Corresponding author.} \\%
Maynooth University \\%
{\footnotesize \url{cormacmonaghan@protonmail.com}} \and
{\large Rafael de Andrade Moral~\orcidlink{0000-0002-0875-3563}}%
 \\%
Maynooth University \\%
{\footnotesize \url{}} \and
{\large Michelle Kelly~\orcidlink{0000-0001-6159-6707}}%
 \\%
National College of Ireland \\%
{\footnotesize \url{}} \and
{\large Joanna McHugh Power~\orcidlink{0000-0002-7387-3107}}%
 \\%
Maynooth University \\%
{\footnotesize \url{}} \and
}

\date{}

% Typeset URLs in the same font as their parent environment
%
% This has to come at the end of the preamble, after any biblatex stuff because 
% some biblatex styles (like APA) define their own \urlstyle{}
\usepackage{url}
\urlstyle{same}

% ---------------------------
% END END-OF-PREAMBLE STUFF
% ---------------------------
\begin{document}
% ---------------
% TITLE SECTION
% ---------------
\published{\textbf{} \\ {\scriptsize Access the code, data, and analysis
at \url{https://github.com/C-Monaghan/crisp-clam}}}

\maketitle

\begin{abstract}
Cognitive decline is a global health concern, making the identification
of early, modifiable risk factors essential. While apathy is a
recognized prodromal marker, procrastination may also signal early
executive dysfunction. We used longitudinal secondary data from the
United States Health and Retirement Study among adults aged 60+
\((n = 549; \bar{x} = 69.70;\sigma = 7.58)\). Cognitive function,
procrastination, and a proxy measure of apathy were assessed.
Transitions between normative cognitive function, mild cognitive
impairment (MCI), and dementia were modeled using a discrete-time
first-order Markov model. Procrastination scores were higher among
individuals with MCI or dementia than those with normative cognitive
function. Procrastination also interacted with age, disproportionately
increasing the risk of decline in the oldest participants.
Procrastination was associated with cognitive impairment and predicted
transitions to MCI, suggesting it may serve as both an early behavioral
marker and compounding risk factor.
\end{abstract}
\vskip 3em

\begin{keywords}
\def\sep{;\ }
Dementia risk\sep Mild cognitive
impairment\sep Procrastination\sep Apathy\sep Executive dysfunction\sep 
Aging;
\end{keywords}
% -------------------
% END TITLE SECTION
% -------------------

Dementia is characterized by progressive decline of cognitive function,
leading to memory loss and difficulties with daily living
(\citeproc{ref-prince2013}{Prince et al., 2013};
\citeproc{ref-sanz-blasco2022}{Sanz-Blasco et al., 2022}). The global
burden of dementia is substantial, with cases projected to rise to 152.8
million by 2050 (\citeproc{ref-nichols2022}{Nichols et al., 2022}).
Identifying and addressing modifiable risk factors is crucial to
mitigate this projected rise in prevalence. Two particularly vulnerable
groups are older adults and those with mild cognitive impairment (MCI),
of whom up to 80\% progress to dementia within six years Tschanz et al.
(\citeproc{ref-tschanz2006}{2006}).

The Lancet Commission on Dementia identified fourteen causal risk
factors including smoking, hypertension, and diabetes, which account for
approximately 45\% of dementia cases worldwide
(\citeproc{ref-livingston2024}{Livingston et al., 2024}). Prodromal
markers, on the other hand, reflect early disease processes and may
signal limited opportunities for prevention
(\citeproc{ref-teipel2025}{Teipel et al., 2025}). One such prodromal
marker is apathy, or the significant loss of motivation, as distinct
from depression and cognitive impairment
(\citeproc{ref-fresnais2023}{Fresnais et al., 2023};
\citeproc{ref-richard2012}{Richard et al., 2012}). Apathy is prevalent
in both MCI and dementia subgroups Salem et al.
(\citeproc{ref-salem2023}{2023}). Apathy is a significant predictor of
the transition from MCI to dementia van Dalen et al.
(\citeproc{ref-vandalen2018}{2018}). While apathy has been
well-established as a prodromal marker and risk factor for dementia
progression, emerging evidence suggests that procrastination may also
relate to early cognitive decline.

Though superficially similar, apathy and procrastination are
behaviorally distinct. Apathy reflects reduced internal drive and
emotional engagement, impairing initiation of action
(\citeproc{ref-fresnais2023}{Fresnais et al., 2023}), whereas
procrastination reflects intact intention but delayed execution
(\citeproc{ref-steel2007}{Steel, 2007}). While both reflect impairments
in executive functioning, procrastination may additionally capture
unique aspects of emotional and motivational dysregulation . For
instance, person A may feel no motivation to attend an exercise class
and therefore never books it (apathy). Person B, however, books the
class and intends to go, but postpones at the last minute
(procrastination). Both apathy and procrastination have been linked to
dysfunction in prefrontal regions implicated in dementia
(\citeproc{ref-fahed2021}{Fahed \& Steffens, 2021};
\citeproc{ref-friden2020}{Fridén, 2020};
\citeproc{ref-joseph2021}{Joseph et al., 2021};
\citeproc{ref-zhang2010}{Y. F. Zhang et al., 2010}).

These parallels raise the possibility that procrastination could serve
as an early marker of cognitive impairment or even a modifiable risk
factor. Chronic procrastination may exacerbate decline by limiting
engagement in cognitively stimulating activities such as physical
exercise, problem-solving, and goal-setting, which build cognitive
resilience and reduce dementia risk
(\citeproc{ref-chowdhary2022}{Chowdhary et al., 2022};
\citeproc{ref-kelly2021}{Kelly \& Walton, 2021};
\citeproc{ref-livingston2024}{Livingston et al., 2024};
\citeproc{ref-mohammadibytamar2020}{Mohammadi Bytamar et al., 2020}).
Reduced engagement may contribute to a cycle of cognitive disengagement,
accelerating decline.

Importantly, procrastination is responsive to intervention through
cognitive-behavioral strategies and self-regulation training
(\citeproc{ref-rozental2014}{Rozental \& Carlbring, 2014};
\citeproc{ref-vaneerde2018}{van Eerde \& Klingsieck, 2018}). Crucially,
procrastination may identify individuals with motivational or executive
dysfunction who do not yet meet the long-term symptom threshold for
apathy, offering a potentially valuable and earlier marker (and possible
target) for neurodegenerative risk.

Since age remains the strongest predictor of MCI and dementia, it is
important to consider how procrastination operates across the lifespan
(i.e.~interaction with age). Many established modifiable risk factors
such as hypertension, hearing loss, smoking, and social isolation, exert
age-dependent effects, with certain factors carrying more weight at
midlife than in late life (\citeproc{ref-livingston2020}{Livingston et
al., 2020}, \citeproc{ref-livingston2024}{2024}). Understanding how
procrastination interacts with age may help clarify its role in the
aetiology of cognitive decline and identify windows of opportunity for
targeted intervention.

To our knowledge, no studies have directly examined procrastination as a
predictor of cognitive decline or dementia progression. Accordingly, the
present study aimed to a) assess differences in procrastination levels
across three groups: individuals with dementia, individuals with MCI,
and individuals with neither dementia nor MCI, and b) test whether
higher procrastination scores predict transition from normative
cognition function to dementia or MCI to dementia.

\section{Method}\label{method}

\subsection{Data and study population}\label{data-and-study-population}

Analyses were conducted using a secondary data source; a multi-wave
prospective cohort study called the Health and Retirement Study (HRS;
(\citeproc{ref-fisher2018}{Fisher \& Ryan, 2018}), which tracks the
health, economic, and social well-being of over \(18,000\) American
adults primarily aged \(50+\). The HRS is managed by the Institute for
Social Research at the University of Michigan, with data collected every
two years. Initial data collection of a participant is conducted through
a face-to-face interview, with follow-up biennial interviews conducted
either by phone or face-to-face. The average retention rate ranges from
\(68.8\%\) to \(92.3\%\) (\citeproc{ref-HRS2017}{Health and Retirement
Study, 2017}). At the time of writing, fifteen years of HRS data are
currently archived.

We focused on four waves of HRS data from 2016 to 2022. Specifically,
our study sample consisted of respondents who participated in an
experimental module assessing procrastination during the 2020 wave.
These experimental modules, administered at the end of the core HRS
interview, consist of concise questionnaires designed to explore new
topics or supplement existing core survey data
(\citeproc{ref-juster1995}{Juster \& Suzman, 1995}). Each respondent
receives only one experimental module, with sample sizes for each module
constituting approximately 10\% of the core sample. As a result, while
the core HRS sample includes approximately \(18,000\) respondents, our
initial sample of interest consisted of \(1,368\) respondents. We
excluded respondents with missing cognitive assessment data for any wave
\((n = 419)\), those under 60 years of age, (as cognitive symptoms
typically occur around this age \(n = 398\)), and those with complete
missing values across the procrastination measure \((n = 2)\). This
resulted in a final analytic sample of \(549\) respondents.

\subsection{Measures}\label{measures}

\subsubsection{Outcome: Cognitive Function and Cognitive
Category}\label{outcome-cognitive-function-and-cognitive-category}

Cognitive function in the HRS is assessed using a series of tests
adapted from the Telephone Interview for Cognitive Status (TICS;
(\citeproc{ref-brandt1988}{Brandt et al., 1988};
\citeproc{ref-fong2009}{Fong et al., 2009}). These tests include an
immediate and delayed \(10\)-noun free recall test (to assess episodic
memory), a serial seven subtraction test (to assess working memory), and
a backward count from \(20\) test (to assess mental processing). Based
on these assessments, Crimmins et al.
(\citeproc{ref-crimmins2011}{2011}) developed both a \(27\)-point
cognitive scale and validated cut-off points to assess and classify
cognitive status. Using these points, respondents who scored \(12 - 27\)
were classified as having normal cognition, \(7 - 11\) as having MCI,
and \(0 - 6\) as having dementia.

\subsubsection{Predictor:
Procrastination}\label{predictor-procrastination}

Procrastination was measured using the Pure Procrastination Scale
(\citeproc{ref-steel2010}{Steel, 2010}), a psychometrically validated
scale for evaluating procrastination when conceptualized as a
dysfunctional delay. The PPS consists of 12 items rated on a Likert
scale ranging from \(1\) (strongly disagree) to \(5\) (strongly agree).
In responding to the scale items, participants were asked to reflect on
their general behavioural tendencies, with no specific time-frame
provided. Total procrastination scores range from \(12\) to \(60\), with
higher scores indicating greater tendency to procrastinate. In the HRS,
the PPS was administered only in wave 3 (2020) of the respective waves.
As such, we use the wave 3 measure as both a retrospective and
prospective proxy for procrastination scores across all waves in the
analysis. This approach assumes relative temporal stability in
procrastination over the study period. In this sample, the PSS had a
Cronbach's \(\alpha\) score of 0.92, indicating high internal
consistency. In responding to the scale items, participants were asked
to reflect on their general behavioural tendencies, with no specific
time-frame provided. An example of a question from the scale includes
``I delay making decisions until it's too late''.

\subsubsection{Covariate: Apathy}\label{covariate-apathy}

While no direct measure of apathy exists within the HRS, we utilised two
questions from the eight-item version of the Center for Epidemiological
Studies Depression (CES-D8) scale (\citeproc{ref-briggs2018}{Briggs et
al., 2018}) as proxies for apathy: ``You felt that everything you did
was an effort'' and ``You could not get going''. Both items capture core
features of apathy (behavioural and motivational disengagement) and,
while not a comprehensive measure of apathy, provide a valid and
pragmatic approximation for modelling purposes. Each item was measured
on a binary ``yes/no'' scale with total scores ranging from \$0\$ to
\$2\$.

\subsubsection{Confounders}\label{confounders}

To account for potential confounding, we controlled for demographic
variables with established associations with both cognitive function and
procrastination (\citeproc{ref-vanderweele2019}{VanderWeele, 2019}).
These included measures of age, sex, and educational attainment
(\citeproc{ref-abner2012}{Abner et al., 2012};
\citeproc{ref-freedman2024}{Freedman \& Cornman, 2024};
\citeproc{ref-yu2013}{Yu et al., 2013}). Educational attainment was
classified into three categories: no formal education, GED (General
Educational Development)/high school diploma, and college/further
education.

\clearpage

\subsection{Data Analysis}\label{data-analysis}

All data analysis was carried out in R (\citeproc{ref-rcoreteam2025}{R
Core Team, 2025}). To model transitions in cognitive states over time,
we employed a first order discrete-time Markov model, a class of
stochastic processes that satisfy the Markov property
(\citeproc{ref-zhang2010}{Y. F. Zhang et al., 2010}), which can be
formally expressed as:

\[P(X_{t+1} = j \vert X_t = i, X_{t-1} = i_{t-1}, \dots X_0 = i_0) = P(X_{t+1} = j \vert X_t = i)\]

This property asserts that the probability of transitioning from state
\$X\_t = i\$ to a future state \$X\_\{t+1\} = j\$ depends only on the
current state \$X\_t\$, and not on the full history of preceding states.
Here, there are three potential states: normative cognitive function,
mild cognitive impairment, and dementia. We consider dementia to be an
absorbing state, i.e.~once an individual reaches this state, they cannot
return to the other states in a future time point.

Unlike continuous-time models, discrete-time Markov models are not
readily supported by a dedicated R package for deriving transition
probabilities. Therefore, we implemented the model manually using
multinomial logistic regression from the \emph{nnet} package
(\citeproc{ref-venables2002}{Venables \& Ripley, 2002}). This approach
estimates the log-odds of transitioning to each non-reference state as a
linear function of covariates, relative to a chosen reference category.
For a system with \(K\) cognitive states (with state \(K\) as the
reference), the model takes the form:

\[log \frac{P(Y = j \; \vert \; x)}{P(Y = K \vert x)} = \beta_0 + \beta_j^\top x \qquad \text{for } j = 1, \dots K-1\]

From these equations, the predicted transition probabilities for the
non-reference states \(j = 1, \dots, K - 1\) are derived as:

\[P(Y = j \vert x) = \frac{e^{\beta_{0j} + \beta_j^\top x}}{1 + \sum^{K - 1}_{k = 1} e^{\beta_{0k} + \beta_k^ \top x}}\]

and for the reference state \(K\) as:

\[ P(Y = K \; \vert \; x) = \frac{1}{1 + \sum^{K - 1}_{k = 1} e^{\beta_{0k} + \beta_j^ \top x }} \]

To assess model fit and guide model selection for covariates in the
\(\beta_j\) vectors, we conducted likelihood-ratio tests comparing full
and reduced models, whose test statistic is defined as:

\[ D = 2 \times (\ell_{\text{reduced}} - \ell_{\text{full}}) \]

where \(\ell_{\text{full}}\) and \(\ell_{\text{reduced}}\) denote the
log-likelihoods of the respective models. Under the null hypothesis that
the full model does not represent an improved fit when compared to the
reduced model, \(D\) has an asymptotic \(\chi^2\) distribution with
degrees of freedom corresponding to the difference between the numbers
of parameters estimated by the two models.

\section{Results}\label{results}

Our final analytic sample comprised \(549\) respondents with the
following age distribution: 60 - 70 \((n = 186)\), 71 - 80
\((n = 203)\), 81 - 90 \((n = 142)\), and 90+ \((n = 18)\). Descriptive
statistics for the full sample, as well as data stratified by cognitive
status (normative cognitive function, MCI, and dementia), are presented
in Table~\ref{tbl-descriptives}. Both Figure~\ref{fig-river} and
supplementary figure S1 capture the unconditional transitions and
transition probabilities (respectively) between wave one and two (first
transition), wave two and three (second transition), and wave three and
four (third transition), yielding a total of \(1,647\) observed
transitions over time.

\begin{table}[H]

\caption{\label{tbl-descriptives}Baseline descriptives for the study
sample and stratified by cognitive status category.}

\centering{

\centering
\begin{talltblr}[         %% tabularray outer open
entry=none,label=none,
note{}={\emph{Note:} Descriptives for continuous and categorical variables are represented
using means ± standard deviations and percentages and frequencies respectively.
NC = Normative cognitive function; MCI = Mild cognitive impairment;
GED = General Educational Development.},
note{a}={Procrastination scores were collected in 2020 (Wave 3) and are presented
here for descriptive comparison, although they were not measured at
baseline.},
]                     %% tabularray outer close
{                     %% tabularray inner open
colspec={Q[]Q[]Q[]Q[]Q[]},
hline{2}={1-5}{solid, black, 0.05em},
hline{1}={1-5}{solid, black, 0.1em},
hline{12}={1-5}{solid, black, 0.1em},
cell{1-2,4-5,7-11}{2}={}{halign=c},
cell{1-2,4-5,7-11}{3}={}{halign=c},
cell{1-2,4-5,7-11}{4}={}{halign=c},
cell{1-2,4-5,7-11}{5}={}{halign=c},
cell{3,6}{1}={c=5}{halign=l},
cell{3,6}{2}={}{halign=l},
cell{3,6}{3}={}{halign=l},
cell{3,6}{4}={}{halign=l},
cell{3,6}{5}={}{halign=l},
}                     %% tabularray inner close
& Full sample (n = 549) & NC (n = 452) & MCI (n = 86) & Dementia (n = 11) \\
Age (years) & 69.70 ± 7.58 & 69.70 ± 7.55 & 69.60 ± 7.55 & 70.00 ± 9.89 \\
\textbf{Sex} & **Sex** & **Sex** & **Sex** & **Sex** \\
Male & 38.43\% (n = 211) & 39.82\% (n =180) & 32.56\% (n = 28) & 27.27\% (n = 3) \\
Female & 61.57\% (n = 338) & 60.18\% (n = 272) & 67.44\% (n = 58) & 72.73\% (n = 8) \\
\textbf{Education} & **Education** & **Education** & **Education** & **Education** \\
No degree & 16.24\% (n = 98) & 10.56\% (n = 47) & 39.53\% (n = 34) & 63.54\% (n = 7) \\
GED & 51.58\% (n = 279) & 51.91\% (n = 231) & 52.33\% (n = 45) & 27.27\% (n = 3) \\
Further education & 32.29\% (n = 175) & 37.53\% (n = 167) & 8.14\% (n = 7) & 9.09\% (n = 1) \\
Apathy & 0.37 ± 0.63 & 0.31 ± 0.59 & 0.65 ± 0.76 & 0.36 ± 0.51 \\
Procrastination\textsuperscript{a} & 28.60 ± 12.00 & 27.70 ± 11.30 & 32.10 ± 13.80 & 39.00 ± 16.00 \\
\end{talltblr}

}

\end{table}%

\begin{figure}[H]

\centering{

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{manuscript_files/figure-pdf/fig-river-1.pdf}

}

\caption{\label{fig-river}Frequency of cognitive status transitions
between HRS waves.}

\end{figure}%

\subsection{Cross sectional
differences}\label{cross-sectional-differences}

Initially, a Kruskal-Wallis test was conducted to examine differences in
procrastination scores (measured in 2020) across three cognitive status
groups after Levene's test indicated violation of homogeneity of
variance \((p = 0.039)\). The analysis revealed a statistically
significant effect of cognitive status,
\((\chi^2(2) = 17.54, p < .001)\), indicating that procrastination
levels differed significantly between at least two of the groups.
Post-hoc analysis using a pairwise Wilcoxon rank-sum test with a
Benjamini-Hochberg correction showed that participants with normal
cognition \((M = 27.7, SD = 11.7)\) reported significantly lower levels
of procrastination than those with both MCI
\((M = 32.1; SD = 11.3; p = 0.004)\) and dementia
\((M = 36.2; SD = 14.8; p = 0.005)\). No significance difference was
found between those with MCI and dementia \((p = 0.334)\).
Figure~\ref{fig-cross-sectional} displays the distribution of
procrastination scores across groups, with both boxplots and dotplots
showing median values and individual data points. Significance bars
indicate the pairwise differences described above.

\begin{figure}

\centering{

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{manuscript_files/figure-pdf/fig-cross-sectional-1.pdf}

}

\caption{\label{fig-cross-sectional}Group differences in 2020
procrastination scores according to cognitive status.}

\end{figure}%

\subsection{Markov analysis}\label{markov-analysis}

Results from the discrete-time Markov analysis, showed that all
covariates significantly influenced the likelihood of transitioning
between cognitive states (see Figure~\ref{fig-markov}). Notably,
procrastination interacted significantly with \emph{age} to affect two
key transitions: increasing the likelihood of transitioning from
normative cognitive function to MCI \((\text{OR} = 1.001; p < 0.001)\)
and decreasing the likelihood of reverting from MCI to normative
cognitive function \((\text{OR} = 0.999; p < 0.001)\). \emph{Women} were
significantly less likely than men to transition from both normative
cognitive function to dementia \((\text{OR} = 0.068; p < 0.001)\) and
from MCI to dementia \((\text{OR} = 0.70; p < 0.001\)).

For \emph{education attainment}, individuals with a GED were less likely
to transition from normative cognitive function to either MCI
\((\text{OR} = 0.49; p < 0.001)\) or dementia
\((\text{OR} = 0.29; p < 0.001)\) and from MCI to dementia
\((\text{OR} = 0.64; p < 0.001)\). They were also more likely to back
transition from MCI to normative cognitive function
\((\text{OR} = 2.18; p < 0.001)\). Those with a college level education
or higher demonstrated a significantly reduced likelihood of
transitioning from normative cognitive function to either MCI
\((\text{OR} = 0.32; p < 0.001)\) or dementia
\((\text{OR} = 0.07; p < 0.001)\) and from MCI to dementia
\((\text{OR} = 0.22; p < 0.001)\). They were also more likely to back
transition from MCI to normative cognitive function
\((\text{OR} = 3.13; p < 0.001)\).

Finally, higher levels of \emph{apathy} were associated with an
increased likelihood of transitioning from normative cognitive function
to both MCI \((\text{OR} = 1.35; p = 0.005)\) and dementia
\((\text{OR} = 1.36; p<0.001)\) and a decreased likelihood of
transitioning from MCI back to normative cognitive function
\((\text{OR} = 0.74; p = 0.003)\).

\begin{figure}

\centering{

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{manuscript_files/figure-pdf/fig-markov-1.pdf}

}

\caption{\label{fig-markov}Odds ratios (95\% Confidence Intervals) of
transitioning from one cognitive state to another calculated using
generalised logit models. \emph{Note}. Odds ratios significantly
different from 1 at a 5\% significance rate are presented in blue
(positive) or orange (negative), otherwise they are presented in gray.
NC = Normative cognitive function; MCI = Mild cognitive impairment.}

\end{figure}%

Figure~\ref{fig-age-preds} presents the predicted transition
probabilities across varying levels of age and procrastination. These
estimates illustrate how the interaction between age and procrastination
influences the likelihood of progressing between cognitive states over
time. Notably, while transition probabilities remain relatively stable
at very low levels of procrastination, substantial shifts emerge as both
age and procrastination increase. In particular, older individuals with
higher procrastination scores show an elevated probability of cognitive
decline transitioning from normative cognitive function to mild
cognitive impairment (MCI) and a reduced likelihood of transitioning
back from MCI to normative cognitive function, highlighting the
compounded risk posed by these two variables in later life.

\begin{figure}

\centering{

\includegraphics[width=1\linewidth,height=\textheight,keepaspectratio]{manuscript_files/figure-pdf/fig-age-preds-1.pdf}

}

\caption{\label{fig-age-preds}Predicted transition probabilities by
procrastination and age. \emph{Note.} Each curve represents a different
age profile, ranging from the minimum (62 years) to the maximum (97
years) of the dataset. Specific ages of interest (70, 80, and 90) are
highlighted.}

\end{figure}%

\section{Discussion}\label{discussion}

Dementia poses a growing global health burden, with modifiable risk
factors and prodromal markers offering important targets for
intervention (\citeproc{ref-livingston2020}{Livingston et al., 2020},
\citeproc{ref-livingston2024}{2024}). While apathy has been well
established as a prodromal marker and predictor of dementia
(\citeproc{ref-donovan2014}{Donovan et al., 2014};
\citeproc{ref-fresnais2023}{Fresnais et al., 2023};
\citeproc{ref-salem2023}{Salem et al., 2023};
\citeproc{ref-vandalen2018}{van Dalen et al., 2018}), emerging evidence
suggests that procrastination may also signal early cognitive
dysfunction (\citeproc{ref-friden2020}{Fridén, 2020};
\citeproc{ref-zhang2019}{S. Zhang et al., 2019}). This study examined
whether procrastination, particularly in later life, could serve as a
predictor of cognitive decline. We assessed both cross-sectional
differences in procrastination across three cognitive groups (normative
cognitive function, MCI and dementia) and longitudinal associations with
transitions in cognitive status, to clarify whether procrastination
functions as a novel and age-sensitive marker of emerging cognitive
impairment.

Our analysis revealed significant group differences in procrastination,
with individuals experiencing cognitive impairment, both MCI and
dementia, reporting higher procrastination scores than those with
normative cognitive function. These findings support the hypothesis that
procrastination may be associated with cognitive decline and aligns with
emerging evidence linking procrastination to cognitive dysfunction
(\citeproc{ref-friden2020}{Fridén, 2020}; \citeproc{ref-zhang2019}{S.
Zhang et al., 2019}). Interestingly, while both MCI and dementia groups
exhibited elevated procrastination scores, no significant difference
emerged between these two groups. This suggests that increases in
procrastination may occur relatively early in the neurodegenerative
process, potentially preceding or paralleling the emergence of more
overt cognitive symptoms.

However, it should be noted that the number of participants classified
as having MCI \((n = 67)\) or dementia \((n = 27)\) was relatively small
compared to those with normative cognitive function \((n = 455)\). This
imbalance in group sizes may have limited the sensitivity of our
comparisons and raises the possibility that subtle differences between
MCI and dementia groups were obscured by statistical power constraints.
Future research should aim to replicate these findings in more evenly
distributed samples to better determine whether procrastination
continues to increase across progressive stages of impairment or
plateaus following early decline.

Beyond these cross-sectional differences, our discrete-time Markov
analysis offered insight into how procrastination alongside other
demographic and psychological factors, influences the likelihood of
transitioning between cognitive states over time. Notably,
procrastination emerged as a dynamic predictor of cognitive change,
interacting with age to significantly increase the odds of transitioning
from normative cognitive function to MCI, while simultaneously
decreasing the odds of moving from MCI back to normative cognitive
function.

These findings imply that procrastination may serve dual roles, both as
a marker of early cognitive dysfunction and, potentially, as a
behavioral impediment to moving from MCI state to the normative
cognitive state. One plausible pathway involves the reinforcement of
maladaptive behaviors such as reduced cognitive engagement or reduced
participation in cognitively protective activities, such as physical
exercise, social interaction or medical adherence
(\citeproc{ref-hajek2025}{Hajek et al., 2025};
\citeproc{ref-kelly2021}{Kelly \& Walton, 2021};
\citeproc{ref-sirois2007}{Sirois, 2007}; \citeproc{ref-stead2010}{Stead
et al., 2010}). In parallel, procrastination has been associated with
chronic stress and elevated cortisol levels
(\citeproc{ref-sirois2023a}{Sirois, 2023}), which may accelerate
hippocampal atrophy, β-amyloid plaque deposition, and brain inflammation
(\citeproc{ref-franks2021}{Franks et al., 2021};
\citeproc{ref-wallensten2023}{Wallensten et al., 2023}), further
undermining cognitive resilience. These behavioral and biological
mechanisms align with self-regulation theories suggesting that
procrastination reflects executive dysfunction
(\citeproc{ref-rozental2014}{Rozental \& Carlbring, 2014};
\citeproc{ref-steel2007}{Steel, 2007}), an early feature of
neurodegenerative progression (\citeproc{ref-clark2012}{Clark et al.,
2012}).

The observed interaction with age (see Figure~\ref{fig-age-preds})
suggests that the influence of procrastination on cognitive trajectories
may grow stronger with advancing age, a period during which
neuro-plasticity diminishes, and behavioral risk factors exert greater
influence (\citeproc{ref-livingston2020}{Livingston et al., 2020},
\citeproc{ref-livingston2024}{2024}). This effect was particularly
pronounced among the oldest participants. As illustrated in
Figure~\ref{fig-age-preds}, the effect of procrastination is relatively
modest for younger-older adults (those age 70). However, the slope of
these transitions steepens considerably for individuals aged 80 and
especially for those age 90. These findings indicate that
procrastination is increasingly associated with a higher risk of
cognitive decline and a lower likelihood of improvement in cognitive
status in late old age.

This pattern underscores the possibility that procrastination functions
as a compounding risk factor in older adulthood, particularly among the
oldest individuals, by both reflecting emerging cognitive difficulties
and potentially accelerating their progression. In younger-old adults,
cognitive reserve and compensatory mechanisms
(\citeproc{ref-gooijers2024}{Gooijers et al., 2024}) may buffer against
the impact of poor self-regulation. However, as individuals reach more
advanced ages, these protective systems weaken
(\citeproc{ref-roberts2015}{Roberts et al., 2015}). Consequently,
maladaptive behavioral tendencies like procrastination can exert a
disproportionate toll on cognitive function. These findings underscore
the potential value of addressing behavioral regulation and
self-management in older adults as part of broader dementia
risk-reduction strategies, particularly for those of a much older age.

By revealing a significant interaction between age and procrastination,
this study highlights the importance of broadening dementia risk models
to incorporate dynamic lifestyle and psychological variables.
Procrastination, as both a modifiable and measurable behavioral tendency
(\citeproc{ref-vaneerde2018}{van Eerde \& Klingsieck, 2018}), represents
a promising target for low-cost, non-invasive interventions aimed at
enhancing cognitive resilience in older populations. Tracking
self-regulatory behavioral tendencies such as procrastination may offer
an early warning system for emerging cognitive risk, opening avenues for
preventative action before irreversible decline takes hold.

\subsection{Limitations}\label{limitations}

Despite the insights offered by this study, several limitations should
be considered. Most notably, although our longitudinal Markov modelling
approach captured changes in cognitive status, a key constraint was the
measurement of procrastination at only a single time point (2020). As a
result, we assumed temporal stability in procrastination scores across
time and used this measure as both a retrospective and prospective proxy
for procrastination scores. However, this precluded analysis of
within-person changes in this behavior over time and whether such
changes affect cognitive transitions. It should be noted that this
limitation is inherent to the HRS dataset rather than a methodological
oversight (Juster \& Suzman, 1995). Future studies should prioritise
repeated assessments of procrastination to determine whether increases
in procrastination precedes, accompanies, or follows cognitive decline.
Moreover, our use of a proxy measure of apathy may not have fully
captured the multidimensional nature of apathy. Future research should
incorporate more validated apathy assessments to strengthen the
behavioral inferences drawn.

Additionally, as noted earlier, the relatively small number of
individuals in the MCI and dementia groups may have constrained
statistical power for certain comparisons and increased the potential
for misclassification bias. Finally, while the models adjusted for
several key demographic and psychological covariates, unmeasured
confounding (undiagnosed medical conditions, medication use, or sleep
quality) could also have influenced cognitive outcomes. Future research
should seek to replicate and extend these findings using more clinically
diverse samples.

\subsection{Conclusions}\label{conclusions}

In summary, this study offers preliminary evidence that procrastination
may function as an early behavioral tendency marker of cognitive
decline, particularly in older age. Individuals with MCI and dementia
reported higher procrastination scores, and longitudinal modelling
revealed that procrastination, especially when coupled with advancing
age, was predictive of increased cognitive decline. These findings
underscore the importance of everyday self-regulatory behaviors in
dementia risk and resilience. As a modifiable and measurable construct,
procrastination holds promise as a target for early detection and
preventative intervention strategies aimed at sustaining cognitive
health in ageing populations.

\section*{References}\label{references}
\addcontentsline{toc}{section}{References}

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-abner2012}
Abner, E. L., Kryscio, R. J., Cooper, G. E., Fardo, D. W., Jicha, G. A.,
Mendiondo, M. S., Van Eldik, L. J., Wan, L., \& Schmitt, F. A. (2012).
Mild cognitive impairment: Statistical models of transition using
longitudinal clinical data. \emph{International Journal of Alzheimer's
Disease}, \emph{2012}(1), 291920.
\url{https://doi.org/10.1155/2012/291920}

\bibitem[\citeproctext]{ref-brandt1988}
Brandt, J., Spencer, M., \& Folstein, Marshal. (1988). The telephone
interview for cognitive status. \emph{Neuropsychiatry, Neuropsychology,
and Behavioral Neurology}, \emph{1}(2), 111--117.

\bibitem[\citeproctext]{ref-briggs2018}
Briggs, R., Carey, D., O'Halloran, A. M., Kenny, R. A., \& Kennelly, S.
P. (2018). Validation of the 8-item centre for epidemiological studies
depression scale in a cohort of community-dwelling older people: Data
from the irish longitudinal study on ageing ({TILDA}). \emph{European
Geriatric Medicine}, \emph{9}, 121--126.
\url{https://doi.org/10.1007/s41999-017-0016-0}

\bibitem[\citeproctext]{ref-chowdhary2022}
Chowdhary, N., Barbui, C., Anstey, K. J., Kivipelto, M., Barbera, M.,
Peters, R., Zheng, L., Kulmala, J., Stephen, R., Ferri, C. P., et al.
(2022). Reducing the risk of cognitive decline and dementia: {WHO}
recommendations. \emph{Frontiers in Neurology}, \emph{12}, 765584.
\url{https://doi.org/10.3389/fneur.2021.765584}

\bibitem[\citeproctext]{ref-clark2012}
Clark, L. R., Schiehser, D. M., Weissberger, G. H., Salmon, D. P.,
Delis, D. C., \& Bondi, M. W. (2012). Specific measures of executive
function predict cognitive decline in older adults. \emph{Journal of the
International Neuropsychological Society}, \emph{18}(1), 118--127.
\url{https://doi.org/10.1017/s1355617711001524}

\bibitem[\citeproctext]{ref-cooper2015}
Cooper, C., Sommerlad, A., Lyketsos, C. G., \& Livingston, G. (2015).
Modifiable predictors of dementia in mild cognitive impairment: A
systematic review and meta-analysis. \emph{American Journal of
Psychiatry}, \emph{172}(4), 323--334.
\url{https://doi.org/10.1176/appi.ajp.2014.14070878}

\bibitem[\citeproctext]{ref-crimmins2011}
Crimmins, E. M., Kim, J. K., Langa, K. M., \& Weir, D. R. (2011).
Assessment of cognition using surveys and neuropsychological assessment:
The health and retirement study and the aging, demographics, and memory
study. \emph{Journals of Gerontology Series B: Psychological Sciences
and Social Sciences}, \emph{66}, i162--i171.
\url{https://doi.org/10.1093/geronb/gbr048}

\bibitem[\citeproctext]{ref-donovan2014}
Donovan, N. J., Wadsworth, L. P., Lorius, N., Locascio, J. J., Rentz, D.
M., Johnson, K. A., Sperling, R. A., Marshall, G. A., Initiative, A. D.
N., et al. (2014). Regional cortical thinning predicts worsening apathy
and hallucinations across the {Alzheimer} disease spectrum. \emph{The
American Journal of Geriatric Psychiatry}, \emph{22}(11), 1168--1179.
\url{https://doi.org/10.1016/j.jagp.2013.03.006}

\bibitem[\citeproctext]{ref-fahed2021}
Fahed, M., \& Steffens, D. C. (2021). Apathy: Neurobiology, assessment
and treatment. \emph{Clinical Psychopharmacology and Neuroscience},
\emph{19}(2), 181. \url{https://doi.org/10.9758/cpn.2021.19.2.181}

\bibitem[\citeproctext]{ref-fisher2018}
Fisher, G. G., \& Ryan, L. H. (2018). Overview of the health and
retirement study and introduction to the special issue. \emph{Work,
Aging and Retirement}, \emph{4}(1), 1--9.
\url{https://doi.org/10.1093/workar/wax032}

\bibitem[\citeproctext]{ref-fong2009}
Fong, T. G., Fearing, M. A., Jones, R. N., Shi, P., Marcantonio, E. R.,
Rudolph, J. L., Yang, F. M., Kiely, D. K., \& Inouye, S. K. (2009).
Telephone interview for cognitive status: {Creating} a crosswalk with
the mini-mental state examination. \emph{Alzheimer's \& Dementia},
\emph{5}(6), 492--497. \url{https://doi.org/10.1016/j.jalz.2009.02.007}

\bibitem[\citeproctext]{ref-franks2021}
Franks, K. H., Bransby, L., Saling, M. M., \& Pase, M. P. (2021).
Association of stress with risk of dementia and mild cognitive
impairment: A systematic review and meta-analysis. \emph{Journal of
Alzheimer's Disease}, \emph{82}(4), 1573--1590.
\url{https://doi.org/10.3233/jad-210094}

\bibitem[\citeproctext]{ref-freedman2024}
Freedman, V. A., \& Cornman, J. C. (2024). Dementia prevalence,
incidence, and mortality trends among {US} adults ages 72 and older,
2011--2021. \emph{The Journals of Gerontology, Series A: Biological
Sciences and Medical Sciences}, \emph{79}, S22--s31.
\url{https://doi.org/10.1093/gerona/glae105}

\bibitem[\citeproctext]{ref-fresnais2023}
Fresnais, D., Humble, M. B., Bejerot, S., Meehan, A. D., \& Fure, B.
(2023). Apathy as a predictor for conversion from mild cognitive
impairment to dementia: A systematic review and meta-analysis of
longitudinal studies. \emph{Journal of Geriatric Psychiatry and
Neurology}, \emph{36}(1), 3--17.
\url{https://doi.org/10.1177/08919887221093361}

\bibitem[\citeproctext]{ref-friden2020}
Fridén, I. (2020). \emph{Procrastination as a form of {Self-regulation
Failure}: {A} review of the cognitive and neural underpinnings}.

\bibitem[\citeproctext]{ref-gooijers2024}
Gooijers, J., Pauwels, L., Hehl, M., Seer, C., Cuypers, K., \& Swinnen,
S. P. (2024). Aging, brain plasticity, and motor learning. \emph{Ageing
Research Reviews}, 102569.
\url{https://doi.org/10.1016/j.arr.2024.102569}

\bibitem[\citeproctext]{ref-hajek2025}
Hajek, A., Gyasi, R. M., Pengpid, S., Kostev, K., Soysal, P., Veronese,
N., Smith, L., Jacob, L., König, H.-H., \& Peltzer, K. (2025).
Associations of procrastination with loneliness, social isolation, and
social withdrawal. \emph{Journal of Public Health}, 1--9.
\url{https://doi.org/10.1007/s10389-025-02419-y}

\bibitem[\citeproctext]{ref-HRS2017}
Health and Retirement Study. (2017). \emph{Sample sizes and response
rates}.

\bibitem[\citeproctext]{ref-joseph2021}
Joseph, S., Knezevic, D., Zomorrodi, R., Blumberger, D. M., Daskalakis,
Z. J., Mulsant, B. H., Pollock, B. G., Voineskos, A., Wang, W., Rajji,
T. K., et al. (2021). Dorsolateral prefrontal cortex excitability
abnormalities in {Alzheimer}'s {Dementia}: {Findings} from transcranial
magnetic stimulation and electroencephalography study.
\emph{International Journal of Psychophysiology}, \emph{169}, 55--62.
\url{https://doi.org/10.1016/j.ijpsycho.2021.08.008}

\bibitem[\citeproctext]{ref-juster1995}
Juster, F. T., \& Suzman, R. (1995). An overview of the health and
retirement study. \emph{Journal of Human Resources}, S7--s56.
\url{https://doi.org/10.2307/146277}

\bibitem[\citeproctext]{ref-kelly2021}
Kelly, S. M., \& Walton, H. R. (2021). {``{I}'ll work out tomorrow''}:
The procrastination in exercise scale. \emph{Journal of Health
Psychology}, \emph{26}(13), 2613--2625.
\url{https://doi.org/10.1177/1359105320916541}

\bibitem[\citeproctext]{ref-livingston2024}
Livingston, G., Huntley, J., Liu, K. Y., Costafreda, S. G., Selbæk, G.,
Alladi, S., Ames, D., Banerjee, S., Burns, A., Brayne, C., Fox, N. C.,
Ferri, C. P., Gitlin, L. N., Howard, R., Kales, H. C., Kivimäki, M.,
Larson, E. B., Nakasujja, N., Rockwood, K., \ldots{} Mukadam, N. (2024).
Dementia prevention, intervention, and care: 2024 report of the {Lancet}
standing {Commission}. \emph{The Lancet}, \emph{404}(10452), 572--628.
\url{https://doi.org/10.1016/s0140-6736(24)01296-0}

\bibitem[\citeproctext]{ref-livingston2020}
Livingston, G., Huntley, J., Sommerlad, A., Ames, D., Ballard, C.,
Banerjee, S., Brayne, C., Burns, A., Cohen-Mansfield, J., Cooper, C., et
al. (2020). Dementia prevention, intervention, and care: 2020 report of
the {Lancet Commission}. \emph{The Lancet}, \emph{396}(10248), 413--446.
\url{https://doi.org/10.1016/s0140-6736(20)30367-6}

\bibitem[\citeproctext]{ref-mohammadibytamar2020}
Mohammadi Bytamar, J., Saed, O., \& Khakpoor, S. (2020). Emotion
regulation difficulties and academic procrastination. \emph{Frontiers in
Psychology}, \emph{11}, 524588.
\url{https://doi.org/10.3389/fpsyg.2020.524588}

\bibitem[\citeproctext]{ref-nichols2022}
Nichols, E., Steinmetz, J. D., Vollset, S. E., Fukutaki, K., Chalek, J.,
Abd-Allah, F., Abdoli, A., Abualhasan, A., Abu-Gharbieh, E., Akram, T.
T., et al. (2022). Estimation of the global prevalence of dementia in
2019 and forecasted prevalence in 2050: An analysis for the {Global
Burden} of {Disease Study} 2019. \emph{The Lancet Public Health},
\emph{7}(2), e105--e125.
\url{https://doi.org/10.1016/s2468-2667(21)00249-8}

\bibitem[\citeproctext]{ref-prince2013}
Prince, M., Bryce, R., Albanese, E., Wimo, A., Ribeiro, W., \& Ferri, C.
P. (2013). The global prevalence of dementia: {A} systematic review and
meta analysis. \emph{Alzheimer's \& Dementia}, \emph{9}(1), 63--75.
\url{https://doi.org/10.1016/j.jalz.2012.11.007}

\bibitem[\citeproctext]{ref-rcoreteam2025}
R Core Team. (2025). \emph{R: A language and environment for statistical
computing} {[}Manual{]}. R Foundation for Statistical Computing.

\bibitem[\citeproctext]{ref-richard2012}
Richard, E., Schmand, B., Eikelenboom, P., Yang, S., Ligthart, S., Moll
van Charante, E., Van Gool, W., \& Initiative, A. D. N. (2012). Symptoms
of apathy are associated with progression from mild cognitive impairment
to {Alzheimer}'s disease in non-depressed subjects. \emph{Dementia and
Geriatric Cognitive Disorders}, \emph{33}(2--3), 204--209.
\url{https://doi.org/10.1159/000338239}

\bibitem[\citeproctext]{ref-roberts2015}
Roberts, R. O., Cha, R. H., Mielke, M. M., Geda, Y. E., Boeve, B. F.,
Machulda, M. M., Knopman, D. S., \& Petersen, R. C. (2015). Risk and
protective factors for cognitive impairment in persons aged 85 years and
older. \emph{Neurology}, \emph{84}(18), 1854--1861.
\url{https://doi.org/10.1212/wnl.0000000000001537}

\bibitem[\citeproctext]{ref-rozental2014}
Rozental, Alexander., \& Carlbring, Per. (2014). Understanding and
treating procrastination: {A} review of a common self-regulatory
failure. \emph{Psychology (Savannah, Ga.)}, \emph{5}(13), 1488.
\url{https://doi.org/10.4236/psych.2014.513160}

\bibitem[\citeproctext]{ref-ruthirakuhan2019}
Ruthirakuhan, M., Herrmann, N., Vieira, D., Gallagher, D., \& Lanctôt,
K. L. (2019). The roles of apathy and depression in predicting
{Alzheimer} disease: A longitudinal analysis in older adults with mild
cognitive impairment. \emph{The American Journal of Geriatric
Psychiatry}, \emph{27}(8), 873--882.
\url{https://doi.org/10.1016/j.jagp.2019.02.003}

\bibitem[\citeproctext]{ref-salem2023}
Salem, H., Suchting, R., Gonzales, M. M., Seshadri, S., \& Teixeira, A.
L. (2023). Apathy as a predictor of conversion from mild cognitive
impairment to {Alzheimer}'s disease: A {Texas Alzheimer}'s research and
care consortium ({TARCC}) cohort-based analysis. \emph{Journal of
Alzheimer's Disease}, \emph{92}(1), 129--139.
\url{https://doi.org/10.3233/jad-220826}

\bibitem[\citeproctext]{ref-sanz-blasco2022}
Sanz-Blasco, R., Ruiz-Sánchez de León, J. M., Ávila-Villanueva, M.,
Valentı́-Soler, M., Gómez-Ramı́rez, J., \& Fernández-Blázquez, M. A.
(2022). Transition from mild cognitive impairment to normal cognition:
Determining the predictors of reversion with multi-state {Markov}
models. \emph{Alzheimer's \& Dementia}, \emph{18}(6), 1177--1185.
\url{https://doi.org/10.1002/alz.12448}

\bibitem[\citeproctext]{ref-shigemizu2020}
Shigemizu, D., Akiyama, S., Higaki, S., Sugimoto, T., Sakurai, T.,
Boroevich, K. A., Sharma, A., Tsunoda, T., Ochiya, T., Niida, S., et al.
(2020). Prognosis prediction model for conversion from mild cognitive
impairment to {Alzheimer}'s disease created by integrative analysis of
multi-omics data. \emph{Alzheimer's Research \& Therapy}, \emph{12},
1--12. \url{https://doi.org/10.1186/s13195-020-00716-0}

\bibitem[\citeproctext]{ref-sirois2007}
Sirois, F. M. (2007). {``{I}'ll look after my health, later''}: {A}
replication and extension of the procrastination--health model with
community-dwelling adults. \emph{Personality and Individual
Differences}, \emph{43}(1), 15--26.
\url{https://doi.org/10.1016/j.paid.2006.11.003}

\bibitem[\citeproctext]{ref-sirois2023a}
Sirois, F. M. (2023). Procrastination and {Stress}: {A Conceptual
Review} of {Why Context Matters}. \emph{International Journal of
Environmental Research and Public Health}, \emph{20}(6), 5031.
\url{https://doi.org/10.3390/ijerph20065031}

\bibitem[\citeproctext]{ref-stead2010}
Stead, R., Shanahan, M. J., \& Neufeld, R. W. (2010). {``{I}'ll go to
therapy, eventually''}: {Procrastination}, stress and mental health.
\emph{Personality and Individual Differences}, \emph{49}(3), 175--180.
\url{https://doi.org/10.1016/j.paid.2010.03.028}

\bibitem[\citeproctext]{ref-steel2007}
Steel, P. (2007). The nature of procrastination: A meta-analytic and
theoretical review of quintessential self-regulatory failure.
\emph{Psychological Bulletin}, \emph{3}(1).
\url{https://doi.org/10.1037/0033-2909.133.1.65}

\bibitem[\citeproctext]{ref-steel2010}
Steel, P. (2010). Arousal, avoidant and decisional procrastinators: {Do}
they exist? \emph{Personality and Individual Differences}, \emph{48}(8),
926--934. \url{https://doi.org/10.1016/j.paid.2010.02.025}

\bibitem[\citeproctext]{ref-teipel2025}
Teipel, S., Akmatov, M., Michalowsky, B., Riedel-Heller, S., Bohlken,
J., \& Holstiege, J. (2025). Timing of risk factors, prodromal features,
and comorbidities of dementia from a large health claims case--control
study. \emph{Alzheimer's Research \& Therapy}, \emph{17}(1), 22.

\bibitem[\citeproctext]{ref-tschanz2006}
Tschanz, J., Welsh-Bohmer, K., Lyketsos, C., Corcoran, C., Green, R. C.,
Hayden, K., Norton, M. C., Zandi, P., Toone, L., West, N., et al.
(2006). Conversion to dementia from mild cognitive disorder: The {Cache
County Study}. \emph{Neurology}, \emph{67}(2), 229--234.
\url{https://doi.org/10.1212/01.wnl.0000224748.48011.84}

\bibitem[\citeproctext]{ref-vandalen2018}
van Dalen, J. W., van Wanrooij, L. L., van Charante, E. P. M., Brayne,
C., van Gool, W. A., \& Richard, E. (2018). Association of apathy with
risk of incident dementia: A systematic review and meta-analysis.
\emph{JAMA Psychiatry}, \emph{75}(10), 1012--1021.
\url{https://doi.org/10.1001/jamapsychiatry.2018.1877}

\bibitem[\citeproctext]{ref-vaneerde2018}
van Eerde, W., \& Klingsieck, K. B. (2018). Overcoming procrastination?
{A} meta-analysis of intervention studies. \emph{Educational Research
Review}, \emph{25}, 73--85.
\url{https://doi.org/10.1016/j.edurev.2018.09.002}

\bibitem[\citeproctext]{ref-vanderweele2019}
VanderWeele, T. J. (2019). Principles of confounder selection.
\emph{European Journal of Epidemiology}, \emph{34}(3), 211--219.
\url{https://doi.org/10.1007/s10654-019-00494-6}

\bibitem[\citeproctext]{ref-venables2002}
Venables, W. N., \& Ripley, B. D. (2002). \emph{Modern applied
statistics with s-{PLUS}} (4th ed.). Springer.

\bibitem[\citeproctext]{ref-wallensten2023}
Wallensten, J., Ljunggren, G., Nager, A., Wachtler, C., Bogdanovic, N.,
Petrovic, P., \& Carlsson, A. C. (2023). Stress, depression, and risk of
dementia--a cohort study in the total population between 18 and 65 years
old in {Region Stockholm}. \emph{Alzheimer's Research \& Therapy},
\emph{15}(1), 161. \url{https://doi.org/10.1186/s13195-023-01308-4}

\bibitem[\citeproctext]{ref-yu2013}
Yu, H., Yang, S., Gao, J., Zhou, L., Liang, R., \& Qu, C. (2013).
Multi-state {Markov} model in outcome of mild cognitive impairments
among community elderly residents in {Mainland China}.
\emph{International Psychogeriatrics}, \emph{25}(5), 797--804.
\url{https://doi.org/10.1017/s1041610212002220}

\bibitem[\citeproctext]{ref-zhang2019}
Zhang, S., Liu, P., \& Feng, T. (2019). To do it now or later: {The}
cognitive mechanisms and neural substrates underlying procrastination.
\emph{Wiley Interdisciplinary Reviews: Cognitive Science}, \emph{10}(4),
e1492. \url{https://doi.org/10.1002/wcs.1492}

\bibitem[\citeproctext]{ref-zhang2010}
Zhang, Y. F., Zhang, Q. F., \& Yu, R. H. (2010). Markov property of
{Markov} chains and its test. \emph{2010 International Conference on
Machine Learning and Cybernetics}, \emph{4}, 1864--1867.
\url{https://doi.org/10.1109/icmlc.2010.5580952}

\end{CSLReferences}

\end{document}

### 2025-10-13 manuscript/manuscript.qmd
Updating manuscript details

@@
 title: "Procrastination as a Marker of Cognitive Decline: Evidence from Longitudinal Transitions in the Older Adult Population"
+date: "2025-09-25"
 author: 
@@
   - Executive dysfunction
-  - Aging;
+  - Aging
+published: "This manuscript is a preprint and is currently under review at *Alzheimer's & Dementia: Diagnosis, Assessment & Disease Monitoring*"
 code-repo: "Access the code, data, and analysis at <https://github.com/C-Monaghan/crisp-clam>"
@@
 
+\clearpage
+
 # Discussion

### 2025-11-07 manuscript/manuscript.qmd
Using new shifu template

@@
 
-This property asserts that the probability of transitioning from state \$X_t =
-i\$ to a future state \$X\_{t+1} = j\$ depends only on the current state
-\$X_t\$, and not on the full history of preceding states. Here, there are three
-potential states: normative cognitive function, mild cognitive impairment, and
-dementia. We consider dementia to be an absorbing state, i.e. once an individual
-reaches this state, they cannot return to the other states in a future time
-point.
+This property asserts that the probability of transitioning from state $X_t = i$
+to a future state $X\_{t+1} = j$ depends only on the current state $X_t$, and
+not on the full history of preceding states. Here, there are three potential
+states: normative cognitive function, mild cognitive impairment, and dementia.
+We consider dementia to be an absorbing state, i.e. once an individual reaches
+this state, they cannot return to the other states in a future time point.
 
@@
 tbl_descriptives <- tibble::tribble(
-  ~"", ~Full, ~NC, ~MCI, ~Dementia,
-  "Age (years)", "69.70 ± 7.58", "69.70 ± 7.55",	"69.60 ± 7.55",	"70.00 ± 9.89",
-  "Male", "38.43% (n = 211)", "39.82% (n =180)", "32.56% (n = 28)", "27.27% (n = 3)",
-  "Female", "61.57% (n = 338)", "60.18% (n = 272)", "67.44% (n = 58)",	"72.73% (n = 8)",
-  "No degree",	"16.24% (n = 98)", "10.56% (n = 47)",	"39.53% (n = 34)",	"63.54% (n = 7)",
-  "GED", "51.58% (n = 279)",	"51.91% (n = 231)",	"52.33% (n = 45)",	"27.27% (n = 3)",
-  "Further education", "32.29% (n = 175)", "37.53% (n = 167)", "8.14% (n = 7)",	"9.09% (n = 1)",
-  "Apathy",	"0.37 ± 0.63", "0.31 ± 0.59",	"0.65 ± 0.76",	"0.36 ± 0.51",
-  "Procrastination",	"28.60 ± 12.00", "27.70 ± 11.30",	"32.10 ± 13.80", "39.00 ± 16.00"
+  ~""                 , ~Full              , ~NC                , ~MCI              , ~Dementia        ,
+  "Age (years)"       , "69.70 ± 7.58"    , "69.70 ± 7.55"    , "69.60 ± 7.55"   , "70.00 ± 9.89"  ,
+  "Male"              , "38.43% (n = 211)" , "39.82% (n =180)"  , "32.56% (n = 28)" , "27.27% (n = 3)" ,
+  "Female"            , "61.57% (n = 338)" , "60.18% (n = 272)" , "67.44% (n = 58)" , "72.73% (n = 8)" ,
+  "No degree"         , "16.24% (n = 98)"  , "10.56% (n = 47)"  , "39.53% (n = 34)" , "63.54% (n = 7)" ,
+  "GED"               , "51.58% (n = 279)" , "51.91% (n = 231)" , "52.33% (n = 45)" , "27.27% (n = 3)" ,
+  "Further education" , "32.29% (n = 175)" , "37.53% (n = 167)" , "8.14% (n = 7)"   , "9.09% (n = 1)"  ,
+  "Apathy"            , "0.37 ± 0.63"     , "0.31 ± 0.59"     , "0.65 ± 0.76"    , "0.36 ± 0.51"   ,
+  "Procrastination"   , "28.60 ± 12.00"   , "27.70 ± 11.30"   , "32.10 ± 13.80"  , "39.00 ± 16.00"
 )
@@
 caption <- glue::glue(
-  "*Note:* Descriptives for continuous and categorical variables are represented 
+  "Note: Descriptives for continuous and categorical variables are represented
   using means ± standard deviations and percentages and frequencies respectively.
@@
   ) |>
-   group_tt(i = list("**Sex**" = 2, "**Education**" = 4)) |>
+   group_tt(i = list("Sex" = 2, "Education" = 4)) |>
    style_tt(j = 2:5, align = "c") |>
-   style_tt(i = c(2, 5), align = "l") |>
+   style_tt(i = c(2, 5), align = "l", bold = TRUE) |>
    format_tt(escape = TRUE) |>
    format_tt("notes", markdown = TRUE) |>
-   format_tt(j = 1, markdown = TRUE) |>
    style_tt(
@@
      )
-
 ```
@@
 ```
-
-\clearpage
 
@@
 
+\newpage
+
+::: {.content-visible when-format="html"}
 # References
+:::

### 2025-11-09 manuscript/manuscript.qmd
Shifu update

@@
     orcid: 0000-0001-9012-3060
-    affiliation: Maynooth University
+    affiliation:
+      - id: 1
+        name: Maynooth University
     attributes:
@@
     orcid: 0000-0002-0875-3563
-    affiliation: Maynooth University
+    affiliation:
+      - id: 1
+        name: Maynooth University
   - name: Michelle Kelly
@@
     orcid: 0000-0001-6159-6707
-    affiliation: National College of Ireland
+    affiliation:
+      id: 2
+      name: National College of Ireland
   - name: Joanna McHugh Power
@@
     orcid: 0000-0002-7387-3107
-    affiliation: Maynooth University
-correspondence-prefix: "Correspondence concerning this article should be addressed to"
+    affiliation:
+      id: 1
+      name: Maynooth University
+correspondence-prefix: "Corresponding author"
 abstract: |
@@
 csl: csl/apa.csl
+block-author: false
 link-citations: true

### 2025-11-25 analysis/data-analysis.qmd
Finalising qmd files

@@
 ---
-title: "Analysis Plan"
+title: "Analysis plan"
 format:
@@
     toc: false
+bibliography: manuscript/references.bib
+csl: manuscript/csl/apa.csl
+editor:
+  markdown:
+    wrap: 80
 ---
@@
 
-To examine whether procrastination scores differed by cognitive status, we employed the **Kruskal–Wallis test**, a non-parametric alternative to one-way ANOVA.
-This test is appropriate when the assumption of homogeneity of variance is violated.
-The Kruskal–Wallis test evaluates whether the distribution of ranks differs across groups and is formally expressed as:
+To examine whether procrastination scores differed by cognitive status, we
+employed the **Kruskal–Wallis test**, a non-parametric alternative to one-way
+ANOVA. This test is appropriate when the assumption of homogeneity of variance
+is violated. The Kruskal–Wallis test evaluates whether the distribution of ranks
+differs across groups and is formally expressed as:
 
@@
 
-where $k$ denotes the number of groups, $n_i$ the number of participants in group $i$, $n$ the total sample size, $\bar{R}$ the mean rank sum of each group, $E_R$ the expected value of the rankings and $\sigma^2$ the rank variance.
-Under the null hypothesis that all groups have the same distribution, $H$ approximately follows a $\chi^2$ distribution with $k-1$ degrees of freedom.
+where $k$ denotes the number of groups, $n_i$ the number of participants in
+group $i$, $n$ the total sample size, $\bar{R}$ the mean rank sum of each group,
+$E_R$ the expected value of the rankings and $\sigma^2$ the rank variance. Under
+the null hypothesis that all groups have the same distribution, $H$
+approximately follows a $\chi^2$ distribution with $k-1$ degrees of freedom.
 
@@
 
-To model transitions in cognitive states over time, we employed a **first order discrete-time Markov model**, a class of stochastic processes that satisfy the Markov property, which can be formally expressed as:
+To model transitions in cognitive states over time, we employed a **first order
+discrete-time Markov model**, a class of stochastic processes that satisfy the
+Markov property, which can be formally expressed as:
 
@@
 
-This property asserts that the probability of transitioning from state $X_t = i$ to a future state $X_{t+1} = j$ depends only on the current state $X_t$, and not on the full history of preceding states.
+This property asserts that the probability of transitioning from state $X_t = i$
+to a future state $X_{t+1} = j$ depends only on the current state $X_t$, and not
+on the full history of preceding states.
 
-Unlike continuous-time models, discrete-time Markov models are not readily supported by a dedicated R package for deriving transition probabilities.
-Therefore, we implemented the model manually using multinomial logistic regression from the *nnet* package.
-This approach estimates the log-odds of transitioning to each non-reference state as a linear function of covariates, relative to a chosen reference state.
-For a system with $K$ cognitive states (with state $K$ as the reference), the model takes the form:
+Unlike continuous-time models, discrete-time Markov models are not readily
+supported by a dedicated R package for deriving transition probabilities.
+Therefore, we implemented the model manually using multinomial logistic
+regression from the `nnet` package [@venables2002]. This approach estimates the log-odds of
+transitioning to each non-reference state as a linear function of covariates,
+relative to a chosen reference state. For a system with $K$ cognitive states
+(with state $K$ as the reference), the model takes the form:
 
@@
 
-Additional, to assess model fit and guide model selection for the covariates in the $\beta_j$ vectors, we conducted likelihood-ratio tests comparing full and reduced models. Such tests are defined as:
+Additional, to assess model fit and guide model selection for the covariates in
+the $\beta_j$ vectors, we conducted likelihood-ratio tests comparing full and
+reduced models. Such tests are defined as:
 
@@
 
-where $\ell_{full}$ and $\ell_{reduced}$ denote the log-likelihoods of the respective models.
-Under the null hypothesis that the full model does not represent an improved fit when compared to the reduced model, $D$ has an asymptotic $\chi^2$ distribution with degrees of freedom corresponding to the difference between the numbers of parameters estimated by the two models.
+where $\ell_{full}$ and $\ell_{reduced}$ denote the log-likelihoods of the
+respective models. Under the null hypothesis that the full model does not
+represent an improved fit when compared to the reduced model, $D$ has an
+asymptotic $\chi^2$ distribution with degrees of freedom corresponding to the
+difference between the numbers of parameters estimated by the two models.

### 2025-11-25 analysis/discrete-markov.qmd
Finalising qmd files

@@
 ---
-title: "Discrete Time Markov Model"
+title: "Markov Model"
 ---
@@
 #| code-summary: "Check out my code"
-
-rm(list = ls())
 
@@
 
-theme_set(
-  theme_minimal() +
+theme_clean <- function(){
+  theme_bw() +
   theme(
@@
     strip.text = element_text(size = 10, colour = colour, face = "bold"),
-      legend.title = element_text(hjust = 0.5, colour = colour, face = "bold"),
-      ))
+    legend.title = element_text(hjust = 0.5, colour = colour, face = "bold")
+  )
+}
 
@@
   paste0("apathy_", seq(2016, 2022, by = 2)),
+  paste0("mean_anxiety_", seq(16, 22, by = 2)),
   "Total_p")
@@
 
-Following this, we can now use the `extract_years()` function to extract the cognitive function data for the years 2016, 2018, and 2020.
+Following this, we can now use the `extract_years()` function to extract the cognitive function data for the years 2016, 2018, 2020, and 2022.
 
@@
   # Make age time varying
-  mutate(Age = Age - (2022 - as.numeric(as.character(wave))))
+    mutate(Age = Age - (2022 - as.numeric(as.character(wave)))) |>
+    ungroup()
 
@@
 # Modelling
-## Markov Process Fundamentals
-
-Discrete-time Markov models belong to a class of stochastic processes that satisfy the **Markov property**, which can be formally expressed as:
-
-$$
-P(X_{t+1} = j \vert X_t = i, X_{t-1} = i_{t-1}, \dots X_0 = i_0) = P(X_{t+1} = j \vert X_t = i)
-$$ {#eq-markov}
-
-This property establishes that the future state $X_{t+1}$ depends only on the current state $X_t$, not on the entire history of states.
-
-### Multinomial Logistic Regression Formulation
-
-We model the transition probabilities using **multinomial logistic regression**, where the log-odds of each transition relative to a reference state are linear functions of covariates.
-
-For a system with $K$ states (using state $K$ as reference), we have:
-
-$$
-log \left( \frac{P(Y = j \vert x)}{P(Y = k \vert x)} \right) = \beta_0 + \beta_j^Tx \qquad \text{for } j = 1, \dots K-1
-$$ {#eq-multinomial}
-
-For non-reference states $j = 1, \dots, K - 1$
-
-$$
-P(Y = j \vert x) = \frac{e^{\beta_{0j} + \beta_j^Tx}}{1 + \sum^{k - 1}_{k = 1} e^{\beta_{0k} + \beta_k^Tx}}
-$$ {#eq-non-reference}
-
-For the reference state $K$:
-
-$$
-P(Y = k \vert x) = \frac{1}{1 + \sum^{k - 1}_{k = 1} e^{\beta_{0k} + \beta_j^Tx }}
-$$ {#eq-reference}
-
-### Absorbing State Specification
-
-Our primary analysis assumes the transition matrix remains fixed:
-
-$$
-P^{(t)} \equiv P \qquad \forall \; t
-$$ {#eq-assumption}
-
-However, we model Dementia as an absorbing state:
-
-$$
-p_{3j} = \begin{cases} 
-1 \qquad \text{if } j = 3 \\
-0 \qquad \text{otherwise} \end{cases}
-$$ {#eq-dementia}
-
-yielding the constrained transition matrix:
-
-$$
-P = \begin{bmatrix} 
-p_{11} & p_{12} & p_{13} \\
-p_{21} & p_{22} & p_{23} \\
-0      & 0      & 1      \\
-\end{bmatrix}
-$$ {#eq-t-matrix-constrained}
-
-## Our model
 
@@
 fit_a <- nnet::multinom(
-  status ~ Gender + Education_tri + Depression + (Total_p * Age) + status_prev, 
+  status ~ Gender + Education_tri + Apathy + (Total_p * Age) + status_prev,
   family = multinomial,
@@
 fit_b <- nnet::multinom(
-  status ~ Gender + Education_tri + Depression + (Total_p * Age) + status_prev, 
+  status ~ Gender + Education_tri + Apathy + (Total_p * Age) + status_prev,
   family = multinomial,
@@
 
-Below we present the estimated parameters for the interaction model (@tbl-interaction-output). We will transform the estimates to odds ratios and colour them based on significance.
+Below we present the estimated parameters for our model (@tbl-interaction-output). We will transform the estimates to odds ratios and colour them based on significance.
 
@@
                  "High school degree vs. No education",
-                 "Further education vs. No education", "Depression",
+                 "Further education vs. No education",
+                 "Apathy Scores", "Anxiety Scores",
                  "Procrastination (2020)",
@@
 #| code-summary: "Check out my code"
-#| fig-cap: "Estimated odds ratios for the interaction model."
+#| fig-cap: "Estimated odds ratios for the Markov model."
 #| message: false
@@
   ggh4x::facet_manual(
-    ~ term, scales = "free_x", 
-    design = design, labeller = labeller(
-     term = c("Depression" = "Apathy")
-    )) +
+    ~ term, scales = "free_x", design = design) +
   coord_cartesian(clip = "off") +
-  theme_bw() +
-  theme(
-    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
-    axis.title = element_text(size = 12, face = "bold"),
-    strip.text = element_text(size = 10, face = "bold"),
-    panel.grid.major.y = element_blank(),
-    panel.grid.minor.y = element_blank())
+  theme_clean() +
+  theme(panel.grid = element_blank())
 
@@
 #| code-summary: "Check out my code"
-#| fig-cap: "Predicted transition probabilities from the interaction model."
+#| fig-cap: "Predicted transition probabilities from the Markov model."
 #| warning: false
@@
   Education_tri = factor(0),
-  Depression = mean(data_stack$Depression, na.rm = TRUE),
+  Apathy = mean(data_stack$Apathy, na.rm = TRUE),
   status_prev = levels(data_stack$status_prev),
@@
   labs(
-    title = "Predicted transition probabilities from the interaction model",
     x = "Total Procrastination", y = "Age",
@@
   facet_grid(status ~ status_prev) +
-  theme_classic() +
-  theme(
-    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
-    axis.title = element_text(size = 12, face = "bold"),
-    strip.text = element_text(size = 10, face = "bold"),
-    panel.grid.major.y = element_blank(),
-    panel.grid.minor.y = element_blank(),
-    legend.title = element_text(size = 10, face = "bold"))
+  theme_clean() +
+  theme(panel.grid = element_blank())
 
@@
   )) +
-  theme_bw() +
+  theme_clean() +
   theme(
-    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
-    axis.title = element_text(size = 12, face = "bold"),
-    strip.text = element_text(size = 10, face = "bold"),
-    panel.grid.major.y = element_blank(),
-    panel.grid.minor.y = element_blank(),
+    panel.grid = element_blank()
     legend.title = element_text(size = 10, face = "bold"),

### 2025-11-25 analysis/hypothesis.qmd
Finalising qmd files

@@
 ---
+
+This study was **not pre-registered** on the [Open Science Framework](https://osf.io/). However, we had two main hypotheses.
 
@@
 
-In order to test our hypotheses we will be using both a **non-parametric ANOVA** and a post-hoc test to compare procrastination scores between cognitive groups, and a **discrete time Markov Model** to assess the association between procrastination and cognitive decline transitions.
+In order to test our hypotheses we will be using both a **non-parametric ANOVA** and a post-hoc test to compare procrastination scores between cognitive groups, and a **discrete time Markov Model** to assess the association between procrastination and cognitive decline transitions (controlling for known covariates).

### 2025-11-25 analysis/kruskal-wallis.qmd
Finalising qmd files

@@
     toc: false
+editor:
+  markdown:
+    wrap: 80
 ---
@@
 
-To examine whether procrastination scores differed across cognitive status groups in 2020, we first tested the assumption of homogeneity of variances using Levene’s test, which was violated ($p = 0.039$).
-Consequently, we employed the **Kruskal–Wallis test**, a rank-based non-parametric alternative to one-way ANOVA, to evaluate whether procrastination scores differed significantly among participants with normative cognitive function, mild cognitive impairment (MCI), and dementia.
+To examine whether procrastination scores differed across cognitive status
+groups in 2020, we first tested the assumption of homogeneity of variances using
+Levene’s test, which was violated ($p = 0.039$). Consequently, we employed the
+**Kruskal–Wallis test**, a rank-based non-parametric alternative to one-way
+ANOVA, to evaluate whether procrastination scores differed significantly among
+participants with normative cognitive function, mild cognitive impairment (MCI),
+and dementia.
 
@@
 ```
-The test revealed a significant effect of cognitive status on procrastination scores ($\chi^2(2) = 17.54, \; p < 0.001$), suggesting that at least two groups differ.
+
+The test revealed a significant effect of cognitive status on procrastination
+scores ($\chi^2(2) = 17.54, \; p < 0.001$), suggesting that at least two groups
+differ.
 
@@
 
-To identify where these differences lay, we conducted **pairwise Wilcoxon rank-sum tests with Benjamini–Hochberg (BH) correction** for multiple comparisons:
+To identify where these differences lay, we conducted **pairwise Wilcoxon
+rank-sum tests with Benjamini–Hochberg (BH) correction** for multiple
+comparisons:
 
@@
 
-- Normative cognitive function participants reported significantly lower procrastination scores than those with MCI ($p = 0.004$).
-- Normative cognitive function participants also reported significantly lower scores than those with dementia ($p = 0.005$).
-- No significant difference was found between the MCI and dementia groups ($p = 0.334$).
+-   Normative cognitive function participants reported significantly lower
+    procrastination scores than those with MCI ($p = 0.004$).
+-   Normative cognitive function participants also reported significantly lower
+    scores than those with dementia ($p = 0.005$).
+-   No significant difference was found between the MCI and dementia groups
+    ($p = 0.334$).
 
@@
 
-Figure fig-kruskal illustrates the distribution of procrastination scores across cognitive status groups.
-Individual scores are displayed as dotplots (left side of each box), and boxplots summarize the group distributions (right side).
-Significance bars represent the results of the pairwise comparisons described above.
+Figure fig-kruskal illustrates the distribution of procrastination scores across
+cognitive status groups. Individual scores are displayed as dotplots (left side
+of each box), and boxplots summarize the group distributions (right side).
+Significance bars represent the results of the pairwise comparisons described
+above.
 

### 2025-11-25 R/pivot-and-factorise.R
Modify helper functions :ambulance:

@@ -6,41 +6,54 @@ pivot_and_factorise <- function(data, time_vary = FALSE) {
   #   - data: Dataset containing cognitive function variables in wide format
   # Returns:
   #   - Long-format dataset with factorized variables, ordered by ID and wave
-  # 
-  
-  if(time_vary == TRUE) {
-    # Pivoting depression variable
-    dep_scores <- data |>
+  #
+
+  if (time_vary == TRUE) {
+    # Pivoting apathy variable
+    apathy_scores <- data |>
       select(ID, starts_with("apathy")) |>
       pivot_longer(
         cols = !ID,
         names_to = "remove",
-        values_to = "Depression") |>
-      select(Depression)
-    
+        values_to = "Apathy"
+      ) |>
+      select(Apathy)
+
+    # Pivot anxiety
+    anxiety_scores <- data |>
+      select(ID, starts_with("mean_anxiety")) |>
+      pivot_longer(
+        cols = !ID,
+        names_to = "remove",
+        values_to = "Anxiety"
+      ) |>
+      select(Anxiety)
+
     # Pivoting whole dataset
     data <- data |>
-      select(!starts_with("apathy")) |>
+      select(!starts_with("apathy") & !starts_with("mean_anxiety")) |>
       pivot_longer(
         cols = starts_with("cogfunction"),
         names_to = "wave",
         names_prefix = "cogfunction",
-        values_to = "status")
-    
+        values_to = "status"
+      )
+
     # Adding in depression scores
-    data <- data |> cbind(dep_scores)
-  } else{
+    data <- data |> cbind(apathy_scores) |> cbind(anxiety_scores)
+  } else {
     # No pivoting of depression
     data <- data |>
       tidyr::pivot_longer(
-        cols = starts_with("cogfunction"), 
+        cols = starts_with("cogfunction"),
         names_to = "wave",
-        names_prefix = "cogfunction", 
-        values_to = "status") |>
+        names_prefix = "cogfunction",
+        values_to = "status"
+      ) |>
       rename(Depression = Total_dep_2016) |>
       select(!starts_with("Total_dep_"))
   }
-  
+
   # Factorizing
   data <- data |>
     mutate(
@@ -50,10 +63,11 @@ pivot_and_factorise <- function(data, time_vary = FALSE) {
       status = factor(
         status,
         levels = c("Normal Cognition", "MCI", "Dementia"),
-        labels = c(1, 2, 3))) |>
[…]

### 2025-11-25 R/tidy-output.R

@@ -3,42 +3,53 @@ tidy_output <- function(fit, multiplicative = FALSE) {
     rename(transition = y.level) |>
     filter(term != "(Intercept)") |>
     mutate(
-      estimate = exp(estimate), conf.low = exp(conf.low), conf.high = exp(conf.high),
+      estimate = exp(estimate),
+      conf.low = exp(conf.low),
+      conf.high = exp(conf.high),
       colour = case_when(
         estimate > 1 & p.value < 0.05 ~ "Positive",
         estimate < 1 & p.value < 0.05 ~ "Negative",
-        TRUE ~ "NS"),
+        TRUE ~ "NS"
+      ),
       colour = factor(colour, levels = c("Positive", "Negative", "NS"))
-      ) |>
+    ) |>
     mutate(across(c(estimate:p.value), ~ round(x = ., digits = 3)))
-    
-  
-  if(any(output$transition == "1")) {
+
+  if (any(output$transition == "1")) {
     output <- output |>
-      mutate(transition = case_when(
-        transition == "1" ~ "MCI - NC",
-        transition == "3" ~ "MCI - Dementia"))
+      mutate(
+        transition = case_when(
+          transition == "1" ~ "MCI - NC",
+          transition == "3" ~ "MCI - Dementia"
+        )
+      )
   } else {
     output <- output |>
-      mutate(transition = case_when(
-        transition == "2" ~ "NC - MCI",
-        transition == "3" ~ "NC - Dementia"))
+      mutate(
+        transition = case_when(
+          transition == "2" ~ "NC - MCI",
+          transition == "3" ~ "NC - Dementia"
+        )
+      )
   }
-  
-  if(multiplicative == FALSE) {
+
+  if (multiplicative == FALSE) {
     output <- output |>
       mutate(
         term = case_when(
           term == "Gender1" ~ "Being female",
           term == "Education_tri1" ~ "High school degree vs. No education",
           term == "Education_tri2" ~ "Further education vs. No education",
-          term == "Total_dep_2016" ~ "Depression Scores (2016)",
+          term == "Apathy" ~ "Apathy Scores",
+          term == "Anxiety" ~ "Anxiety Scores",
           term == "Total_p" ~ "Procrastination (2020)",
           term == "status_prev2" ~ "Previous state: MCI",
           term == "status_prev3" ~ "Previous state: Dementia",
-          term == "Age:Total_p" ~ "Age x Procrastination",
[…]

### 2025-11-25 manuscript/manuscript.qmd
adding manuscript revisions

@@
 title: "Procrastination as a Marker of Cognitive Decline: Evidence from Longitudinal Transitions in the Older Adult Population"
-date: "2025-09-25"
+date: "2025-11-25"
 author:
@@
 impairment [@fresnais2023; @richard2012]. Apathy is prevalent in both MCI and
-dementia subgroups [@donovan2014; @fresnais2023, @salem2023; @vandalen2018].
+dementia subgroups [@donovan2014; @fresnais2023; @salem2023; @vandalen2018].
 Apathy is a significant predictor of the transition from MCI to dementia
-[@ruthirakuhan2019; @salem2023, @vandalen2018]. While apathy has been
+[@ruthirakuhan2019; @salem2023; @vandalen2018]. While apathy has been
 well-established as a prodromal marker and risk factor for dementia progression,
@@
 
-Though superficially similar, apathy and procrastination are behaviorally
-distinct. Apathy reflects reduced internal drive and emotional engagement,
-impairing initiation of action [@fresnais2023], whereas procrastination reflects
-intact intention but delayed execution [@steel2007]. While both reflect
-impairments in executive functioning, procrastination may additionally capture
-unique aspects of emotional and motivational dysregulation . For instance,
-person A may feel no motivation to attend an exercise class and therefore never
-books it (apathy). Person B, however, books the class and intends to go, but
-postpones at the last minute (procrastination). Both apathy and procrastination
-have been linked to dysfunction in prefrontal regions implicated in dementia
-[@friden2020; @zhang2010; @joseph2021; @fahed2021].
+Although apathy and procrastination can appear similar in that both lead to
+reduced engagement in everyday activities, the underlying behavioural processes
+are distinct. Apathy reflects reduced internal drive and emotional engagement,
+impairing initiation of action [@fresnais2023]. In contrast, procrastination is
+a self-regulatory failure defined by the voluntary delay of an important and
+intended course of action despite expecting negative consequences [@sirois2013;
+@klingsieck2013]. This distinction suggests that individuals who procrastinate
+typically form intentions but fail to execute them in a timely manner. This
+failure can often be understood through an affective framework, where delayed
+action may stem from the avoidance of negative emotions (i.e., stress or
+anxiety) associated with the task [@sirois2013]. Furthermore, while also
+conceptualised as a stable trait linked to specific personality factors
+[@meng2024; @sirois2017c], these factors can shift in later life and may
+interact with emerging cognitive or motivational changes to influence
+procrastination behaviour in older adult [@oi2024; @donnellan2008].
+
+Both apathy and procrastination have been linked to dysfunction in prefrontal
+regions implicated in dementia [@friden2020; @zhang2010; @joseph2021;
+@fahed2021]. Procrastination has been associated with alterations in brain
+circuitry supporting cognitive control and emotional regulation. Research has
+identified both structural and functional differences primarily in the
+prefrontal cortex (PFC) and parahippocampal cortex [@zhang2019]. Specifically,
+high procrastination is associated with decreased gray matter volume in the
+dorsolateral PFC (dlPFC) and increased spontaneous activity in the ventromedial
+PFC (vmPFC) [@liu2017]. Additionally, functional connectivity studies indicate
+altered couplings between both the vmPFC and dlPFC, along with the vmPFC and
+frontal gyri [@zhang2019]. Critically, these same networks are affected in
+neurodegenerative conditions, including Alzheimer’s disease [@joseph2021].
 
@@
 As a result, while the core HRS sample includes approximately $18,000$
-respondents, our initial sample of interest consisted of $1,368$ respondents. We
-excluded respondents with missing cognitive assessment data for any wave
-$(n = 419)$, those under 60 years of age, (as cognitive symptoms typically occur
-around this age $n = 398$), and those with complete missing values across the
-procrastination measure $(n = 2)$. This resulted in a final analytic sample of
-$549$ respondents.
+respondents, our inclusion criteria (i.e., having completed the experimental
+module) identified an initial sample of $1,368$ respondents. We then applied the
+following exclusion criteria: respondents with missing cognitive assessment data
+for any $(n = 419)$, those under 60 years of age, (as cognitive symptoms
+typically occur around this age $n = 398$), and those with complete missing
+values across the procrastination measure $(n = 2)$. The application of these
+exclusion criteria resulted in a final analytic sample of $549$ respondents, of
+whom $61.57\% (n = 338)$ were female.
 
@@
 approximation for modelling purposes. Each item was measured on a binary
-“yes/no” scale with total scores ranging from \$0\$ to \$2\$.
+“yes/no” scale with total scores ranging from $0$ to $2$.
 
@@
 
-\clearpage
+\newpage
 
@@
 
+\newpage
+
 # Results
@@
   "Age (years)"       , "69.70 ± 7.58"    , "69.70 ± 7.55"    , "69.60 ± 7.55"   , "70.00 ± 9.89"  ,
-  "Male"              , "38.43% (n = 211)" , "39.82% (n =180)"  , "32.56% (n = 28)" , "27.27% (n = 3)" ,
-  "Female"            , "61.57% (n = 338)" , "60.18% (n = 272)" , "67.44% (n = 58)" , "72.73% (n = 8)" ,
-  "No degree"         , "16.24% (n = 98)"  , "10.56% (n = 47)"  , "39.53% (n = 34)" , "63.54% (n = 7)" ,
-  "GED"               , "51.58% (n = 279)" , "51.91% (n = 231)" , "52.33% (n = 45)" , "27.27% (n = 3)" ,
-  "Further education" , "32.29% (n = 175)" , "37.53% (n = 167)" , "8.14% (n = 7)"   , "9.09% (n = 1)"  ,
+  "Male"              , "38.43%<br>(n = 211)" , "39.82%<br>(n = 180)"  , "32.56%<br>(n = 28)" , "27.27%<br>(n = 3)" ,
+  "Female"            , "61.57%<br>(n = 338)" , "60.18%<br>(n = 272)" , "67.44%<br>(n = 58)" , "72.73%<br>(n = 8)" ,
+  "No degree"         , "16.24%<br>(n = 98)"  , "10.56%<br>(n = 47)"  , "39.53%<br>(n = 34)" , "63.54%<br>(n = 7)" ,
+  "GED"               , "51.58%<br>(n = 279)" , "51.91%<br>(n = 231)" , "52.33%<br>(n = 45)" , "27.27%<br>(n = 3)" ,
+  "Further education" , "32.29%<br>(n = 175)" , "37.53%<br>(n = 167)" , "8.14%<br>(n = 7)"   , "9.09%<br>(n = 1)"  ,
   "Apathy"            , "0.37 ± 0.63"     , "0.31 ± 0.59"     , "0.65 ± 0.76"    , "0.36 ± 0.51"   ,
@@
 tbl_descriptives |>
-  setNames(c(" ", "Full sample (n = 549)", "NC (n = 452)", "MCI (n = 86)", "Dementia (n = 11)")) |>
-  tt(notes = list(
-    caption,
+  setNames(c(" ", "Full sample<br>(n = 549)", "NC<br>(n = 452)", "MCI<br>(n = 86)", "Dementia<br>(n = 11)")) |>
+  tt(notes = list(caption,
   a = list(i = 10, j = 1, text = footnote))
@@
    style_tt(j = 2:5, align = "c") |>
+   style_tt(j = 1, align = "l") |>
    style_tt(i = c(2, 5), align = "l", bold = TRUE) |>
-   format_tt(escape = TRUE) |>
+   # format_tt(escape = TRUE) |>
+   format_tt(linebreak = "<br>") |>
    format_tt("notes", markdown = TRUE) |>
@@
 assessments of procrastination to determine whether increases in procrastination
-precedes, accompanies, or follows cognitive decline. Moreover, our use of a
-proxy measure of apathy may not have fully captured the multidimensional nature
-of apathy. Future research should incorporate more validated apathy assessments
-to strengthen the behavioral inferences drawn.
+precedes, accompanies, or follows cognitive decline.
 
+A further limitation concerns the classification of cognitive status. As our
+analyses rely on the HRS, we are constrained by its operational definitions of
+MCI and dementia [@crimmins2011], which are based on threshold scores on the
+TICS. Although widely used [@langa2017; @davydow2015; @wei2018], these cut-offs
+may not fully align with other clinical diagnostic criteria or
+neuropsychological assessments, raising the possibility of misclassification.
+Future research using more comprehensive cognitive batteries or clinical
+diagnoses would help validate and refine the patterns observed here.
+
+Additionally, negative emotional states (e.g., anxiety or depression) are
+important psychological factors linked to both cognitive ageing and
+procrastination. However, we were unable to adjust for anxiety in our models due
+to substantial missingness in the HRS anxiety measure (approximately 61% missing
+for our sample). Similarly, depression was not included as a covariate because
+two CES-D items were used to construct our proxy measure of apathy, leaving a
+non-standard six-item score that would not provide a valid index of depressive
+symptoms. Future research using datasets with more complete emotional-health
+assessments may help clarify the role of negative emotions in these
+associations.
+
+sMoreover, our use of a proxy measure of apathy may not have fully captured the
+multidimensional nature of apathy. Future research should incorporate more
+validated apathy assessments to strengthen the behavioral inferences drawn.
 Additionally, as noted earlier, the relatively small number of individuals in
@@
 clinically diverse samples.
+
+\newpage
 
@@
 
-\newpage
-
 ::: {.content-visible when-format="html"}

### 2025-11-25 analysis/data-analysis.qmd
Update

@@
     toc: false
-bibliography: manuscript/references.bib
-csl: manuscript/csl/apa.csl
+bibliography: ../manuscript/references.bib
+csl: ../manuscript/csl/apa.csl
 editor:

### 2025-11-25 analysis/discrete-markov.qmd
Update

@@
   theme(
-    panel.grid = element_blank()
+    panel.grid = element_blank(),
     legend.title = element_text(size = 10, face = "bold"),

### 2025-12-02 R/pivot-and-factorise.R
Expanding for depression

@@ -19,7 +19,17 @@ pivot_and_factorise <- function(data, time_vary = FALSE) {
       ) |>
       select(Apathy)
 
-    # Pivot anxiety
+    # Pivoting depression variable
+    depression_scores <- data |>
+      select(ID, starts_with("depression")) |>
+      pivot_longer(
+        cols = !ID,
+        names_to = "remove",
+        values_to = "Depression"
+      ) |>
+      select(Depression)
+
+    # Pivot anxiety variable
     anxiety_scores <- data |>
       select(ID, starts_with("mean_anxiety")) |>
       pivot_longer(
@@ -31,7 +41,11 @@ pivot_and_factorise <- function(data, time_vary = FALSE) {
 
     # Pivoting whole dataset
     data <- data |>
-      select(!starts_with("apathy") & !starts_with("mean_anxiety")) |>
+      select(
+        !starts_with("apathy") &
+          !starts_with("mean_anxiety") &
+          !starts_with("depression")
+      ) |>
       pivot_longer(
         cols = starts_with("cogfunction"),
         names_to = "wave",
@@ -40,7 +54,10 @@ pivot_and_factorise <- function(data, time_vary = FALSE) {
       )
 
     # Adding in depression scores
-    data <- data |> cbind(apathy_scores) |> cbind(anxiety_scores)
+    data <- data |>
+      cbind(apathy_scores) |>
+      cbind(depression_scores) |>
+      cbind(anxiety_scores)
   } else {
     # No pivoting of depression
     data <- data |>

### 2025-12-02 R/tidy-output.R

@@ -40,8 +40,9 @@ tidy_output <- function(fit, multiplicative = FALSE) {
           term == "Gender1" ~ "Being female",
           term == "Education_tri1" ~ "High school degree vs. No education",
           term == "Education_tri2" ~ "Further education vs. No education",
-          term == "Apathy" ~ "Apathy Scores",
-          term == "Anxiety" ~ "Anxiety Scores",
+          term == "Apathy" ~ "Apathy",
+          term == "Anxiety" ~ "Anxiety",
+          term == "Depression" ~ "Depression",
           term == "Total_p" ~ "Procrastination (2020)",
           term == "status_prev2" ~ "Previous state: MCI",
           term == "status_prev3" ~ "Previous state: Dementia",

### 2025-12-02 manuscript/manuscript.qmd
Update manuscript

@@
 title: "Procrastination as a Marker of Cognitive Decline: Evidence from Longitudinal Transitions in the Older Adult Population"
-date: "2025-11-25"
+date: "2025-12-02"
 author:
@@
     wrap: 80
-keep-tex: true
 ---
@@
 impairment [@fresnais2023; @richard2012]. Apathy is prevalent in both MCI and
-dementia subgroups [@donovan2014; @fresnais2023; @salem2023; @vandalen2018].
-Apathy is a significant predictor of the transition from MCI to dementia
-[@ruthirakuhan2019; @salem2023; @vandalen2018]. While apathy has been
-well-established as a prodromal marker and risk factor for dementia progression,
-emerging evidence suggests that procrastination may also relate to early
-cognitive decline.
+dementia subgroups [@donovan2014; @fresnais2023; @salem2023; @vandalen2018], and
+is a robust predictor of progression from MCI to dementia [@ruthirakuhan2019;
+@salem2023; @vandalen2018]. While apathy has been well-established as a
+prodromal marker and risk factor for dementia progression, emerging evidence
+suggests that procrastination may also relate to early cognitive decline.
 
@@
 conceptualised as a stable trait linked to specific personality factors
-[@meng2024; @sirois2017c], these factors can shift in later life and may
-interact with emerging cognitive or motivational changes to influence
-procrastination behaviour in older adult [@oi2024; @donnellan2008].
+[@meng2024], these factors can shift in later life and may interact with
+emerging cognitive or motivational changes to influence procrastination
+behaviour in older adult [@oi2024; @donnellan2008].
 
 Both apathy and procrastination have been linked to dysfunction in prefrontal
-regions implicated in dementia [@friden2020; @zhang2010; @joseph2021;
-@fahed2021]. Procrastination has been associated with alterations in brain
-circuitry supporting cognitive control and emotional regulation. Research has
-identified both structural and functional differences primarily in the
-prefrontal cortex (PFC) and parahippocampal cortex [@zhang2019]. Specifically,
-high procrastination is associated with decreased gray matter volume in the
-dorsolateral PFC (dlPFC) and increased spontaneous activity in the ventromedial
-PFC (vmPFC) [@liu2017]. Additionally, functional connectivity studies indicate
-altered couplings between both the vmPFC and dlPFC, along with the vmPFC and
-frontal gyri [@zhang2019]. Critically, these same networks are affected in
-neurodegenerative conditions, including Alzheimer’s disease [@joseph2021].
+regions implicated in dementia [@zhang2010; @joseph2021]. Procrastination has
+been associated with alterations in brain circuitry supporting cognitive control
+and emotional regulation. Research has identified both structural and functional
+differences primarily in the prefrontal cortex (PFC) and parahippocampal cortex
+[@zhang2019]. Specifically, high procrastination is associated with decreased
+gray matter volume in the dorsolateral PFC (dlPFC) and increased spontaneous
+activity in the ventromedial PFC (vmPFC). Additionally, functional connectivity
+studies indicate altered couplings between both the vmPFC and dlPFC, along with
+the vmPFC and frontal gyri [@zhang2019]. Critically, these same networks are
+affected in neurodegenerative conditions, including Alzheimer’s disease
+[@joseph2021].
 
@@
 goal-setting, which build cognitive resilience and reduce dementia risk
-[@kelly2021; @mohammadibytamar2020; @chowdhary2022; @livingston2024]. Reduced
-engagement may contribute to a cycle of cognitive disengagement, accelerating
-decline.
+[@kelly2021; @livingston2024]. Reduced engagement may contribute to a cycle of
+cognitive disengagement, accelerating decline.
 
@@
 
+\newpage
+
 ### Predictor: Procrastination
@@
 
+### Covariate: Depression
+
+Depressive symptoms were included in our model based on their associations with
+both procrastination [@klingsieck2013] and cognitive ageing [@livingston2020].
+As mentioned, the HRS uses the CESD-8 as a measure of depressive symptomatology.
+The scale consists of 8 items that are scored as either 0 (no) or 1 (yes), with
+items 4 and 6 being reversed scored. Since we used two items from this scale as
+proxies for apathy, we scored the remaining 6 items, giving a total score
+ranging from 0 to 6, with higher scores indicating higher levels of depressive
+symptoms. Across the waves, this 6-item scale showed reliable internal
+consistency $(\alpha \in [0.81-0.83])$. An example of a question from the scale
+includes “Much of the time during the past week, your sleep was restless”.
+
 ### Confounders
@@
 [@vanderweele2019]. These included measures of age, sex, and educational
-attainment [@abner2012; @freedman2024; @yu2013]. Educational attainment was
-classified into three categories: no formal education, GED (General Educational
+attainment [@livingston2020; @steel2013]. Educational attainment was classified
+into three categories: no formal education, GED (General Educational
 Development)/high school diploma, and college/further education.
@@
 cognitive states over time, we employed a first order discrete-time Markov
-model, a class of stochastic processes that satisfy the Markov property
-[@zhang2010], which can be formally expressed as:
+model, a class of stochastic processes that satisfy the Markov property, which
+can be formally expressed as:
 
@@
 stratified by cognitive status (normative cognitive function, MCI, and
-dementia), are presented in @tbl-descriptives. Both @fig-river and supplementary
-figure S1 capture the unconditional transitions and transition probabilities
-(respectively) between wave one and two (first transition), wave two and three
-(second transition), and wave three and four (third transition), yielding a
-total of $1,647$ observed transitions over time.
+dementia), are presented in @tbl-descriptives. Both @fig-river and Figure S1 in
+the supplement capture the unconditional transitions and transition
+probabilities (respectively) between wave one and two (first transition), wave
+two and three (second transition), and wave three and four (third transition),
+yielding a total of $1,647$ observed transitions over time.
 
@@
   "Apathy"            , "0.37 ± 0.63"        , "0.31 ± 0.59"        , "0.65 ± 0.76"       , "0.36 ± 0.51"      ,
+  "Depression"        , "0.98 ± 1.48"        , "0.88 ± 1.41"        , "1.41 ± 1.71"       , "1.09 ± 1.87"      ,
   "Procrastination"   , "28.60 ± 12.00"      , "27.70 ± 11.30"      , "32.10 ± 13.80"     , "39.00 ± 16.00"
@@
   tt(notes = list(caption,
-  a = list(i = 10, j = 1, text = footnote))
+  a = list(i = 11, j = 1, text = footnote))
   ) |>
@@
 significantly influenced the likelihood of transitioning between cognitive
-states (see @fig-markov). Notably, procrastination interacted significantly with
-*age* to affect two key transitions: increasing the likelihood of transitioning
-from normative cognitive function to MCI $(\text{OR} = 1.001; p < 0.001)$ and
-decreasing the likelihood of reverting from MCI to normative cognitive function
-$(\text{OR} = 0.999; p < 0.001)$. *Women* were significantly less likely than
-men to transition from both normative cognitive function to dementia
-$(\text{OR} = 0.068; p < 0.001)$ and from MCI to dementia
-$(\text{OR} = 0.70; p < 0.001$).
+states (see @fig-markov). Notably, *procrastination* interacted significantly
+with *age* to affect two key transitions: increasing the likelihood of
+transitioning from normative cognitive function to MCI
+$(\text{OR} = 1.001; p < 0.001)$ and decreasing the likelihood of reverting from
+MCI to normative cognitive function $(\text{OR} = 0.999; p < 0.001)$. *Women*
+were significantly less likely than men to transition from both normative
+cognitive function to dementia $(\text{OR} = 0.67; p < 0.001)$ and from MCI to
+dementia $(\text{OR} = 0.71; p < 0.001$).
 
@@
 transition from normative cognitive function to either MCI
-$(\text{OR} = 0.49; p < 0.001)$ or dementia $(\text{OR} = 0.29; p < 0.001)$ and
+$(\text{OR} = 0.46; p < 0.001)$ or dementia $(\text{OR} = 0.29; p < 0.001)$ and
 from MCI to dementia $(\text{OR} = 0.64; p < 0.001)$. They were also more likely
 to back transition from MCI to normative cognitive function
-$(\text{OR} = 2.18; p < 0.001)$. Those with a college level education or higher
+$(\text{OR} = 2.24; p < 0.001)$. Those with a college level education or higher
 demonstrated a significantly reduced likelihood of transitioning from normative
-cognitive function to either MCI $(\text{OR} = 0.32; p < 0.001)$ or dementia
+cognitive function to either MCI $(\text{OR} = 0.31; p < 0.001)$ or dementia
 $(\text{OR} = 0.07; p < 0.001)$ and from MCI to dementia
 $(\text{OR} = 0.22; p < 0.001)$. They were also more likely to back transition
-from MCI to normative cognitive function $(\text{OR} = 3.13; p < 0.001)$.
+from MCI to normative cognitive function $(\text{OR} = 3.23; p < 0.001)$.
 
-Finally, higher levels of *apathy* were associated with an increased likelihood
-of transitioning from normative cognitive function to both MCI
-$(\text{OR} = 1.35; p = 0.005)$ and dementia $(\text{OR} = 1.36; p<0.001)$ and a
-decreased likelihood of transitioning from MCI back to normative cognitive
-function $(\text{OR} = 0.74; p = 0.003)$.
+Higher levels of *depression* were associated with an increased likelihood of
+transitioning from normative cognitive function to MCI
+$(\text{OR} = 1.14; p = 0.013)$ and a reduced likelihood of transitioning back
+from MCI to normal cognition $\text{OR} = 0.88; p = 0.013)$. Finally, higher
+levels of *apathy* were associated with an increased likelihood of transitioning
+from normative cognitive function to dementia $(\text{OR} = 1.20; p < 0.001)$
 
@@
 @vandalen2018], emerging evidence suggests that procrastination may also signal
-early cognitive dysfunction [@friden2020; @zhang2019]. This study examined
+early cognitive dysfunction [@joseph2021; @zhang2019]. This study examined
 whether procrastination, particularly in later life, could serve as a predictor
@@
 with cognitive decline and aligns with emerging evidence linking procrastination
-to cognitive dysfunction [@friden2020; @zhang2019]. Interestingly, while both
+to cognitive dysfunction [@joseph2021; @zhang2019]. Interestingly, while both
 MCI and dementia groups exhibited elevated procrastination scores, no
@@
 subtle differences between MCI and dementia groups were obscured by statistical
-power constraints. Future research should aim to replicate these findings in
-more evenly distributed samples to better determine whether procrastination
-continues to increase across progressive stages of impairment or plateaus
-following early decline.
+power constraints. In addition, our analytic sample contained a higher
+proportion of women than men $(61.57\% \text{female})$ reflecting the
+demographic composition of the available HRS respondents who met eligibility
+criteria. Although this imbalance was not the result of a sampling decision, it
+may influence the generalisability of our findings, as women in older cohorts
+differ from men in both dementia risk profiles and health-seeking behaviours
+[@berve2025]. Future research should aim to replicate these findings in more
+evenly distributed samples to better determine whether these results are
+consistent across men and women and whether procrastination continues to
+increase across progressive stages of impairment or plateaus following early
+decline.
 
@@
 self-regulation theories suggesting that procrastination reflects executive
-dysfunction [@rozental2014; @steel2007], an early feature of neurodegenerative
-progression [@clark2012].
+dysfunction [@rozental2014], an early feature of neurodegenerative progression
+[@clark2012]. At the same time, these results align with both the affective and
+trait-based perspectives on procrastination. Higher procrastination in
+individuals with cognitive impairment may partly reflect affective dysregulation
+(e.g., anxiety or stress-related avoidance) [@sirois2013] or age-related shifts
+in stable behavioural tendencies that accompany neurodegenerative change
+[@oi2024]. While our study cannot differentiate between these frameworks, our
+results highlight that procrastination is a multifaceted construct encompassing
+motivational, affective, and self-regulatory dimensions.
 
@@
 that this limitation is inherent to the HRS dataset rather than a methodological
-oversight (Juster & Suzman, 1995). Future studies should prioritise repeated
-assessments of procrastination to determine whether increases in procrastination
-precedes, accompanies, or follows cognitive decline.
+oversight [@juster1995]. Future studies should prioritise repeated assessments
+of procrastination to determine whether increases in procrastination precedes,
+accompanies, or follows cognitive decline.
 
@@
 MCI and dementia [@crimmins2011], which are based on threshold scores on the
-TICS. Although widely used [@langa2017; @davydow2015; @wei2018], these cut-offs
-may not fully align with other clinical diagnostic criteria or
-neuropsychological assessments, raising the possibility of misclassification.
-Future research using more comprehensive cognitive batteries or clinical
-diagnoses would help validate and refine the patterns observed here.
+TICS [@brandt1988; @fong2009]. Developed as a telephone-based analogue of the
+Mini-Mental State Examination, the TICS is a widely adopted and well-validated
+cognitive screener [@fong2009]. However, like all brief screening measures,
+TICS-based classifications may not fully align with more detailed clinical
+assessments, raising the possibility of diagnostic misclassification. Future
+research using more comprehensive cognitive batteries or clinical diagnoses
+would help validate and refine the patterns observed here.
 
-Additionally, negative emotional states (e.g., anxiety or depression) are
-important psychological factors linked to both cognitive ageing and
-procrastination. However, we were unable to adjust for anxiety in our models due
-to substantial missingness in the HRS anxiety measure (approximately 61% missing
-for our sample). Similarly, depression was not included as a covariate because
-two CES-D items were used to construct our proxy measure of apathy, leaving a
-non-standard six-item score that would not provide a valid index of depressive
-symptoms. Future research using datasets with more complete emotional-health
-assessments may help clarify the role of negative emotions in these
-associations.
+Additionally, negative emotional states (e.g., anxiety) are important
+psychological factors linked to both cognitive ageing and procrastination
+[@livingston2020; @sirois2013]. However, we were unable to robustly adjust for
+anxiety in our models due to substantial systematic missingness in the HRS
+anxiety measure (see Figure S2 and Figure S3 in the supplement). Moreover, our
+use of a proxy measure of apathy may not have fully captured the
+multidimensional nature of apathy. Additionally, it meant that only six of the
+original eight CESD-8 items were used for scoring depression, possibly
+introducing measurement error. Future research should incorporate more validated
+apathy assessments to strengthen the behavioral inferences drawn.
 
-sMoreover, our use of a proxy measure of apathy may not have fully captured the
-multidimensional nature of apathy. Future research should incorporate more
-validated apathy assessments to strengthen the behavioral inferences drawn.
-Additionally, as noted earlier, the relatively small number of individuals in
-the MCI and dementia groups may have constrained statistical power for certain
-comparisons and increased the potential for misclassification bias. Finally,
-while the models adjusted for several key demographic and psychological
+Finally, while the models adjusted for several key demographic and psychological
 covariates, unmeasured confounding (undiagnosed medical conditions, medication
@@
 clinically diverse samples.
-
-\newpage
 

### 2025-12-03 analysis/discrete-markov.qmd
Update :fire:

--- 2025-11-25 analysis/discrete-markov.qmd
@@
   paste0("apathy_", seq(2016, 2022, by = 2)),
+  paste0("depression_", seq(2016, 2022, by = 2)),
   paste0("mean_anxiety_", seq(16, 22, by = 2)),
@@
 fit_a <- nnet::multinom(
-  status ~ Gender + Education_tri + Apathy + (Total_p * Age) + status_prev,
+  status ~ Gender + Education_tri + Apathy + Depression + (Total_p * Age) + status_prev,
   family = multinomial,
@@
 fit_b <- nnet::multinom(
-  status ~ Gender + Education_tri + Apathy + (Total_p * Age) + status_prev,
+  status ~ Gender + Education_tri + Apathy + Depression + (Total_p * Age) + status_prev,
   family = multinomial,
@@
                  "Further education vs. No education",
-                 "Apathy Scores", "Anxiety Scores",
+                 "Apathy", "Anxiety", "Depression",
                  "Procrastination (2020)",
@@
   DDEEFF
-  ##GG##
+  #GGHH#
 "
@@
   Apathy = mean(data_stack$Apathy, na.rm = TRUE),
+  Depression = mean(data_stack$Depression, na.rm = TRUE),
   status_prev = levels(data_stack$status_prev),
@@
   theme(
-    panel.grid = element_blank()
+    panel.grid = element_blank(),
     legend.title = element_text(size = 10, face = "bold"),
-    legend.position = "none")
+    legend.position = "none") -> new_fig
 ```

### 2025-12-03 analysis/discrete-markov.qmd
Update :fire:

@@
     legend.title = element_text(size = 10, face = "bold"),
-    legend.position = "none") -> new_fig
+    legend.position = "none")
 ```
